### 2019-05-06 exploratory/fall_rotation_2018/MouseSensspecFigures_2018-12-19.Rmd
Make code executable

---
title: "OptifitLabMeetingPlots"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse)
```

## Generation of figures for end of semester rotation presentation

This is the code used to generate the figures presented at lab meeting on 12/12/18. The data was generated from the last version of the bash scripts, not from snakemake. The sensspec file here was generated automatically, except for the final 60 lines containing purely OptiClust MCC data which I added manually for a 100% of data OptiClust comparison.

For all plots here, a single point is the MCC score from a single run of OptiFit or OptiClust.

First we import the data and create factors (so things don't just plot alphabetically) and full text labels for later use in our plots.

```{r cars}
mice_labmeeting <- read_tsv("mice.sensspec.labmeeting")
mice_labmeeting$method <- factor(mice_labmeeting$method, levels=c("none", "samp.abund", "samp.dist",
                                                                    "ref.dist", "ref.abund", "OptiClust"))
mice_labmeeting$type <- factor(mice_labmeeting$type, levels = c("REF", "SAMP",
                                                                  "SAMP_C_REF", "SAMP_O_REF", "SAMP_C_NOREF" ,"SAMP_O_NOREF"))
mice_labmeeting$label <- sapply(mice_labmeeting$type, function(x)  c("REF", "SAMP", "Closed, with ref",
                                                       "Open, with ref", "Closed, no ref", "Open, no ref")[x])
```

The first comparison plot we want to make is for the reference alone using OptiClust. This just lets us see how the MCC score varies using different sized datasets. As noted in the legend, the weighting method for sampling here is "none" meaning sequences were completely randomly chosen to be a part of the reference. We can see that the MCC seems to converge as the number of sequences increases.

```{r mice_ref}
mice_ref <- filter(mice_labmeeting, method %in% c("none") & type %in% c("REF"))

mice_ref_plot <- ggplot(data=mice_ref, aes(x = refp, y = mcc, color = method)) +
  facet_wrap(~ label, ncol=2) +
  geom_jitter(width = 1) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  ylab("MCC") +
  xlab("Percent of data used as reference (%)")

mice_ref_plot 
```

Next is a comparison of weighting methods when choosing sequences to go into the reference. First we look at random sampling vs sampling by abundance. We can see that in almost every case, MCC scores obtained when sampling by abundance are superior to sampling completely randomly.
```{r mice_abund}
mice_by_abund <- filter(mice_labmeeting, method %in% c("ref.abund", "none") & !(type %in% c("REF", "SAMP")))
mice_abund_plot <- ggplot(data=mice_by_abund, aes(x = refp, y = mcc, color = method)) +
  facet_wrap(~ label, ncol=2) +
  stat_summary(fun.y = "mean", geom = "point", size = 4) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 1, size = 1) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  ylab("MCC") +
  xlab("Percent of data used as reference (%)")

mice_abund_plot
```

We also want to look at weighting our reference selection by the number of pairwise connections (distances <3%) per sequence. Here we can see that when dropping the reference from the final output, weighting by connections gives lower MCC scores, but when including the reference it appears to give higher MCC scores when compared to completely random sampling.

```{r mice_dist}
mice_by_dist <- filter(mice_labmeeting, method %in% c("ref.dist", "none") & !(type %in% c("REF", "SAMP")))
mice_dist_plot <- ggplot(data=mice_by_dist, aes(x = refp, y = mcc, color = method)) +
  facet_wrap(~ label, ncol=2) +
  stat_summary(fun.y = "mean", geom = "point", size = 4) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 1, size = 1) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  ylab("MCC") +
  xlab("Percent of data used as reference (%)")

mice_dist_plot
```

Now for a direct comparison of sampling by abundance and sampling by connections. Except in the case of small references when closed fitting with reference, sampling by abundance produces consistently higher MCC scores.

```{r mice_dist_abund}
mice_by_dist_abund <- filter(mice_labmeeting, method %in% c("ref.dist", "ref.abund") & !(type %in% c("REF", "SAMP")))
mice_dist_abund_plot <- ggplot(data=mice_by_dist_abund, aes(x = refp, y = mcc, color = method)) +
  facet_wrap(~ label, ncol=2) +
  stat_summary(fun.y = "mean", geom = "point", size = 4) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 1, size = 1) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  ylab("MCC") +
  xlab("Percent of data used as reference (%)")

mice_dist_abund_plot
```

To look at random variation, we look back at our first plot when sampling by abundance, but look at the individual data points. We can see that while the MCC scores generally cluster to a single value, sometimes they randomly drop out to a lower MCC. Overall this effect seems to be rare though, especially when sampling by abundance. Statistical quantification of this "drop-out" rate may be necessary.

```{r abund_variation}
mice_abund_variation_plot <- ggplot(data=mice_by_abund, aes(x = refp, y = mcc, color = method)) +
  facet_wrap(~ label, ncol=2) +
  geom_jitter(width = 1) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  ylab("MCC") +
  xlab("Percent of data used as reference (%)")

mice_abund_variation_plot
```

Finally, we wanted to compare OptiFit's results to OptiClust to ensure that the algorithm tends to give similar MCC scores in similar scenarios. This helps verify that the algorithm is doing what we think it should be doing, and gives confidence that we are getting MCC scores as close to optimal as possible. We can see that OptiFit does indeed give MCC scores almost equivalent to OptiClust, with most scores be ~.1% less than OptiClust. Even the random "drop-outs" are only around 2% worse than OptiClust, indicating overall good MCC scores when comparing the two methods.

```{r opticlust}
mice_opticlust <- filter(mice_labmeeting, method %in% c("ref.abund", "OptiClust") & !(type %in% c("REF", "SAMP",
                                                                                                   "SAMP_C_REF", "SAMP_C_NOREF",
                                                                                                   "SAMP_O_NOREF")))
opticlust_plot <- ggplot(data=mice_opticlust, aes(x = refp, y = mcc, color = method)) +
  facet_wrap(~ label, ncol=2) +
  geom_jitter(width = 1) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  ylab("MCC") +
  xlab("Percent of data used as reference (%)")

opticlust_plot
```

### 2020-05-28 exploratory/2020/README.Rmd
Add exploratory plots

---
title: "plots.Rmd"
author: "Kelly Sovacool"
date: "5/27/2020"
output: 
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      fig.path = "figures/"
                      )
```
```{r lib}
library(here)
library(tidyverse)
theme_set(theme_classic())
```

# Fitting datasets to a reference database vs _de novo_ clustering

```{r fit_db_sensspec}
sensspec <- read_tsv(here('subworkflows/2_fit_reference_db/results/merged.sensspec')) %>% 
  mutate(method=fct_relevel(method, c("open", "closed", "de_novo")))
sensspec %>% group_by(dataset, method) %>% 
  ggplot(aes(x=method, y=mcc, fill=dataset)) +
  geom_jitter()
```

```{r fit_db_benchmarks}
benchmarks <- read_tsv(here('subworkflows/2_fit_reference_db/results/benchmarks.txt')) %>% 
  mutate(method=fct_relevel(method, c("open", "closed", "de_novo")))
benchmarks %>% group_by(dataset, method) %>% 
  ggplot(aes(x=method, y=s, fill=dataset)) +
  geom_jitter() +
  labs(y='seconds')
```

```{r render, echo=FALSE, eval=FALSE}
rmarkdown::render(here::here("exploratory/2020/README.Rmd"), 
                  output_dir = "exploratory/2020/")
```

### 2020-05-28 exploratory/2020/README.Rmd
Use functions

@@
 ---
-title: "plots.Rmd"
-author: "Kelly Sovacool"
+title: "Exploratory Plots"
 date: "5/27/2020"
@@
 # Fitting datasets to a reference database vs _de novo_ clustering
+```{r functions}
+relevel_method <- function(df) {
+  df %>%
+    mutate(method = fct_relevel(method, c("open", "closed", "de_novo")))
+}
+plot_jitter <- function(df, y) {
+  df %>% group_by(dataset, method) %>% 
+      ggplot(aes(x=method, y={{ y }}, fill=dataset)) +
+  geom_jitter()
+}
+```
 
@@
 sensspec <- read_tsv(here('subworkflows/2_fit_reference_db/results/merged.sensspec')) %>% 
-  mutate(method=fct_relevel(method, c("open", "closed", "de_novo")))
-sensspec %>% group_by(dataset, method) %>% 
-  ggplot(aes(x=method, y=mcc, fill=dataset)) +
-  geom_jitter()
+  relevel_method()
+sensspec %>% 
+  plot_jitter(mcc)
 ```
@@
 benchmarks <- read_tsv(here('subworkflows/2_fit_reference_db/results/benchmarks.txt')) %>% 
-  mutate(method=fct_relevel(method, c("open", "closed", "de_novo")))
-benchmarks %>% group_by(dataset, method) %>% 
-  ggplot(aes(x=method, y=s, fill=dataset)) +
-  geom_jitter() +
+  relevel_method
+benchmarks %>% 
+  plot_jitter(s) +
   labs(y='seconds')

### 2020-05-28 exploratory/2020/README.Rmd
Plot fraction of reads mapped

@@
 ```{r fit_db_sensspec}
-sensspec <- read_tsv(here('subworkflows/2_fit_reference_db/results/merged.sensspec')) %>% 
+sensspec <- read_tsv(here('subworkflows/2_fit_reference_db/results/sensspec.txt')) %>% 
   relevel_method()
@@
 
+```{r fraction_reads_mapped}
+df_fraction_mapped <- read_tsv(here('subworkflows/2_fit_reference_db/results/fraction_reads_mapped.txt'))
+df_fraction_mapped %>% 
+  ggplot(aes(dataset, fraction_mapped)) +
+  geom_jitter()
+```
+
 ```{r render, echo=FALSE, eval=FALSE}

### 2020-05-28 exploratory/2020/README.Rmd
Don't need dist file for sub2

@@
 ```{r fraction_reads_mapped}
-df_fraction_mapped <- read_tsv(here('subworkflows/2_fit_reference_db/results/fraction_reads_mapped.txt'))
-df_fraction_mapped %>% 
+fractions <- read_tsv(here('subworkflows/2_fit_reference_db/results/fraction_reads_mapped.txt'))
+fractions %>% 
   ggplot(aes(dataset, fraction_mapped)) +

### 2020-05-28 subworkflows/2_fit_reference_db/debug.sh
Forcerun last merge job

@@ -35,4 +35,4 @@
 #SBATCH --mail-user=sovacool@umich.edu
 #SBATCH --mail-type=BEGIN,END
 
-time snakemake --profile config/slurm --latency-wait 90 --forcerun results/soil_silva_bact_v4/optifit/method_open/printref_f/seed_0/soil_silva.bact_v4.all.unique.optifit_mcc.sensspec
+time snakemake --profile config/slurm --latency-wait 90 --forcerun merge_mapped_stats

### 2020-05-29 exploratory/2020/README.Rmd
Update plots with marine data

@@
   df %>% group_by(dataset, method) %>% 
-      ggplot(aes(x=method, y={{ y }}, fill=dataset)) +
+      ggplot(aes(x=method, y={{ y }}, color=dataset)) +
   geom_jitter()
@@
 
+```{r fit_db_input_size}
+
+```
+
 ```{r fraction_reads_mapped}
@@
 fractions %>% 
-  ggplot(aes(dataset, fraction_mapped)) +
-  geom_jitter()
+  ggplot(aes(x=dataset, y=fraction_mapped, color=dataset)) +
+  geom_jitter() +
+  ylim(0, 1)
 ```

### 2020-05-29 exploratory/2020/README.Rmd
Pick color palette and use arg forwarding

--- 2020-05-28 exploratory/2020/README.Rmd
@@
 theme_set(theme_classic())
+color_palette <- RColorBrewer::brewer.pal(4, "Dark2")
+dataset_colors <- list(human = color_palette[[3]],
+                       marine = color_palette[[1]],
+                       mouse = color_palette[[4]],
+                       soil = color_palette[[2]]
+                       )
 ```
@@
 }
-plot_jitter <- function(df, y) {
-  df %>% group_by(dataset, method) %>% 
-      ggplot(aes(x=method, y={{ y }}, fill=dataset)) +
-  geom_jitter()
+plot_jitter <- function(.df, ...) {
+  .df %>% 
+    ggplot(aes(...)) +
+    geom_jitter() +
+    scale_color_manual(values=dataset_colors)
 }
@@
 sensspec %>% 
-  plot_jitter(mcc)
+  group_by(dataset, method) %>% 
+  plot_jitter(x = method, y = mcc, color = dataset) +
+  ylim(0, 1)
 ```
@@
 benchmarks %>% 
-  plot_jitter(s) +
+  group_by(dataset, method) %>% 
+  plot_jitter(x = method, y = s, color = dataset) +
   labs(y='seconds')
+```
+
+```{r fit_db_input_size}
+
 ```
@@
 fractions %>% 
-  ggplot(aes(dataset, fraction_mapped)) +
-  geom_jitter()
+  plot_jitter(x=dataset, y=fraction_mapped, color=dataset) +
+  ylim(0, 1)
 ```

### 2020-06-09 exploratory/2020/README.Rmd

@@
 }
-plot_jitter <- function(.df, ...) {
-  .df %>% 
-    ggplot(aes(...)) +
-    geom_jitter() +
-    scale_color_manual(values=dataset_colors)
-}
 ```
@@
 ```{r fit_db_sensspec}
-sensspec <- read_tsv(here('subworkflows/2_fit_reference_db/results/sensspec.txt')) %>% 
+sensspec <-
+  read_tsv(here('subworkflows/2_fit_reference_db/results/sensspec.tsv')) %>%
   relevel_method()
@@
   group_by(dataset, method) %>%
-  plot_jitter(x = method, y = mcc, color = dataset) +
-  ylim(0, 1)
+  ggplot(aes(x = method, y = mcc, color = dataset)) +
+  geom_jitter(alpha = 0.5) +
+  scale_color_manual(values = dataset_colors) +
+  ylim(0, 1) +
+  facet_wrap("ref") 
 ```
 
-```{r fit_db_benchmarks}
-benchmarks <- read_tsv(here('subworkflows/2_fit_reference_db/results/benchmarks.txt')) %>% 
+```{r fit_db_benchmarks, warning=FALSE}
+benchmarks <-
+  read_tsv(here('subworkflows/2_fit_reference_db/results/benchmarks.tsv')) %>%
   relevel_method
@@
   group_by(dataset, method) %>%
-  plot_jitter(x = method, y = s, color = dataset) +
-  labs(y='seconds')
+  ggplot(aes(x = method, y = s, color = dataset)) +
+  geom_boxplot() +
+  scale_color_manual(values = dataset_colors) +
+  facet_wrap("ref") +
+  labs(y = 'seconds')
 ```
@@
 ```{r fit_db_input_size}
-
+input_sizes <-
+  read_tsv(here('subworkflows/2_fit_reference_db/results/input_sizes.tsv'))
 ```
 
+Only for closed-reference OptiFit
+
 ```{r fraction_reads_mapped}
-fractions <- read_tsv(here('subworkflows/2_fit_reference_db/results/fraction_reads_mapped.txt'))
+fractions <- read_tsv(here('subworkflows/2_fit_reference_db/results/fraction_reads_mapped.tsv'))
 fractions %>% 
-  plot_jitter(x=dataset, y=fraction_mapped, color=dataset) +
+  ggplot(aes(x=dataset, y=fraction_mapped, color=dataset)) +
+  geom_jitter(alpha = 0.5) +
+  scale_color_manual(values = dataset_colors) +
+  facet_wrap("ref") +
   ylim(0, 1)

### 2020-06-09 exploratory/2020/README.Rmd
Show input size tables

--- 2020-05-29 exploratory/2020/README.Rmd
@@
 theme_set(theme_classic())
+color_palette <- RColorBrewer::brewer.pal(4, "Dark2")
+dataset_colors <- c(
+  human = color_palette[[3]],
+  marine = color_palette[[1]],
+  mouse = color_palette[[4]],
+  soil = color_palette[[2]]
+)
 ```
@@
 }
-plot_jitter <- function(df, y) {
-  df %>% group_by(dataset, method) %>% 
-      ggplot(aes(x=method, y={{ y }}, color=dataset)) +
-  geom_jitter()
-}
 ```
@@
 ```{r fit_db_sensspec}
-sensspec <- read_tsv(here('subworkflows/2_fit_reference_db/results/sensspec.txt')) %>% 
+sensspec <-
+  read_tsv(here('subworkflows/2_fit_reference_db/results/sensspec.tsv')) %>%
   relevel_method()
 sensspec %>%
-  plot_jitter(mcc)
+  group_by(dataset, method) %>%
+  ggplot(aes(x = method, y = mcc, color = dataset)) +
+  geom_boxplot(alpha = 0.5) +
+  scale_color_manual(values = dataset_colors) +
+  #ylim(0, 1) +
+  facet_grid(dataset ~ ref)
 ```
 
-```{r fit_db_benchmarks}
-benchmarks <- read_tsv(here('subworkflows/2_fit_reference_db/results/benchmarks.txt')) %>% 
+```{r fit_db_benchmarks, warning=FALSE}
+benchmarks <-
+  read_tsv(here('subworkflows/2_fit_reference_db/results/benchmarks.tsv')) %>%
   relevel_method
 benchmarks %>%
-  plot_jitter(s) +
-  labs(y='seconds')
+  group_by(dataset, method) %>%
+  ggplot(aes(x = method, y = s, color = dataset)) +
+  geom_boxplot() +
+  scale_color_brewer(palette = "Dark2") +
+  #scale_color_manual(values = dataset_colors) +  # bug in geom_boxplot with manual colors?
+  facet_wrap("ref") +
+  labs(y = 'seconds')
 ```
 
-```{r fit_db_input_size}
+```{r fit_db_ref_size}
+ref_sizes <-
+  read_tsv(here('subworkflows/2_fit_reference_db/results/ref_sizes.tsv'))
+ref_sizes
+```
 
+Potential problem: started with 174k reference sequences in RDP; filter with the hard parameter leaves only ~6k sequences remaining.
+
+```{r fit_db_dataset_size}
+dataset_sizes <-
+  read_tsv(here('subworkflows/1_prep_samples/results/dataset_sizes.tsv'))
+dataset_sizes
 ```
@@
 ```{r fraction_reads_mapped}
-fractions <- read_tsv(here('subworkflows/2_fit_reference_db/results/fraction_reads_mapped.txt'))
+fractions <- read_tsv(here('subworkflows/2_fit_reference_db/results/fraction_reads_mapped.tsv'))
 fractions %>% 
   ggplot(aes(x=dataset, y=fraction_mapped, color=dataset)) +
-  geom_jitter() +
-  ylim(0, 1)
+  geom_boxplot(alpha = 0.5) +
+  scale_color_manual(values = dataset_colors) +
+  facet_wrap("ref") +
+  ylim(0, 1) +
+  labs(title="Sequences mapped during closed-reference OptiFit")
 ```

### 2020-06-09 exploratory/2020/README.Rmd
Use knitr::kable for tables

@@
 library(here)
+library(knitr)
 library(tidyverse)
@@
   read_tsv(here('subworkflows/2_fit_reference_db/results/ref_sizes.tsv'))
-ref_sizes
+kable(ref_sizes)
 ```
@@
   read_tsv(here('subworkflows/1_prep_samples/results/dataset_sizes.tsv'))
-dataset_sizes
+kable(dataset_sizes)
 ```

### 2020-06-10 exploratory/2020/README.Rmd
Add subheadings

@@
 
+### Performance as measured by MCC
 ```{r fit_db_sensspec}
@@
 
+### Performance as measured by runtime
 ```{r fit_db_benchmarks, warning=FALSE}
@@
 
+### Reference & dataset sizes
 ```{r fit_db_ref_size}
@@
 ```
-
-Potential problem: started with 174k reference sequences in RDP; filter with the hard parameter leaves only ~6k sequences remaining.
 
@@
 
+### Fraction of sequences that map to the reference
 ```{r fraction_reads_mapped}

### 2020-06-10 exploratory/2020/README.Rmd
Swap order of ref & dataset

@@
 ### Reference & dataset sizes
+```{r fit_db_dataset_size}
+dataset_sizes <-
+  read_tsv(here('subworkflows/1_prep_samples/results/dataset_sizes.tsv'))
+kable(dataset_sizes)
+```
+
 ```{r fit_db_ref_size}
@@
 kable(ref_sizes)
-```
-
-```{r fit_db_dataset_size}
-dataset_sizes <-
-  read_tsv(here('subworkflows/1_prep_samples/results/dataset_sizes.tsv'))
-kable(dataset_sizes)
 ```

### 2020-06-10 exploratory/2020/README.Rmd
Use manual colors

@@
   geom_boxplot() +
-  scale_color_brewer(palette = "Dark2") +
-  #scale_color_manual(values = dataset_colors) +  # bug in geom_boxplot with manual colors?
+  scale_color_manual(values = dataset_colors) +  
   facet_wrap("ref") +

### 2020-06-20 exploratory/2020/README.Rmd
Update sensspec & benchmark boxplots with cowplot

@@
 ```{r lib}
+library(cowplot)
 library(here)
@@
 }
+plot_box_mcc <- function(df) {
+  df %>%
+    group_by(dataset, method) %>%
+    ggplot(aes(x = method, y = mcc, color = dataset)) +
+    geom_boxplot(alpha = 0.5) +
+    scale_color_manual(values = dataset_colors) +
+    ylim(0.6, 1) +
+    facet_grid(dataset ~ ref) +
+    theme(legend.position = "None",
+          axis.title.x = element_blank())
+}
+plot_box_time <- function(df) {
+  df %>%
+    group_by(dataset, method) %>%
+    ggplot(aes(x = method, y = s, color = dataset)) +
+    geom_boxplot() +
+    scale_color_manual(values = dataset_colors) +
+    ylim(0, 1000) +
+    facet_wrap("ref") +
+    labs(y = 'seconds') +
+    theme(axis.title.x = element_blank())
+}
 ```
@@
 ```{r fit_db_sensspec}
-sensspec <-
-  read_tsv(here('subworkflows/2_fit_reference_db/results/sensspec.tsv')) %>%
-  relevel_method()
-sensspec %>%
-  group_by(dataset, method) %>%
-  ggplot(aes(x = method, y = mcc, color = dataset)) +
-  geom_boxplot(alpha = 0.5) +
-  scale_color_manual(values = dataset_colors) +
-  #ylim(0, 1) +
-  facet_grid(dataset ~ ref)
+sensspec_fit <-
+  read_tsv(here('subworkflows/2_fit_reference_db/results/sensspec.tsv')) 
+sensspec_clust <- read_tsv(here('subworkflows/1_prep_samples/results/sensspec.tsv'))
+#sensspec <- bind_rows(sensspec_fit, sensspec_clust)
+
+mcc_plot_fit <- sensspec_fit %>% 
+  plot_box_mcc() +
+  theme(axis.text.y = element_blank(),
+        axis.title.y = element_blank())
+mcc_plot_clust <- sensspec_clust %>% 
+  plot_box_mcc() + 
+  theme(strip.background.y = element_blank(),
+        strip.text.y = element_blank())
+
+plot_grid(mcc_plot_clust, mcc_plot_fit, 
+          align = "h", rel_widths = c(1,3))
 ```
@@
 ```{r fit_db_benchmarks, warning=FALSE}
-benchmarks <-
-  read_tsv(here('subworkflows/2_fit_reference_db/results/benchmarks.tsv')) %>%
-  relevel_method
-benchmarks %>%
-  group_by(dataset, method) %>%
-  ggplot(aes(x = method, y = s, color = dataset)) +
-  geom_boxplot() +
-  scale_color_manual(values = dataset_colors) +  
-  facet_wrap("ref") +
-  labs(y = 'seconds')
+benchmarks_fit <-
+  read_tsv(here('subworkflows/2_fit_reference_db/results/benchmarks.tsv'))
+benchmarks_clust <- read_tsv(here('subworkflows/1_prep_samples/results/benchmarks.tsv'))
+
+time_plot_fit <- benchmarks_fit %>% 
+  plot_box_time() +
+  theme(axis.text.y = element_blank(),
+        axis.title.y = element_blank())
+time_plot_clust <- benchmarks_clust %>% 
+  plot_box_time() +
+  theme(legend.position = "None")
+
+plot_grid(time_plot_clust, time_plot_fit,
+          align = "h", rel_widths = c(1,3))
 ```

### 2020-06-23 exploratory/2020/README.Rmd
Fix ylims

@@
     scale_color_manual(values = dataset_colors) +
-    ylim(0.6, 1) +
+    ylim(0.5, 1) +
     facet_grid(dataset ~ ref) +
@@
     scale_color_manual(values = dataset_colors) +
-    ylim(0, 1000) +
     facet_wrap("ref") +
+    ylim(0,2100) +
     labs(y = 'seconds') +

### 2020-06-30 exploratory/2020/README.Rmd
Add hline to mcc plot for de novo

@@
 library(cowplot)
+library(ggtext)
 library(here)
@@
     ylim(0.5, 1) +
+    #facet_wrap("ref") +
     facet_grid(dataset ~ ref) +
@@
   read_tsv(here('subworkflows/2_fit_reference_db/results/sensspec.tsv')) 
-sensspec_clust <- read_tsv(here('subworkflows/1_prep_samples/results/sensspec.tsv'))
+sensspec_clust <- read_tsv(here('subworkflows/1_prep_samples/results/sensspec.tsv')) %>% 
+  select(-ref) %>% 
+  group_by(dataset) %>% 
+  summarize(mean_mcc = mean(mcc))
 #sensspec <- bind_rows(sensspec_fit, sensspec_clust)
 
-mcc_plot_fit <- sensspec_fit %>% 
+sensspec_fit %>% 
   plot_box_mcc() +
-  theme(axis.text.y = element_blank(),
-        axis.title.y = element_blank())
-mcc_plot_clust <- sensspec_clust %>% 
-  plot_box_mcc() + 
-  theme(strip.background.y = element_blank(),
-        strip.text.y = element_blank())
-
-plot_grid(mcc_plot_clust, mcc_plot_fit, 
-          align = "h", rel_widths = c(1,3))
+  geom_hline(aes(yintercept = mean_mcc), sensspec_clust) +
+  labs(caption = "The black line represents the *de novo* clustering mcc sccore.") +
+  theme(plot.caption = element_markdown())
 ```

### 2020-06-30 exploratory/2020/README.Rmd
Put boxplot behind hline for mcc plot

@@
 }
-plot_box_mcc <- function(df) {
-  df %>%
-    group_by(dataset, method) %>%
-    ggplot(aes(x = method, y = mcc, color = dataset)) +
-    geom_boxplot(alpha = 0.5) +
-    scale_color_manual(values = dataset_colors) +
-    ylim(0.5, 1) +
-    #facet_wrap("ref") +
-    facet_grid(dataset ~ ref) +
-    theme(legend.position = "None",
-          axis.title.x = element_blank())
-}
+
 plot_box_time <- function(df) {
@@
 #sensspec <- bind_rows(sensspec_fit, sensspec_clust)
-
 sensspec_fit %>%
-  plot_box_mcc() +
+  group_by(dataset, method) %>%
+  ggplot(aes(x = method, y = mcc, color = dataset)) +
   geom_hline(aes(yintercept = mean_mcc), sensspec_clust) +
+  geom_boxplot(alpha = 0.5) +
+  scale_color_manual(values = dataset_colors) +
+  ylim(0.5, 1) +
+  facet_grid(dataset ~ ref) +
   labs(caption = "The black line represents the *de novo* clustering mcc sccore.") +
-  theme(plot.caption = element_markdown())
+  theme(legend.position = "None",
+        axis.title.x = element_blank(),
+        plot.caption = element_markdown())
 ```

### 2020-06-30 exploratory/2020/README.Rmd
Fix typo

@@
   facet_grid(dataset ~ ref) +
-  labs(caption = "The black line represents the *de novo* clustering mcc sccore.") +
+  labs(caption = "The black line represents the *de novo* clustering mcc score.") +
   theme(legend.position = "None",

### 2020-08-24 exploratory/2020/sub3_fit_subsets.Rmd
Setup notebook for subworkflow 3

---
title: "Exploratory Plots"
date: "Aug. 2020"
output: 
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      fig.path = "figures/"
                      )
```
```{r lib}
library(cowplot)
library(ggtext)
library(here)
library(knitr)
library(tidyverse)
theme_set(theme_classic())
color_palette <- RColorBrewer::brewer.pal(4, "Dark2")
dataset_colors <- c(
  human = color_palette[[3]],
  marine = color_palette[[1]],
  mouse = color_palette[[4]],
  soil = color_palette[[2]]
)
```
```{r}
sensspec_fit <-
  read_tsv(here('subworkflows/3_fit_sample_subset/results/sensspec.tsv'))
```

## Subsetting datasets with different reference sizes

```{r fit_ref_frac}
sensspec_fit %>%
  group_by(dataset, method) %>%
  ggplot(aes(x = ref_frac, y = mcc, color = dataset)) +
  geom_point(alpha = 0.5) +
  scale_color_manual(values = dataset_colors) +
  ylim(0.5, 1) +
  facet_grid(method ~ ref_weight)
```

sample fraction = 0.2

## Runtime
```{r}
benchmarks_fit <-
  read_tsv(here('subworkflows/2_fit_reference_db/results/benchmarks.tsv'))
benchmarks_fit %>% 
    group_by(dataset, method) %>%
    ggplot(aes(x = method, y = s, color = dataset)) +
    geom_boxplot() +
    scale_color_manual(values = dataset_colors) +
    #facet_wrap("ref_weight") +
    ylim(0,2100) +
    labs(y = 'seconds') +
    theme(axis.title.x = element_blank())
```

## Fraction of sequences that map to the reference
```{r fraction_reads_mapped}
fractions <- read_tsv(here('subworkflows/3_fit_sample_subset/results/fraction_reads_mapped.tsv'))
fractions %>% 
  ggplot(aes(x=dataset, y=fraction_mapped, color=dataset)) +
  geom_boxplot(alpha = 0.5) +
  scale_color_manual(values = dataset_colors) +
  facet_wrap("ref_weight") +
  ylim(0, 1) +
  labs(title="Sequences mapped during closed-reference OptiFit")
```

### 2020-08-24 subworkflows/3_fit_sample_subset/code/debug.sh
More seeds & ref fractions

@@ -1,0 +1,38 @@
+#!/bin/bash
+
+##################
+####  Slurm preamble
+
+#### #### ####  These are the most frequently changing options
+
+####  Job name
+#SBATCH --job-name=sub3
+
+####  Request resources here
+####    These are typically, number of processors, amount of memory,
+####    an the amount of time a job requires.  May include processor
+####    type, too.
+
+#SBATCH --nodes=1
+#SBATCH --ntasks-per-node=1
+#SBATCH --cpus-per-task=1
+#SBATCH --mem-per-cpu=100MB
+#SBATCH --time=48:00:00
+
+#SBATCH --output=log/hpc/slurm-%j_%x.out
+
+####  Slurm account and partition specification here
+####    These will change if you work on multiple projects, or need
+####    special hardware, like large memory nodes or GPUs.
+
+#SBATCH --account=pschloss99
+#SBATCH --partition=standard
+
+#### #### ####  These are the least frequently changing options
+
+####  Your e-mail address and when you want e-mail
+
+#SBATCH --mail-user=sovacool@umich.edu
+#SBATCH --mail-type=BEGIN,END
+
+time snakemake --profile config/slurm --latency-wait 90 --forcerun merge_results3

### 2020-09-04 exploratory/2020/sub3_fit_subsets.Rmd
Update plots with results from 1 seed & fixed bug in fraction mapped code

@@
 fractions %>% 
-  ggplot(aes(x=dataset, y=fraction_mapped, color=dataset)) +
+  ggplot(aes(x=ref_weight, y=fraction_mapped, color=dataset)) +
   geom_boxplot(alpha = 0.5) +
   scale_color_manual(values = dataset_colors) +
-  facet_wrap("ref_weight") +
+  facet_wrap("dataset", nrow = 1) + 
   ylim(0, 1) +

### 2020-09-04 exploratory/2020/sub3_fit_subsets.Rmd
Update plots

--- 2020-08-24 exploratory/2020/sub3_fit_subsets.Rmd
@@
   read_tsv(here('subworkflows/2_fit_reference_db/results/benchmarks.tsv'))
+# TODO: show number of unique seqs 
 benchmarks_fit %>% 
@@
     scale_color_manual(values = dataset_colors) +
-    #facet_wrap("ref_weight") +
+    #facet_grid(ref_weight ~ .) +
     ylim(0,2100) +
@@
 fractions %>% 
-  ggplot(aes(x=dataset, y=fraction_mapped, color=dataset)) +
-  geom_boxplot(alpha = 0.5) +
+  ggplot(aes(x=ref_frac, y=fraction_mapped, color=dataset)) +
+  geom_point(alpha = 0.5) +
   scale_color_manual(values = dataset_colors) +
-  facet_wrap("ref_weight") +
+  facet_grid(ref_weight ~ dataset) +
   ylim(0, 1) +

### 2020-09-04 exploratory/2020/sub3_fit_subsets.Rmd
Update fraction plot

--- 2020-08-24 exploratory/2020/sub3_fit_subsets.Rmd
@@
   read_tsv(here('subworkflows/2_fit_reference_db/results/benchmarks.tsv'))
+# TODO: show number of unique seqs, use as x axis?
 benchmarks_fit %>% 
@@
 fractions %>% 
-  ggplot(aes(x=dataset, y=fraction_mapped, color=dataset)) +
-  geom_boxplot(alpha = 0.5) +
+  group_by(dataset, ref_weight, ref_frac) %>% 
+  ggplot(aes(x=ref_frac, y=fraction_mapped, color=dataset)) +
+  geom_point(alpha = 0.5) +
   scale_color_manual(values = dataset_colors) +

### 2020-09-09 code/R/render.R
Set up paper & code to render it

@@ -1,0 +1,4 @@
+source(snakemake@input[["fcns"]])
+
+log_smk()
+rmarkdown::render(here::here(snakemake@input[["Rmd"]]))

### 2020-09-09 paper/paper.tex
Set up paper & code to render it

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={OptiFit: a fast method for fitting amplicon sequences to existing OTUs},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1.0in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}

\usepackage[none]{hyphenat}

\usepackage{setspace}
\doublespacing
\setlength{\parskip}{1em}

\usepackage{lineno}

\usepackage{pdfpages}
\usepackage{comment}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {}%
  {\par}

\title{\textbf{OptiFit: a fast method for fitting amplicon sequences to
existing OTUs}}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle

\begin{center}
\vspace{25mm}

Kelly L. Sovacool${^1}$, Sarah L. Westcott${^2}$, M. Brodie Mumphrey${^1}$,
Gabrielle A. Dotson${^1}$, Patrick D. Schloss${^2\dagger}$

$1$ Department of Computational Medicine and Bioinformatics,
University of Michigan, Ann Arbor, MI 48109

$2$ Department of Microbiology and Immunology,
University of Michigan, Ann Arbor, MI 48109

\vspace{30mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

\end{center}

\newpage
\linenumbers

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

Assigning amplicon sequences to Operational Taxonomic Units (OTUs) is an
important step in characterizing the composition of microbial
communities across large datasets. OptiClust, a \emph{de novo} OTU
clustering method in the mothur program, has been shown to produce
higher quality OTU assignments than other methods and at comparable or
faster speeds (1, 2). A notable difference between \emph{de novo}
clustering and database-dependent methods is that OTU assignments
clustered with \emph{de novo} methods are not stable when new sequences
are added to a dataset (3). However, in some cases one may wish to
incorporate new samples into a previously clustered dataset without
performing clustering again on all sequences, such as when deploying a
machine learning model where OTUs are features (4). To provide an
efficient and robust method to fit amplicon sequence data to existing
OTUs, we developed the OptiFit algorithm as a new component of the
mothur program. To benchmark the OptiFit algorithm against \emph{de
novo} clustering with the OptiClust algorithm, we used four published
datasets isolated from soil (5), marine (6), mouse (7), and human (8)
samples. For each dataset, a subset of sequences was randomly selected
and clustered into OTUs with OptiClust, then the remaining sequences
were fit to the existing OTUs using the OptiFit algorithm. This was
repeated with subsets of varying sizes ranging from 10 to 90\% of
sequences in order to evaluate the bounds of the dataset size required
for OptiFit. Separately, all sequences were clustered with OptiClust to
provide a baseline of OTU assignment quality and runtime performance.
Each of these routines was repeated 10 times with different random seeds
to produce results that are robust to random variation. OTU quality was
evaluated using the Matthews Correlation Coefficient (MCC) with a
sequence similarity threshold of 97\% as described previously (3, 9). On
average, fitting sequences into existing OTUs with OptiFit performed 10
times faster than \emph{de novo} clustering with OptiClust, while the
average MCC scores produced were nearly indistinguishable across each
dataset. The OptiFit results across subset sizes ranging from 10 to 90\%
of sequences were also very similar, with slightly higher MCC scores for
larger subset sizes. Thus, OptiFit is an efficient way to fit new
sequences to existing OTUs yet without sacrificing the quality of OTU
assignments.

\hypertarget{importance}{%
\subsubsection{Importance}\label{importance}}

\newpage

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

\hypertarget{results}{%
\subsection{Results}\label{results}}

\hypertarget{discussion}{%
\subsection{Discussion}\label{discussion}}

\hypertarget{materials-and-methods}{%
\subsection{Materials and Methods}\label{materials-and-methods}}

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

KLS received support from the NIH Training Program in Bioinformatics
(T32 GM070449).

The funders had no role in study design, data collection and
interpretation, or the decision to submit the work for publication.

\hypertarget{references}{%
\subsection*{References}\label{references}}
\addcontentsline{toc}{subsection}{References}

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-westcott_opticlust_2017}{}%
1. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
Improved Method for Assigning Amplicon-Based Sequence Data to
Operational Taxonomic Units. mSphere \textbf{2}:e00073--17.
doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.

\leavevmode\hypertarget{ref-schloss_introducing_2009}{}%
2. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Van Horn DJ}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-source, platform-independent,
community-supported software for describing and comparing microbial
communities. Applied and Environmental Microbiology
\textbf{75}:7537--7541.
doi:\href{https://doi.org/10.1128/AEM.01541-09}{10.1128/AEM.01541-09}.

\leavevmode\hypertarget{ref-westcott_novo_2015}{}%
3. \textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo clustering
methods outperform reference-based methods for assigning 16S rRNA gene
sequences to operational taxonomic units. PeerJ \textbf{3}:e1487.
doi:\href{https://doi.org/10.7717/peerj.1487}{10.7717/peerj.1487}.

\leavevmode\hypertarget{ref-topcuoglu_framework_2020}{}%
4. \textbf{Topçuoğlu BD}, \textbf{Lesniak NA}, \textbf{Ruffin MT},
\textbf{Wiens J}, \textbf{Schloss PD}. 2020. A Framework for Effective
Application of Machine Learning to Microbiome-Based Classification
Problems. mBio \textbf{11}.
doi:\href{https://doi.org/10.1128/mBio.00434-20}{10.1128/mBio.00434-20}.

\leavevmode\hypertarget{ref-johnston_metagenomics_2016}{}%
5. \textbf{Johnston ER}, \textbf{Rodriguez-R LM}, \textbf{Luo C},
\textbf{Yuan MM}, \textbf{Wu L}, \textbf{He Z}, \textbf{Schuur EAG},
\textbf{Luo Y}, \textbf{Tiedje JM}, \textbf{Zhou J},
\textbf{Konstantinidis KT}. 2016. Metagenomics Reveals Pervasive
Bacterial Populations and Reduced Community Diversity across the Alaska
Tundra Ecosystem. Front Microbiol \textbf{7}.
doi:\href{https://doi.org/10.3389/fmicb.2016.00579}{10.3389/fmicb.2016.00579}.

\leavevmode\hypertarget{ref-henson_artificial_2016}{}%
6. \textbf{Henson MW}, \textbf{Pitre DM}, \textbf{Weckhorst JL},
\textbf{Lanclos VC}, \textbf{Webber AT}, \textbf{Thrash JC}. 2016.
Artificial Seawater Media Facilitate Cultivating Members of the
Microbial Majority from the Gulf of Mexico. mSphere \textbf{1}.
doi:\href{https://doi.org/10.1128/mSphere.00028-16}{10.1128/mSphere.00028-16}.

\leavevmode\hypertarget{ref-schloss_stabilization_2012}{}%
7. \textbf{Schloss PD}, \textbf{Schubert AM}, \textbf{Zackular JP},
\textbf{Iverson KD}, \textbf{Young VB}, \textbf{Petrosino JF}. 2012.
Stabilization of the murine gut microbiome following weaning. Gut
Microbes \textbf{3}:383--393.
doi:\href{https://doi.org/10.4161/gmic.21008}{10.4161/gmic.21008}.

\leavevmode\hypertarget{ref-baxter_microbiota-based_2016}{}%
8. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
\textbf{Schloss PD}. 2016. Microbiota-based model improves the
sensitivity of fecal immunochemical test for detecting colonic lesions.
Genome Med \textbf{8}:37.
doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.

\leavevmode\hypertarget{ref-schloss_application_2016}{}%
9. \textbf{Schloss PD}. 2016. Application of a Database-Independent
Approach To Assess the Quality of Operational Taxonomic Unit Picking
Methods. mSystems \textbf{1}:e00027--16.
doi:\href{https://doi.org/10.1128/mSystems.00027-16}{10.1128/mSystems.00027-16}.
\end{cslreferences}

\end{document}

### 2020-09-10 subworkflows/3_fit_sample_subset/code/debug.sh
Update slurm scripts

@@ -35,4 +35,4 @@
 #SBATCH --mail-user=sovacool@umich.edu
 #SBATCH --mail-type=BEGIN,END
 
-time snakemake --profile ../../config/slurm --latency-wait 90 --configfile ../../config/config.yaml --forcerun merge_all_inputs
+time snakemake --profile ../../config/slurm --latency-wait 90 --configfile ../../config/config.yaml --forcerun merge_benchmarks

### 2020-09-10 subworkflows/3_fit_sample_subset/code/submit_sub3.sh

@@ -16,7 +16,7 @@
 #SBATCH --nodes=1
 #SBATCH --ntasks-per-node=1
 #SBATCH --cpus-per-task=1
-#SBATCH --mem-per-cpu=500MB
+#SBATCH --mem-per-cpu=10GB
 #SBATCH --time=48:00:00
 
 #SBATCH --output=log/hpc/slurm-%j_%x.out
@@ -35,4 +35,4 @@
 #SBATCH --mail-user=sovacool@umich.edu
 #SBATCH --mail-type=BEGIN,END
 
-time snakemake --profile ../../config/slurm --latency-wait 90
+time snakemake --profile ../../config/slurm --latency-wait 90 --forcerun merge_sensspec

### 2020-09-10 exploratory/2020/sub3_fit_subsets.Rmd
Use boxplot & jitter geoms

--- 2020-09-04 exploratory/2020/sub3_fit_subsets.Rmd
@@
 
-```{r fit_ref_frac}
+```{r fit_ref_frac_open}
 sensspec_fit %>%
-  group_by(dataset, method) %>%
-  ggplot(aes(x = ref_frac, y = mcc, color = dataset)) +
-  geom_point(alpha = 0.5) +
+  filter(method == 'open') %>% 
+  group_by(dataset, ref_weight) %>%
+  ggplot(aes(x = as.factor(ref_frac), y = mcc, color = dataset)) +
+  geom_boxplot() +
   scale_color_manual(values = dataset_colors) +
   ylim(0.5, 1) +
-  facet_grid(method ~ ref_weight)
+  facet_grid(dataset ~ ref_weight) +
+  labs(title='Open-reference OptiFit Performance',
+       x='reference fraction')
+```
+
+```{r fit_ref_frac_closed}
+sensspec_fit %>%
+  filter(method == 'closed') %>% 
+  group_by(dataset, ref_weight) %>%
+  ggplot(aes(x = as.factor(ref_frac), y = mcc, color = dataset)) +
+  geom_boxplot() +
+  scale_color_manual(values = dataset_colors) +
+  ylim(0.5, 1) +
+  facet_grid(dataset ~ ref_weight) +
+  labs(title='Closed-reference OptiFit Performance',
+       x='reference fraction')
 ```
@@
 ## Runtime
-```{r}
-benchmarks_fit <-
-  read_tsv(here('subworkflows/2_fit_reference_db/results/benchmarks.tsv'))
-benchmarks_fit %>% 
-    group_by(dataset, method) %>%
-    ggplot(aes(x = method, y = s, color = dataset)) +
-    geom_boxplot() +
+```{r runtime_ref_size}
+runtime <- full_join(
+  read_tsv(here('subworkflows/3_fit_sample_subset/results/benchmarks.tsv')),
+  read_tsv(here('subworkflows/3_fit_sample_subset/results/input_sizes.tsv'))
+  )
+runtime %>% ggplot(aes(x=num_ref_seqs, y=s, color=dataset)) +
+  geom_point(alpha = 0.5) +
   scale_color_manual(values = dataset_colors) +
-    #facet_wrap("ref_weight") +
-    ylim(0,2100) +
-    labs(y = 'seconds') +
-    theme(axis.title.x = element_blank())
+  labs(title = 'Runtime over reference size',
+       x = '# sequences in reference',
+       y = 'seconds')
+```
+
+```{r runtime_sample_size}
+runtime %>% ggplot(aes(x=num_sample_seqs, y=s, color=dataset)) +
+  geom_point(alpha = 0.5) +
+  scale_color_manual(values = dataset_colors) +
+  labs(title = 'Runtime over sample size',
+       x = '# sequences in sample',
+       y = 'seconds')
 ```
@@
 fractions %>% 
-  ggplot(aes(x=ref_weight, y=fraction_mapped, color=dataset)) +
-  geom_boxplot(alpha = 0.5) +
+  group_by(dataset, ref_weight, ref_frac) %>% 
+  ggplot(aes(x=ref_frac, y=fraction_mapped, color=dataset)) +
+  geom_jitter(alpha = 0.5, width = 0.01, size=1) +
   scale_color_manual(values = dataset_colors) +
-  facet_wrap("dataset", nrow = 1) + 
+  facet_wrap("ref_weight") +
   ylim(0, 1) +
-  labs(title="Sequences mapped during closed-reference OptiFit")
+  labs(title="Sequences mapped during closed-reference OptiFit",
+       x='reference fraction',
+       y='fraction mapped')
 ```

### 2020-09-10 exploratory/2020/sub3_fit_subsets.Rmd
Update plots with 10 seeds & 8 ref fracs

@@
 runtime %>% ggplot(aes(x=num_ref_seqs, y=s, color=dataset)) +
-  geom_point(alpha = 0.5) +
+  geom_point(alpha = 0.3) +
   scale_color_manual(values = dataset_colors) +
@@
 
-```{r runtime_sample_size}
-runtime %>% ggplot(aes(x=num_sample_seqs, y=s, color=dataset)) +
+```{r runtime_total_size}
+runtime %>% 
+  mutate(num_total_seqs = num_ref_seqs + num_sample_seqs) %>% 
+  ggplot(aes(x=num_total_seqs, y=s, color=dataset)) +
   geom_point(alpha = 0.5) +
   scale_color_manual(values = dataset_colors) +
-  labs(title = 'Runtime over sample size',
-       x = '# sequences in sample',
+  xlim(0, 300000) +
+  labs(title = 'Runtime over sample + reference size',
+       x = '# sequences total',
        y = 'seconds')
@@
   group_by(dataset, ref_weight, ref_frac) %>% 
-  ggplot(aes(x=ref_frac, y=fraction_mapped, color=dataset)) +
+  ggplot(aes(x=ref_frac, y=fraction_mapped, color=dataset, shape=ref_weight)) +
   geom_jitter(alpha = 0.5, width = 0.01, size=1) +

### 2020-09-11 exploratory/2020/sub3_fit_subsets.Rmd
Remove shapes

@@
   group_by(dataset, ref_weight, ref_frac) %>% 
-  ggplot(aes(x=ref_frac, y=fraction_mapped, color=dataset, shape=ref_weight)) +
+  ggplot(aes(x=ref_frac, y=fraction_mapped, color=dataset)) +
   geom_jitter(alpha = 0.5, width = 0.01, size=1) +

### 2020-09-11 code/py/count_gaps.py
Count total chars too

@@ -3,16 +3,16 @@
 
 def main():
     with open(snakemake.input.fasta, "r") as infile:
-        n_gaps = count_gaps(infile)
+        n_gaps, total_chars = count_gaps(infile)
     with open(snakemake.output.txt, "w") as outfile:
-        outfile.write("dataset\tn_gaps\n")
-        outfile.write(f"{snakemake.wildcards.dataset}\t{n_gaps}\n")
+        outfile.write("dataset\tn_gaps\ttotal_chars\n")
+        outfile.write(f"{snakemake.wildcards.dataset}\t{n_gaps}\t{total_chars}\n")
 
-
-def count_gaps(input_filename, gap_char="-", header_char=">"):
-    return sum(
-        line.count(gap_char) for line in infile if not line.startswith(header_char)
-    )
+def count_gaps(infile, gap_char="-", header_char=">"):
+    seqs_cat = ''.join(line.strip() for line in infile if not line.startswith(header_char))
+    n_gaps = seqs_cat.count(gap_char)
+    total_chars = len(seqs_cat)
+    return n_gaps, seqs_cat
 
 
 main()

### 2020-09-13 exploratory/2020/sub3_fit_subsets.Rmd
Write rule to merge gap counts
More runtime plots

--- 2020-09-10 exploratory/2020/sub3_fit_subsets.Rmd
@@
 title: "Exploratory Plots"
-date: "Aug. 2020"
+date: "Sept. 2020"
 output: 
@@
 ## Runtime
-```{r runtime_ref_size}
+```{r runtime_ref_seqs}
 runtime <- full_join(
@@
   read_tsv(here('subworkflows/3_fit_sample_subset/results/input_sizes.tsv'))
-  )
+  ) %>% 
+  full_join(read_tsv(here('subworkflows/3_fit_sample_subset/results/gap_counts.tsv'))) %>% 
+  mutate(num_total_seqs = num_ref_seqs + num_sample_seqs,
+         gaps_frac = n_gaps / total_chars)
 runtime %>% ggplot(aes(x=num_ref_seqs, y=s, color=dataset)) +
-  geom_point(alpha = 0.5) +
+  geom_point(alpha = 0.3) +
+  facet_grid(method ~ ref_weight) +
   scale_color_manual(values = dataset_colors) +
@@
 
-```{r runtime_sample_size}
-runtime %>% ggplot(aes(x=num_sample_seqs, y=s, color=dataset)) +
+```{r runtime_total_seqs}
+runtime %>% 
+  ggplot(aes(x=num_total_seqs, y=s, color=dataset)) +
   geom_point(alpha = 0.5) +
+  facet_grid(method ~ ref_weight) +
   scale_color_manual(values = dataset_colors) +
-  labs(title = 'Runtime over sample size',
-       x = '# sequences in sample',
+  xlim(0, 300000) +
+  labs(title = 'Runtime over sample + reference size',
+       x = '# sequences total',
+       y = 'seconds')
+```
+
+```{r runtime_gaps}
+runtime %>% 
+  ggplot(aes(x=n_gaps, y=s, color=dataset)) +
+  geom_point(alpha = 0.5) +
+  facet_grid(method ~ ref_weight) +
+  scale_color_manual(values = dataset_colors) +
+  labs(title = 'Runtime over number of gaps',
+       x = '# gaps',
+       y = 'seconds')
+```
+
+```{r runtime_numotus}
+runtime %>% full_join(sensspec_fit) %>% 
+  ggplot(aes(x=numotus, y=s, color=dataset, shape = ref_weight)) +
+  geom_point(alpha = 0.5) +
+  facet_grid(method ~ ref_weight) +
+  scale_color_manual(values = dataset_colors) +
+  labs(title = 'Runtime over number of OTUs',
+       x = '# OTUs',
        y = 'seconds')

### 2020-09-13 exploratory/2020/sub3_fit_subsets.Rmd
Don't need quite so many runtime plots

@@
 
-```{r runtime_total_seqs}
-runtime %>% 
-  ggplot(aes(x=num_total_seqs, y=s, color=dataset)) +
-  geom_point(alpha = 0.5) +
-  facet_grid(method ~ ref_weight) +
-  scale_color_manual(values = dataset_colors) +
-  xlim(0, 300000) +
-  labs(title = 'Runtime over sample + reference size',
-       x = '# sequences total',
-       y = 'seconds')
-```
-
-```{r runtime_gaps}
-runtime %>% 
-  ggplot(aes(x=n_gaps, y=s, color=dataset)) +
-  geom_point(alpha = 0.5) +
-  facet_grid(method ~ ref_weight) +
-  scale_color_manual(values = dataset_colors) +
-  labs(title = 'Runtime over number of gaps',
-       x = '# gaps',
-       y = 'seconds')
-```
-
 ```{r runtime_numotus}

### 2020-09-14 exploratory/2020/sub3_fit_subsets.Rmd
Show open & closed performance on one plot

@@
 
-```{r fit_ref_frac_open}
+```{r fit_ref_frac}
 sensspec_fit %>%
-  filter(method == 'open') %>% 
-  group_by(dataset, ref_weight) %>%
   ggplot(aes(x = as.factor(ref_frac), y = mcc, color = dataset)) +
@@
   ylim(0.5, 1) +
-  facet_grid(dataset ~ ref_weight) +
-  labs(title='Open-reference OptiFit Performance',
-       x='reference fraction')
-```
-
-```{r fit_ref_frac_closed}
-sensspec_fit %>%
-  filter(method == 'closed') %>% 
-  group_by(dataset, ref_weight) %>%
-  ggplot(aes(x = as.factor(ref_frac), y = mcc, color = dataset)) +
-  geom_boxplot() +
-  scale_color_manual(values = dataset_colors) +
-  ylim(0.5, 1) +
-  facet_grid(dataset ~ ref_weight) +
-  labs(title='Closed-reference OptiFit Performance',
+  facet_grid(method ~ ref_weight) +
+  labs(title='OptiFit Performance',
        x='reference fraction')

### 2020-09-14 exploratory/2020/sub3_fit_subsets.Rmd
Clean up x axis labels

@@
   scale_color_manual(values = dataset_colors) +
+  scale_x_continuous(breaks = seq(0, 200000, 50000),
+                     labels = c('0', '50k', '100k', '150k', '200k')) +
   labs(title = 'Runtime over reference size',
@@
   scale_color_manual(values = dataset_colors) +
+  scale_x_continuous(breaks = seq(0, 40000, 10000),
+                     labels = c('0', '10k', '20k', '30k', '40k')) +
   labs(title = 'Runtime over number of OTUs',

### 2020-09-14 exploratory/2020/sub3_fit_subsets.Rmd
Use stat_summary

@@
 sensspec_fit %>%
-  ggplot(aes(x = as.factor(ref_frac), y = mcc, color = dataset)) +
-  geom_boxplot() +
+  ggplot(aes(x = ref_frac, y = mcc, color = dataset)) +
+  geom_jitter(size = 1, alpha = 0.3, width = 0.01) +
+  stat_summary(fun = mean, geom = 'crossbar') +
   scale_color_manual(values = dataset_colors) +
-  ylim(0.5, 1) +
   facet_grid(method ~ ref_weight) +
@@
 ## Runtime
+
 ```{r runtime_ref_seqs}
@@
          gaps_frac = n_gaps / total_chars)
+
 runtime %>% ggplot(aes(x=num_ref_seqs, y=s, color=dataset)) +

### 2020-09-14 exploratory/2020/sub3_fit_subsets.Rmd
MOAR wordz

@@
 
+Sample fraction of 0.2 with 20 different seeds.
+
 ```{r fit_ref_frac}
@@
 
-sample fraction = 0.2
+The performance is very consistent across reference fractions when weighting sequences by abundance or not at all (simple).
+When weighting by distance, the performance decreases as the reference fraction increases.
+This is expected because a larger fraction means **more** sequences with **fewer** 
+pairwise distances under the 0.03 threshold are included in the reference.
 
@@
 
+Runtime is also fairly consistent regardless of weighting method.
+For abundance and simple weighting, runtime seems to continue increasing roughly linearly (maybe a little worse than linear).
+When weighting by distance, runtime reaches a plateau sooner than other selection methods as the size of the reference increases.
+Perhaps the sequences with the most similarities are the most influential
+over OTU clustering, 
+so after you reach some critical number of similarities, the runtime doesn't increase as much.
+I'm not sure why the human dataset takes so much longer than the others.
+
 ```{r runtime_numotus}
@@
 ```
+
+Open-reference clustering has more OTUs because it will cluster any sequences 
+_de novo_ that do not fit into existing OTUs, 
+while closed-reference clustering throws them out.
+This plot is confounded by the number of sequences in the reference, 
+perhaps I should re-plot it as the ratio of OTUs to reference sequences?
 
@@
 ```
+
+The fraction of sequences that are able to be fit to the reference in closed-reference
+clustering increases as the reference size increases when weighting sequences 
+by abundance or not at all for reference selection.
+It's also remarkably stable between abundance & simple weighting, and 
+between all seeds.
+Weighting by distance performs poorly, likely because the top n sequences with 
+the most pairwise similarities are included in the reference, 
+leaving sequences with few similarities to then be fit to the highly-connected reference.

### 2020-10-01 exploratory/2020/sub4_fitpercent.Rmd
Use built-in fitpercent parameter

---
title: "Exploratory Plots"
date: "Oct. 2020"
output: 
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      fig.path = "figures/"
                      )
```
```{r lib}
library(cowplot)
library(ggtext)
library(glue)
library(here)
library(knitr)
library(tidyverse)
theme_set(theme_classic())
color_palette <- RColorBrewer::brewer.pal(4, "Dark2")
dataset_colors <- c(
  human = color_palette[[3]],
  marine = color_palette[[1]],
  mouse = color_palette[[4]],
  soil = color_palette[[2]]
)
```

## OptiFit performance with varied reference sizes

sample frac = 1 - reference frac

```{r read_join}
dirpath <- 'subworkflows/3_fit_sample_subset/results/all-seqs/'
results <- list.files(here(dirpath)) %>% 
  (function(x) { here(glue('{dirpath}{x}')) }) %>% 
  map(read_tsv) %>% 
  reduce(full_join) %>% 
  mutate(num_total_seqs = num_ref_seqs + num_sample_seqs,
         gaps_frac = n_gaps / total_chars)
head(results)
dim(results)
```

```{r sensspec}
results %>% 
  ggplot(aes(x = ref_frac, y = mcc, color = dataset)) +
  geom_jitter(size = 1, alpha = 0.3, width = 0.01) +
  stat_summary(fun = mean, geom = 'crossbar') +
  scale_color_manual(values = dataset_colors) +
  facet_grid(method ~ ref_weight) +
  labs(title='OptiFit Performance',
       x='reference fraction')
```

```{r runtime}
results %>% ggplot(aes(x=num_ref_seqs, y=s, color=dataset)) +
  geom_point(alpha = 0.3) +
  facet_grid(method ~ ref_weight) +
  scale_color_manual(values = dataset_colors) +
  scale_x_continuous(breaks = seq(0, 200000, 50000),
                     labels = c('0', '50k', '100k', '150k', '200k')) +
  labs(title = 'Runtime over reference size',
       x = '# sequences in reference',
       y = 'seconds')
results %>% 
  ggplot(aes(x=numotus, y=s, color=dataset)) +
  geom_point(alpha = 0.5) +
  facet_grid(method ~ ref_weight) +
  scale_color_manual(values = dataset_colors) +
  #scale_x_continuous(breaks = seq(0, 40000, 10000),
  #                   labels = c('0', '10k', '20k', '30k', '40k')) +
  labs(title = 'Runtime over number of OTUs',
       x = '# OTUs',
       y = 'seconds')
```

```{r fraction_mapped}
results %>% 
  group_by(dataset, ref_weight, ref_frac) %>% 
  ggplot(aes(x=ref_frac, y=fraction_mapped, color=dataset)) +
  geom_jitter(alpha = 0.5, width = 0.01, size=1) +
  scale_color_manual(values = dataset_colors) +
  facet_wrap("ref_weight") +
  ylim(0, 1) +
  labs(title="Sequences mapped during closed-reference OptiFit",
       x='reference fraction',
       y='fraction mapped')
```

### 2020-10-01 subworkflows/4_fitpercent/code/debug.sh

@@ -1,0 +1,38 @@
+#!/bin/bash
+
+##################
+####  Slurm preamble
+
+#### #### ####  These are the most frequently changing options
+
+####  Job name
+#SBATCH --job-name=debug
+
+####  Request resources here
+####    These are typically, number of processors, amount of memory,
+####    an the amount of time a job requires.  May include processor
+####    type, too.
+
+#SBATCH --nodes=1
+#SBATCH --ntasks-per-node=1
+#SBATCH --cpus-per-task=1
+#SBATCH --mem-per-cpu=500MB
+#SBATCH --time=48:00:00
+
+#SBATCH --output=log/hpc/slurm-%j_%x.out
+
+####  Slurm account and partition specification here
+####    These will change if you work on multiple projects, or need
+####    special hardware, like large memory nodes or GPUs.
+
+#SBATCH --account=pschloss99
+#SBATCH --partition=standard
+
+#### #### ####  These are the least frequently changing options
+
+####  Your e-mail address and when you want e-mail
+
+#SBATCH --mail-user=sovacool@umich.edu
+#SBATCH --mail-type=BEGIN,END
+
+time snakemake --profile ../../config/slurm --latency-wait 90 --configfile ../../config/config_test.yaml

### 2020-10-01 exploratory/2020/sub4_fitpercent.Rmd
Use full results tsv

@@
 ```{r read_join}
-dirpath <- 'subworkflows/3_fit_sample_subset/results/all-seqs/'
-results <- list.files(here(dirpath)) %>% 
-  (function(x) { here(glue('{dirpath}{x}')) }) %>% 
-  map(read_tsv) %>% 
-  reduce(full_join) %>% 
-  mutate(num_total_seqs = num_ref_seqs + num_sample_seqs,
-         gaps_frac = n_gaps / total_chars)
+results <- read_tsv('subworkflows/4_fitpercent/results/full_results.tsv')
 head(results)

### 2020-10-05 exploratory/2020/sub3_fit_subsets_fixed.Rmd
Pin input files to 'MOAR wordz' commit

---
title: "Exploratory Plots"
date: "Sept. 2020"
output: 
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      fig.path = "figures/"
                      )
```
```{r lib}
library(cowplot)
library(ggtext)
library(here)
library(knitr)
library(tidyverse)
theme_set(theme_classic())
color_palette <- RColorBrewer::brewer.pal(4, "Dark2")
dataset_colors <- c(
  human = color_palette[[3]],
  marine = color_palette[[1]],
  mouse = color_palette[[4]],
  soil = color_palette[[2]]
)
```
```{r download_data}
dat <- c('https://raw.githubusercontent.com/SchlossLab/OptiFitAnalysis/4fa7127bb6bacaa63657f0c18c0a8c6b6d18ebe7/subworkflows/3_fit_sample_subset/results/sensspec.tsv?token=AEHR6TOTYRF3XTCACDHGFRK7QRYH2',
              'https://raw.githubusercontent.com/SchlossLab/OptiFitAnalysis/4fa7127bb6bacaa63657f0c18c0a8c6b6d18ebe7/subworkflows/3_fit_sample_subset/results/benchmarks.tsv?token=AEHR6TLJWOKI4YC6OJ6BRMC7QRYXY',
              'https://raw.githubusercontent.com/SchlossLab/OptiFitAnalysis/4fa7127bb6bacaa63657f0c18c0a8c6b6d18ebe7/subworkflows/3_fit_sample_subset/results/input_sizes.tsv?token=AEHR6TJVM5FBTZRRY4USOYS7QRY6U',
              'https://raw.githubusercontent.com/SchlossLab/OptiFitAnalysis/4fa7127bb6bacaa63657f0c18c0a8c6b6d18ebe7/subworkflows/3_fit_sample_subset/results/gap_counts.tsv?token=AEHR6TMQKLHG2RNGUZQI35C7QRY5G',
              'https://raw.githubusercontent.com/SchlossLab/OptiFitAnalysis/4fa7127bb6bacaa63657f0c18c0a8c6b6d18ebe7/subworkflows/3_fit_sample_subset/results/fraction_reads_mapped.tsv?token=AEHR6TIY56KPVNGRKCOAYI27QRZBW') %>% 
  lapply(read_tsv) %>% 
  reduce(full_join) %>% 
  mutate(num_total_seqs = num_ref_seqs + num_sample_seqs,
         gaps_frac = n_gaps / total_chars)
```

## Subsetting datasets with different reference sizes

Sample fraction of 0.2 with 20 different seeds.

```{r fit_ref_frac}
dat %>%
  ggplot(aes(x = ref_frac, y = mcc, color = dataset)) +
  geom_jitter(size = 1, alpha = 0.3, width = 0.01) +
  stat_summary(fun = mean, geom = 'crossbar') +
  scale_color_manual(values = dataset_colors) +
  facet_grid(method ~ ref_weight) +
  labs(title='OptiFit Performance',
       x='reference fraction')
```

The performance is very consistent across reference fractions when weighting sequences by abundance or not at all (simple).
When weighting by distance, the performance decreases as the reference fraction increases.
This is expected because a larger fraction means **more** sequences with **fewer** 
pairwise distances under the 0.03 threshold are included in the reference.

## Runtime

```{r runtime_ref_seqs}
dat %>% ggplot(aes(x=num_ref_seqs, y=s, color=dataset)) +
  geom_point(alpha = 0.3) +
  facet_grid(method ~ ref_weight) +
  scale_color_manual(values = dataset_colors) +
  scale_x_continuous(breaks = seq(0, 200000, 50000),
                     labels = c('0', '50k', '100k', '150k', '200k')) +
  labs(title = 'Runtime over reference size',
       x = '# sequences in reference',
       y = 'seconds')
```

Runtime is also fairly consistent regardless of weighting method.
For abundance and simple weighting, runtime seems to continue increasing roughly linearly (maybe a little worse than linear).
When weighting by distance, runtime reaches a plateau sooner than other selection methods as the size of the reference increases.
Perhaps the sequences with the most similarities are the most influential
over OTU clustering, 
so after you reach some critical number of similarities, the runtime doesn't increase as much.
I'm not sure why the human dataset takes so much longer than the others.

```{r runtime_numotus}
dat %>%
  ggplot(aes(x=numotus, y=s, color=dataset, shape = ref_weight)) +
  geom_point(alpha = 0.5) +
  facet_grid(method ~ ref_weight) +
  scale_color_manual(values = dataset_colors) +
  scale_x_continuous(breaks = seq(0, 40000, 10000),
                     labels = c('0', '10k', '20k', '30k', '40k')) +
  labs(title = 'Runtime over number of OTUs',
       x = '# OTUs',
       y = 'seconds')
```

Open-reference clustering has more OTUs because it will cluster any sequences 
_de novo_ that do not fit into existing OTUs, 
while closed-reference clustering throws them out.
This plot is confounded by the number of sequences in the reference, 
perhaps I should re-plot it as the ratio of OTUs to reference sequences?

## Fraction of sequences that map to the reference
```{r fraction_reads_mapped}
dat %>% 
  filter(method == 'closed') %>% 
  group_by(dataset, ref_weight, ref_frac) %>% 
  ggplot(aes(x=ref_frac, y=fraction_mapped, color=dataset)) +
  geom_jitter(alpha = 0.5, width = 0.01, size=1) +
  scale_color_manual(values = dataset_colors) +
  facet_wrap("ref_weight") +
  ylim(0, 1) +
  labs(title="Sequences mapped during closed-reference OptiFit",
       x='reference fraction',
       y='fraction mapped')
```

The fraction of sequences that are able to be fit to the reference in closed-reference
clustering increases as the reference size increases when weighting sequences 
by abundance or not at all for reference selection.
It's also remarkably stable between abundance & simple weighting, and 
between all seeds.
Weighting by distance performs poorly, likely because the top n sequences with 
the most pairwise similarities are included in the reference, 
leaving sequences with few similarities to then be fit to the highly-connected reference.

### 2020-10-05 exploratory/2020/sub3_fit_subsets_fixed.Rmd
Plot de novo results alongside optifit

@@
 )
+full_join_files <- function(filenames) {
+  filenames %>% 
+  lapply(read_tsv) %>% 
+  reduce(full_join)
+}
 ```
@@
               'https://raw.githubusercontent.com/SchlossLab/OptiFitAnalysis/4fa7127bb6bacaa63657f0c18c0a8c6b6d18ebe7/subworkflows/3_fit_sample_subset/results/fraction_reads_mapped.tsv?token=AEHR6TIY56KPVNGRKCOAYI27QRZBW') %>%
-  lapply(read_tsv) %>% 
-  reduce(full_join) %>% 
+  full_join_files() %>% 
   mutate(num_total_seqs = num_ref_seqs + num_sample_seqs,
          gaps_frac = n_gaps / total_chars)
+
+denovo <- c('https://raw.githubusercontent.com/SchlossLab/OptiFitAnalysis/4fa7127bb6bacaa63657f0c18c0a8c6b6d18ebe7/subworkflows/1_prep_samples/results/sensspec.tsv?token=AEHR6TOXMAB75QDAO7CGTA27QR4KQ',
+            'https://raw.githubusercontent.com/SchlossLab/OptiFitAnalysis/4fa7127bb6bacaa63657f0c18c0a8c6b6d18ebe7/subworkflows/1_prep_samples/results/benchmarks.tsv?token=AEHR6TI35D2TNWARGGL37W27QR4MM',
+            'https://raw.githubusercontent.com/SchlossLab/OptiFitAnalysis/4fa7127bb6bacaa63657f0c18c0a8c6b6d18ebe7/subworkflows/1_prep_samples/results/dataset_sizes.tsv?token=AEHR6TJKE2C7SRDUZSGRTCC7QR4MG') %>%
+  full_join_files() %>% 
+  mutate(ref_frac = 1)
+
+denovo_sum <- denovo %>% group_by(dataset) %>% summarize(mean_mcc=mean(mcc), 
+                                                         sd_mcc=sd(mcc),
+                                                         mean_sec=mean(s),
+                                                         sd_sec=sd(s),
+                                                         num_ref_seqs = mean(num_seqs))
+denovo_sum
 ```
@@
        x='reference fraction')
+dat %>%
+  ggplot(aes(x = ref_frac, y = mcc, color = dataset)) +
+  geom_jitter(size = 1, alpha = 0.3, width = 0.01) +
+  stat_summary(fun = mean, geom = 'crossbar') +
+  geom_hline(aes(yintercept = mean_mcc, color = dataset), denovo_sum) +
+  scale_color_manual(values = dataset_colors) +
+  facet_grid(method ~ ref_weight) +
+  labs(title='OptiFit Performance',
+       x='reference fraction',
+       caption='Sample fraction = 0.2 for OptiFit runs. The horizontal line represents _de novo_ clustering for the whole dataset.') +
+  theme(plot.caption = element_markdown())
+dat %>%
+  ggplot(aes(x = ref_frac, y = mcc, color = ref_weight)) +
+  geom_jitter(size = 1, alpha = 0.3, width = 0.01) +
+  stat_summary(fun = mean, geom = 'crossbar') +
+  geom_hline(aes(yintercept = mean_mcc), denovo_sum) +
+  facet_grid(dataset ~ method) +
+  labs(title='OptiFit Performance',
+       x='reference fraction',
+       caption='Sample fraction = 0.2 for OptiFit runs. The horizontal line represents _de novo_ clustering for the whole dataset.') +
+  theme(plot.caption = element_markdown())
 ```
@@
        y = 'seconds')
+dat %>% group_by(num_ref_seqs, dataset, ref_weight, method) %>% 
+  summarize(mean_sec=mean(s)) %>% 
+  ggplot(aes(x=num_ref_seqs, y=mean_sec, color=ref_weight)) +
+  geom_point(alpha = 0.5) +
+  geom_point(aes(x=num_ref_seqs, y=mean_sec, color=NULL), denovo_sum) +
+  facet_grid(dataset ~ method, scales = 'free') +
+  #scale_color_manual(values = dataset_colors) +
+  scale_x_continuous(breaks = seq(0, 200000, 50000),
+                     labels = c('0', '50k', '100k', '150k', '200k')) +
+  labs(title = 'Runtime over reference size',
+       x = '# sequences in reference',
+       y = 'seconds',
+       caption = 'Sample fraction = 0.2 for OptiFit runs. The black dot represents _de novo_ clustering for the whole dataset') +
+  theme(plot.caption = element_markdown())
 ```
@@
        y='fraction mapped')
+dat %>% 
+  filter(method == 'closed') %>% 
+  group_by(dataset, ref_weight, ref_frac) %>% 
+  ggplot(aes(x=ref_frac, y=fraction_mapped, color=ref_weight)) +
+  geom_jitter(alpha = 0.5, width = 0.01, size=1) +
+  #scale_color_manual(values = dataset_colors) +
+  facet_wrap("dataset", nrow=1) +
+  ylim(0, 1) +
+  labs(title="Sequences mapped during closed-reference OptiFit",
+       x='reference fraction',
+       y='fraction mapped')
 ```

### 2020-10-05 exploratory/2020/sub3_fit_subsets_fixed.Rmd
Try using newline in caption

@@
        x='reference fraction',
-       caption='Sample fraction = 0.2 for OptiFit runs. The horizontal line represents _de novo_ clustering for the whole dataset.') +
+       caption='Sample fraction = 0.2 for OptiFit runs.\nThe horizontal line represents _de novo_ clustering for the whole dataset.') +
   theme(plot.caption = element_markdown())
@@
        y = 'seconds',
-       caption = 'Sample fraction = 0.2 for OptiFit runs. The black dot represents _de novo_ clustering for the whole dataset') +
+       caption = 'Sample fraction = 0.2 for OptiFit runs.\nThe black dot represents _de novo_ clustering for the whole dataset') +
   theme(plot.caption = element_markdown())

### 2020-10-06 subworkflows/3_fit_sample_subset/code/debug.sh
Update test slurm scripts

@@ -6,7 +6,7 @@
 #### #### ####  These are the most frequently changing options
 
 ####  Job name
-#SBATCH --job-name=debug
+#SBATCH --job-name=debug3
 
 ####  Request resources here
 ####    These are typically, number of processors, amount of memory,
@@ -35,4 +35,5 @@
 #SBATCH --mail-user=sovacool@umich.edu
 #SBATCH --mail-type=BEGIN,END
 
-time snakemake --profile ../../config/slurm --latency-wait 90 --configfile ../../config/config_test.yaml
+which mothur
+#time snakemake --profile ../../config/slurm --latency-wait 90 --configfile ../../config/config_test.yaml

### 2020-10-06 subworkflows/3_fit_sample_subset/code/test.sh

@@ -35,4 +35,4 @@
 #SBATCH --mail-user=sovacool@umich.edu
 #SBATCH --mail-type=BEGIN,END
 
-time snakemake --profile ../../config/slurm --latency-wait 90 --configfile ../../config/config_test.yaml 
+time snakemake --profile ../../config/slurm --latency-wait 90 --configfile ../../config/config_test.yaml --forceall

### 2020-10-06 exploratory/2020/sub3_fit_subsets_fixed.Rmd

@@
        x='reference fraction',
-       caption='Sample fraction = 0.2 for OptiFit runs. The horizontal line represents _de novo_ clustering for the whole dataset.') +
+       caption='Sample fraction = 0.2 for OptiFit runs. Black line: _de novo_ clustering on the entire dataset') +
   theme(plot.caption = element_markdown())
@@
        y = 'seconds',
-       caption = 'Sample fraction = 0.2 for OptiFit runs.\nThe black dot represents _de novo_ clustering for the whole dataset') +
+       caption = 'Sample fraction = 0.2 for OptiFit runs. Black dot: _de novo_ clustering on the entire dataset') +
   theme(plot.caption = element_markdown())

### 2020-10-07 exploratory/2020/sub3_fit_all-seqs.Rmd
Create plots with query seqs as complement of ref seqs"""""" """""""""""""""""""

---
title: "Exploratory Plots"
date: "Sept. 2020"
output: 
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      fig.path = "figures/"
                      )
```
```{r lib}
library(cowplot)
library(ggtext)
library(here)
library(knitr)
library(tidyverse)
theme_set(theme_classic())
color_palette <- RColorBrewer::brewer.pal(4, "Dark2")
dataset_colors <- c(
  human = color_palette[[3]],
  marine = color_palette[[1]],
  mouse = color_palette[[4]],
  soil = color_palette[[2]]
)
full_join_files <- function(filenames) {
  filenames %>% 
  lapply(read_tsv) %>% 
  reduce(full_join)
}
```
```{r download_data}
dat <- c('subworkflows/3_fit_sample_subset/results/all-seqs/sensspec.tsv',
         'subworkflows/3_fit_sample_subset/results/all-seqs/benchmarks.tsv',
         'subworkflows/3_fit_sample_subset/results/all-seqs/input_sizes.tsv',
         'subworkflows/3_fit_sample_subset/results/all-seqs/sensspec.tsv',
         'subworkflows/3_fit_sample_subset/results/all-seqs/fraction_reads_mapped.tsv') %>% 
  here() %>% 
  full_join_files() %>% 
  mutate(num_total_seqs = num_ref_seqs + num_sample_seqs,
         gaps_frac = n_gaps / total_chars)

denovo <- c('https://raw.githubusercontent.com/SchlossLab/OptiFitAnalysis/4fa7127bb6bacaa63657f0c18c0a8c6b6d18ebe7/subworkflows/1_prep_samples/results/sensspec.tsv?token=AEHR6TOXMAB75QDAO7CGTA27QR4KQ',
            'https://raw.githubusercontent.com/SchlossLab/OptiFitAnalysis/4fa7127bb6bacaa63657f0c18c0a8c6b6d18ebe7/subworkflows/1_prep_samples/results/benchmarks.tsv?token=AEHR6TI35D2TNWARGGL37W27QR4MM',
            'https://raw.githubusercontent.com/SchlossLab/OptiFitAnalysis/4fa7127bb6bacaa63657f0c18c0a8c6b6d18ebe7/subworkflows/1_prep_samples/results/dataset_sizes.tsv?token=AEHR6TJKE2C7SRDUZSGRTCC7QR4MG') %>% 
  full_join_files() %>% 
  mutate(ref_frac = 1)

denovo_sum <- denovo %>% group_by(dataset) %>% summarize(mean_mcc=mean(mcc), 
                                                         sd_mcc=sd(mcc),
                                                         mean_sec=mean(s),
                                                         sd_sec=sd(s),
                                                         num_ref_seqs = mean(num_seqs))
denovo_sum
```

## OptiFit performance with varied reference sizes

The query sequences are the complement of the reference sequences. i.e. $query\_frac = 1 - reference\_frac$

```{r mcc_complement}
dat %>%
  ggplot(aes(x = ref_frac, y = mcc, color = dataset)) +
  geom_jitter(size = 1, alpha = 0.3, width = 0.01) +
  stat_summary(fun = mean, geom = 'crossbar') +
  scale_color_manual(values = dataset_colors) +
  facet_grid(method ~ ref_weight) +
  labs(title='OptiFit Performance',
       x='reference fraction')
dat %>%
  ggplot(aes(x = ref_frac, y = mcc, color = dataset)) +
  geom_jitter(size = 1, alpha = 0.3, width = 0.01) +
  stat_summary(fun = mean, geom = 'crossbar') +
  geom_hline(aes(yintercept = mean_mcc, color = dataset), denovo_sum, linetype='dashed') +
  scale_color_manual(values = dataset_colors) +
  facet_grid(method ~ ref_weight) +
  labs(title='OptiFit Performance',
       x='reference fraction',
       caption='Dashed line: _de novo_ clustering on the whole dataset.') +
  theme(plot.caption = element_markdown())

dat %>%
  ggplot(aes(x = ref_frac, y = mcc, color = ref_weight)) +
  geom_jitter(size = 1, alpha = 0.3, width = 0.01) +
  stat_summary(fun = mean, geom = 'crossbar', alpha = 0.5) +
  geom_hline(aes(yintercept = mean_mcc), denovo_sum) +
  facet_grid(dataset ~ method, scales = 'free') +
  labs(title='OptiFit Performance',
       x='reference fraction',
       caption='Black line: _de novo_ clustering on the whole dataset') +
  theme(plot.caption = element_markdown())
dat %>%
  ggplot(aes(x = ref_frac, y = mcc, color = ref_weight)) +
  geom_jitter(size = 1, alpha = 0.3, width = 0.01) +
  stat_summary(fun = mean, geom = 'crossbar', alpha = 0.5) +
  geom_hline(aes(yintercept = mean_mcc), denovo_sum) +
  facet_grid(dataset ~ method) +
  ylim(0.5, 1) +
  labs(title='OptiFit Performance',
       x='reference fraction',
       caption='Black line: _de novo_ clustering on the whole dataset') +
  theme(plot.caption = element_markdown())
```

Closed- and open-reference clustering perform virtually the same as _de novo_ clustering.

The performance is very consistent across reference fractions when weighting sequences by abundance or not at all (simple).
When weighting by distance, the performance decreases slightly as the reference fraction increases.
This is expected because a larger fraction means **more** sequences with **fewer** 
pairwise distances under the 0.03 threshold are included in the reference.

## Runtime

```{r runtime_nseqs_complement}
dat %>% ggplot(aes(x=num_ref_seqs, y=s, color=dataset)) +
  geom_point(alpha = 0.3) +
  geom_point(aes(x=num_ref_seqs, y=mean_sec), denovo_sum, shape='triangle') +
  facet_grid(method ~ ref_weight) +
  scale_color_manual(values = dataset_colors) +
  scale_x_continuous(breaks = seq(0, 300000, 50000),
                     labels = c('0', '50k', '100k', '150k', '200k', '250k', '300k')) +
  labs(title = 'Runtime over reference size',
       x = '# sequences in reference',
       y = 'seconds',
       caption = 'Triangles: mean runtime for _de novo_ clustering on the entire dataset') +
  theme(plot.caption = element_markdown())
dat %>% group_by(num_ref_seqs, dataset, ref_weight, method) %>% 
  summarize(mean_sec=mean(s)) %>% 
  ggplot(aes(x=num_ref_seqs, y=mean_sec, color=ref_weight)) +
  geom_point(alpha = 0.5) +
  geom_point(aes(x=num_ref_seqs, y=mean_sec, color=NULL), denovo_sum, shape='triangle') +
  facet_grid(dataset ~ method, scales = 'free') +
  #scale_color_manual(values = dataset_colors) +
  scale_x_continuous(breaks = seq(0, 300000, 50000),
                     labels = c('0', '50k', '100k', '150k', '200k', '250k', '300k')) +
  labs(title = 'Mean Runtime over reference size',
       x = '# sequences in reference',
       y = 'seconds',
       caption = 'Black triangles: mean runtime for _de novo_ clustering on the entire dataset') +
  theme(plot.caption = element_markdown())
```

```{r runtime_numotus_complement}
dat %>%
  ggplot(aes(x=numotus, y=s, color=dataset)) +
  geom_point(alpha = 0.3) +
  facet_grid(method ~ ref_weight) +
  scale_color_manual(values = dataset_colors) +
  scale_x_continuous(breaks = seq(0, 80000, 20000),
                     labels = c('0', '20k', '40k', '60k', '80k')) +
  labs(title = 'Runtime over number of OTUs',
       x = '# OTUs',
       y = 'seconds')
dat %>%
  ggplot(aes(x=numotus, y=s, color=ref_weight)) +
  geom_point(alpha = 0.3) +
  facet_grid(dataset ~ method, scales = 'free') +
  labs(title = 'Runtime over number of OTUs',
       x = '# OTUs',
       y = 'seconds')
```

Open-reference clustering has more OTUs because it will cluster any sequences 
_de novo_ that do not fit into existing OTUs, 
while closed-reference clustering throws them out.
This plot is confounded by the number of sequences in the reference, 
perhaps I should re-plot it as the ratio of OTUs to reference sequences?

## Fraction of sequences that map to the reference
```{r fraction_mapped_complement}
dat %>% 
  filter(method == 'closed') %>% 
  group_by(dataset, ref_weight, ref_frac) %>% 
  ggplot(aes(x=ref_frac, y=fraction_mapped, color=dataset)) +
  geom_jitter(alpha = 0.5, width = 0.01, size=1) +
  scale_color_manual(values = dataset_colors) +
  facet_wrap("ref_weight") +
  ylim(0, 1) +
  labs(title="Sequences mapped during closed-reference OptiFit",
       x='reference fraction',
       y='fraction mapped')
dat %>% 
  filter(method == 'closed') %>% 
  group_by(dataset, ref_weight, ref_frac) %>% 
  ggplot(aes(x=ref_frac, y=fraction_mapped, color=ref_weight)) +
  geom_jitter(alpha = 0.5, width = 0.01, size=1) +
  facet_wrap("dataset", nrow=1) +
  ylim(0, 1) +
  labs(title="Sequences mapped during closed-reference OptiFit",
       x='reference fraction',
       y='fraction mapped')
```

The fraction of sequences that are able to be fit to the reference in closed-reference
clustering increases as the reference size increases when weighting sequences 
by abundance or by a simple random sample.
It's also remarkably stable between abundance & simple weighting, and 
between all seeds.
Weighting by distance performs poorly, likely because the top n sequences with 
the most pairwise similarities are included in the reference, 
leaving sequences with few similarities to then be fit to the highly-connected reference.

### 2020-10-07 exploratory/2020/sub4_fitpercent.Rmd
Rename chunks to avoid conflicts

@@
 
-```{r sensspec}
+```{r sensspec_fitpercent}
 results %>% 
@@
 
-```{r runtime}
+```{r runtime_fitpercent}
 results %>% ggplot(aes(x=num_ref_seqs, y=s, color=dataset)) +
@@
 
-```{r fraction_mapped}
+```{r fraction_mapped_fitpercent}
 results %>% 

### 2020-10-08 exploratory/2020/sub3_fit_all-seqs.Rmd
Paste filepaths

@@
 
-denovo <- c('https://raw.githubusercontent.com/SchlossLab/OptiFitAnalysis/4fa7127bb6bacaa63657f0c18c0a8c6b6d18ebe7/subworkflows/1_prep_samples/results/sensspec.tsv?token=AEHR6TOXMAB75QDAO7CGTA27QR4KQ',
-            'https://raw.githubusercontent.com/SchlossLab/OptiFitAnalysis/4fa7127bb6bacaa63657f0c18c0a8c6b6d18ebe7/subworkflows/1_prep_samples/results/benchmarks.tsv?token=AEHR6TI35D2TNWARGGL37W27QR4MM',
-            'https://raw.githubusercontent.com/SchlossLab/OptiFitAnalysis/4fa7127bb6bacaa63657f0c18c0a8c6b6d18ebe7/subworkflows/1_prep_samples/results/dataset_sizes.tsv?token=AEHR6TJKE2C7SRDUZSGRTCC7QR4MG') %>%
+denovo <- c('sensspec.tsv?token=AEHR6TOXMAB75QDAO7CGTA27QR4KQ',
+            'benchmarks.tsv?token=AEHR6TI35D2TNWARGGL37W27QR4MM',
+            'dataset_sizes.tsv?token=AEHR6TJKE2C7SRDUZSGRTCC7QR4MG') %>% 
+  lapply(function(x) { paste0('https://raw.githubusercontent.com/SchlossLab/OptiFitAnalysis/4fa7127bb6bacaa63657f0c18c0a8c6b6d18ebe7/subworkflows/1_prep_samples/results/', x)}) %>% 
   full_join_files() %>% 

### 2020-10-08 exploratory/2020/sub3_fit_all-seqs.Rmd
Plot memory usage

@@
   mutate(num_total_seqs = num_ref_seqs + num_sample_seqs,
-         gaps_frac = n_gaps / total_chars)
+         gaps_frac = n_gaps / total_chars,
+         mem_mb = max_rss,
+         mem_gb = mem_mb / 1024)
 
@@
                                                          sd_sec=sd(s),
-                                                         num_ref_seqs = mean(num_seqs))
+                                                         num_ref_seqs = mean(num_seqs),
+                                                         mean_mem_gb=mean(max_rss/1024),
+                                                         sd_mem_gb=sd(max_rss/1024))
 denovo_sum
@@
 
+## Memory usage
+
+```{r mem}
+dat %>% group_by(num_ref_seqs, dataset, ref_weight, method) %>% 
+  summarize(mean_mem_gb=mean(mem_gb)) %>% 
+  ggplot(aes(x=num_ref_seqs, y=mean_mem_gb, color=dataset)) +
+  geom_point(alpha = 0.5) +
+  geom_point(aes(x=num_ref_seqs, y=mean_mem_gb), denovo_sum, shape='triangle') +
+  facet_grid(method ~ ref_weight) +
+  scale_color_manual(values = dataset_colors) +
+  scale_x_continuous(breaks = seq(0, 300000, 50000),
+                     labels = c('0', '50k', '100k', '150k', '200k', '250k', '300k')) +
+  labs(title = 'Memory usage over reference size',
+       x = '# sequences in reference',
+       y = 'MB',
+       caption = 'Triangles: mean runtime for _de novo_ clustering on the entire dataset') +
+  theme(plot.caption = element_markdown())
+dat %>% group_by(num_ref_seqs, dataset, ref_weight, method) %>% 
+  summarize(mean_mem_gb=mean(mem_gb)) %>% 
+  ggplot(aes(x=num_ref_seqs, y=mean_mem_gb, color=ref_weight)) +
+  geom_point(alpha = 0.5) +
+  geom_point(aes(x=num_ref_seqs, y=mean_mem_gb, color=NULL), denovo_sum, shape='triangle') +
+  facet_grid(dataset ~ method, scales = 'free') +
+  scale_x_continuous(breaks = seq(0, 300000, 50000),
+                     labels = c('0', '50k', '100k', '150k', '200k', '250k', '300k')) +
+  labs(title = 'Memory usage over reference size',
+       x = '# sequences in reference',
+       y = 'MB',
+       caption = 'Black triangles: mean runtime for _de novo_ clustering on the entire dataset') +
+  theme(plot.caption = element_markdown())
+```
+
 ## Fraction of sequences that map to the reference

### 2020-10-08 exploratory/2020/sub3_fit_all-seqs.Rmd
Set random seed

@@
 }
+set.seed(2018)
 ```

### 2020-10-16 subworkflows/4_fitpercent/code/debug.sh
Commit old 4_fitpercent stuff

@@ -35,4 +35,4 @@
 #SBATCH --mail-user=sovacool@umich.edu
 #SBATCH --mail-type=BEGIN,END
 
-time snakemake --profile ../../config/slurm --latency-wait 90 --configfile ../../config/config_test.yaml --forceall
+time snakemake --profile ../../config/slurm --latency-wait 90 --configfile ../../config/config_test.yaml results/soil/refweight_none/fitpercent_10/seed_10/method_closed/printref_f/soil.optifit_mcc_denovo.best.list

### 2020-10-16 subworkflows/4_fitpercent/code/debug.sh
Remove fitpercent from GLC

@@ -1,38 +1,0 @@
-#!/bin/bash
-
-##################
-####  Slurm preamble
-
-#### #### ####  These are the most frequently changing options
-
-####  Job name
-#SBATCH --job-name=debug4
-
-####  Request resources here
-####    These are typically, number of processors, amount of memory,
-####    an the amount of time a job requires.  May include processor
-####    type, too.
-
-#SBATCH --nodes=1
-#SBATCH --ntasks-per-node=1
-#SBATCH --cpus-per-task=1
-#SBATCH --mem-per-cpu=500MB
-#SBATCH --time=48:00:00
-
-#SBATCH --output=log/hpc/slurm-%j_%x.out
-
-####  Slurm account and partition specification here
-####    These will change if you work on multiple projects, or need
-####    special hardware, like large memory nodes or GPUs.
-
-#SBATCH --account=pschloss1
-#SBATCH --partition=standard
-
-#### #### ####  These are the least frequently changing options
-
-####  Your e-mail address and when you want e-mail
-
-#SBATCH --mail-user=sovacool@umich.edu
-#SBATCH --mail-type=BEGIN,END
-
-time snakemake --profile ../../config/slurm --latency-wait 90 --configfile ../../config/config_test.yaml results/soil/refweight_none/fitpercent_10/seed_10/method_closed/printref_f/soil.optifit_mcc_denovo.best.list

### 2020-10-16 subworkflows/4_fitpercent/code/test.sh

@@ -1,38 +1,0 @@
-#!/bin/bash
-
-##################
-####  Slurm preamble
-
-#### #### ####  These are the most frequently changing options
-
-####  Job name
-#SBATCH --job-name=test4
-
-####  Request resources here
-####    These are typically, number of processors, amount of memory,
-####    an the amount of time a job requires.  May include processor
-####    type, too.
-
-#SBATCH --nodes=1
-#SBATCH --ntasks-per-node=1
-#SBATCH --cpus-per-task=1
-#SBATCH --mem-per-cpu=500MB
-#SBATCH --time=48:00:00
-
-#SBATCH --output=log/hpc/slurm-%j_%x.out
-
-####  Slurm account and partition specification here
-####    These will change if you work on multiple projects, or need
-####    special hardware, like large memory nodes or GPUs.
-
-#SBATCH --account=pschloss1
-#SBATCH --partition=standard
-
-#### #### ####  These are the least frequently changing options
-
-####  Your e-mail address and when you want e-mail
-
-#SBATCH --mail-user=sovacool@umich.edu
-#SBATCH --mail-type=BEGIN,END
-
-time snakemake --profile ../../config/slurm --latency-wait 90 --configfile ../../config/config_test.yaml 

### 2020-10-16 exploratory/2020/sub3_fit_all-seqs.Rmd

@@
        x = '# sequences in reference',
-       y = 'MB',
+       y = 'GB',
        caption = 'Triangles: mean runtime for _de novo_ clustering on the entire dataset') +
@@
        x = '# sequences in reference',
-       y = 'MB',
+       y = 'GB',
        caption = 'Black triangles: mean runtime for _de novo_ clustering on the entire dataset') +

### 2020-10-23 subworkflows/4_vsearch/code/biom_to_list.R
Replace hyphens with underscores in seq ids

@@ -21,8 +21,7 @@ biom_to_list <- function(biom_file_name, list_file_name, label = 0.03) {
   list_data <- paste(paste(c("label", "numOTUs", otu_names),
     collapse = "\t"
   ),
-  paste(c(label, n_otus, otu_assignments %>%
-    gsub("_", ".", .)), # revert seq names to originals with dots
+  paste(c(label, n_otus, otu_assignments), # revert seq names to originals with dots
   collapse = "\t"
   ),
   sep = "\n"

### 2020-10-23 subworkflows/4_vsearch/code/prep_count_table.R

@@ -1,5 +1,6 @@
 library(dplyr)
 readr::read_tsv(snakemake@input[["count"]]) %>%
-  mutate(dummyRefGroup = total) %>%
+  mutate(dummyRefGroup = total,
+         Representative_Sequence = gsub('[\\.-]', '_', Representative_Sequence)) %>%
   select(Representative_Sequence, dummyRefGroup) %>%
   readr::write_tsv(snakemake@output[["count"]])

### 2020-10-23 exploratory/2020/sub4_vsearch.Rmd
Plot vsearch with soil silva v4

---
title: "VSEARCH clustering"
date: "Oct. 2020"
output: 
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      fig.path = "figures/"
                      )
```
```{r lib}
library(cowplot)
library(ggtext)
library(glue)
library(here)
library(knitr)
library(tidyverse)
theme_set(theme_classic())
color_palette <- RColorBrewer::brewer.pal(4, "Dark2")
dataset_colors <- c(
  human = color_palette[[3]],
  marine = color_palette[[1]],
  mouse = color_palette[[4]],
  soil = color_palette[[2]]
)
set.seed(2018)

optifit <- c('subworkflows/2_fit_reference_db/results/sensspec.tsv',
             'subworkflows/2_fit_reference_db/results/benchmarks.tsv',
             'subworkflows/2_fit_reference_db/results/fraction_reads_mapped.tsv'
             ) %>% 
  here() %>% 
  map(read_tsv) %>%
  reduce(full_join) 
opticlust <- c('subworkflows/1_prep_samples/results/sensspec.tsv',
               'subworkflows/1_prep_samples/results/benchmarks.tsv') %>% 
  here() %>% 
  map(read_tsv) %>% 
  reduce(full_join) 
vsearch <- read_tsv(here('subworkflows/4_vsearch/results/vsearch_results.tsv'))

dat <- bind_rows(optifit, opticlust) %>% 
  mutate(tool = "mothur") %>% 
  bind_rows(vsearch) %>% 
  mutate(mem_mb = max_rss,
         mem_gb = mem_mb / 1024,
         sec = s)
```

For now, just going to show silva v4 for reference-clustering with the soil dataset.
Will expand to include more variables later.
```{r silva-soil-only}
dat_filt <- dat %>% filter(ref %in% c('silva', NA), 
                           region %in% c('bact_v4', NA),
                           dataset == 'soil') %>% 
  select(dataset, ref, region, tool, method, mcc, sec, mem_gb, numotus, fraction_mapped) %>% 
  mutate(tool_method = glue("{tool}_{method}"))

dat_filt %>% pivot_longer(c(mcc, sec, mem_gb), names_to = 'metric') %>% 
  ggplot(aes(x = method, y = value, color = tool)) +
  geom_boxplot() +
  facet_wrap('metric', scales = 'free')
```

## OTU Quality

```{r mcc}
dat_filt %>% ggplot(aes(x = tool, y = mcc, color = dataset)) +
  geom_boxplot() +
  facet_wrap('method') +
  scale_color_manual(values = dataset_colors) +
  ylim(0, 1)
```

## Performance

```{r runtime}
dat_filt %>% ggplot(aes(x = tool, y = sec, color = dataset)) +
  geom_boxplot() +
  facet_wrap('method') +
  scale_color_manual(values = dataset_colors)
dat_filt %>% ggplot(aes(x = tool, y = mem_gb, color = dataset)) +
  geom_boxplot() +
  facet_wrap('method') +
  scale_color_manual(values = dataset_colors)
```

## Fraction reads mapped during closed-reference clustering

```{r fraction}
dat_filt %>% filter(method == 'closed') %>% 
  ggplot(aes(x = tool, y = fraction_mapped, color = dataset)) +
  geom_boxplot() +
  scale_color_manual(values = dataset_colors) +
  ylim(0, 1)
```

### 2020-10-23 subworkflows/4_vsearch/code/prep_count_table.R
🎨 Style R code ⬡

@@ -1,6 +1,8 @@
 library(dplyr)
 readr::read_tsv(snakemake@input[["count"]]) %>%
-  mutate(dummyRefGroup = total,
-         Representative_Sequence = gsub('[\\.-]', '_', Representative_Sequence)) %>%
+  mutate(
+    dummyRefGroup = total,
+    Representative_Sequence = gsub("[\\.-]", "_", Representative_Sequence)
+  ) %>%
   select(Representative_Sequence, dummyRefGroup) %>%
   readr::write_tsv(snakemake@output[["count"]])

### 2020-10-26 exploratory/2020/sub3_fit_all-seqs.Rmd
Update plots before lab meeting

@@
 
-denovo <- c('sensspec.tsv?token=AEHR6TOXMAB75QDAO7CGTA27QR4KQ',
-            'benchmarks.tsv?token=AEHR6TI35D2TNWARGGL37W27QR4MM',
-            'dataset_sizes.tsv?token=AEHR6TJKE2C7SRDUZSGRTCC7QR4MG') %>% 
-  lapply(function(x) { paste0('https://raw.githubusercontent.com/SchlossLab/OptiFitAnalysis/4fa7127bb6bacaa63657f0c18c0a8c6b6d18ebe7/subworkflows/1_prep_samples/results/', x)}) %>% 
+denovo <- c('sensspec.tsv?token=AEHR6TKSSN7MCV5HE2X7GVK7T5ZH2',
+            'benchmarks.tsv?token=AEHR6TPJKCME4ZASVBZ3QYK7T5ZL6',
+            'dataset_sizes.tsv?token=AEHR6TNFM52JSRD3LU45DB27T5ZN2') %>% 
+  lapply(function(x) { paste0('https://raw.githubusercontent.com/SchlossLab/OptiFitAnalysis/master/subworkflows/1_prep_samples/results/', x)}) %>% 
   full_join_files() %>% 
@@
   facet_grid(dataset ~ method, scales = 'free') +
-  #scale_color_manual(values = dataset_colors) +
   scale_x_continuous(breaks = seq(0, 300000, 50000),
@@
        caption = 'Black triangles: mean runtime for _de novo_ clustering on the entire dataset') +
+  theme(plot.caption = element_markdown())
+dat %>% ggplot(aes(x=num_ref_seqs, y=s)) +
+  geom_point(aes(color = ref_weight), alpha = 0.3) +
+  geom_point(aes(x=num_ref_seqs, y=mean_sec), denovo_sum, shape='triangle') +
+  facet_grid(dataset ~ method, scales = 'free') +
+  scale_x_continuous(breaks = seq(0, 300000, 50000),
+                     labels = c('0', '50k', '100k', '150k', '200k', '250k', '300k')) +
+  labs(title = 'Runtime over reference size',
+       x = '# sequences in reference',
+       y = 'seconds',
+       caption = 'Triangles: mean runtime for _de novo_ clustering on the entire dataset') +
   theme(plot.caption = element_markdown())

### 2020-10-26 exploratory/2020/sub4_vsearch.Rmd
Update plots before lab meeting

@@
   geom_boxplot() +
-  facet_wrap('metric', scales = 'free')
+  facet_wrap('metric', scales = 'free') + 
+  labs(caption = "For reference clustering, the soil dataset was fit to the silva db.") +
+  theme(axis.title = element_blank())
 ```

### 2020-10-27 exploratory/2020/sub4_vsearch.Rmd
Update vsearch plots with gg db; newer optifit version

--- 2020-10-23 exploratory/2020/sub4_vsearch.Rmd
@@
 
-For now, just going to show silva v4 for reference-clustering with the soil dataset.
-Will expand to include more variables later.
-```{r silva-soil-only}
-dat_filt <- dat %>% filter(ref %in% c('silva', NA), 
+Preliminary results with just the soil dataset vs greengenes below.
+Will expand to include more datasets later.
+```{r vsearch-soil-only}
+dat_filt <- dat %>% filter(ref %in% c('gg', NA), 
                            region %in% c('bact_v4', NA),
@@
   geom_boxplot() +
-  facet_wrap('metric', scales = 'free')
+  facet_wrap('metric', scales = 'free') + 
+  labs(caption = "For reference clustering, the soil dataset was fit to the greengenes db.") +
+  theme(axis.title = element_blank())
 ```
@@
 
-```{r mcc}
+```{r vsearch_mcc}
 dat_filt %>% ggplot(aes(x = tool, y = mcc, color = dataset)) +
@@
 
-```{r runtime}
+```{r vsearch_runtime}
 dat_filt %>% ggplot(aes(x = tool, y = sec, color = dataset)) +
@@
 
-```{r fraction}
+```{r vsearch_fraction}
 dat_filt %>% filter(method == 'closed') %>% 

### 2020-10-28 exploratory/2020/sub4_vsearch.Rmd

@@
          mem_gb = mem_mb / 1024,
-         sec = s)
+         sec = s) %>% 
+  filter(ref %in% c('gg', NA))
 ```
@@
 ```{r vsearch_mcc}
-dat_filt %>% ggplot(aes(x = tool, y = mcc, color = dataset)) +
+mcc <- dat %>% ggplot(aes(x = method, y = mcc, color = tool)) +
   geom_boxplot() +
-  facet_wrap('method') +
-  scale_color_manual(values = dataset_colors) +
-  ylim(0, 1)
+  facet_wrap('dataset', ncol = 1) +
+  ylim(0, 1) + 
+  labs(x = '')
+mcc
 ```
@@
 ```{r vsearch_runtime}
-dat_filt %>% ggplot(aes(x = tool, y = sec, color = dataset)) +
+sec <- dat %>% ggplot(aes(x = method, y = sec, color = tool)) +
   geom_boxplot() +
-  facet_wrap('method') +
-  scale_color_manual(values = dataset_colors)
-dat_filt %>% ggplot(aes(x = tool, y = mem_gb, color = dataset)) +
+  facet_wrap('dataset', scales = 'free', ncol = 1)  +
+  labs(x = '')
+sec
+mem <- dat %>% ggplot(aes(x = method, y = mem_gb, color = tool)) +
   geom_boxplot() +
-  facet_wrap('method') +
-  scale_color_manual(values = dataset_colors)
+  facet_wrap('dataset', scales = 'free', ncol = 1)  +
+  labs(x = '')
+mem
 ```
@@
 ```{r vsearch_fraction}
-dat_filt %>% filter(method == 'closed') %>% 
-  ggplot(aes(x = tool, y = fraction_mapped, color = dataset)) +
+dat %>% filter(method == 'closed') %>% 
+  ggplot(aes(x = method, y = fraction_mapped, color = tool)) +
   geom_boxplot() +
-  scale_color_manual(values = dataset_colors) +
-  ylim(0, 1)
+  facet_wrap('dataset') +
+  ylim(0, 1) +
+  labs(caption = 'Sequences were aligned to the greengenes database.')
 ```
+
+## All metrics summarized on one plot
+
+```{r vsearch_summary}
+dat %>% pivot_longer(c(mcc, sec, mem_gb), names_to = 'metric') %>% 
+  ggplot(aes(x = method, y = value, color = tool)) +
+  geom_boxplot() +
+  facet_wrap(dataset ~ metric, scales = 'free', ncol = 3, nrow = 4) + 
+  labs(caption = "For reference clustering, the soil dataset was fit to the greengenes db.") +
+  theme(axis.title = element_blank())
+```
+
+```{r vsearch_cowplot}
+legend <- get_legend(mcc)
+pgrid <- plot_grid(mcc + theme(legend.position = 'none'), 
+          sec + theme(legend.position = 'none'), 
+          mem + theme(legend.position = 'none'), 
+          align = 'h', nrow = 1)
+plot_grid(pgrid, legend, rel_widths = c(3, .4))
+```

### 2020-11-08 code/R/cbind_optifit_seed.R
Fix typo: use cbind not rbind

@@ -22,4 +22,4 @@ mutate_columns <- function(df_orig) {
     )
 }
 
-rbind_all('tsv')
+cbind_all('tsv')

### 2020-11-08 exploratory/2020-10/sub2_fit_dbs.Rmd
Reorganize notebooks

--- 2020-06-30 exploratory/2020/README.Rmd
@@
 title: "Exploratory Plots"
-date: "5/27/2020"
+date: "May 2020"
 output: 
@@
   group_by(dataset, method) %>%
-  ggplot(aes(x = method, y = mcc, color = dataset)) +
+  ggplot(aes(x = method, y = mcc, color = ref)) +
   geom_hline(aes(yintercept = mean_mcc), sensspec_clust) +
   geom_boxplot(alpha = 0.5) +
-  scale_color_manual(values = dataset_colors) +
-  ylim(0.5, 1) +
-  facet_grid(dataset ~ ref) +
+  scale_color_brewer(palette = 'Dark2') +
+  #ylim(0.5, 1) +
+  facet_wrap('dataset', scales = 'free') +
   labs(caption = "The black line represents the *de novo* clustering mcc score.") +
-  theme(legend.position = "None",
+  theme(#legend.position = "None",
         axis.title.x = element_blank(),
@@
 ### Fraction of sequences that map to the reference
-```{r fraction_reads_mapped}
+```{r fraction_reads_mapped_dbs}
 fractions <- read_tsv(here('subworkflows/2_fit_reference_db/results/fraction_reads_mapped.tsv'))
 fractions %>% 
-  ggplot(aes(x=dataset, y=fraction_mapped, color=dataset)) +
+  ggplot(aes(x=dataset, y=fraction_mapped, color=ref)) +
   geom_boxplot(alpha = 0.5) +
-  scale_color_manual(values = dataset_colors) +
-  facet_wrap("ref") +
+  scale_color_brewer(palette = 'Dark2') +
+  facet_wrap("dataset", scales = 'free') +
   ylim(0, 1) +

### 2020-11-08 exploratory/2020-11/plots.Rmd

---
date: 'Nov. 2020'
output: 
  github_document:
    html_preview: false
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      fig.path = "figures/"
                      )
```
```{r lib}
library(cowplot)
library(ggtext)
library(glue)
library(here)
library(knitr)
library(tidyverse)
theme_set(theme_classic())
color_palette <- RColorBrewer::brewer.pal(4, "Dark2")
dataset_colors <- c(
  human = color_palette[[3]],
  marine = color_palette[[1]],
  mouse = color_palette[[4]],
  soil = color_palette[[2]]
)
set.seed(2018)
```

## Fit to reference databases

```{r}
ref_sizes <- read_tsv(here('subworkflows/2_fit_reference_db/results/ref_sizes.tsv'))
optifit_dbs <- read_tsv(here('subworkflows/2_fit_reference_db/results/optifit_dbs_results.tsv'))
```

```{r}
optifit_dbs
```

### 2020-11-11 exploratory/2020-11/plots.Rmd
Initial plots from updated results

@@
 ```{r lib}
+set.seed(2018)
 library(cowplot)
@@
 library(tidyverse)
+
 theme_set(theme_classic())
@@
 )
-set.seed(2018)
+
+mutate_perf <- function(dat) {
+  dat %>% 
+    mutate(mem_mb = max_rss,
+           mem_gb = mem_mb / 1024) %>% 
+    rename(sec = s,
+           num_otus = sobs)
+}
 ```
 
-## Fit to reference databases
+## _de novo_ clustering
 
-```{r}
+```{r load_opticlust}
+opticlust <- read_tsv(here('subworkflows/1_prep_samples/results/opticlust_results.tsv')) %>% 
+  full_join(read_tsv(here('subworkflows/1_prep_samples/results/dataset_sizes.tsv'))) %>% 
+  mutate_perf()
+sum_opticlust <- opticlust %>% 
+  group_by(dataset) %>% 
+  summarize(mean_mcc = mean(mcc),
+            sd_mcc = sd(mcc),
+            mean_sec = mean(sec),
+            sd_sec = sd(sec),
+            mean_npshannon = mean(npshannon))
+```
+```{r mcc_opticlust}
+opticlust %>% 
+  ggplot(aes(dataset, mcc)) +
+  geom_jitter(alpha = 0.5) +
+  ylim(0.5, 1)
+```
+```{r runtime_opticlust}
+opticlust %>% 
+  ggplot(aes(num_seqs, sec, color = dataset)) +
+  geom_point(alpha = 0.5)
+opticlust %>% 
+  ggplot(aes(num_otus, sec, color = dataset)) +
+  geom_point(alpha = 0.5)
+opticlust %>% 
+  ggplot(aes(npshannon, sec, color = dataset)) +
+  geom_point(alpha = 0.5)
+```
+## fit to reference databases
+
+```{r load_fit-dbs}
 ref_sizes <- read_tsv(here('subworkflows/2_fit_reference_db/results/ref_sizes.tsv'))
-optifit_dbs <- read_tsv(here('subworkflows/2_fit_reference_db/results/optifit_dbs_results.tsv'))
+optifit_dbs <- read_tsv(here('subworkflows/2_fit_reference_db/results/optifit_dbs_results.tsv')) %>% 
+  mutate_perf()
+head(optifit_dbs)
 ```
 
-```{r}
-optifit_dbs
+```{r mcc_fit-db}
+optifit_dbs %>% 
+  ggplot(aes(x=method, y=mcc, color=ref)) +
+  geom_hline(aes(yintercept = mean_mcc), sum_opticlust) +
+  geom_jitter(alpha = 0.5) + 
+  facet_wrap('dataset') +
+  ylim(0, 1) +
+  labs(title='OTU Quality: OptiFit with reference databases',
+       caption='Black line: _de novo_ clustering') +
+  theme(plot.caption = element_markdown())
 ```
+
+```{r fraction-mapped_fit-db}
+optifit_dbs %>% filter(method == 'closed') %>% 
+  ggplot(aes(x=dataset, y=fraction_mapped, color=ref)) +
+  geom_jitter(alpha = 0.5) + 
+  ylim(0, 1) +
+  labs(title="Sequences mapped during closed-reference OptiFit") +
+  theme(plot.caption = element_markdown())
+```
+```{r runtime_fit-db}
+optifit_dbs %>% 
+  ggplot(aes(x=method, y=sec, color=ref)) +
+  geom_jitter(alpha = 0.5) +
+  geom_hline(aes(yintercept = mean_sec), sum_opticlust) +
+  facet_wrap('dataset', scales = 'free') +
+  labs(title='OptiFit runtime with reference databases',
+       caption='Black line: _de novo_ clustering on the whole dataset') +
+  theme(plot.caption = element_markdown())
+```
+
+## fit split datasets
+
+```{r load_fit-split}
+optifit_split <- read_tsv(here('subworkflows/3_fit_sample_split/results/optifit_split_results.tsv')) %>% 
+  mutate_perf()
+```
+
+```{r mcc_fit-split}
+optifit_split %>% 
+  ggplot(aes(x = ref_frac, y = mcc, color = ref_weight)) +
+  geom_jitter(size = 1, alpha = 0.3, width = 0.01) +
+  stat_summary(fun = mean, geom = 'crossbar', alpha = 0.5) +
+  geom_hline(aes(yintercept = mean_mcc), sum_opticlust) +
+  facet_grid(dataset ~ method) +
+  ylim(0, 1) +
+  labs(title='OTU Quality: OptiFit with split datasets',
+       x='reference fraction',
+       caption='Black line: _de novo_ clustering on the whole dataset') +
+  theme(plot.caption = element_markdown())
+```
+
+```{r runtime_fit-split}
+optifit_split %>% 
+  ggplot(aes(x = ref_frac, y = sec, color = ref_weight)) +
+  geom_jitter(size = 1, alpha = 0.3, width = 0.01) +
+  stat_summary(fun = mean, geom = 'crossbar', alpha = 0.5) +
+  geom_hline(aes(yintercept = mean_sec), sum_opticlust) +
+  facet_grid(dataset ~ method, scales = 'free') +
+  labs(title='Runtime: OptiFit with split datasets',
+       x='reference fraction',
+       caption='Black line: _de novo_ clustering on the whole dataset') +
+  theme(plot.caption = element_markdown())
+```
+
+```{r diversity_fit-split}
+optifit_split %>% 
+  ggplot(aes(x = ref_frac, y = npshannon, color = ref_weight)) +
+  geom_jitter(size = 1, alpha = 0.3, width = 0.01) +
+  stat_summary(fun = mean, geom = 'crossbar', alpha = 0.5) +
+  geom_hline(aes(yintercept = mean_npshannon), sum_opticlust) +
+  facet_grid(dataset ~ method, scales = 'free') +
+  labs(title='Diversity: OptiFit with split datasets',
+       x='reference fraction',
+       caption='Black line: _de novo_ clustering on the whole dataset') +
+  theme(plot.caption = element_markdown())
+```
+
+## vsearch
+
+For reference-based clustering, datasets were fit to the greengenes database.
+
+```{r load_vsearch}
+vsearch <- read_tsv(here('subworkflows/4_vsearch/results/vsearch_results.tsv')) %>% 
+  mutate_perf() %>% 
+  full_join(opticlust) %>% 
+  full_join(optifit_dbs %>% filter(ref == 'gg'))
+```
+
+```{r mcc_vsearch}
+vsearch %>% 
+  ggplot(aes(x = method, y = mcc, color = tool)) +
+  geom_boxplot() +
+  facet_wrap('dataset') +
+  ylim(0, 1) + 
+  labs(x = '')
+```
+
+```{r sec_vsearch}
+vsearch %>% 
+  ggplot(aes(x = method, y = sec, color = tool)) +
+  geom_boxplot() +
+  facet_wrap('dataset') +
+  labs(x = '')
+```
+
+```{r diversity_vsearch}
+vsearch %>% 
+  ggplot(aes(x = method, y = npshannon, color = tool)) +
+  geom_boxplot() +
+  facet_wrap('dataset') +
+  labs(x = '')
+```
+
+```{r fraction-mapped_vsearch}
+vsearch %>% filter(method == 'closed') %>% 
+  ggplot(aes(x = method, y = fraction_mapped, color = tool)) +
+  geom_boxplot() + 
+  facet_wrap('dataset') +
+  ylim(0, 1) +
+  labs(title="Sequences mapped during closed-reference OptiFit") +
+  theme(plot.caption = element_markdown())
+```

### 2020-11-11 exploratory/2020-11/plots.Rmd
Plot diversity for fit-dbs

@@
   geom_point(alpha = 0.5)
-opticlust %>% 
-  ggplot(aes(num_otus, sec, color = dataset)) +
-  geom_point(alpha = 0.5)
-opticlust %>% 
-  ggplot(aes(npshannon, sec, color = dataset)) +
-  geom_point(alpha = 0.5)
 ```
@@
   labs(title='OptiFit runtime with reference databases',
-       caption='Black line: _de novo_ clustering on the whole dataset') +
+       caption='Black line: _de novo_ clustering') +
+  theme(plot.caption = element_markdown())
+```
+```{r diversity_fit-db}
+optifit_dbs %>% 
+  ggplot(aes(x=method, y=npshannon, color=ref)) +
+  geom_jitter(alpha = 0.5) +
+  geom_hline(aes(yintercept = mean_npshannon), sum_opticlust) +
+  facet_wrap('dataset', scales = 'free') +
+  labs(title='OptiFit runtime with reference databases',
+       caption='Black line: _de novo_ clustering') +
   theme(plot.caption = element_markdown())
@@
 vsearch %>% filter(method == 'closed') %>% 
-  ggplot(aes(x = method, y = fraction_mapped, color = tool)) +
+  ggplot(aes(x = dataset, y = fraction_mapped, color = tool)) +
   geom_boxplot() + 
-  facet_wrap('dataset') +
   ylim(0, 1) +

### 2020-11-30 paper/paper.Rmd
Create rmarkdown::distill output

---
title: "**OptiFit: a fast method for fitting amplicon sequences to existing OTUs**"
date: '`r Sys.Date()`'
author:
  - first_name: "Kelly"
    last_name: "Sovacool"
    url: https://github.com/kelly-sovacool
    affiliation: Department of Computational Medicine and Bioinformatics, University of Michigan, Ann Arbor, MI
    orcid_id: 0000-0003-3283-829X
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
    includes:
      in_header: header.tex
csl: msystems.csl
fontsize: 12pt
geometry: margin=1.0in
---

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  tidy = TRUE,
  echo = FALSE,
  eval = TRUE,
  warning = FALSE,
  cache = FALSE
)

inline_hook <- function(x){
	print(x)
	if(is.numeric(x)){
		if(abs(x - round(x)) < .Machine$double.eps^0.5){
			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
		} else {
			paste(format(x,big.mark=',', digits=2, nsmall=2, scientific=FALSE))
		}

	} else {
    	paste(x)      
	}
}
knitr::knit_hooks$set(inline=inline_hook)
```

\begin{center}
\vspace{25mm}

Kelly L. Sovacool${^1}$, Sarah L. Westcott${^2}$, M. Brodie Mumphrey${^1}$,
Gabrielle A. Dotson${^1}$, Patrick D. Schloss${^2\dagger}$

$1$ Department of Computational Medicine and Bioinformatics,
University of Michigan, Ann Arbor, MI 48109

$2$ Department of Microbiology and Immunology,
University of Michigan, Ann Arbor, MI 48109

\vspace{30mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

\end{center}

\newpage
\linenumbers

## Abstract

Assigning amplicon sequences to Operational Taxonomic Units (OTUs) is an
important step in characterizing the composition of microbial communities across
large datasets. OptiClust, a _de novo_ OTU clustering method in the mothur
program, has been shown to produce higher quality OTU assignments than other
methods and at comparable or faster speeds [@westcott_opticlust_2017;
@schloss_introducing_2009]. A notable difference between _de novo_ clustering
and database-dependent methods is that OTU assignments clustered with _de novo_
methods are not stable when new sequences are added to a dataset
[@westcott_novo_2015]. However, in some cases one may wish to incorporate new
samples into a previously clustered dataset without performing clustering again
on all sequences, such as when deploying a machine learning model where OTUs are
features [@topcuoglu_framework_2020]. To provide an efficient and robust method
to fit amplicon sequence data to existing OTUs, we developed the OptiFit
algorithm as a new component of the mothur program. To benchmark the OptiFit
algorithm against _de novo_ clustering with the OptiClust algorithm, we used
four published datasets isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse [@schloss_stabilization_2012], and human
[@baxter_microbiota-based_2016] samples. For each dataset, a subset of sequences
was randomly selected and clustered into OTUs with OptiClust, then the remaining
sequences were fit to the existing OTUs using the OptiFit algorithm. This was
repeated with subsets of varying sizes ranging from 10 to 90% of sequences in
order to evaluate the bounds of the dataset size required for OptiFit.
Separately, all sequences were clustered with OptiClust to provide a baseline of
OTU assignment quality and runtime performance. Each of these routines was
repeated 10 times with different random seeds to produce results that are robust
to random variation. OTU quality was evaluated using the Matthews Correlation
Coefficient (MCC) with a sequence similarity threshold of 97% as described
previously [@schloss_application_2016; @westcott_novo_2015]. On average, fitting
sequences into existing OTUs with OptiFit performed 10 times faster than _de
novo_ clustering with OptiClust, while the average MCC scores produced were
nearly indistinguishable across each dataset. The OptiFit results across subset
sizes ranging from 10 to 90% of sequences were also very similar, with slightly
higher MCC scores for larger subset sizes. Thus, OptiFit is an efficient way to
fit new sequences to existing OTUs yet without sacrificing the quality of OTU
assignments.

### Importance

\newpage

## Introduction

## Results

## Discussion

## Materials and Methods

## Acknowledgements

KLS received support from the NIH Training Program in Bioinformatics (T32 GM070449).

The funders had no role in study design, data collection and interpretation, 
or the decision to submit the work for publication.

## References

### 2020-11-30 code/R/render.R

@@ -1,4 +1,8 @@
 source(snakemake@input[["fcns"]])
 
 log_smk()
-rmarkdown::render(here::here(snakemake@input[["Rmd"]]))
+rmarkdown::render(here::here(snakemake@input[["Rmd"]]),
+                  output_format = 'pdf_document')
+rmarkdown::render(here::here(snakemake@input[["Rmd"]]), 
+                  output_format = 'distill::distill_article', 
+                  output_dir = 'docs/')

### 2020-11-30 code/R/render.R
🎨 Style R code ⬡

@@ -2,7 +2,9 @@ source(snakemake@input[["fcns"]])
 
 log_smk()
 rmarkdown::render(here::here(snakemake@input[["Rmd"]]),
-                  output_format = 'pdf_document')
-rmarkdown::render(here::here(snakemake@input[["Rmd"]]), 
-                  output_format = 'distill::distill_article', 
-                  output_dir = 'docs/')
+  output_format = "pdf_document"
+)
+rmarkdown::render(here::here(snakemake@input[["Rmd"]]),
+  output_format = "distill::distill_article",
+  output_dir = "docs/"
+)

### 2020-11-30 code/R/uc_to_list.R

@@ -1,32 +1,42 @@
 # original source: https://github.com/SchlossLab/Schloss_Cluster_PeerJ_2015/blob/master/code/uc_to_list.R
 # Modified Nov. 2020 by KLS
 library(tidyverse)
-uc_to_list <- function(clustered_file_name, 
-                       list_file_name, 
-                       label = 0.03){
-
-	clustered <- read.table(file = clustered_file_name, 
-	                        stringsAsFactors = FALSE) %>% 
-	  mutate(seq_id = str_replace(V9, '(.*);size=\\d+$', '\\1'), 
-	         otu_id = paste0('OTU_', V2),
-	         record_type = V1) %>%
-	  filter(record_type %in% c('S', 'H')) %>% 
-	  select(seq_id, otu_id)
-	otu_ids <- clustered %>% pull(otu_id) %>% unique()
-	num_otus <- length(otu_ids)
-	otus <- sapply(otu_ids, function(x) { clustered %>% 
-	    filter(otu_id == x) %>% 
-	    pull(seq_id) %>% 
-	    paste(collapse = ',')})
-	# storing this in memory as a giant string is a bad idea, 
-	# but I only have to run it once per vsearch clustering job,
-	# so I don't really care.
-	list_data <- paste(paste(c("label", "numOTUs", otu_ids),
-	                         collapse = "\t"),
-	                   paste(c(label, num_otus, otus),
-	                         collapse = "\t"),
-	                   sep = "\n")
-	write(list_data, list_file_name)
+uc_to_list <- function(clustered_file_name,
+                       list_file_name,
+                       label = 0.03) {
+  clustered <- read.table(
+    file = clustered_file_name,
+    stringsAsFactors = FALSE
+  ) %>%
+    mutate(
+      seq_id = str_replace(V9, "(.*);size=\\d+$", "\\1"),
+      otu_id = paste0("OTU_", V2),
+      record_type = V1
+    ) %>%
+    filter(record_type %in% c("S", "H")) %>%
+    select(seq_id, otu_id)
+  otu_ids <- clustered %>%
+    pull(otu_id) %>%
+    unique()
+  num_otus <- length(otu_ids)
+  otus <- sapply(otu_ids, function(x) {
+    clustered %>%
+      filter(otu_id == x) %>%
+      pull(seq_id) %>%
+      paste(collapse = ",")
+  })
+  # storing this in memory as a giant string is a bad idea,
[…]

### 2020-11-30 code/py/rbind_files.py
🎨 Style Python code 🐍

@@ -1,11 +1,11 @@
 # perform row bind, but don't hold data in memory.
 # assumes all tsv files have the same columns.
 
-with open(snakemake.output.tsv, 'w') as outfile:
-    header_row = ''
+with open(snakemake.output.tsv, "w") as outfile:
+    header_row = ""
     for infilename in snakemake.input.tsv:
-        with open(infilename, 'r') as infile:
-            line = next(infile) # throw out header row except first time
+        with open(infilename, "r") as infile:
+            line = next(infile)  # throw out header row except first time
             if not header_row:
                 header_row = line
                 outfile.write(header_row)

### 2020-11-30 paper/paper.Rmd
Render paper to docs/index.html

@@
 author:
-  - first_name: "Kelly"
-    last_name: "Sovacool"
+  - name: "Kelly L. Sovacool"
     url: https://github.com/kelly-sovacool
@@
 geometry: margin=1.0in
+repository_url: https
 ---

### 2020-11-30 code/R/render.R

@@ -5,4 +5,4 @@ rmarkdown::render(here::here(snakemake@input[["Rmd"]]),
                   output_format = 'pdf_document')
 rmarkdown::render(here::here(snakemake@input[["Rmd"]]), 
                   output_format = 'distill::distill_article', 
-                  output_dir = 'docs/')
+                  output_file = here::here('docs','index.html'))

### 2020-11-30 code/R/render.R
🎨 Style R code ⬡

@@ -2,7 +2,9 @@ source(snakemake@input[["fcns"]])
 
 log_smk()
 rmarkdown::render(here::here(snakemake@input[["Rmd"]]),
-                  output_format = 'pdf_document')
-rmarkdown::render(here::here(snakemake@input[["Rmd"]]), 
-                  output_format = 'distill::distill_article', 
-                  output_file = here::here('docs','index.html'))
+  output_format = "pdf_document"
+)
+rmarkdown::render(here::here(snakemake@input[["Rmd"]]),
+  output_format = "distill::distill_article",
+  output_file = here::here("docs", "index.html")
+)

### 2020-11-30 code/R/render.R
Render pdf to docs/

@@ -2,7 +2,8 @@ source(snakemake@input[["fcns"]])
 
 log_smk()
 rmarkdown::render(here::here(snakemake@input[["Rmd"]]),
-                  output_format = 'pdf_document')
+                  output_format = 'pdf_document',
+                  output_file = here::here('docs', 'paper.pdf'))
 rmarkdown::render(here::here(snakemake@input[["Rmd"]]), 
                   output_format = 'distill::distill_article', 
                   output_file = here::here('docs','index.html'))

### 2020-11-30 paper/paper.Rmd
Fun with distill metadata
https://rstudio.github.io/distill/metadata.html

@@
     orcid_id: 0000-0003-3283-829X
+  - name: "Sarah L. Westcott"
+    url: https://github.com/mothur-westcott
+    affiliation: Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI
+  - name: "M. Brodie Mumphrey"
+    url: https://github.com/MBMumphrey
+    affiliation: Department of Computational Medicine and Bioinformatics, University of Michigan, Ann Arbor, MI
+  - name: "Gabrielle A. Dotson"
+    url: https://github.com/dotsonga
+    affiliation: Department of Computational Medicine and Bioinformatics, University of Michigan, Ann Arbor, MI
+  - name: "Patrick D. Schloss"
+    url: https://github.com/pschloss
+    affiliation: Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI
 bibliography: references.bib
@@
   pdf_document:
-    keep_tex: true
+    keep_tex: false
     includes:
@@
 geometry: margin=1.0in
-repository_url: https
+repository_url: https://github.com/SchlossLab/OptiFitAnalysis
+creative_commons: CC BY-SA
+twitter:
+  creator: "@kelly_sovacool"
+  site: "@PatSchloss"
 ---

### 2020-12-01 paper/paper.Rmd
Use distill article

@@
 date: '`r Sys.Date()`'
-author:
+authors:
   - name: "Kelly L. Sovacool"
@@
       in_header: header.tex
+  distill::distill_article:
+    toc: true
+    toc_depth: 2
 csl: msystems.csl
@@
 
-## Acknowledgements
+## Acknowledgements {.appendix}
 
@@
 
+## Author Contributions {.appendix}
+
 ## References

### 2020-12-04 paper/paper.Rmd
Reorganize paper I/O

@@
 output:
+  distill::distill_article:
+    toc: true
+    toc_float: true
+    toc_depth: 2
   pdf_document:
@@
     includes:
-      in_header: header.tex
-  distill::distill_article:
-    toc: true
-    toc_depth: 2
+      in_header: preamble.tex
+      before_body: head.tex
+      after_body: tail.tex
 csl: msystems.csl
@@
 ```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
-knitr::opts_chunk$set(
-  tidy = TRUE,
-  echo = FALSE,
-  eval = TRUE,
-  warning = FALSE,
-  cache = FALSE
-)
-
-inline_hook <- function(x){
-	print(x)
-	if(is.numeric(x)){
-		if(abs(x - round(x)) < .Machine$double.eps^0.5){
-			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
-		} else {
-			paste(format(x,big.mark=',', digits=2, nsmall=2, scientific=FALSE))
-		}
-
-	} else {
-    	paste(x)      
-	}
-}
-knitr::knit_hooks$set(inline=inline_hook)
+mothuR::set_knitr_opts()
 ```
 
-\begin{center}
-\vspace{25mm}
-
-Kelly L. Sovacool${^1}$, Sarah L. Westcott${^2}$, M. Brodie Mumphrey${^1}$,
-Gabrielle A. Dotson${^1}$, Patrick D. Schloss${^2\dagger}$
-
-$1$ Department of Computational Medicine and Bioinformatics,
-University of Michigan, Ann Arbor, MI 48109
-
-$2$ Department of Microbiology and Immunology,
-University of Michigan, Ann Arbor, MI 48109
-
-\vspace{30mm}
-
-$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu
-
-\end{center}
-
-\newpage
-\linenumbers
-
 ## Abstract
-
-Assigning amplicon sequences to Operational Taxonomic Units (OTUs) is an
-important step in characterizing the composition of microbial communities across
-large datasets. OptiClust, a _de novo_ OTU clustering method in the mothur
-program, has been shown to produce higher quality OTU assignments than other
-methods and at comparable or faster speeds [@westcott_opticlust_2017;
-@schloss_introducing_2009]. A notable difference between _de novo_ clustering
-and database-dependent methods is that OTU assignments clustered with _de novo_
-methods are not stable when new sequences are added to a dataset
-[@westcott_novo_2015]. However, in some cases one may wish to incorporate new
-samples into a previously clustered dataset without performing clustering again
-on all sequences, such as when deploying a machine learning model where OTUs are
-features [@topcuoglu_framework_2020]. To provide an efficient and robust method
-to fit amplicon sequence data to existing OTUs, we developed the OptiFit
-algorithm as a new component of the mothur program. To benchmark the OptiFit
-algorithm against _de novo_ clustering with the OptiClust algorithm, we used
-four published datasets isolated from soil [@johnston_metagenomics_2016], marine
-[@henson_artificial_2016], mouse [@schloss_stabilization_2012], and human
-[@baxter_microbiota-based_2016] samples. For each dataset, a subset of sequences
-was randomly selected and clustered into OTUs with OptiClust, then the remaining
-sequences were fit to the existing OTUs using the OptiFit algorithm. This was
-repeated with subsets of varying sizes ranging from 10 to 90% of sequences in
-order to evaluate the bounds of the dataset size required for OptiFit.
-Separately, all sequences were clustered with OptiClust to provide a baseline of
-OTU assignment quality and runtime performance. Each of these routines was
-repeated 10 times with different random seeds to produce results that are robust
-to random variation. OTU quality was evaluated using the Matthews Correlation
-Coefficient (MCC) with a sequence similarity threshold of 97% as described
-previously [@schloss_application_2016; @westcott_novo_2015]. On average, fitting
-sequences into existing OTUs with OptiFit performed 10 times faster than _de
-novo_ clustering with OptiClust, while the average MCC scores produced were
-nearly indistinguishable across each dataset. The OptiFit results across subset
-sizes ranging from 10 to 90% of sequences were also very similar, with slightly
-higher MCC scores for larger subset sizes. Thus, OptiFit is an efficient way to
-fit new sequences to existing OTUs yet without sacrificing the quality of OTU
-assignments.
 
@@
 ## Author Contributions {.appendix}
-
-## References

### 2020-12-04 code/R/render.R

@@ -1,11 +1,6 @@
 source(snakemake@input[["fcns"]])
-
 log_smk()
 rmarkdown::render(here::here(snakemake@input[["Rmd"]]),
-  output_format = "pdf_document",
-  output_file = here::here("docs", "paper.pdf")
-)
-rmarkdown::render(here::here(snakemake@input[["Rmd"]]),
-  output_format = "distill::distill_article",
-  output_file = here::here("docs", "index.html")
+  output_format = snakemake@params[['format']],
+  output_file = here::here(snakemake@output[['file']])
 )

### 2020-12-04 code/R/render.R
🎨 Style R code ⬡

@@ -1,6 +1,6 @@
 source(snakemake@input[["fcns"]])
 log_smk()
 rmarkdown::render(here::here(snakemake@input[["Rmd"]]),
-  output_format = snakemake@params[['format']],
-  output_file = here::here(snakemake@output[['file']])
+  output_format = snakemake@params[["format"]],
+  output_file = here::here(snakemake@output[["file"]])
 )

### 2020-12-04 paper/paper.Rmd
Override knit button to put output in docs/ dir

@@
       after_body: tail.tex
+knit: (function(inputFile, encoding) {
+  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs", output_format = "all") })
 csl: msystems.csl

### 2020-12-08 paper/paper.Rmd
Independent db outline resutls

@@
 
+- 16S rRNA gene sequence datasets from human gut, mouse gut, marine, and soil environments were downloaded, processed with mothur, and clustered _de novo_ with OptiClust.
+- Independent reference databases (greengenes, silva, and RDP) were clustered _de novo_, then query datasets were fit to the _de novo_ OTUs using OptiFit.
+  - In open-reference mode, OTU quality as measured by the MCC score was similar between fitting the datasets to reference OTUs with OptiFit and clustering the datasets _de novo_ with OptiClust. 
+  - However, in closed-reference mode, OTU quality was slightly worse when fitting to greengenes and silva, and much worse when fitting to RDP as compared to OptiClust.
+OptiFit was able to map more query sequences to reference OTUs created with the greengenes and silva databases than with RDP.
+  - These results held true for all four datasets and all three reference databases.
+- Datasets were randomly split into a reference fraction and a query fraction.
+  - Reference sequences were clustered _de novo_ with OptiClust.
+  - Query sequences were then fit to the _de novo_ OTUs with OptiFit.
+- vsearch with greengenes
+
 ## Discussion

### 2020-12-08 paper/paper.Rmd
Outline split datasets results

@@
   - Query sequences were then fit to the _de novo_ OTUs with OptiFit.
+  - Reference sizes from 10% to 80% of the sequences were created, with the remaining sequences used for the query.
+  - OTU quality from fitting split datasets was highly similar to that from _de novo_ clustering the whole dataset.
+  - Closed-reference OptiFit with split datasets was faster than OptiClust on whole datasets.
+    - OptiClust performed faster than open-reference OptiFit only when the OptiFit reference fraction was 30% or less.
+  - Different methods for selecting the sequences to be used as the reference were tested; simple random sample, weighting sequences by relative abundance, and weighting by similarity to other sequences in the dataset.
+    - OTU quality was similar with the simple and abundance-weighted sampling, but slightly worse with similarity-weighted sampling.
 - vsearch with greengenes

### 2020-12-08 paper/paper.Rmd
Outline vsearch results

@@
 OptiFit was able to map more query sequences to reference OTUs created with the greengenes and silva databases than with RDP.
+  - In terms of runtime, closed-reference OptiFit outperformed OptiClust, while OptiClust out-performed open-reference OptiFit.
   - These results held true for all four datasets and all three reference databases.
 - Datasets were randomly split into a reference fraction and a query fraction.
-  - Reference sequences were clustered _de novo_ with OptiClust.
-  - Query sequences were then fit to the _de novo_ OTUs with OptiFit.
-  - Reference sizes from 10% to 80% of the sequences were created, with the remaining sequences used for the query.
+Reference sizes from 10% to 80% of the sequences were created, with the remaining sequences used for the query.
+Reference sequences were clustered _de novo_ with OptiClust.
+Query sequences were then fit to the _de novo_ OTUs with OptiFit.
   - OTU quality from fitting split datasets was highly similar to that from _de novo_ clustering the whole dataset.
@@
     - OptiClust performed faster than open-reference OptiFit only when the OptiFit reference fraction was 30% or less.
-  - Different methods for selecting the sequences to be used as the reference were tested; simple random sample, weighting sequences by relative abundance, and weighting by similarity to other sequences in the dataset.
+  - Different methods for selecting the sequences to be used as the reference were tested; a simple random sample, weighting sequences by relative abundance, and weighting by similarity to other sequences in the dataset.
     - OTU quality was similar with the simple and abundance-weighted sampling, but slightly worse with similarity-weighted sampling.
-- vsearch with greengenes
+    - TODO: fraction mapped
+- To compare to existing software, vsearch was used to cluster OTUs _de novo_ or with reference-based clustering to the greengenes database.
+  - For all datasets and clustering methods (_de novo_, open reference, and closed reference), mothur's clustering algorithms produced higher quality OTUs than vsearch.
+  - When closed-reference clustering against the greengenes database, vsearch was able to map more query sequences to the reference than mothur's OptiFit algorithm.
+  - In terms of runtime, OptiFit genearlly performed faster than vsearch when reference clustering, while vsearch _de novo_ clustering outperformed OptiClust.
 

### 2020-12-08 paper/paper.Rmd
Add headings

@@
 - 16S rRNA gene sequence datasets from human gut, mouse gut, marine, and soil environments were downloaded, processed with mothur, and clustered _de novo_ with OptiClust.
+
+## Reference clustering with independent databases
+
 - Independent reference databases (greengenes, silva, and RDP) were clustered _de novo_, then query datasets were fit to the _de novo_ OTUs using OptiFit.
@@
   - These results held true for all four datasets and all three reference databases.
+
+## Reference clustering with split datasets
+
 - Datasets were randomly split into a reference fraction and a query fraction.
@@
     - TODO: fraction mapped
+
+## Comparison to vsearch
+
 - To compare to existing software, vsearch was used to cluster OTUs _de novo_ or with reference-based clustering to the greengenes database.

### 2020-12-08 exploratory/2020-11/plots.Rmd
Outline fraction mapped for split datasets

@@
 ---
-date: 'Nov. 2020'
+date: 'Nov. - Dec. 2020'
 output: 
@@
 ```
+```{r fraction-mapped_fit-split}
+optifit_split %>% 
+  filter(method == 'closed', tool == 'mothur', is.na(ref)) %>% 
+  #group_by(dataset, ref_weight, ref_frac) %>% 
+  ggplot(aes(x=ref_frac, y=fraction_mapped, color=ref_weight)) +
+  geom_jitter(alpha = 0.5, width = 0.01, size=1) +
+  facet_wrap("dataset", nrow=1) +
+  ylim(0, 1) +
+  labs(title="Sequences mapped during closed-reference OptiFit",
+       x='reference fraction',
+       y='fraction mapped')
+```
 

### 2020-12-08 paper/paper.Rmd
Smaller subheadings
Outline fraction mapped for split datasets

@@
 
-- 16S rRNA gene sequence datasets from human gut, mouse gut, marine, and soil environments were downloaded, processed with mothur, and clustered _de novo_ with OptiClust.
+- 16S rRNA gene sequence datasets from human gut, mouse gut, marine, and soil environments were processed with mothur and clustered _de novo_ with mothur's OptiClust algorithm.
+- All clustering was performed as a distance threshold of 0.03.
+- The Matthews Correlation Coefficient was calculated to serve as a measure of OTU quality.
 
-## Reference clustering with independent databases
+### Reference clustering with public databases
 
-- Independent reference databases (greengenes, silva, and RDP) were clustered _de novo_, then query datasets were fit to the _de novo_ OTUs using OptiFit.
-  - In open-reference mode, OTU quality as measured by the MCC score was similar between fitting the datasets to reference OTUs with OptiFit and clustering the datasets _de novo_ with OptiClust. 
+- Public reference databases (greengenes, silva, and RDP) were clustered _de novo_ using OptiClust, then query datasets were fit to the _de novo_ OTUs using OptiFit.
+  - In open-reference mode, OTU quality was similar between fitting the datasets to reference OTUs with OptiFit and clustering the datasets _de novo_ with OptiClust. 
   - However, in closed-reference mode, OTU quality was slightly worse when fitting to greengenes and silva, and much worse when fitting to RDP as compared to OptiClust.
@@
 
-## Reference clustering with split datasets
+### Reference clustering with split datasets
 
@@
     - OTU quality was similar with the simple and abundance-weighted sampling, but slightly worse with similarity-weighted sampling.
-    - TODO: fraction mapped
+    - The fraction of query sequences that are able to be fit to the reference OTUs in closed-reference mode decreases as the reference fraction increases.
 
-## Comparison to vsearch
+### Comparison to vsearch
 
@@
 
+### The OptiFit Algorithm
+
+### Analysis Workflow
+
 ## Acknowledgements {.appendix}

### 2020-12-08 paper/paper.Rmd
Describe dbs further up

@@
 
-- 16S rRNA gene sequence datasets from human gut, mouse gut, marine, and soil environments were processed with mothur and clustered _de novo_ with mothur's OptiClust algorithm.
-- All clustering was performed as a distance threshold of 0.03.
+- 16S rRNA gene sequence datasets from human gut, mouse gut, marine, and soil environments as well as the greengenes, silva, and RDP public databases were processed with mothur and clustered _de novo_ with mothur's OptiClust algorithm.
+- All clustering was performed at a distance threshold of 0.03 and repeated with 100 different random seeds.
 - The Matthews Correlation Coefficient was calculated to serve as a measure of OTU quality.
@@
 
-- Public reference databases (greengenes, silva, and RDP) were clustered _de novo_ using OptiClust, then query datasets were fit to the _de novo_ OTUs using OptiFit.
+- Public reference databases were clustered _de novo_ using OptiClust, then query datasets were fit to the _de novo_ OTUs using OptiFit.
   - In open-reference mode, OTU quality was similar between fitting the datasets to reference OTUs with OptiFit and clustering the datasets _de novo_ with OptiClust. 

### 2020-12-08 exploratory/2020-11/plots.Rmd
Create fit split fraction mapped plot

--- 2020-11-11 exploratory/2020-11/plots.Rmd
@@
 
+### fraction mapped
+
+```{r fraction-mapped_fit-split}
+optifit_split %>% 
+  filter(method == 'closed', tool == 'mothur', is.na(ref)) %>% 
+  #group_by(dataset, ref_weight, ref_frac) %>% 
+  ggplot(aes(x=ref_frac, y=fraction_mapped, color=ref_weight)) +
+  geom_jitter(alpha = 0.5, width = 0.01, size=1) +
+  facet_wrap("dataset", nrow=1) +
+  ylim(0, 1) +
+  labs(title="Sequences mapped during closed-reference OptiFit",
+       x='reference fraction',
+       y='fraction mapped')
+```
+
+It looks like one result per parameter set is left over from before I re-ran 
+everything after revering to the older silva version.
+[Here's what the plot looked like with the newest (buggy) silva](https://github.com/SchlossLab/OptiFitAnalysis/blob/master/exploratory/2020-05/sub3_fit_all-seqs.md#fraction-of-sequences-that-map-to-the-reference).
+Need to go back and figure out why those files didn't get removed/overwritten
+by Snakemake when I re-ran everything.
+
 ```{r diversity_fit-split}

### 2020-12-08 paper/paper.Rmd
Minor edits

@@
   - In open-reference mode, OTU quality was similar between fitting the datasets to reference OTUs with OptiFit and clustering the datasets _de novo_ with OptiClust.
+This held true for all datasets and reference databases.
   - However, in closed-reference mode, OTU quality was slightly worse when fitting to greengenes and silva, and much worse when fitting to RDP as compared to OptiClust.
@@
   - In terms of runtime, closed-reference OptiFit outperformed OptiClust, while OptiClust out-performed open-reference OptiFit.
-  - These results held true for all four datasets and all three reference databases.
 
@@
 Reference sizes from 10% to 80% of the sequences were created, with the remaining sequences used for the query.
-Reference sequences were clustered _de novo_ with OptiClust.
-Query sequences were then fit to the _de novo_ OTUs with OptiFit.
+Reference sequences were clustered _de novo_ with OptiClust, then query sequences were fit to the _de novo_ OTUs with OptiFit.
   - OTU quality from fitting split datasets was highly similar to that from _de novo_ clustering the whole dataset.

### 2020-12-10 exploratory/2020-11/plots.Rmd
Plot fraction mapped with open-ref as sanity check

@@
 It looks like one result per parameter set is left over from before I re-ran 
-everything after revering to the older silva version.
+everything after reverting to the older silva version.
 [Here's what the plot looked like with the newest (buggy) silva](https://github.com/SchlossLab/OptiFitAnalysis/blob/master/exploratory/2020-05/sub3_fit_all-seqs.md#fraction-of-sequences-that-map-to-the-reference).
@@
 by Snakemake when I re-ran everything.
+
+Let's try a sanity check with open-reference -- should all be 1.
+
+```{r fraction-mapped-open_fit-split}
+optifit_split %>% 
+  filter(method == 'open', tool == 'mothur', is.na(ref)) %>% 
+  #group_by(dataset, ref_weight, ref_frac) %>% 
+  ggplot(aes(x=ref_frac, y=fraction_mapped, color=ref_weight)) +
+  geom_jitter(alpha = 0.5, width = 0.01, size=1) +
+  facet_wrap("dataset", nrow=1) +
+  ylim(0, 1) +
+  labs(title="Sequences mapped during open-reference OptiFit",
+       x='reference fraction',
+       y='fraction mapped')
+```
+
+Did mothur's list file format change in the latest dev version?
 

### 2020-12-10 paper/head.tex

--- 2020-09-09 paper/paper.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  12pt,
-]{article}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={OptiFit: a fast method for fitting amplicon sequences to existing OTUs},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage[margin=1.0in]{geometry}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-\usepackage{helvet} % Helvetica font
-\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
-\usepackage[T1]{fontenc}
-
-\usepackage[none]{hyphenat}
-
-\usepackage{setspace}
-\doublespacing
-\setlength{\parskip}{1em}
-
-\usepackage{lineno}
-
-\usepackage{pdfpages}
-\usepackage{comment}
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newenvironment{cslreferences}%
-  {}%
-  {\par}
-
-\title{\textbf{OptiFit: a fast method for fitting amplicon sequences to
-existing OTUs}}
-\author{}
-\date{\vspace{-2.5em}}
-
-\begin{document}
-\maketitle
-
 \begin{center}
@@
 
-Kelly L. Sovacool${^1}$, Sarah L. Westcott${^2}$, M. Brodie Mumphrey${^1}$,
-Gabrielle A. Dotson${^1}$, Patrick D. Schloss${^2\dagger}$
+Kelly L. Sovacool${^1}$, Sarah L. Westcott${^2}$, M. Brodie Mumphrey${^1}$, Gabrielle A. Dotson${^1}$, Patrick D. Schloss${^2\dagger}$
 
@@
 \linenumbers
-
-\hypertarget{abstract}{%
-\subsection{Abstract}\label{abstract}}
-
-Assigning amplicon sequences to Operational Taxonomic Units (OTUs) is an
-important step in characterizing the composition of microbial
-communities across large datasets. OptiClust, a \emph{de novo} OTU
-clustering method in the mothur program, has been shown to produce
-higher quality OTU assignments than other methods and at comparable or
-faster speeds (1, 2). A notable difference between \emph{de novo}
-clustering and database-dependent methods is that OTU assignments
-clustered with \emph{de novo} methods are not stable when new sequences
-are added to a dataset (3). However, in some cases one may wish to
-incorporate new samples into a previously clustered dataset without
-performing clustering again on all sequences, such as when deploying a
-machine learning model where OTUs are features (4). To provide an
-efficient and robust method to fit amplicon sequence data to existing
-OTUs, we developed the OptiFit algorithm as a new component of the
-mothur program. To benchmark the OptiFit algorithm against \emph{de
-novo} clustering with the OptiClust algorithm, we used four published
-datasets isolated from soil (5), marine (6), mouse (7), and human (8)
-samples. For each dataset, a subset of sequences was randomly selected
-and clustered into OTUs with OptiClust, then the remaining sequences
-were fit to the existing OTUs using the OptiFit algorithm. This was
-repeated with subsets of varying sizes ranging from 10 to 90\% of
-sequences in order to evaluate the bounds of the dataset size required
-for OptiFit. Separately, all sequences were clustered with OptiClust to
-provide a baseline of OTU assignment quality and runtime performance.
-Each of these routines was repeated 10 times with different random seeds
-to produce results that are robust to random variation. OTU quality was
-evaluated using the Matthews Correlation Coefficient (MCC) with a
-sequence similarity threshold of 97\% as described previously (3, 9). On
-average, fitting sequences into existing OTUs with OptiFit performed 10
-times faster than \emph{de novo} clustering with OptiClust, while the
-average MCC scores produced were nearly indistinguishable across each
-dataset. The OptiFit results across subset sizes ranging from 10 to 90\%
-of sequences were also very similar, with slightly higher MCC scores for
-larger subset sizes. Thus, OptiFit is an efficient way to fit new
-sequences to existing OTUs yet without sacrificing the quality of OTU
-assignments.
-
-\hypertarget{importance}{%
-\subsubsection{Importance}\label{importance}}
-
-\newpage
-
-\hypertarget{introduction}{%
-\subsection{Introduction}\label{introduction}}
-
-\hypertarget{results}{%
-\subsection{Results}\label{results}}
-
-\hypertarget{discussion}{%
-\subsection{Discussion}\label{discussion}}
-
-\hypertarget{materials-and-methods}{%
-\subsection{Materials and Methods}\label{materials-and-methods}}
-
-\hypertarget{acknowledgements}{%
-\subsection{Acknowledgements}\label{acknowledgements}}
-
-KLS received support from the NIH Training Program in Bioinformatics
-(T32 GM070449).
-
-The funders had no role in study design, data collection and
-interpretation, or the decision to submit the work for publication.
-
-\hypertarget{references}{%
-\subsection*{References}\label{references}}
-\addcontentsline{toc}{subsection}{References}
-
-\hypertarget{refs}{}
-\begin{cslreferences}
-\leavevmode\hypertarget{ref-westcott_opticlust_2017}{}%
-1. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
-Improved Method for Assigning Amplicon-Based Sequence Data to
-Operational Taxonomic Units. mSphere \textbf{2}:e00073--17.
-doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.
-
-\leavevmode\hypertarget{ref-schloss_introducing_2009}{}%
-2. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
-\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
-\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
-\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
-\textbf{Thallinger GG}, \textbf{Van Horn DJ}, \textbf{Weber CF}. 2009.
-Introducing mothur: Open-source, platform-independent,
-community-supported software for describing and comparing microbial
-communities. Applied and Environmental Microbiology
-\textbf{75}:7537--7541.
-doi:\href{https://doi.org/10.1128/AEM.01541-09}{10.1128/AEM.01541-09}.
-
-\leavevmode\hypertarget{ref-westcott_novo_2015}{}%
-3. \textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo clustering
-methods outperform reference-based methods for assigning 16S rRNA gene
-sequences to operational taxonomic units. PeerJ \textbf{3}:e1487.
-doi:\href{https://doi.org/10.7717/peerj.1487}{10.7717/peerj.1487}.
-
-\leavevmode\hypertarget{ref-topcuoglu_framework_2020}{}%
-4. \textbf{Topçuoğlu BD}, \textbf{Lesniak NA}, \textbf{Ruffin MT},
-\textbf{Wiens J}, \textbf{Schloss PD}. 2020. A Framework for Effective
-Application of Machine Learning to Microbiome-Based Classification
-Problems. mBio \textbf{11}.
-doi:\href{https://doi.org/10.1128/mBio.00434-20}{10.1128/mBio.00434-20}.
-
-\leavevmode\hypertarget{ref-johnston_metagenomics_2016}{}%
-5. \textbf{Johnston ER}, \textbf{Rodriguez-R LM}, \textbf{Luo C},
-\textbf{Yuan MM}, \textbf{Wu L}, \textbf{He Z}, \textbf{Schuur EAG},
-\textbf{Luo Y}, \textbf{Tiedje JM}, \textbf{Zhou J},
-\textbf{Konstantinidis KT}. 2016. Metagenomics Reveals Pervasive
-Bacterial Populations and Reduced Community Diversity across the Alaska
-Tundra Ecosystem. Front Microbiol \textbf{7}.
-doi:\href{https://doi.org/10.3389/fmicb.2016.00579}{10.3389/fmicb.2016.00579}.
-
-\leavevmode\hypertarget{ref-henson_artificial_2016}{}%
-6. \textbf{Henson MW}, \textbf{Pitre DM}, \textbf{Weckhorst JL},
-\textbf{Lanclos VC}, \textbf{Webber AT}, \textbf{Thrash JC}. 2016.
-Artificial Seawater Media Facilitate Cultivating Members of the
-Microbial Majority from the Gulf of Mexico. mSphere \textbf{1}.
-doi:\href{https://doi.org/10.1128/mSphere.00028-16}{10.1128/mSphere.00028-16}.
-
-\leavevmode\hypertarget{ref-schloss_stabilization_2012}{}%
-7. \textbf{Schloss PD}, \textbf{Schubert AM}, \textbf{Zackular JP},
-\textbf{Iverson KD}, \textbf{Young VB}, \textbf{Petrosino JF}. 2012.
-Stabilization of the murine gut microbiome following weaning. Gut
-Microbes \textbf{3}:383--393.
-doi:\href{https://doi.org/10.4161/gmic.21008}{10.4161/gmic.21008}.
-
-\leavevmode\hypertarget{ref-baxter_microbiota-based_2016}{}%
-8. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
-\textbf{Schloss PD}. 2016. Microbiota-based model improves the
-sensitivity of fecal immunochemical test for detecting colonic lesions.
-Genome Med \textbf{8}:37.
-doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
-
-\leavevmode\hypertarget{ref-schloss_application_2016}{}%
-9. \textbf{Schloss PD}. 2016. Application of a Database-Independent
-Approach To Assess the Quality of Operational Taxonomic Unit Picking
-Methods. mSystems \textbf{1}:e00027--16.
-doi:\href{https://doi.org/10.1128/mSystems.00027-16}{10.1128/mSystems.00027-16}.
-\end{cslreferences}
-
-\end{document}

### 2020-12-10 paper/paper.Rmd

@@
 
+TODO: Describe OptiClust and vsearch.
+
 ## Results
+
+TODO: brief description of OptiFit algorithm. the method we came up with is the result. OptiClust paper had toy example to walk through the algorithm.
+
+TODO: Paragraph at beginning to describe datasets: why picked, how processed
 
@@
 
+TODO: compare optifit performance for split dataset vs public database
+TODO: double-check with Sarah that MCCs are from full OTU dataset or just query sequnce OTUs
+TODO: double-check reference fractions aren't flipped (fraction mapped plot looks weird)
+
 - Datasets were randomly split into a reference fraction and a query fraction.
@@
 ### Comparison to vsearch
+
+TODO: move this into first section with public database.
+vsearch is more permissive than mothur (radius vs diameter with 0.03 dissimilarity threshold).
 

### 2020-12-10 code/py/fraction_reads_mapped.py
Debugging -- list file only outputs reference seq ids

@@ -28,14 +28,16 @@ def main():
     ref_seqs = parse_seqs(snakemake.input.ref)
     mapped_seqs = parse_list(snakemake.input.list)
 
+    print('map=ref',mapped_seqs == ref_seqs)
+    print('ref in map', len(ref_seqs.intersection(mapped_seqs)))
+    print('len(map)', len(mapped_seqs))
     # remove reference seqs from mapped seqs
     mapped_seqs = mapped_seqs - ref_seqs
-
-    print('map=ref',mapped_seqs == ref_seqs)
+    print('len(map) after remove ref seqs', len(mapped_seqs))
 
     # mapped reads should not contain reads not in query
     assert not (mapped_seqs - query_seqs)
-
+    
     fraction_mapped = round(
         len(mapped_seqs.intersection(query_seqs)) / len(query_seqs), 3
     )

### 2020-12-10 exploratory/2020-11/plots.Rmd
Use accnos instead of list file

@@
 
-Did mothur's list file format change in the latest dev version?
+Did mothur's list file format change in the latest dev version? The fraction
+mapped values are the same as the reference fraction when the ref weight is
+simple.
 
+```{r fraction-mapped-open_fit-split}
+optifit_split %>% 
+  filter(method == 'open', tool == 'mothur', is.na(ref)) %>% 
+  mutate()
+  ggplot(aes(x=ref_frac, y=fraction_mapped, color=ref_weight)) +
+  geom_jitter(alpha = 0.5, width = 0.01, size=1) +
+  facet_wrap("dataset", nrow=1) +
+  ylim(0, 1) +
+  labs(title="Sequences mapped during open-reference OptiFit",
+       x='reference fraction',
+       y='fraction mapped')
+```
 ```{r diversity_fit-split}

### 2020-12-10 code/py/fraction_reads_mapped.py

@@ -12,21 +12,10 @@ def parse_seqs(infilename):
         seqs = {line.split("\t")[0].strip() for line in count_file}
     return seqs
 
-
-def parse_list(infilename):
-    with open(infilename, "r") as list_file:
-        # third column onward of second line
-        # each seq in each OTU delimited by comma
-        next(list_file)
-        line = next(list_file)
-        mapped_seqs = {seq for column in line.split()[2:] for seq in column.split(",")}
-    return mapped_seqs
-
-
 def main():
-    query_seqs = parse_seqs(snakemake.input.count)
+    query_seqs = parse_seqs(snakemake.input.query)
     ref_seqs = parse_seqs(snakemake.input.ref)
-    mapped_seqs = parse_list(snakemake.input.list)
+    mapped_seqs = parse_seqs(snakemake.input.mapped)
 
     # remove reference seqs from mapped seqs
     mapped_seqs = mapped_seqs - ref_seqs

### 2020-12-10 code/py/fraction_reads_mapped.py
Fix typo; use ref accnos for ref

@@ -17,7 +17,7 @@ def main():
     ref_seqs = parse_seqs(snakemake.input.ref)
     mapped_seqs = parse_seqs(snakemake.input.mapped)
 
-    print('map=ref',mapped_seqs == ref_seqs)
+    print('map==ref',mapped_seqs == ref_seqs)
     print('ref in map', len(ref_seqs.intersection(mapped_seqs)))
     print('len(map)', len(mapped_seqs))
     # remove reference seqs from mapped seqs
@@ -26,7 +26,7 @@ def main():
 
     # mapped reads should not contain reads not in query
     assert not (mapped_seqs - query_seqs)
-    
+
     fraction_mapped = round(
         len(mapped_seqs.intersection(query_seqs)) / len(query_seqs), 3
     )

### 2020-12-10 paper/paper.Rmd
Paste some text from ASM abstract

--- 2020-12-08 paper/paper.Rmd
@@
 
+Assigning amplicon sequences to Operational Taxonomic Units (OTUs) is an
+important step in characterizing the composition of microbial communities across
+large datasets. OptiClust, a _de novo_ OTU clustering method in the mothur
+program, has been shown to produce higher quality OTU assignments than other
+methods and at comparable or faster speeds [@westcott_opticlust_2017;
+@schloss_introducing_2009]. A notable difference between _de novo_ clustering
+and database-dependent methods is that OTU assignments clustered with _de novo_
+methods are not stable when new sequences are added to a dataset
+[@westcott_novo_2015]. However, in some cases one may wish to incorporate new
+samples into a previously clustered dataset without performing clustering again
+on all sequences, such as when deploying a machine learning model where OTUs are
+features [@topcuoglu_effective_2019]. To provide an efficient and robust method
+to fit amplicon sequence data to existing OTUs, we developed the OptiFit
+algorithm as a new component of the mothur program.
+
+- TODO: Describe OptiClust and vsearch.
+
 ## Results
+
+- TODO: brief description of OptiFit algorithm. the method we came up with is the result. OptiClust paper had toy example to walk through the algorithm. 
+- TODO: Paragraph at beginning to describe datasets: why picked, how processed.
+
+We used four published datasets isolated from soil
+[@johnston_metagenomics_2016], marine [@henson_artificial_2016], mouse
+[@schloss_stabilization_2012], and human [@baxter_microbiota-based_2016]
+samples.
 
@@
 - All clustering was performed at a distance threshold of 0.03 and repeated with 100 different random seeds.
-- The Matthews Correlation Coefficient was calculated to serve as a measure of OTU quality.
+- The Matthews Correlation Coefficient was calculated to serve as a measure of OTU quality. OTU quality was evaluated using the Matthews Correlation
+Coefficient (MCC) with a sequence dissimilarity threshold of 0.03% as described
+previously [@schloss_application_2016; @westcott_novo_2015].
 
@@
 
+- TODO: compare optifit performance for split dataset vs public database
+- TODO: double-check with Sarah that MCCs are from full OTU dataset or just query sequnce OTUs
+- TODO: double-check reference fractions aren't flipped (fraction mapped plot looks weird)
+
 - Datasets were randomly split into a reference fraction and a query fraction.
@@
 ### Comparison to vsearch
+
+- TODO: move this into first section with public database.
+- vsearch is more permissive than mothur (radius vs diameter with 0.03 dissimilarity threshold).
 

### 2020-12-10 paper/paper.Rmd
Add code availability statement

@@
 
-### The OptiFit Algorithm
+### Sequence Data Processing Steps
 
-### Analysis Workflow
+### Benchmarking
+
+### Data and Code Availability
+
+TODO: cite all these dependencies
+
+We implemented the analysis workflow in Snakemake and relied on Python, R, and
+GNU bash. Dependencies include mothur v1.45.0, vsearch v2.13.3, numpy, R
+Markdown, the Tidyverse metapackage, and the conda environment manager. A
+reproducible version of the workflow, manuscript, and conda environment is
+available at **TODO: UPDATED REPO LINK**.
 

### 2020-12-10 paper/paper.Rmd
Increase ToC depth

@@
     toc_float: true
-    toc_depth: 2
+    toc_depth: 3
   pdf_document:

### 2020-12-11 code/slurm/submit_sub2.sh
Forcerun optifit

@@ -35,4 +35,4 @@
 #SBATCH --mail-user=sovacool@umich.edu
 #SBATCH --mail-type=BEGIN,END
 
-time snakemake --profile config/slurm --latency-wait 90 --configfile config/config.yaml
+time snakemake --profile config/slurm --latency-wait 90 --configfile config/config.yaml --forcerun fit_to_ref_db

### 2020-12-11 exploratory/2020-11/plots.Rmd

--- 2020-12-10 exploratory/2020-11/plots.Rmd
@@
   ggplot(aes(x=method, y=sec, color=ref)) +
-  geom_jitter(alpha = 0.5) +
+  geom_boxplot(alpha = 0.5) +
   geom_hline(aes(yintercept = mean_sec), sum_opticlust) +
@@
 
-Did mothur's list file format change in the latest dev version?
+Did mothur's list file format change in the latest dev version? The fraction
+mapped values are the same as the reference fraction when the ref weight is
+simple.
 
@@
 vsearch %>% 
-  ggplot(aes(x = method, y = npshannon, color = tool)) +
+  ggplot(aes(x = method, y = shannon, color = tool)) +
   geom_boxplot() +

### 2020-12-11 paper/paper.Rmd

@@
 ## Author Contributions {.appendix}
+
+KLS wrote the analysis code, evaluated the algorithm, and wrote the original draft of the manuscript.
+SLW designed and implemented the OptiFit algorithm and assisted in debugging the analysis code.
+MBM and GAD contributed analysis code.
+PDS conceived the study, supervised the project, and assisted in debugging the analysis code.
+All authors reviewed and edited the manuscript.

### 2020-12-11 exploratory/2020-11/plots.Rmd
Compare all mothur clustering strategies

@@
 )
-
+tri_colors <-  c("darkorange","darkorchid","cyan4") # https://allisonhorst.github.io/palmerpenguins/articles/intro.html
 mutate_perf <- function(dat) {
@@
             mean_sec = mean(sec),
-            sd_sec = sd(sec),
-            mean_npshannon = mean(npshannon))
+            sd_sec = sd(sec))#,
+            #mean_shannon = mean(shannon))
 ```
@@
 ```{r fraction-mapped_fit-db}
-optifit_dbs %>% filter(method == 'closed') %>% 
+optifit_dbs %>% #filter(method == 'closed') %>% 
   ggplot(aes(x=dataset, y=fraction_mapped, color=ref)) +
   geom_jitter(alpha = 0.5) + 
+  facet_wrap('method') +
   ylim(0, 1) +
@@
 ```
+
+Plotted the fraction mapped for open-reference clustering too as a sanity check.
+Final plots to be included in the paper won't include open-ref.
+
 ```{r runtime_fit-db}
@@
 ```
-```{r diversity_fit-db}
+```{r diversity_fit-db, eval=FALSE}
 optifit_dbs %>% 
-  ggplot(aes(x=method, y=npshannon, color=ref)) +
+  ggplot(aes(x=method, y=shannon, color=ref)) +
   geom_jitter(alpha = 0.5) +
-  geom_hline(aes(yintercept = mean_npshannon), sum_opticlust) +
+  geom_hline(aes(yintercept = mean_shannon), sum_opticlust) +
   facet_wrap('dataset', scales = 'free') +
@@
 
-```{r fraction-mapped-open_fit-split}
+```{r fraction-mapped-grid_fit-split}
 optifit_split %>% 
-  filter(method == 'open', tool == 'mothur', is.na(ref)) %>% 
-  #group_by(dataset, ref_weight, ref_frac) %>% 
+  filter(tool == 'mothur', is.na(ref)) %>% 
+  filter(dataset == 'soil') %>% 
   ggplot(aes(x=ref_frac, y=fraction_mapped, color=ref_weight)) +
   geom_jitter(alpha = 0.5, width = 0.01, size=1) +
-  facet_wrap("dataset", nrow=1) +
+  facet_grid(dataset ~ method) +
   ylim(0, 1) +
@@
 
-```{r diversity_fit-split}
+```{r diversity_fit-split, eval = FALSE}
 optifit_split %>% 
-  ggplot(aes(x = ref_frac, y = npshannon, color = ref_weight)) +
+  ggplot(aes(x = ref_frac, y = shannon, color = ref_weight)) +
   geom_jitter(size = 1, alpha = 0.3, width = 0.01) +
   stat_summary(fun = mean, geom = 'crossbar', alpha = 0.5) +
-  geom_hline(aes(yintercept = mean_npshannon), sum_opticlust) +
+  geom_hline(aes(yintercept = mean_shannon), sum_opticlust) +
   facet_grid(dataset ~ method, scales = 'free') +
@@
   theme(plot.caption = element_markdown())
+```
+
+## compare all mothur clustering strategies
+
+```{r join_all-opti}
+opti_all <- list(optifit_dbs %>% mutate(strategy = glue('database_{ref}')),
+                 optifit_split %>% mutate(strategy = 'self-split'),
+                 opticlust %>% mutate(strategy = 'de_novo')) %>% 
+  reduce(full_join)
+
+sum_opti_all <- opti_all %>% 
+  group_by(dataset, strategy, method) %>% 
+  summarize(n = n(),
+            mcc_median = median(mcc),  # TODO: tidy way to avoid this repetitiveness?
+            sec_median = median(sec),
+            mem_gb_median = median(mem_gb),
+            frac_map_median = median(fraction_mapped))
+head(sum_opti_all)
+```
+
+```{r mcc_all-opti}
+sum_opti_all %>% 
+  ggplot(aes(strategy, mcc_median, color = method)) +
+  geom_point() +
+  facet_wrap(dataset ~ .) +
+  scale_color_manual(values = tri_colors) +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+```
+```{r runtime_all-opti}
+sum_opti_all %>% 
+  ggplot(aes(strategy, sec_median, color = method)) +
+  geom_point() +
+  facet_wrap(dataset ~ .) +
+  scale_color_manual(values = tri_colors) +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+```
+```{r memory_all-opti}
+sum_opti_all %>% 
+  ggplot(aes(strategy, mem_gb_median, color = method)) +
+  geom_point() +
+  facet_wrap(dataset ~ .) +
+  scale_color_manual(values = tri_colors) +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+```
+
+```{r fraction-mapped_all-opti}
+sum_opti_all %>% 
+  filter(strategy != 'de_novo') %>% 
+  ggplot(aes(strategy, frac_map_median, color = method)) +
+  geom_point() +
+  facet_wrap(dataset ~ .) +
+  scale_color_manual(values = tri_colors) +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
 ```
@@
 ```{r fraction-mapped_vsearch}
-vsearch %>% filter(method == 'closed') %>% 
+vsearch %>% #filter(method == 'closed') %>% 
   ggplot(aes(x = dataset, y = fraction_mapped, color = tool)) +
@@
   ylim(0, 1) +
+  facet_wrap('method') +
   labs(title="Sequences mapped during closed-reference OptiFit") +
@@
 ```
+
+The fraction mapped for _de novo_ open-reference is shown here as a sanity check.
+Final plots will only include closed-reference.

### 2020-12-12 code/py/check_split.py
Write code to validate accnos files from split_weighted_subsample

@@ -1,0 +1,22 @@
+
+parse_accnos(filename):
+    with open(filename, 'r') as file:
+        seqs = {line.strip() for line in file}
+    return seqs
+
+
+def main():
+    ref = parse_accnos(snakemake.input.ref)
+    query = parse_accnos(snakemake.input.query)
+    all = parse_accnos(snakemake.intput.all)
+
+    assert not ref.intersection(query)
+    assert all - ref.intersection(all)
+    assert all - query.intersection(all)
+    assert ref + query == all
+
+    with open(snakemake.output.txt, 'w') as file:
+        file.write('check_split_passed\n')
+        file.write(True)
+
+main()

### 2020-12-12 exploratory/2020-11/plots.Rmd
Weird pattern in incorrect fraction mapped results

@@
 
+### is there a pattern?
+
+```{r debug_frac_map}
+debug_dat <- optifit_split %>% 
+  filter(method == 'open') %>% 
+  mutate(is_correct = fraction_mapped == 1) %>% 
+  select(dataset, seed, method, fraction_mapped, is_correct, 
+         sample_frac, ref_frac, ref_weight) %>%
+  group_by(dataset, seed, ref_frac, ref_weight, is_correct)
+
+debug_dat %>% filter(seed ==1, ref_weight == 'distance', dataset == 'soil') %>% pull(is_correct)
+debug_dat %>% filter(seed ==2, ref_weight == 'distance', dataset == 'soil') %>% pull(is_correct)
+
+debug_dat %>% filter(seed ==1, ref_weight == 'distance', dataset == 'human') %>% pull(is_correct)
+debug_dat %>% filter(seed ==2, ref_weight == 'distance', dataset == 'human') %>% pull(is_correct)
+
+```
+
+## diversity
+
 ```{r diversity_fit-split, eval = FALSE}

### 2020-12-12 code/py/check_split.py
Fix syntax

@@ -1,5 +1,5 @@
 
-parse_accnos(filename):
+def parse_accnos(filename):
     with open(filename, 'r') as file:
         seqs = {line.strip() for line in file}
     return seqs

### 2020-12-12 code/py/check_split.py
Fix typo

@@ -8,7 +8,7 @@ def parse_accnos(filename):
 def main():
     ref = parse_accnos(snakemake.input.ref)
     query = parse_accnos(snakemake.input.query)
-    all = parse_accnos(snakemake.intput.all)
+    all = parse_accnos(snakemake.input.all)
 
     assert not ref.intersection(query)
     assert all - ref.intersection(all)

### 2020-12-12 exploratory/2020-11/plots.Rmd
Use set union method

@@
 
-debug_dat %>% filter(seed ==1, ref_weight == 'distance', dataset == 'human') %>% pull(is_correct)
-debug_dat %>% filter(seed ==2, ref_weight == 'distance', dataset == 'human') %>% pull(is_correct)
+debug_dat %>% filter(seed ==1, ref_weight == 'simple', dataset == 'human') %>% pull(is_correct)
+debug_dat %>% filter(seed ==2, ref_weight == 'simple', dataset == 'human') %>% pull(is_correct)
 

### 2020-12-12 code/py/check_split.py

@@ -13,7 +13,7 @@ def main():
     assert not ref.intersection(query)
     assert all - ref.intersection(all)
     assert all - query.intersection(all)
-    assert ref + query == all
+    assert ref.union(query) == all
 
     with open(snakemake.output.txt, 'w') as file:
         file.write('check_split_passed\n')

### 2020-12-12 code/py/check_split.py
🎨 Style Python code 🐍

@@ -1,6 +1,5 @@
-
 def parse_accnos(filename):
-    with open(filename, 'r') as file:
+    with open(filename, "r") as file:
         seqs = {line.strip() for line in file}
     return seqs
 
@@ -15,8 +14,9 @@ def main():
     assert all - query.intersection(all)
     assert ref.union(query) == all
 
-    with open(snakemake.output.txt, 'w') as file:
-        file.write('check_split_passed\n')
+    with open(snakemake.output.txt, "w") as file:
+        file.write("check_split_passed\n")
         file.write(True)
+
 
 main()

### 2020-12-12 code/py/check_split.py
Fix syntax

@@ -16,7 +16,7 @@ def main():
 
     with open(snakemake.output.txt, "w") as file:
         file.write("check_split_passed\n")
-        file.write(True)
+        file.write("True")
 
 
 main()

### 2020-12-12 exploratory/2020-11/plots.Rmd
Confirmed I fixed the bug on test config
removing the pick_best_cluster_seed rule worked!

--- 2020-12-11 exploratory/2020-11/plots.Rmd
@@
   ggplot(aes(x=ref_frac, y=fraction_mapped, color=ref_weight)) +
-  geom_jitter(alpha = 0.5, width = 0.01, size=1) +
+  geom_point(alpha = 0.5, size=1) +
   facet_wrap("dataset", nrow=1) +
@@
   ggplot(aes(x=ref_frac, y=fraction_mapped, color=ref_weight)) +
-  geom_jitter(alpha = 0.5, width = 0.01, size=1) +
+  geom_point(alpha = 0.5, size=1) +
   facet_grid(dataset ~ method) +
@@
 simple.
+
+### is there a pattern?
+
+```{r debug_frac_map}
+debug_dat <- optifit_split %>% 
+  filter(method == 'open') %>% 
+  mutate(is_correct = fraction_mapped == 1) %>% 
+  select(dataset, seed, method, fraction_mapped, is_correct, 
+         sample_frac, ref_frac, ref_weight) %>%
+  group_by(dataset, seed, ref_frac, ref_weight, is_correct)
+
+for (s in 1:10) {
+  debug_dat %>% 
+    filter(seed == s, ref_weight == 'simple', dataset == 'soil') %>% 
+    pull(is_correct) %>% 
+    print()
+}
+
+```
+
+Pattern of correctness among ref fracs is flipped between seed 1 and 2
+
+## diversity
 

### 2020-12-13 exploratory/2020-11_Nov-Dec/plots.Rmd
Rename exploratory subdirs for clarity

--- 2020-12-12 exploratory/2020-11/plots.Rmd
@@
 
-for (s in 1:10) {
+for (s in 1:2) {
   debug_dat %>% 
@@
 
-## diversity
+### diversity
 
@@
 ```
+
+## vsearch
+
+For reference-based clustering, datasets were fit to the greengenes database.
+
+```{r load_vsearch}
+vsearch <- read_tsv(here('subworkflows/4_vsearch/results/vsearch_results.tsv')) %>% 
+  mutate_perf()
+vsearch_join <- vsearch %>% 
+  full_join(opticlust) %>% 
+  full_join(optifit_dbs %>% filter(ref == 'gg'))
+```
+
+```{r mcc_vsearch}
+vsearch_join 
+  ggplot(aes(x = method, y = mcc, color = tool)) +
+  geom_boxplot() +
+  facet_wrap('dataset') +
+  ylim(0, 1) + 
+  labs(x = '')
+```
+
+```{r sec_vsearch}
+vsearch_join 
+  ggplot(aes(x = method, y = sec, color = tool)) +
+  geom_boxplot() +
+  facet_wrap('dataset') +
+  labs(x = '')
+```
+
+```{r diversity_vsearch}
+vsearch_join 
+  ggplot(aes(x = method, y = shannon, color = tool)) +
+  geom_boxplot() +
+  facet_wrap('dataset') +
+  labs(x = '')
+```
+
+```{r fraction-mapped_vsearch}
+vsearch_join #filter(method == 'closed') %>% 
+  ggplot(aes(x = dataset, y = fraction_mapped, color = tool)) +
+  geom_boxplot() + 
+  ylim(0, 1) +
+  facet_wrap('method') +
+  labs(title="Sequences mapped during closed-reference OptiFit") +
+  theme(plot.caption = element_markdown())
+```
+
+The fraction mapped for _de novo_ open-reference is shown here as a sanity check.
+Final plots will only include closed-reference.
 
@@
   scale_color_manual(values = tri_colors) +
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+  ylim(0, 1) +
+  #theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
+  coord_flip() +
+  labs(title = 'OTU Quality', x = '', y = 'Median MCC')
 ```
@@
   geom_point() +
-  facet_wrap(dataset ~ .) +
+  facet_wrap(dataset ~ ., scales = 'free_y') +
   scale_color_manual(values = tri_colors) +
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
 ```
@@
   geom_point() +
-  facet_wrap(dataset ~ .) +
+  facet_wrap(dataset ~ ., scales = 'free_y') +
   scale_color_manual(values = tri_colors) +
@@
 
-## vsearch
+## compare mothur & vsearch with just the human dataset
 
-For reference-based clustering, datasets were fit to the greengenes database.
+```{r join_tools_human}
+all_human <- list(opti_all %>% 
+                    filter(dataset == 'human', 
+                           ref == 'gg' | method == 'de_novo'),
+                 vsearch %>% filter(dataset == 'human')
+                 ) %>% 
+  reduce(full_join) %>% 
+  mutate(strategy = case_when(
+    method == 'de_novo' ~ method,
+    TRUE ~ as.character(glue('{method}_ref_{ref}'))))
 
-```{r load_vsearch}
-vsearch <- read_tsv(here('subworkflows/4_vsearch/results/vsearch_results.tsv')) %>% 
-  mutate_perf() %>% 
-  full_join(opticlust) %>% 
-  full_join(optifit_dbs %>% filter(ref == 'gg'))
+sum_all_human <- all_human %>% 
+  group_by(tool, strategy, method) %>% 
+  summarize(n = n(),
+            mcc_median = median(mcc),  # TODO: tidy way to avoid this repetitiveness?
+            sec_median = median(sec),
+            mem_gb_median = median(mem_gb),
+            frac_map_median = median(fraction_mapped))
+head(sum_all_human)
 ```
 
-```{r mcc_vsearch}
-vsearch %>% 
-  ggplot(aes(x = method, y = mcc, color = tool)) +
-  geom_boxplot() +
-  facet_wrap('dataset') +
+```{r mcc_all-human}
+sum_all_human %>% 
+  ggplot(aes(strategy, mcc_median, color = tool)) +
+  geom_point() +
+  #facet_wrap(tool ~ .) +
   ylim(0, 1) +
-  labs(x = '')
+  scale_color_manual(values = tri_colors) +
+  #theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
+  coord_flip()
 ```
 
-```{r sec_vsearch}
-vsearch %>% 
-  ggplot(aes(x = method, y = sec, color = tool)) +
-  geom_boxplot() +
-  facet_wrap('dataset') +
-  labs(x = '')
+```{r sum_all-human}
+sum_all_human %>% 
+  pivot_longer(c(mcc_median, sec_median, mem_gb_median, frac_map_median), names_to = 'metric') %>% 
+  ggplot(aes(strategy, value, color = tool)) +
+  geom_point() +
+  facet_wrap('metric', scales = 'free')
+
 ```
 
-```{r diversity_vsearch}
-vsearch %>% 
-  ggplot(aes(x = method, y = shannon, color = tool)) +
+```{r metrics_all-human}
+all_human %>% 
+  pivot_longer(c(mcc, sec, mem_gb, fraction_mapped), names_to = 'metric') %>% 
+  ggplot(aes(strategy, value, color = tool)) +
   geom_boxplot() +
-  facet_wrap('dataset') +
-  labs(x = '')
+  facet_wrap('metric', scales = 'free') +
+  labs(x='', y='', 
+       title="mothur vs. vsearch", 
+       caption = 'The human dataset is shown here. The patterns are similar with other datasets.')
 ```
-
-```{r fraction-mapped_vsearch}
-vsearch %>% #filter(method == 'closed') %>% 
-  ggplot(aes(x = dataset, y = fraction_mapped, color = tool)) +
-  geom_boxplot() + 
-  ylim(0, 1) +
-  facet_wrap('method') +
-  labs(title="Sequences mapped during closed-reference OptiFit") +
-  theme(plot.caption = element_markdown())
-```
-
-The fraction mapped for _de novo_ open-reference is shown here as a sanity check.
-Final plots will only include closed-reference.

### 2020-12-13 exploratory/2020-11_Nov-Dec/plots.Rmd
Use boxplots instead of jitters for fit-db

@@
   geom_hline(aes(yintercept = mean_mcc), sum_opticlust) +
-  geom_jitter(alpha = 0.5) + 
+  geom_boxplot() + 
   facet_wrap('dataset') +
+  scale_color_manual(values=tri_colors) +
   ylim(0, 1) +
@@
 ```{r fraction-mapped_fit-db}
-optifit_dbs %>% #filter(method == 'closed') %>% 
+optifit_dbs %>% filter(method == 'closed') %>% 
   ggplot(aes(x=dataset, y=fraction_mapped, color=ref)) +
-  geom_jitter(alpha = 0.5) + 
+  geom_boxplot() + 
   facet_wrap('method') +
@@
 ```{r mcc_vsearch}
-vsearch_join 
+vsearch_join %>% 
   ggplot(aes(x = method, y = mcc, color = tool)) +
@@
 ```{r sec_vsearch}
-vsearch_join 
+vsearch_join %>% 
   ggplot(aes(x = method, y = sec, color = tool)) +
@@
 ```{r diversity_vsearch}
-vsearch_join 
+vsearch_join %>% 
   ggplot(aes(x = method, y = shannon, color = tool)) +
@@
 ```{r fraction-mapped_vsearch}
-vsearch_join #filter(method == 'closed') %>% 
+vsearch_join %>% #filter(method == 'closed') %>% 
   ggplot(aes(x = dataset, y = fraction_mapped, color = tool)) +
@@
   labs(x='', y='',  
-       title="mothur vs. vsearch", 
-       caption = 'The human dataset is shown here. The patterns are similar with other datasets.')
+       caption = 'Only the human dataset is shown here. The patterns are similar with other datasets.')
 ```

### 2020-12-13 code/py/split_weighted_subsample.py
🎨 Style Python code 🐍

@@ -13,8 +13,10 @@ def main():
     np.random.seed(int(snakemake.wildcards.seed))
 
     all_seqs = SeqList.from_files(
-        snakemake.input.fasta, snakemake.input.count, snakemake.input.dist,
-        threshold = snakemake.params.dissim_thresh
+        snakemake.input.fasta,
+        snakemake.input.count,
+        snakemake.input.dist,
+        threshold=snakemake.params.dissim_thresh,
     )
     num_all_seqs = len(all_seqs)
     ref_frac = float(snakemake.wildcards.ref_frac)

### 2020-12-14 exploratory/2020-11_Nov-Dec/plots.Rmd
More plots with debugging split fraction mapped

@@
             mean_sec = mean(sec),
-            sd_sec = sd(sec))#,
+            sd_sec = sd(sec),
+            mean_mem_gb = mean(mem_gb))#,
             #mean_shannon = mean(shannon))
@@
   ylim(0, 1) +
+  scale_color_manual(values=tri_colors) +
   labs(title="Sequences mapped during closed-reference OptiFit") +
@@
 ```
-
-Plotted the fraction mapped for open-reference clustering too as a sanity check.
-Final plots to be included in the paper won't include open-ref.
 
@@
   facet_wrap('dataset', scales = 'free') +
+  scale_color_manual(values=tri_colors) +
   labs(title='OptiFit runtime with reference databases',
@@
 ```
+
+```{r memory_fit-db}
+optifit_dbs %>% 
+  ggplot(aes(x=method, y=mem_gb, color=ref)) +
+  geom_boxplot(alpha = 0.5) +
+  geom_hline(aes(yintercept = mean_mem_gb), sum_opticlust) +
+  facet_wrap('dataset', scales = 'free') +
+  scale_color_manual(values=tri_colors) +
+  labs(title='OptiFit runtime with reference databases',
+       caption='Black line: _de novo_ clustering') +
+  theme(plot.caption = element_markdown())
+```
+
 ```{r diversity_fit-db, eval=FALSE}
@@
   ylim(0, 1) +
-  labs(title="Sequences mapped during open-reference OptiFit",
+  labs(title="Sequences mapped during OptiFit",
        x='reference fraction',
-       y='fraction mapped')
+       y='fraction mapped',
+       caption=glue("Ran with {optifit_split %>% pull(seed) %>% unique() %>% length()} seeds"))
 ```
@@
 
-```{r mcc_all-opti}
+```{r mcc_all-opti, fig.width=10, fig.height=3.5}
 sum_opti_all %>% 
@@
   geom_point() +
-  facet_wrap(dataset ~ .) +
+  facet_wrap(dataset ~ ., nrow=1) +
   scale_color_manual(values = tri_colors) +
@@
 
-```{r metrics_all-human}
+```{r metrics_all-human, fig.width=10, fig.height=3.5}
 all_human %>% 
@@
   geom_boxplot() +
-  facet_wrap('metric', scales = 'free') +
+  facet_wrap('metric', scales = 'free_y', nrow=1) +
   labs(x='', y='',  
-       caption = 'Only the human dataset is shown here. The patterns are similar with other datasets.')
+       caption = 'Only the human dataset is shown here. The patterns are similar with other datasets.') +
+  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
 ```

### 2020-12-14 docs/slides/2020-12-14_lab-mtg.Rmd
Create slides for lab mtg pres.

---
title: "OptiFit project update"
subtitle: "+ a case study in debugging"
author: "Kelly Sovacool"
date: "2020-12-14 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      #titleSlideClass: ['center', 'middle', 'title-slide']
    chakra: libs/remark-latest.min.js
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE,
                  message = FALSE
                  )
```

```{r chakra, eval = FALSE}
# run once to download remark-lastest
xaringan::summon_remark(to = 'docs/slides/libs')
```

```{r libs}
library(here)
library(knitr)
library(tidyverse)
xaringanExtra::use_share_again()
```

## OTU Clustering methods

- _De novo_ (OptiClust)
  - Cluster sequences based on their similarity.
  - OTU's aren't stable; you'll get different OTU assignments with different random seeds.
- Reference-based (OptiFit)
  - First cluster reference sequences _de novo_, then fit query sequences to those OTUs.

.footnote[
  [Westcott & Schloss (2015) _PeerJ_ ](https://doi.org/10.7717%2Fpeerj.1487);
  [Schloss (2016) _mSystems_ ](https://doi.org/10.1128%2FmSystems.00027-16);
  [Westcott & Schloss (2017) _mSphere_](10.1128/mSphereDirect.00073-17)
]

--

### Reference-based strategies

- Modes
  - Open - sequences that can't be fit to the reference are then clustered _de novo_ to create additional OTUs.
  - Closed - sequences that can't be fit to the reference are thrown out.
- Reference choice
  - External database (Silva, greengenes, RDP).
  - Split the dataset into a reference & query set.

---

# Changes since a couple months ago

- Sarah Westcott fixed some bugs in mothur & made OptiFit faster.
- I re-ran everything with silva v132.
- Ran vsearch on all the datasets
- Discovered a bug in my workflow code (last week!)

---

# Reference clustering against a database

.pull-left[
```{r mcc_fit-db}
include_graphics(here('exploratory', '2020-11_Nov-Dec', 'figures', 'mcc_fit-db-1.png'))
```
]

--

.pull-right[
```{r fraction-mapped_fit-db}
include_graphics(here('exploratory', '2020-11_Nov-Dec', 'figures', 'fraction-mapped_fit-db-1.png'))
```
]

???
- _De novo_ and open-reference perform equally well.
- Closed reference performs slightly worse.
- RDP doesn't perform well.
- Prior to switching to silva v132 & updating mothur, closed-reference artificially outperformed open-reference.

---

# Reference clustering against a database

.pull-left[
```{r runtime_fit-db}
include_graphics(here('exploratory', '2020-11_Nov-Dec', 'figures', 
                      'runtime_fit-db-1.png'))
```
]

--

.pull-right[
```{r memory_fit-db}
include_graphics(here('exploratory', '2020-11_Nov-Dec', 'figures', 
                      'memory_fit-db-1.png'))
```
]

---

## Summarizing mothur results

Whenever there's virtually no variation in the metric.

```{r sum_all_mothur}
include_graphics(here('exploratory', '2020-11_Nov-Dec', 'figures', 
                      'mcc_all-opti-1.png'))
```

---

## mothur vs. vsearch

```{r vsearch}
include_graphics(here('exploratory', '2020-11_Nov-Dec', 'figures', 
                      'metrics_all-human-1.png'))
```

???
- Just showing the human dataset here because results patterns were basically the same across all datasets.
- Only used greengenes for reference because that's what
QIIME does.

---

# Splitting datasets into reference & query sets

.pull-left[
Before reverting to Silva v132

```{r frac-map-old}
include_graphics(here('exploratory', '2020-05_May-Oct', 'figures', 
                      'fraction_mapped_complement-2.png'))
```
]
--
.pull-right[
After reverting to Silva v132:

```{r frac-map-new}
include_graphics(here('exploratory', 
                      '2020-11_Nov-Dec', 
                      'figures', 
                      'debug-fraction-mapped_fit-split-1.png'))
```
]

???
- 1 seed out of every parameter set has the same pattern as the old results
- I thought maybe my snakemake workflow somehow kept one result file from before reverting to silva. Like maybe it accidentally didn't overwrite with the new results?
  - But I checked the timestamps of the results files and they were updated after re-doing with silva v132.
---

## Sanity check: open-ref should have fraction mapped = 1

.pull-left[
```{r}
include_graphics(here('exploratory', 
                      '2020-11_Nov-Dec', 
                      'figures',
                      'debug-fraction-mapped-open_fit-split-1.png'))
```
]

???
- Now that is really weird. For weight=simple, the fraction mapped is equal to the sample frac most of the time. So the fraction mapped is incorrect most of the time.
- Next I thought; maybe there's a bug in mothur where the open-ref output files have one less column than the closed-ref.
  - I brought this up to Sarah W before and she said she fixed it. Maybe I forgot to recompile mothur?
  - Bingo! I hadn't pulled the latest changes. So I recompiled and re-ran with one parameter set & one seed, and it worked! Problem solved! Well, not quite. When re-running with more than one seed, the problem persisted.
---

## is there a pattern?

```{r}
dat_100 <- read_tsv(here('exploratory',
                               '2020-11_Nov-Dec',
                               'debug-split_100-seeds.tsv'))
dat_100 %>% filter(method == 'open') %>% 
  mutate(is_correct = fraction_mapped == 1) %>% 
  select(dataset, seed, is_correct, fraction_mapped,
         sample_frac, ref_frac, ref_weight) %>% 
  filter(is_correct, dataset == 'soil')
```

???
- is there a bug in my code to calculate the fraction mapped? let's check results from fitting to databases

---

## Sanity check: results from fitting to databases

.pull-left[
```{r}
include_graphics(here('exploratory', 
                      '2020-11_Nov-Dec', 
                      'figures',
                      'debug-fraction-mapped-open_fit-split-1.png'))
```
]
.pull-right[
```{r}
include_graphics(here('exploratory', 
                      '2020-11_Nov-Dec', 
                      'figures',
                      'fraction-mapped-open_fit-db-1.png'))
```
]
???
- so there isn't a bug in my fraction mapped code and there isn't a bug in mothur, otherwise these would be wrong too.
- is there a bug in my code to split the datasets into reference & query fractions?
  - I had written assertion statements throughout that code to make sure the query + reference fraction were correct.
  - Double-checked the resulting accnos files anyway just to be safe. They were all correct.
---

## WHAT?!

```{r}
split_2 <- read_tsv(here('exploratory',
                          '2020-11_Nov-Dec',
                          'debug-split_2-seeds.tsv')) %>%
  filter(method == 'open') %>% 
  mutate(is_correct = fraction_mapped == 1) %>% 
  select(dataset, seed, method, fraction_mapped, is_correct, 
         sample_frac, ref_frac, ref_weight)

split_2 %>% 
  filter(dataset == 'soil', ref_weight == 'simple') %>% 
  select(sample_frac, ref_frac, seed, is_correct) %>% 
  pivot_wider(names_from = seed, 
              values_from = is_correct,
              names_glue = 'seed_{seed}_is_correct')
```

Re-ran with only 2 seeds.

---

```{bash, eval = FALSE}
wget https://raw.githubusercontent.com/SchlossLab/OptiFitAnalysis/c900c5fa5f3081b73536a4e568f66817a66b3d0c/figures/rulegraph_fit_split.png?token=AEHR6TPEQM4HG6GCTVOUSTK74DIXK -O exploratory/2020-11_Nov-Dec/figures/rulegraph1.png
```
.pull-left[
Workflow

```{r rulegraph_old, out.width='60%'}
include_graphics(here('exploratory', 
                      '2020-11_Nov-Dec', 
                      'figures',
                      'rulegraph1.png'))
```
]
--
.pull-right[
So far:
- Code shared with  workflow that fits to databases must be correct.
- Code that didn't change between previous and most recent results must be correct.
  - What code did I change between then and now? Did I change anything?
]
???
- Browsed my commit history on github and realized that I had done a little refactoring to clean up the workflow a bit. 
- I also added the rule pick_best_cluster_seed to check off a TODO in the analysis plan. I forgot about that, thought that had been there much longer.
- So I removed that rule and tried re-running the workflow with one parameter set and 2 seeds to check whether that was the problem.

---

## Fixed it!
.pull-left[
```{r rulegraph_new, out.width='60%'}
include_graphics(here('figures',
                      'rulegraph_fit_split.png'))
```
]
.pull-right[
```{r}
include_graphics(here('exploratory', 
                      '2020-11_Nov-Dec', 
                      'figures',
                      'fraction-mapped-grid_fit-split-1.png'))
```
]

---

## What happened here?

.pull-left[
- With `pick_best_cluster_seed`, the OptiClust results from the best seed were used for downstream OptiFit runs with _all seeds_.
  - Same seed variable used to create the data splits.
- When seeds mismatched, query sets were being fit to reference OTUs from different data splits.
- Results were correct whenever the OptiFit seed corresponded to the best OptiClust seed, and incorrect in all other cases.
]

---

## Conclusions

- Don't change more than one thing at a time, even if you think it's "just refactoring".
- Start debugging from the easiest step to fix.
- Have config files with different parameter sets to make debugging quick & easy.
  - default: all datasets & parameters & 100 seeds.
  - debug: one dataset & only one of each parameter choice & 2 seeds.
- Make many small commits to make it easy to figure out where you introduced a bug.
- Test your assumptions
  - Use `assertthat` pkg in R.

### 2020-12-14 docs/slides/2020-12-14_lab-mtg.Rmd
Add links (post-meeting)

@@
   - Use `assertthat` pkg in R.
+
+---
+
+# Useful links
+
+- [these slides (R Markdown)](https://github.com/SchlossLab/OptiFitAnalysis/blob/master/docs/slides/2020-12-14_lab-mtg.Rmd)
+- [these slides (HTML)](http://www.schlosslab.org/OptiFitAnalysis/slides/2020-12-14_lab-mtg.html#1)
+- [my code club on unit testing in R](https://github.com/SchlossLab/intro-testing-r)
+- [`assertthat` R package](https://github.com/hadley/assertthat)
+- [Jenny Bryan’s talk on debugging R code](https://rstudio.com/resources/rstudioconf-2020/object-of-type-closure-is-not-subsettable/)

### 2020-12-15 exploratory/2020-11_Nov-Dec/plots.Rmd
Reflow
Plot de novo stats as hline

@@
   group_by(dataset) %>% 
-  summarize(mean_mcc = mean(mcc),
-            sd_mcc = sd(mcc),
-            mean_sec = mean(sec),
-            sd_sec = sd(sec),
-            mean_mem_gb = mean(mem_gb))#,
-            #mean_shannon = mean(shannon))
+  summarize(mcc_median = median(mcc),
+            sec_median = median(sec),
+            mem_gb_median = median(mem_gb))#,
+            #median_shannon = median(shannon))
 ```
@@
   ggplot(aes(x=method, y=mcc, color=ref)) +
-  geom_hline(aes(yintercept = mean_mcc), sum_opticlust) +
+  geom_hline(aes(yintercept = mcc_median), sum_opticlust) +
   geom_boxplot() + 
@@
   geom_boxplot(alpha = 0.5) +
-  geom_hline(aes(yintercept = mean_sec), sum_opticlust) +
+  geom_hline(aes(yintercept = sec_median), sum_opticlust) +
   facet_wrap('dataset', scales = 'free') +
@@
   geom_boxplot(alpha = 0.5) +
-  geom_hline(aes(yintercept = mean_mem_gb), sum_opticlust) +
+  geom_hline(aes(yintercept = mem_gb_median), sum_opticlust) +
   facet_wrap('dataset', scales = 'free') +
@@
   geom_jitter(alpha = 0.5) +
-  geom_hline(aes(yintercept = mean_shannon), sum_opticlust) +
+  geom_hline(aes(yintercept = median_shannon), sum_opticlust) +
   facet_wrap('dataset', scales = 'free') +
@@
   stat_summary(fun = mean, geom = 'crossbar', alpha = 0.5) +
-  geom_hline(aes(yintercept = mean_mcc), sum_opticlust) +
+  geom_hline(aes(yintercept = mcc_median), sum_opticlust) +
   facet_grid(dataset ~ method) +
@@
   stat_summary(fun = mean, geom = 'crossbar', alpha = 0.5) +
-  geom_hline(aes(yintercept = mean_sec), sum_opticlust) +
+  geom_hline(aes(yintercept = sec_median), sum_opticlust) +
   facet_grid(dataset ~ method, scales = 'free') +
@@
   stat_summary(fun = mean, geom = 'crossbar', alpha = 0.5) +
-  geom_hline(aes(yintercept = mean_shannon), sum_opticlust) +
+  geom_hline(aes(yintercept = median_shannon), sum_opticlust) +
   facet_grid(dataset ~ method, scales = 'free') +
@@
 ```{r join_all-opti}
-opti_all <- list(optifit_dbs %>% mutate(strategy = glue('database_{ref}')),
-                 optifit_split %>% mutate(strategy = 'self-split'),
-                 opticlust %>% mutate(strategy = 'de_novo')) %>% 
+
+plot_denovo_hline <- function(yint, dat = sum_opticlust) {
+  list(geom_hline(data = dat, aes(yintercept = {{yint}})),
+       labs(caption = "Black line: _de novo_ clustering0"),
+       theme(plot.caption = element_markdown())
+  )
+}
+
+optifit_all <- list(optifit_dbs %>% 
+                   mutate(strategy = glue('database_{ref}')),
+                 optifit_split %>% 
+                   mutate(strategy = 'self-split')) %>% 
   reduce(full_join)
 
-sum_opti_all <- opti_all %>% 
+sum_optifit <- optifit_all %>% 
   group_by(dataset, strategy, method) %>% 
@@
             frac_map_median = median(fraction_mapped))
-head(sum_opti_all)
+head(sum_optifit)
 ```
@@
 ```{r mcc_all-opti, fig.width=10, fig.height=3.5}
-sum_opti_all %>% 
+sum_optifit %>% 
   ggplot(aes(strategy, mcc_median, color = method)) +
@@
   coord_flip() +
-  labs(title = 'OTU Quality', x = '', y = 'Median MCC')
+  labs(title = 'OTU Quality', x = '', y = 'Median MCC') + 
+  plot_denovo_hline(yint = mcc_median)
+
 ```
 ```{r runtime_all-opti}
-sum_opti_all %>% 
+sum_optifit %>% 
   ggplot(aes(strategy, sec_median, color = method)) +
@@
   scale_color_manual(values = tri_colors) +
-  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
+  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
+  plot_denovo_hline(sec_median)
 ```
 ```{r memory_all-opti}
-sum_opti_all %>% 
+sum_optifit %>% 
   ggplot(aes(strategy, mem_gb_median, color = method)) +
@@
   scale_color_manual(values = tri_colors) +
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
+  plot_denovo_hline(mem_gb_median)
 ```
@@
 ```{r fraction-mapped_all-opti}
-sum_opti_all %>% 
-  filter(strategy != 'de_novo') %>% 
+sum_optifit %>% 
   ggplot(aes(strategy, frac_map_median, color = method)) +
@@
 ```{r join_tools_human}
-all_human <- list(opti_all %>% 
+all_human <- list(optifit_all %>% 
                     filter(dataset == 'human', 
                            ref == 'gg' | method == 'de_novo'),
+                  opticlust %>% filter(dataset == 'human'),
                   vsearch %>% filter(dataset == 'human')

### 2020-12-15 paper/paper.Rmd
Cite software dependencies

@@
 ---
-title: "**OptiFit: a fast method for fitting amplicon sequences to existing OTUs**"
+title: "OptiFit: a fast method for fitting amplicon sequences to existing OTUs"
 date: '`r Sys.Date()`'
@@
 
-TODO: cite all these dependencies
-
-We implemented the analysis workflow in Snakemake and relied on Python, R, and
-GNU bash. Dependencies include mothur v1.45.0, vsearch v2.13.3, numpy, R
-Markdown, the Tidyverse metapackage, and the conda environment manager. A
-reproducible version of the workflow, manuscript, and conda environment is
-available at **TODO: UPDATED REPO LINK**.
+We implemented the analysis workflow in Snakemake [@koster_snakemake_2012] and
+relied on Python [@van_rossum_python_2009], R [@r_core_team_r_2020], and GNU
+bash. Dependencies include mothur v1.45.0 [@schloss_introducing_2009], vsearch
+v2.13.3 [@rognes_vsearch_2016], numpy [@harris_array_2020], the Tidyverse
+metapackage [@wickham_welcome_2019], R Markdown [@xie_r_2018], and the conda
+environment manager [@noauthor_anaconda_2016]. A reproducible version of the
+workflow, manuscript, and conda environment is available at 
+**TODO: UPDATED REPO LINK**.
 
@@
 KLS received support from the NIH Training Program in Bioinformatics (T32 GM070449).
+
+PDS received support from **TODO: Pat's grant(s)**.
 

### 2020-12-18 exploratory/2020-11_Nov-Dec/debug-split.Rmd

---
date: 'Nov. - Dec. 2020'
output: 
  github_document:
    html_preview: false
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      fig.path = "figures/"
                      )
```
```{r lib}
set.seed(2018)
library(cowplot)
library(ggtext)
library(glue)
library(here)
library(knitr)
library(tidyverse)

theme_set(theme_classic())
color_palette <- RColorBrewer::brewer.pal(4, "Dark2")
dataset_colors <- c(
  human = color_palette[[3]],
  marine = color_palette[[1]],
  mouse = color_palette[[4]],
  soil = color_palette[[2]]
)
tri_colors <-  c("darkorange","darkorchid","cyan4") # https://allisonhorst.github.io/palmerpenguins/articles/intro.html
mutate_perf <- function(dat) {
  dat %>% 
    mutate(mem_mb = max_rss,
           mem_gb = mem_mb / 1024) %>% 
    rename(sec = s,
           num_otus = sobs)
}
```

## _de novo_ clustering

```{r load_opticlust}
opticlust <- read_tsv(here('subworkflows/1_prep_samples/results/opticlust_results.tsv')) %>% 
  full_join(read_tsv(here('subworkflows/1_prep_samples/results/dataset_sizes.tsv'))) %>% 
  mutate_perf()
sum_opticlust <- opticlust %>% 
  group_by(dataset) %>% 
  summarize(mean_mcc = mean(mcc),
            sd_mcc = sd(mcc),
            mean_sec = mean(sec),
            sd_sec = sd(sec),
            mean_mem_gb = mean(mem_gb))#,
            #mean_shannon = mean(shannon))
```

## fit to reference databases

```{r load_fit-dbs}
ref_sizes <- read_tsv(here('subworkflows/2_fit_reference_db/results/ref_sizes.tsv'))
optifit_dbs <- read_tsv(here('subworkflows/2_fit_reference_db/results/optifit_dbs_results.tsv')) %>% 
  mutate_perf()
head(optifit_dbs)
```

Plotted the fraction mapped for open-reference clustering too as a sanity check.
Final plots to be included in the paper won't include open-ref.
```{r fraction-mapped-open_fit-db}

optifit_dbs %>% #filter(method == 'closed') %>% 
  ggplot(aes(x=dataset, y=fraction_mapped, color=ref)) +
  geom_boxplot() + 
  facet_wrap('method') +
  ylim(0, 1) +
  scale_color_manual(values=tri_colors) +
  labs(title="Fit to reference database") +
  theme(plot.caption = element_markdown())
```

## fit split datasets

Using buggy data from commit `824c1306eb464349f9d331f85550d539f8c66c63`.

```{bash, eval = FALSE}
wget https://raw.githubusercontent.com/SchlossLab/OptiFitAnalysis/824c1306eb464349f9d331f85550d539f8c66c63/subworkflows/3_fit_sample_split/results/optifit_split_results.tsv?token=AEHR6TKMLXIG57IGER2VXSS74C5WC -O exploratory/2020-11_Nov-Dec/debug-split_100-seeds.tsv
```

```{r load_fit-split}
optifit_split <- read_tsv(here('exploratory',
                               '2020-11_Nov-Dec',
                               'debug-split_100-seeds.tsv')) %>% 
  mutate_perf()
```

```{r debug-mcc_fit-split}
optifit_split %>% 
  ggplot(aes(x = ref_frac, y = mcc, color = ref_weight)) +
  geom_jitter(size = 1, alpha = 0.3, width = 0.01) +
  stat_summary(fun = mean, geom = 'crossbar', alpha = 0.5) +
  geom_hline(aes(yintercept = mean_mcc), sum_opticlust) +
  facet_grid(dataset ~ method) +
  ylim(0, 1) +
  labs(title='OTU Quality: OptiFit with split datasets',
       x='reference fraction',
       caption='Black line: _de novo_ clustering on the whole dataset') +
  theme(plot.caption = element_markdown())
```

### fraction mapped

```{r debug-fraction-mapped_fit-split}
optifit_split %>% 
  filter(method == 'closed', tool == 'mothur', is.na(ref)) %>% 
  #group_by(dataset, ref_weight, ref_frac) %>% 
  ggplot(aes(x=ref_frac, y=fraction_mapped, color=ref_weight)) +
  geom_point(alpha = 0.5, size=1) +
  facet_wrap("dataset", nrow=1) +
  ylim(0, 1) +
  labs(title="Sequences mapped during closed-reference OptiFit",
       x='reference fraction',
       y='fraction mapped')
```

It looks like one result per parameter set is left over from before I re-ran 
everything after reverting to the older silva version.
[Here's what the plot looked like with the newest (buggy) silva](https://github.com/SchlossLab/OptiFitAnalysis/blob/master/exploratory/2020-05/sub3_fit_all-seqs.md#fraction-of-sequences-that-map-to-the-reference).
Need to go back and figure out why those files didn't get removed/overwritten
by Snakemake when I re-ran everything.

Let's try a sanity check with open-reference -- should all be 1.

```{r debug-fraction-mapped-open_fit-split}
optifit_split %>% 
  filter(tool == 'mothur', is.na(ref)) %>% 
  #filter(dataset == 'soil') %>% 
  ggplot(aes(x=ref_frac, y=fraction_mapped, color=ref_weight)) +
  geom_point(alpha = 0.5, size=1) +
  facet_grid(method ~ dataset) +
  ylim(0, 1) +
  labs(title="Fit dataset splits",
       x='reference fraction',
       y='fraction mapped')
```

Did mothur's list file format change in the latest dev version? The fraction
mapped values are the same as the reference fraction when the ref weight is
simple.

### is there a pattern?

```{r debug-pattern}
debug_dat <- optifit_split %>% 
  filter(method == 'open') %>% 
  mutate(is_correct = fraction_mapped == 1) %>% 
  select(dataset, seed, method, fraction_mapped, is_correct, 
         sample_frac, ref_frac, ref_weight)

debug_dat %>% filter(is_correct)

```

Can't discern any pattern...

## smaller dataset

Using buggy data from commit `1de7b839eae9b222b66f15dd24b0bb54a8a35459`.

```{bash, eval = FALSE}
wget https://raw.githubusercontent.com/SchlossLab/OptiFitAnalysis/1de7b839eae9b222b66f15dd24b0bb54a8a35459/subworkflows/3_fit_sample_split/results/optifit_split_results.tsv?token=AEHR6TJQFZT4XT6XIYWOKNS74C7ME -O exploratory/2020-11_Nov-Dec/debug-split_2-seeds.tsv
```

```{r debug-2_fit-split}
split_2 <- read_tsv(here('exploratory',
                          '2020-11_Nov-Dec',
                          'debug-split_2-seeds.tsv')) %>%
  mutate_perf() %>% 
  filter(method == 'open') %>% 
  mutate(is_correct = fraction_mapped == 1) %>% 
  select(dataset, seed, method, fraction_mapped, is_correct, 
         sample_frac, ref_frac, ref_weight)

split_2 %>% 
  filter(dataset == 'soil', ref_weight == 'simple') %>% 
  select(sample_frac, ref_frac, seed, is_correct) %>% 
  pivot_wider(names_from = seed, 
              values_from = is_correct,
              names_glue = 'seed_{seed}_is_correct')

```

The pattern of correctness among ref fracs is flipped between seed 1 and 2

Whenever the fit seed matches the best cluster seed,
the fraction mapped results are correct.
The bug disappears if you only use 1 seed, because then
all fit runs will use the same seed as the data split.

### 2020-12-18 exploratory/2020-11_Nov-Dec/frac-map-split-bug.tex

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=0.5in]{geometry}
\usepackage{amsfonts}

\begin{document}
\begin{flushleft}
\Large{\textbf{Explaining the fraction mapped bug}} \\
Kelly L. Sovacool \\
16 Dec. 2020
\end{flushleft}

\subsection*{The Bug}
For each seed, the dataset was split into a reference and query fraction.
Then, reference sequences were clustered \textit{de novo} into OTUs with OptiClust.
The seed which produced the best \textit{de novo} OTUs was selected and used as the set of reference OTUs for all downstream OptiFit runs.
For each seed, the query sequences were fit to the reference OTUs from the best clustering seed with OptiFit.
This was a problem because the same seed variable was used for the dataset split, OptiClust, and OptiFit steps.
Whenever the current OptiFit seed corresponded to the best OptiClust seed,
the query and reference sequences were from the same data split and the results were correct;
in all other cases they were from different splits and the results were unexpected.

With open-reference clustering, any query sequences which can't initially be mapped to the reference OTUs are then clustered \textit{de novo},
so that all query sequences end up in the final OptiFit OTUs.
Thus, the fraction of query sequences which mapped to the reference should always be 1 with open-reference clustering.
However, with this bug, the fraction mapped was less than 1 whenever the OptiFit seed was different from the best OptiClust seed.

To fix the bug, I removed the step that picked the best seed from OptiClust for all downstream OptiFit runs.
This step caused query sets to be fit to reference sets from different data splits.
It was not necessary anyway since OptiClust results were remarkably similar across different seeds.
If we wanted to keep this step, an alternate fix would be to have multiple seed variables
(i.e. one each for the data split and clustering steps)
to prevent query sequences from one data split being fit to reference sequences from a different data split.

\subsection*{Why fraction mapped = query fraction}
\begin{itemize}
    \item $q$: fraction of sequences in the query.
    \item $1 - q$: fraction of sequences in the reference.
    \item $m$: the mapped sequences as a fraction of the total sequences.
    \item $\frac{m}{q}$: fraction of query sequences that mapped.
    \item $u$: the unmapped sequences as a fraction of the total.
    \item $\frac{u}{q}$: fraction of query sequences that did not map.
\end{itemize}

Interestingly, with open-reference clustering and a simple random sampling method for splitting the dataset into reference and query fractions,
\textbf{the fraction of query sequences which mapped to the reference was always $q$}.

$$\frac{m}{q} = q $$
$$ m = q^2 $$:
$$\frac{u}{q} = 1 - \frac{m}{q} $$
$$\frac{u}{q} = 1 - \frac{q^2}{q} $$
 $$ u = q ( 1 - q )$$

So we expect the unmapped sequences as a fraction of the total to be $q (1 - q)$
based on the observation that the fraction of query sequences which mapped to the reference was always $q$.

Let $A$ be the event that sequence X is selected for the query.
Let $B$ be the event that sequence X is selected for the reference.
$A$ and $B$ are independent events. When a simple random sample is used,
the probability that sequence X is selected for the query with seed 1 and sequence X is selected for the reference with seed 2 is:

$$ \mathbb{P}(A) \mathbb{P}(B) = q ( 1 - q ) $$

When running OptiFit with the printref parameter set to false, the reference sequences are not written to the output list file.
It seems that this includes any reference sequences that are also in the query.
Therefore, the sequences that are in both the query and reference are counted as unmapped, i.e.

$$ u = q ( 1 - q )  = \mathbb{P}(A) \mathbb{P}(B)$$

\end{document}

### 2020-12-18 exploratory/2020-11_Nov-Dec/plots.Rmd
Knit

@@
 
-It looks like one result per parameter set is left over from before I re-ran 
-everything after reverting to the older silva version.
-[Here's what the plot looked like with the newest (buggy) silva](https://github.com/SchlossLab/OptiFitAnalysis/blob/master/exploratory/2020-05/sub3_fit_all-seqs.md#fraction-of-sequences-that-map-to-the-reference).
-Need to go back and figure out why those files didn't get removed/overwritten
-by Snakemake when I re-ran everything.
-
 Let's try a sanity check with open-reference -- should all be 1.
@@
 ```
-
-Did mothur's list file format change in the latest dev version? The fraction
-mapped values are the same as the reference fraction when the ref weight is
-simple.
-
-### is there a pattern?
-
-```{r debug_frac_map}
-debug_dat <- optifit_split %>% 
-  filter(method == 'open') %>% 
-  mutate(is_correct = fraction_mapped == 1) %>% 
-  select(dataset, seed, method, fraction_mapped, is_correct, 
-         sample_frac, ref_frac, ref_weight) %>%
-  group_by(dataset, seed, ref_frac, ref_weight, is_correct)
-
-for (s in 1:2) {
-  debug_dat %>% 
-    filter(seed == s, ref_weight == 'simple', dataset == 'soil') %>% 
-    pull(is_correct) %>% 
-    print()
-}
-
-```
-
-Pattern of correctness among ref fracs is flipped between seed 1 and 2
 

### 2021-01-22 paper/paper.Rmd
Draft results & discussion

@@
 
-### Importance
-
-\newpage
-
-## Introduction
-
 Assigning amplicon sequences to Operational Taxonomic Units (OTUs) is an
@@
 
-- TODO: Describe OptiClust and vsearch.
+- **TODO: summarize results & conclusion**
+
+### Importance
+
+**TODO**
+
+\newpage
+
+## Introduction
+
+- **TODO: broad context; importance of amplicon sequencing**
+- **TODO: describe reference vs de novo clustering. trade-offs. machine learning use-case.**
+- **TODO: mention OptiClust and vsearch.**
 
@@
 
-- TODO: brief description of OptiFit algorithm. the method we came up with is the result. OptiClust paper had toy example to walk through the algorithm. 
-- TODO: Paragraph at beginning to describe datasets: why picked, how processed.
+### OptiFit algorithm
 
-We used four published datasets isolated from soil
-[@johnston_metagenomics_2016], marine [@henson_artificial_2016], mouse
-[@schloss_stabilization_2012], and human [@baxter_microbiota-based_2016]
-samples.
+OptiFit leverages the method employed by OptiClust of iteratively assigning sequences to OTUs to produce the highest quality OTUs possible, and extends this method for reference-based clustering.
 
-- 16S rRNA gene sequence datasets from human gut, mouse gut, marine, and soil environments as well as the greengenes, silva, and RDP public databases were processed with mothur and clustered _de novo_ with mothur's OptiClust algorithm.
-- All clustering was performed at a distance threshold of 0.03 and repeated with 100 different random seeds.
-- The Matthews Correlation Coefficient was calculated to serve as a measure of OTU quality. OTU quality was evaluated using the Matthews Correlation
-Coefficient (MCC) with a sequence dissimilarity threshold of 0.03% as described
-previously [@schloss_application_2016; @westcott_novo_2015].
+- **TODO: brief description of OptiFit algorithm**
+- **TODO: open & closed reference modes**
+- **TODO: mcc score; only considers query seqs (when printref=f)**
+
+To evaluate the OptiFit algorithm and compare to existing methods, we used four
+published datasets isolated from soil [@johnston_metagenomics_2016], marine
+[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
+gut [@baxter_microbiota-based_2016] samples. There are two strategies for
+generating OTUs with OptiFit: 1) fit sequences to reference OTUs of an
+independent database, or 2) split the dataset into a reference and query
+fraction, then fit the query sequences to OTUs generated by clustering the
+reference sequences _de novo_. For each dataset repeated with 100 random seeds,
+we generated OTUs with OptiFit using both strategies, and also clustered _de
+novo_ OTUs with OptiClust for comparison. All clustering was performed at a
+sequence dissimilarity threshold of 0.03 and OTU quality was evaluated using the
+Matthews Correlation Coefficient (MCC) as described previously
+[@schloss_application_2016; @westcott_novo_2015]. We calculated the fraction of
+query sequences that were fit to existing OTUs in closed-reference mode as an
+additional measure of quality for this mode.
 
@@
 
-- Public reference databases were clustered _de novo_ using OptiClust, then query datasets were fit to the _de novo_ OTUs using OptiFit.
-  - In open-reference mode, OTU quality was similar between fitting the datasets to reference OTUs with OptiFit and clustering the datasets _de novo_ with OptiClust.
-This held true for all datasets and reference databases.
-  - However, in closed-reference mode, OTU quality was slightly worse when fitting to greengenes and silva, and much worse when fitting to RDP as compared to OptiClust.
-OptiFit was able to map more query sequences to reference OTUs created with the greengenes and silva databases than with RDP.
-  - In terms of runtime, closed-reference OptiFit outperformed OptiClust, while OptiClust out-performed open-reference OptiFit.
+To evaluate reference-based clustering with independent databases, we fit each
+dataset to reference OTUs generated by _de novo_ clustering the Greengenes
+database (v13_8_99), Silva non-redundant database (v132), and the Ribosomal
+Database Project (RDP; v16). In open-reference mode, OTU quality was similar
+between fitting the datasets to reference OTUs with OptiFit and clustering the
+datasets _de novo_ with OptiClust. This held true for all datasets and reference
+databases. However, in closed-reference mode, OTU quality was slightly worse
+when fitting to greengenes and silva, and much worse when fitting to RDP as
+compared to OptiClust. No more than half of query sequences were fit to
+reference OTUs in closed-reference mode across any dataset/database combination.
+OptiFit was able to fit more query sequences to reference OTUs created with the
+greengenes and silva databases than with RDP. In terms of runtime,
+closed-reference OptiFit outperformed OptiClust, while OptiClust outperformed
+open-reference OptiFit.
+
+To compare to existing software, vsearch was used to cluster OTUs _de novo_ or
+with reference-based clustering to the greengenes database. For all datasets and
+clustering methods (_de novo_, open reference, and closed reference), mothur's
+clustering algorithms produced higher quality OTUs than vsearch. When
+closed-reference clustering against the greengenes database, vsearch was able to
+map more query sequences to the reference than mothur's OptiFit algorithm. In
+terms of runtime, OptiFit generally performed faster than vsearch when reference
+clustering, while vsearch _de novo_ clustering outperformed OptiClust.
 
@@
 
-- TODO: compare optifit performance for split dataset vs public database
-- TODO: double-check with Sarah that MCCs are from full OTU dataset or just query sequnce OTUs
-- TODO: double-check reference fractions aren't flipped (fraction mapped plot looks weird)
+Datasets were randomly split into a reference fraction and a query fraction.
+Reference sizes from 10% to 80% of the sequences were created, with the
+remaining sequences used for the query. Reference sequences were clustered _de
+novo_ with OptiClust, then query sequences were fit to the _de novo_ OTUs with
+OptiFit.
 
-- Datasets were randomly split into a reference fraction and a query fraction.
-Reference sizes from 10% to 80% of the sequences were created, with the remaining sequences used for the query.
-Reference sequences were clustered _de novo_ with OptiClust, then query sequences were fit to the _de novo_ OTUs with OptiFit.
-  - OTU quality from fitting split datasets was highly similar to that from _de novo_ clustering the whole dataset.
-  - Closed-reference OptiFit with split datasets was faster than OptiClust on whole datasets.
-    - OptiClust performed faster than open-reference OptiFit only when the OptiFit reference fraction was 30% or less.
-  - Different methods for selecting the sequences to be used as the reference were tested; a simple random sample, weighting sequences by relative abundance, and weighting by similarity to other sequences in the dataset.
-    - OTU quality was similar with the simple and abundance-weighted sampling, but slightly worse with similarity-weighted sampling.
-    - The fraction of query sequences that are able to be fit to the reference OTUs in closed-reference mode decreases as the reference fraction increases.
+OTU quality from the split dataset strategy with OptiFit was highly similar to
+that from _de novo_ clustering the whole dataset with OptiClust regardless of
+mode. OTU quality was remarkably stable across 100 different random seeds. In
+terms of runtime, closed-reference OptiFit performed faster than OptiClust on
+whole datasets. In open-reference mode, OptiClust performed faster than OptiFit
+only when the OptiFit reference fraction was 30% or less. The split dataset
+strategy performed just as well as the database strategy in open-reference mode
+regardless of database used, and outperformed the database strategy in
+closed-reference mode.
 
-### Comparison to vsearch
-
-- TODO: move this into first section with public database.
-- vsearch is more permissive than mothur (radius vs diameter with 0.03 dissimilarity threshold).
-
-- To compare to existing software, vsearch was used to cluster OTUs _de novo_ or with reference-based clustering to the greengenes database.
-  - For all datasets and clustering methods (_de novo_, open reference, and closed reference), mothur's clustering algorithms produced higher quality OTUs than vsearch.
-  - When closed-reference clustering against the greengenes database, vsearch was able to map more query sequences to the reference than mothur's OptiFit algorithm.
-  - In terms of runtime, OptiFit genearlly performed faster than vsearch when reference clustering, while vsearch _de novo_ clustering outperformed OptiClust.
+We also tested three methods for selecting the sequences to be used as the
+reference; a simple random sample, weighting sequences by relative abundance,
+and weighting by similarity to other sequences in the dataset. OTU quality was
+similar with the simple and abundance-weighted sampling, but slightly worse with
+similarity-weighted sampling. In closed-reference mode, The fraction of query
+sequences that can be fit to the reference OTUs decreases as the reference
+fraction increases.
 
@@
 
+We developed a new algorithm for fitting sequences to existing OTUs and have
+demonstrated its suitability for reference-based clustering. OptiFit makes the
+iterative method employed by OptiClust available for tasks where
+reference-based clustering is required. We have shown that OTU quality is
+similar between OptiClust and OptiFit in open-reference mode, regardless of
+strategy employed. Open-reference OptiFit does perform slower than OptiClust
+due to the additional _de novo_ clustering step, so users may prefer OptiClust
+for tasks that do not require reference OTUs.
+
+When fitting to public databases, OTU quality dropped in closed-reference mode
+to different degrees depending on the database and dataset source, and no more
+than half of query sequences were able to be fit to OTUs across any
+dataset/database combination. This may reflect limitations of reference
+databases, which are unlikely to contain sequences from rare and novel microbes.
+This drop in quality was most notable with RDP. We recommend users who require
+an independent reference database opt for Greengenes or Silva instead. Since
+OptiClust performs faster than open-reference OptiFit and creates higher quality
+OTUs than closed-reference OptiFit with the database strategy, we recommend
+using OptiClust rather than fitting to a database where possible.
+
+The mothur algorithms produced higher quality OTUs than vsearch in
+open-reference, closed-reference, or _de novo_ modes. However, vsearch was able
+to fit more sequence into OTUs than OptiFit in closed-reference mode. While both
+mothur and vsearch use a dissimilarity threshold for determining how to assign
+sequences into OTUs, vsearch is more permissive than mothur. Mothur requires
+that all pairs of sequences in an OTU are within the dissimilarity threshold
+without penalizing the MCC, while vsearch only requires sequences to be similar
+to one other sequence in the OTU. In this way, vsearch sacrifices OTU quality in
+order to allow more sequences to fit to OTUs. Users who require closed-reference
+clustering may prefer to use vsearch if they wish to maximize the fraction of
+sequences that can be fit at the cost of OTU quality. However, mothur's
+OptiClust or OptiFit are preferred for _de novo_ or open-reference clustering.
+
+When fitting with the split dataset strategy, OTU quality was remarkably similar
+when reference sequences were selected by a simple random sample or weighted by
+abundance, but quality was slightly worse when sequences were weighted by
+similarity. We recommend using a simple random sample since the more
+sophisticated reference selection methods do not offer any benefit. The
+similarity in OTU quality between OptiClust and OptiFit with this strategy
+demonstrates the suitability of using OptiFit to fit sequences to existing OTUs,
+such as when using already-trained machine learning models to make predictions
+on new data.
+
+- **TODO: big picture concluding paragraph**
+
 ## Materials and Methods
 
-### Sequence Data Processing Steps
+### Data Processing Steps
 
@@
 relied on Python [@van_rossum_python_2009], R [@r_core_team_r_2020], and GNU
-bash. Dependencies include mothur v1.45.0 [@schloss_introducing_2009], vsearch
+bash. Software used includes mothur v1.45.0 [@schloss_introducing_2009], vsearch
 v2.13.3 [@rognes_vsearch_2016], numpy [@harris_array_2020], the Tidyverse
 metapackage [@wickham_welcome_2019], R Markdown [@xie_r_2018], and the conda
-environment manager [@noauthor_anaconda_2016]. A reproducible version of the
-workflow, manuscript, and conda environment is available at 
-**TODO: UPDATED REPO LINK**.
+environment manager [@noauthor_anaconda_2016]. The complete workflow,
+manuscript, and conda environment are available at **TODO: UPDATED REPO LINK**.
 

### 2021-01-26 paper/paper.Rmd
Introduce algorithms

@@
 
-- **TODO: broad context; importance of amplicon sequencing**
-- **TODO: describe reference vs de novo clustering. trade-offs. machine learning use-case.**
+Amplicon sequencing has become a mainstay of microbial ecology and
+host-associated microbiome research. Researchers can affordably generate
+millions of sequences to characterize the composition of hundreds of samples
+from culture-independent microbial communities. In a typical analysis pipeline,
+16S rRNA gene sequences are assigned to Operational Taxonomic Units (OTUs) to
+facilitate comparison of taxonomic composition between communities. A distance
+threshold of 3% (or sequence similarity of 97%) is commonly used to cluster
+sequences into OTUs based on either a reference database or pairwise comparisons
+of the sequences within the dataset. The method chosen for clustering affects
+the quality of OTU assignments and thus may impact downstream analyses of 
+community composition [@westcott_opticlust_2017; @schloss_application_2016; @westcott_novo_2015].
+
+Closed-reference methods assign sequences to a set of pre-made OTUs generated
+from reference sequences. If a query sequence is not within the distance
+threshold to any of the reference sequences, it is discarded. _De novo_ methods
+cluster sequences based on their distance to each other, without the use of an
+external reference. Open-reference methods first perform closed-reference
+clustering, then any sequences that cannot be assigned to reference OTUs are
+clustered _de novo_ to create additional OTUs. While reference-based clustering
+is generally fast, it is limited by the diversity represented in the reference
+database. _De novo_ clustering overcomes the limitations of reference databases
+by considering only sequences in the dataset, but is more computationally
+intensive and generates different OTU assignments when new sequences are
+introduced.
+
+- **TODO: clustering methods. describe reference vs de novo clustering. trade-offs. machine learning use-case.**
 - **TODO: mention OptiClust and vsearch.**
+- **TODO:** gap in knowledge/ problem statement.
+acknowledge knock against OTUs, that they're not reproducible across studies.
+idea of ref-clustering is you can use existing OTUs to compare data across studies.
+can we make a ref-based method that's as good as _de novo_.
+- **TODO:** in this study, we fix that problem.
 
@@
 
-OptiFit leverages the method employed by OptiClust of iteratively assigning sequences to OTUs to produce the highest quality OTUs possible, and extends this method for reference-based clustering.
+OptiFit leverages the method employed by OptiClust of iteratively assigning
+sequences to OTUs to produce the highest quality OTUs possible, and extends this
+method for reference-based clustering.
 
@@
 
+- TODO: separate paragraphs for closed & open reference clustering. weave in vsearch within those paragraphs. mention median mcc values.
+- TODO: avoid slightly/much -- use numbers, e.g. X% better/worse.
+- TODO: put the comparison at the beginning of the sentence, so you don't have to know what it is at the end.
+
 To evaluate reference-based clustering with independent databases, we fit each
@@
 reference OTUs in closed-reference mode across any dataset/database combination.
+**TODO: specify exact numbers for fraction mapped.**
 OptiFit was able to fit more query sequences to reference OTUs created with the
@@
 
+- **TODO:** for these data, we don't see a compelling reason to use reference-based clustering over _de novo_.
+it was supposed to speed things up.
+the reason to do reference-based is if you like the ref OTUs -- e.g.
+for ML or downstream tools e.g.
+PiCrust?
+- **TODO:** highlight difference between what we're doing and what previously was done.
+  others use single ref seq to define an otu, while we use all the ref & query seqs to define an otu. highlight why ours is better than previous methods.
+
 We developed a new algorithm for fitting sequences to existing OTUs and have
@@
 using OptiClust rather than fitting to a database where possible.
+(**Pat: if you don't have breadth, closed-ref will suck.**)
 

### 2021-01-29 paper/paper.Rmd
More details

@@
 
-Closed-reference methods assign sequences to a set of pre-made OTUs generated
-from reference sequences. If a query sequence is not within the distance
-threshold to any of the reference sequences, it is discarded. _De novo_ methods
-cluster sequences based on their distance to each other, without the use of an
-external reference. Open-reference methods first perform closed-reference
+There are three main categories of OTU clustering algorithms: closed reference,
+open reference, and _de novo_ clustering. Closed reference methods assign
+sequences to a set of pre-made OTUs generated from reference sequences. If a
+query sequence is not within the distance threshold to any of the reference
+sequences, it is discarded. While reference-based clustering is generally fast,
+it is limited by the diversity represented in the reference database. _De novo_
+methods cluster sequences based on their distance to each other, without the use
+of an external reference. open reference methods first perform closed reference
 clustering, then any sequences that cannot be assigned to reference OTUs are
-clustered _de novo_ to create additional OTUs. While reference-based clustering
-is generally fast, it is limited by the diversity represented in the reference
-database. _De novo_ clustering overcomes the limitations of reference databases
-by considering only sequences in the dataset, but is more computationally
-intensive and generates different OTU assignments when new sequences are
-introduced.
+clustered _de novo_ to create additional OTUs.  _De novo_ clustering overcomes
+the limitations of reference databases by considering only sequences in the
+dataset, but is more computationally intensive and generates different OTU
+assignments when new sequences are introduced. Previous studies found that _de
+novo_ clustering created the highest quality OTU assignments based on the
+Matthews correlation coefficient (MCC) [@schloss_application_2016;
+@westcott_novo_2015].
 
@@
 [@schloss_application_2016; @westcott_novo_2015]. We calculated the fraction of
-query sequences that were fit to existing OTUs in closed-reference mode as an
+query sequences that were fit to existing OTUs in closed reference mode as an
 additional measure of quality for this mode.
@@
 database (v13_8_99), Silva non-redundant database (v132), and the Ribosomal
-Database Project (RDP; v16). In open-reference mode, OTU quality was similar
+Database Project (RDP; v16). In open reference mode, OTU quality was similar
 between fitting the datasets to reference OTUs with OptiFit and clustering the
 datasets _de novo_ with OptiClust. This held true for all datasets and reference
-databases. However, in closed-reference mode, OTU quality was slightly worse
+databases. However, in closed reference mode, OTU quality was slightly worse
 when fitting to greengenes and silva, and much worse when fitting to RDP as
 compared to OptiClust. No more than half of query sequences were fit to
-reference OTUs in closed-reference mode across any dataset/database combination.
+reference OTUs in closed reference mode across any dataset/database combination.
 **TODO: specify exact numbers for fraction mapped.**
@@
 greengenes and silva databases than with RDP. In terms of runtime,
-closed-reference OptiFit outperformed OptiClust, while OptiClust outperformed
-open-reference OptiFit.
+closed reference OptiFit outperformed OptiClust, while OptiClust outperformed
+open reference OptiFit.
 
@@
 clustering algorithms produced higher quality OTUs than vsearch. When
-closed-reference clustering against the greengenes database, vsearch was able to
+closed reference clustering against the greengenes database, vsearch was able to
 map more query sequences to the reference than mothur's OptiFit algorithm. In
@@
 mode. OTU quality was remarkably stable across 100 different random seeds. In
-terms of runtime, closed-reference OptiFit performed faster than OptiClust on
-whole datasets. In open-reference mode, OptiClust performed faster than OptiFit
+terms of runtime, closed reference OptiFit performed faster than OptiClust on
+whole datasets. In open reference mode, OptiClust performed faster than OptiFit
 only when the OptiFit reference fraction was 30% or less. The split dataset
-strategy performed just as well as the database strategy in open-reference mode
+strategy performed just as well as the database strategy in open reference mode
 regardless of database used, and outperformed the database strategy in
-closed-reference mode.
+closed reference mode.
 
@@
 similar with the simple and abundance-weighted sampling, but slightly worse with
-similarity-weighted sampling. In closed-reference mode, The fraction of query
+similarity-weighted sampling. In closed reference mode, The fraction of query
 sequences that can be fit to the reference OTUs decreases as the reference
@@
 reference-based clustering is required. We have shown that OTU quality is
-similar between OptiClust and OptiFit in open-reference mode, regardless of
-strategy employed. Open-reference OptiFit does perform slower than OptiClust
+similar between OptiClust and OptiFit in open reference mode, regardless of
+strategy employed. open reference OptiFit does perform slower than OptiClust
 due to the additional _de novo_ clustering step, so users may prefer OptiClust
@@
 
-When fitting to public databases, OTU quality dropped in closed-reference mode
+When fitting to public databases, OTU quality dropped in closed reference mode
 to different degrees depending on the database and dataset source, and no more
@@
 an independent reference database opt for Greengenes or Silva instead. Since
-OptiClust performs faster than open-reference OptiFit and creates higher quality
-OTUs than closed-reference OptiFit with the database strategy, we recommend
+OptiClust performs faster than open reference OptiFit and creates higher quality
+OTUs than closed reference OptiFit with the database strategy, we recommend
 using OptiClust rather than fitting to a database where possible.
-(**Pat: if you don't have breadth, closed-ref will suck.**)
+(**TODO: "if you don't have breadth, closed ref will suck." - Pat**)
 
 The mothur algorithms produced higher quality OTUs than vsearch in
-open-reference, closed-reference, or _de novo_ modes. However, vsearch was able
-to fit more sequence into OTUs than OptiFit in closed-reference mode. While both
+open reference, closed reference, or _de novo_ modes. However, vsearch was able
+to fit more sequence into OTUs than OptiFit in closed reference mode. While both
 mothur and vsearch use a dissimilarity threshold for determining how to assign
@@
 to one other sequence in the OTU. In this way, vsearch sacrifices OTU quality in
-order to allow more sequences to fit to OTUs. Users who require closed-reference
+order to allow more sequences to fit to OTUs. Users who require closed reference
 clustering may prefer to use vsearch if they wish to maximize the fraction of
 sequences that can be fit at the cost of OTU quality. However, mothur's
-OptiClust or OptiFit are preferred for _de novo_ or open-reference clustering.
+OptiClust or OptiFit are preferred for _de novo_ or open reference clustering.
 

### 2021-02-08 exploratory/2021/plots.Rmd
Re-make plots with RDP v18

--- 2020-12-18 exploratory/2020-11_Nov-Dec/plots.Rmd
@@
 ---
-date: 'Nov. - Dec. 2020'
+date: '`r Sys.Date()`'
 output: 

### 2021-02-08 exploratory/2021/plots.Rmd
Fix header

@@
 ```
+
 ## fit to reference databases
@@
   mutate_perf()
-head(optifit_dbs)
 ```

### 2021-02-15 paper/paper.Rmd
Re-make plots with RDP v18
Capitalization

@@
 the quality of OTU assignments and thus may impact downstream analyses of
-community composition [@westcott_opticlust_2017; @schloss_application_2016; @westcott_novo_2015].
+community composition [@westcott_opticlust_2017; @schloss_application_2016;
+@westcott_novo_2015].
 
@@
 - **TODO: open & closed reference modes**
-- **TODO: mcc score; only considers query seqs (when printref=f)**
+- **TODO: MCC score; only considers query sequences (when printref=f)**
 
@@
 
-- TODO: separate paragraphs for closed & open reference clustering. weave in vsearch within those paragraphs. mention median mcc values.
+- TODO: separate paragraphs for closed & open reference clustering. weave in vsearch within those paragraphs. mention median MCC values.
 - TODO: avoid slightly/much -- use numbers, e.g. X% better/worse.
@@
 databases. However, in closed reference mode, OTU quality was slightly worse
-when fitting to greengenes and silva, and much worse when fitting to RDP as
+when fitting to Greengenes and Silva, and much worse when fitting to RDP as
 compared to OptiClust. No more than half of query sequences were fit to
@@
 OptiFit was able to fit more query sequences to reference OTUs created with the
-greengenes and silva databases than with RDP. In terms of runtime,
+Greengenes and Silva databases than with RDP. In terms of run time,
 closed reference OptiFit outperformed OptiClust, while OptiClust outperformed
@@
 - **TODO:** highlight difference between what we're doing and what previously was done.
-  others use single ref seq to define an otu, while we use all the ref & query seqs to define an otu. highlight why ours is better than previous methods.
+  others use single ref seq to define an OTU, while we use all the ref & query seqs to define an otu. highlight why ours is better than previous methods.
 

### 2021-02-23 paper/paper.Rmd
Problem statement

@@
 sequences, it is discarded. While reference-based clustering is generally fast,
-it is limited by the diversity represented in the reference database. _De novo_
-methods cluster sequences based on their distance to each other, without the use
-of an external reference. open reference methods first perform closed reference
-clustering, then any sequences that cannot be assigned to reference OTUs are
-clustered _de novo_ to create additional OTUs.  _De novo_ clustering overcomes
+it is limited by the diversity represented in the reference database. Rare or
+novel sequences in the sample will be lost if they are not represented in the
+database. _De novo_ methods cluster sequences based on their distance to each
+other, without the use of an external reference ._De novo_ clustering overcomes
 the limitations of reference databases by considering only sequences in the
 dataset, but is more computationally intensive and generates different OTU
-assignments when new sequences are introduced. Previous studies found that _de
-novo_ clustering created the highest quality OTU assignments based on the
+assignments when new sequences are introduced.  Open reference methods take a
+hybrid approach, first performing closed reference clustering, then any
+sequences that cannot be assigned to reference OTUs are clustered _de novo_ to
+create additional OTUs. Previous studies found that the OptiClust _de novo_
+clustering algorithm created the highest quality OTU assignments based on the
 Matthews correlation coefficient (MCC) [@schloss_application_2016;
-@westcott_novo_2015].
+@westcott_novo_2015]. However, a disadvantage of _de novo_ clustering is that
+different input samples will result in different OTUs, making it difficult to
+compare taxonomic composition of communities between studies or to use machine
+learning models trained with _de novo_ OTUs to make predictions on new data.
 
-- **TODO: clustering methods. describe reference vs de novo clustering. trade-offs. machine learning use-case.**
-- **TODO: mention OptiClust and vsearch.**
-- **TODO:** gap in knowledge/ problem statement.
-acknowledge knock against OTUs, that they're not reproducible across studies.
-idea of ref-clustering is you can use existing OTUs to compare data across studies.
-can we make a ref-based method that's as good as _de novo_.
-- **TODO:** in this study, we fix that problem.
+To overcome the limitations of _de novo_ clustering while maintaining OTU
+quality, we developed OptiFit, a reference-based clustering algorithm in the
+mothur program which takes existing OTUs as the reference to fit new sequences
+to. Here, we tested the OptiFit algorithm with databases or _de novo_ OTUs as
+the reference and compared the performance to existing tools.
 

### 2021-02-23 paper/paper.Rmd
Rephrase _de novo_ limitations

@@
 database. _De novo_ methods cluster sequences based on their distance to each
-other, without the use of an external reference ._De novo_ clustering overcomes
+other, without the use of an external reference. _De novo_ clustering overcomes
 the limitations of reference databases by considering only sequences in the
 dataset, but is more computationally intensive and generates different OTU
-assignments when new sequences are introduced.  Open reference methods take a
-hybrid approach, first performing closed reference clustering, then any
-sequences that cannot be assigned to reference OTUs are clustered _de novo_ to
-create additional OTUs. Previous studies found that the OptiClust _de novo_
-clustering algorithm created the highest quality OTU assignments based on the
-Matthews correlation coefficient (MCC) [@schloss_application_2016;
-@westcott_novo_2015]. However, a disadvantage of _de novo_ clustering is that
-different input samples will result in different OTUs, making it difficult to
-compare taxonomic composition of communities between studies or to use machine
-learning models trained with _de novo_ OTUs to make predictions on new data.
+assignments when new sequences are introduced. Unstable OTU assignments make it
+difficult to use _de novo_ clustering to compare taxonomic composition of
+communities between studies or to use machine learning models trained with _de
+novo_ OTUs to make predictions on new data. Open reference methods take a hybrid
+approach, first performing closed reference clustering, then any sequences that
+cannot be assigned to reference OTUs are clustered _de novo_ to create
+additional OTUs. Previous studies found that the OptiClust _de novo_ clustering
+algorithm created the highest quality OTU assignments of all clustering methods
+based on the Matthews correlation coefficient (MCC) [@schloss_application_2016;
+@westcott_novo_2015].
 
@@
 mothur program which takes existing OTUs as the reference to fit new sequences
-to. Here, we tested the OptiFit algorithm with databases or _de novo_ OTUs as
-the reference and compared the performance to existing tools.
+to. Here, we tested the OptiFit algorithm with the reference as databases or _de
+novo_ OTUs and compared the performance to existing tools.
 

### 2021-03-03 docs/slides/2020-12-14_lab-mtg.Rmd
Make slides self-contained so images will show up on gh-pages

@@
     chakra: libs/remark-latest.min.js
+    self_contained: true
 ---

### 2021-03-03 docs/slides/2021-03-11_bistro.Rmd
Create file for BISTRO pres

--- 2020-12-14 docs/slides/2020-12-14_lab-mtg.Rmd
@@
 author: "Kelly Sovacool"
-date: "2020-12-14 (updated: `r Sys.Date()`)"
+date: "2021-03-11 (updated: `r Sys.Date()`)"
 output:
   xaringan::moon_reader:
-    css: [default, default-fonts]
+    css: libs/xaringan-themer.css
     lib_dir: libs
@@
                   )
+# prev slides:
+# - Oct 2020 research talk https://docs.google.com/presentation/d/1GbbCYmlF4jkrO5BGyCTgtFBOL4pfzPDIQLMZBND0Hzg/edit#slide=id.p
+# - Dec 2020 research talk https://github.com/SchlossLab/OptiFitAnalysis/blob/master/docs/slides/2020-12-14_lab-mtg.Rmd
+# - fall 2019 JC https://github.com/kelly-sovacool/bioinf603-fall2019
+
+# TODO: use http urls instead of here filepaths for figures
+```
+
+```{r xaringan-themer, include=FALSE, warning=FALSE}
+library(xaringanthemer)
+style_mono_light(
+  header_font_google = google_font("Helvetica"),
+  text_font_google = google_font("Helvetica")
+)
+# style_mono_light(base_color = "#23395b")
+# style_mono_accent(
+#   base_color = "#1c5253",
+#   header_font_google = google_font("Josefin Sans"),
+#   text_font_google   = google_font("Montserrat", "300", "300i"),
+#   code_font_google   = google_font("Fira Mono")
+# )
 ```
@@
 
-# Changes since a couple months ago
-
-- Sarah Westcott fixed some bugs in mothur & made OptiFit faster.
-- I re-ran everything with silva v132.
-- Ran vsearch on all the datasets
-- Discovered a bug in my workflow code (last week!)
-
----
-
-# Reference clustering against a database
-
-.pull-left[
-```{r mcc_fit-db}
-include_graphics(here('exploratory', '2020-11_Nov-Dec', 'figures', 'mcc_fit-db-1.png'))
-```
-]
-
---
-
-.pull-right[
-```{r fraction-mapped_fit-db}
-include_graphics(here('exploratory', '2020-11_Nov-Dec', 'figures', 'fraction-mapped_fit-db-1.png'))
-```
-]
-
-???
-- _De novo_ and open-reference perform equally well.
-- Closed reference performs slightly worse.
-- RDP doesn't perform well.
-- Prior to switching to silva v132 & updating mothur, closed-reference artificially outperformed open-reference.
-
----
-
-# Reference clustering against a database
-
-.pull-left[
-```{r runtime_fit-db}
-include_graphics(here('exploratory', '2020-11_Nov-Dec', 'figures', 
-                      'runtime_fit-db-1.png'))
-```
-]
-
---
-
-.pull-right[
-```{r memory_fit-db}
-include_graphics(here('exploratory', '2020-11_Nov-Dec', 'figures', 
-                      'memory_fit-db-1.png'))
-```
-]
-
----
-
-## Summarizing mothur results
-
-Whenever there's virtually no variation in the metric.
-
-```{r sum_all_mothur}
-include_graphics(here('exploratory', '2020-11_Nov-Dec', 'figures', 
-                      'mcc_all-opti-1.png'))
-```
-
----
-
-## mothur vs. vsearch
-
-```{r vsearch}
-include_graphics(here('exploratory', '2020-11_Nov-Dec', 'figures', 
-                      'metrics_all-human-1.png'))
-```
-
-???
-- Just showing the human dataset here because results patterns were basically the same across all datasets.
-- Only used greengenes for reference because that's what
-QIIME does.
-
----
-
-# Splitting datasets into reference & query sets
-
-.pull-left[
-Before reverting to Silva v132
-
-```{r frac-map-old}
-include_graphics(here('exploratory', '2020-05_May-Oct', 'figures', 
-                      'fraction_mapped_complement-2.png'))
-```
-]
---
-.pull-right[
-After reverting to Silva v132:
-
-```{r frac-map-new}
-include_graphics(here('exploratory', 
-                      '2020-11_Nov-Dec', 
-                      'figures', 
-                      'debug-fraction-mapped_fit-split-1.png'))
-```
-]
-
-???
-- 1 seed out of every parameter set has the same pattern as the old results
-- I thought maybe my snakemake workflow somehow kept one result file from before reverting to silva. Like maybe it accidentally didn't overwrite with the new results?
-  - But I checked the timestamps of the results files and they were updated after re-doing with silva v132.
----
-
-## Sanity check: open-ref should have fraction mapped = 1
-
-.pull-left[
-```{r}
-include_graphics(here('exploratory', 
-                      '2020-11_Nov-Dec', 
-                      'figures',
-                      'debug-fraction-mapped-open_fit-split-1.png'))
-```
-]
-
-???
-- Now that is really weird. For weight=simple, the fraction mapped is equal to the sample frac most of the time. So the fraction mapped is incorrect most of the time.
-- Next I thought; maybe there's a bug in mothur where the open-ref output files have one less column than the closed-ref.
-  - I brought this up to Sarah W before and she said she fixed it. Maybe I forgot to recompile mothur?
-  - Bingo!
-I hadn't pulled the latest changes.
-So I recompiled and re-ran with one parameter set & one seed, and it worked!
-Problem solved!
-Well, not quite.
-When re-running with more than one seed, the problem persisted.
----
-
-## is there a pattern?
-
-```{r}
-dat_100 <- read_tsv(here('exploratory',
-                               '2020-11_Nov-Dec',
-                               'debug-split_100-seeds.tsv'))
-dat_100 %>% filter(method == 'open') %>% 
-  mutate(is_correct = fraction_mapped == 1) %>% 
-  select(dataset, seed, is_correct, fraction_mapped,
-         sample_frac, ref_frac, ref_weight) %>% 
-  filter(is_correct, dataset == 'soil')
-```
-
-???
-- is there a bug in my code to calculate the fraction mapped? let's check results from fitting to databases
-
----
-
-## Sanity check: results from fitting to databases
-
-.pull-left[
-```{r}
-include_graphics(here('exploratory', 
-                      '2020-11_Nov-Dec', 
-                      'figures',
-                      'debug-fraction-mapped-open_fit-split-1.png'))
-```
-]
-.pull-right[
-```{r}
-include_graphics(here('exploratory', 
-                      '2020-11_Nov-Dec', 
-                      'figures',
-                      'fraction-mapped-open_fit-db-1.png'))
-```
-]
-???
-- so there isn't a bug in my fraction mapped code and there isn't a bug in mothur, otherwise these would be wrong too.
-- is there a bug in my code to split the datasets into reference & query fractions?
-  - I had written assertion statements throughout that code to make sure the query + reference fraction were correct.
-  - Double-checked the resulting accnos files anyway just to be safe. They were all correct.
----
-
-## WHAT?!
-
-```{r}
-split_2 <- read_tsv(here('exploratory',
-                          '2020-11_Nov-Dec',
-                          'debug-split_2-seeds.tsv')) %>%
-  filter(method == 'open') %>% 
-  mutate(is_correct = fraction_mapped == 1) %>% 
-  select(dataset, seed, method, fraction_mapped, is_correct, 
-         sample_frac, ref_frac, ref_weight)
-
-split_2 %>% 
-  filter(dataset == 'soil', ref_weight == 'simple') %>% 
-  select(sample_frac, ref_frac, seed, is_correct) %>% 
-  pivot_wider(names_from = seed, 
-              values_from = is_correct,
-              names_glue = 'seed_{seed}_is_correct')
-```
-
-Re-ran with only 2 seeds.
-
----
-
-```{bash, eval = FALSE}
-wget https://raw.githubusercontent.com/SchlossLab/OptiFitAnalysis/c900c5fa5f3081b73536a4e568f66817a66b3d0c/figures/rulegraph_fit_split.png?token=AEHR6TPEQM4HG6GCTVOUSTK74DIXK -O exploratory/2020-11_Nov-Dec/figures/rulegraph1.png
-```
-.pull-left[
-Workflow
-
-```{r rulegraph_old, out.width='60%'}
-include_graphics(here('exploratory', 
-                      '2020-11_Nov-Dec', 
-                      'figures',
-                      'rulegraph1.png'))
-```
-]
---
-.pull-right[
-So far:
-- Code shared with  workflow that fits to databases must be correct.
-- Code that didn't change between previous and most recent results must be correct.
-  - What code did I change between then and now? Did I change anything?
-]
-???
-- Browsed my commit history on github and realized that I had done a little refactoring to clean up the workflow a bit. 
-- I also added the rule pick_best_cluster_seed to check off a TODO in the analysis plan. I forgot about that, thought that had been there much longer.
-- So I removed that rule and tried re-running the workflow with one parameter set and 2 seeds to check whether that was the problem.
-
----
-
-## Fixed it!
-.pull-left[
-```{r rulegraph_new, out.width='60%'}
-include_graphics(here('figures',
-                      'rulegraph_fit_split.png'))
-```
-]
-.pull-right[
-```{r}
-include_graphics(here('exploratory', 
-                      '2020-11_Nov-Dec', 
-                      'figures',
-                      'fraction-mapped-grid_fit-split-1.png'))
-```
-]
-
----
-
-## What happened here?
-
-.pull-left[
-- With `pick_best_cluster_seed`, the OptiClust results from the best seed were used for downstream OptiFit runs with _all seeds_.
-  - Same seed variable used to create the data splits.
-- When seeds mismatched, query sets were being fit to reference OTUs from different data splits.
-- Results were correct whenever the OptiFit seed corresponded to the best OptiClust seed, and incorrect in all other cases.
-]
-
----
-
-## Conclusions
-
-- Don't change more than one thing at a time, even if you think it's "just refactoring".
-- Start debugging from the easiest step to fix.
-- Have config files with different parameter sets to make debugging quick & easy.
-  - default: all datasets & parameters & 100 seeds.
-  - debug: one dataset & only one of each parameter choice & 2 seeds.
-- Make many small commits to make it easy to figure out where you introduced a bug.
-- Test your assumptions
-  - Use `assertthat` pkg in R.
+![](../../exploratory/2020-11_Nov-Dec/figures/mcc_fit-db-1.png)

### 2021-03-04 paper/paper.Rmd
Write description of algorithm

@@
 it is limited by the diversity represented in the reference database. Rare or
-novel sequences in the sample will be lost if they are not represented in the
+novel sequences in the sample will be lost if they are not represented by a similar sequence in the
 database. _De novo_ methods cluster sequences based on their distance to each
@@
 
-### OptiFit algorithm
+### The OptiFit algorithm
+
+- **TODO: ask Sarah Westcott to check the accuracy of this description**
 
@@
 sequences to OTUs to produce the highest quality OTUs possible, and extends this
-method for reference-based clustering.
-
-- **TODO: brief description of OptiFit algorithm**
-- **TODO: open & closed reference modes**
-- **TODO: MCC score; only considers query sequences (when printref=f)**
+method for reference-based clustering. OptiFit takes as input a list of
+reference OTUs and their sequences, a list of query sequences to assign to the
+reference OTUs, the sequence pairs that are within the distance threshold (e.g.
+0.03), and the metric to assess clustering quality (default: MCC). Query
+sequences are randomly seeded in reference OTUs, then for each sequence the
+algorithm calculates the quality metric based on whether the sequences stays in
+its current OTU or moves to each of the other OTUs. This process is repeated
+until the quality metric stabilizes, changing by no more than 0.0001 by default,
+or until a maximum number of iterations is reached (default: 100). In
+closed-reference mode, any query sequences that cannot be assigned are thrown
+out (**TODO: what determines whether a seq can't be assigned?**), and the
+results only contain OTUs that exist in the original reference. In
+open-reference mode, un-assigned query sequences are clustered _de novo_ using
+OptiClust to generate additional OTUs. The final quality score is reported with
+the best OTU assignments.
 
@@
 novo_ OTUs with OptiClust for comparison. All clustering was performed at a
-sequence dissimilarity threshold of 0.03 and OTU quality was evaluated using the
+sequence distance threshold of 0.03 and OTU quality was evaluated using the
 Matthews Correlation Coefficient (MCC) as described previously
@@
 
-- TODO: separate paragraphs for closed & open reference clustering. weave in vsearch within those paragraphs. mention median MCC values.
-- TODO: avoid slightly/much -- use numbers, e.g. X% better/worse.
-- TODO: put the comparison at the beginning of the sentence, so you don't have to know what it is at the end.
+- **TODO:** separate paragraphs for closed & open reference clustering. weave in vsearch within those paragraphs. mention median MCC values.
+- **TODO:** avoid slightly/much -- use numbers, e.g. X% better/worse.
+- **TODO:** put the comparison at the beginning of the sentence, so you don't have to know what it is at the end.
 
@@
 **TODO: specify exact numbers for fraction mapped.**
-OptiFit was able to fit more query sequences to reference OTUs created with the
+OptiFit was able to fit **X** more query sequences to reference OTUs created with the
 Greengenes and Silva databases than with RDP. In terms of run time,

### 2021-03-04 paper/paper.Rmd
Bold places where I need to insert exact numbers

@@
 mothur program which takes existing OTUs as the reference to fit new sequences
-to. Here, we tested the OptiFit algorithm with the reference as databases or _de
+to. **TODO: more words here?** 
+Here, we tested the OptiFit algorithm with the reference as databases or _de
 novo_ OTUs and compared the performance to existing tools.
@@
 ## Results
+
+**TODO: make a plan for the figures**
 
@@
 
-- **TODO:** separate paragraphs for closed & open reference clustering. weave in vsearch within those paragraphs. mention median MCC values.
-- **TODO:** avoid slightly/much -- use numbers, e.g. X% better/worse.
-- **TODO:** put the comparison at the beginning of the sentence, so you don't have to know what it is at the end.
-
 To evaluate reference-based clustering with independent databases, we fit each
@@
 database (v13_8_99), Silva non-redundant database (v132), and the Ribosomal
-Database Project (RDP; v16). In open reference mode, OTU quality was similar
-between fitting the datasets to reference OTUs with OptiFit and clustering the
-datasets _de novo_ with OptiClust. This held true for all datasets and reference
-databases. However, in closed reference mode, OTU quality was slightly worse
-when fitting to Greengenes and Silva, and much worse when fitting to RDP as
-compared to OptiClust. No more than half of query sequences were fit to
-reference OTUs in closed reference mode across any dataset/database combination.
-**TODO: specify exact numbers for fraction mapped.**
-OptiFit was able to fit **X** more query sequences to reference OTUs created with the
-Greengenes and Silva databases than with RDP. In terms of run time,
-closed reference OptiFit outperformed OptiClust, while OptiClust outperformed
-open reference OptiFit.
+Database Project (RDP; v16). As a comparison to existing software, vsearch was
+used to cluster OTUs _de novo_ or with reference-based clustering to the
+greengenes database. In open reference mode, OTU quality was similar between
+fitting the datasets to reference OTUs with OptiFit and clustering the datasets
+_de novo_ with OptiClust (**TODO: diff. in medians**). This held true for all
+datasets and reference databases. OptiFit produced higher quality OTUs than
+vsearch when open reference clustering against the greengenes database, with
+median MCC scores of **A** and **B%** (respectively).
+**TODO: open ref runtime of vsearch vs optifit vs opticlust**.
 
-To compare to existing software, vsearch was used to cluster OTUs _de novo_ or
-with reference-based clustering to the greengenes database. For all datasets and
-clustering methods (_de novo_, open reference, and closed reference), mothur's
-clustering algorithms produced higher quality OTUs than vsearch. When
-closed reference clustering against the greengenes database, vsearch was able to
-map more query sequences to the reference than mothur's OptiFit algorithm. In
-terms of runtime, OptiFit generally performed faster than vsearch when reference
-clustering, while vsearch _de novo_ clustering outperformed OptiClust.
+In closed reference mode, OptiFit produced lower quality OTUs than OptiClust by
+**X%** when fitting sequences to Greengenes and Silva, and **Y%** worse when
+fitting to RDP. Only up to **Z%** of query sequences were fit to reference OTUs
+in closed reference mode across any dataset/database combination. OptiFit was
+able to fit **X** more query sequences to reference OTUs created with the
+Greengenes and Silva databases than with RDP. vsearch was able to map more query
+sequences to the greengenes reference than OptiFit in closed reference mode. In
+terms of run time, closed reference OptiFit outperformed OptiClust by **X%**,
+while OptiClust outperformed open reference OptiFit by **Y%**. **TODO vsearch
+closed ref run time**.
+
+For all datasets and clustering methods (_de novo_, open reference, and closed
+reference), mothur's clustering algorithms produced higher quality OTUs than
+vsearch.  In terms of runtime, OptiFit generally performed faster than vsearch
+when reference clustering, while vsearch _de novo_ clustering outperformed
+OptiClust.
 
@@
 that from _de novo_ clustering the whole dataset with OptiClust regardless of
-mode. OTU quality was remarkably stable across 100 different random seeds. In
-terms of runtime, closed reference OptiFit performed faster than OptiClust on
-whole datasets. In open reference mode, OptiClust performed faster than OptiFit
-only when the OptiFit reference fraction was 30% or less. The split dataset
-strategy performed just as well as the database strategy in open reference mode
-regardless of database used, and outperformed the database strategy in
-closed reference mode.
+mode (**TODO: diff in MCC medians**). OTU quality was remarkably stable across
+100 different random seeds. In terms of runtime, closed reference OptiFit
+performed faster than OptiClust on whole datasets by **Z%**. In open reference
+mode, OptiClust performed **X to Y%** faster than OptiFit only when the OptiFit
+reference fraction was 30% or less. The split dataset strategy performed just as
+well as the database strategy in open reference mode regardless of database
+used, and outperformed the database strategy in closed reference mode by **W%**.
 
-We also tested three methods for selecting the sequences to be used as the
-reference; a simple random sample, weighting sequences by relative abundance,
-and weighting by similarity to other sequences in the dataset. OTU quality was
-similar with the simple and abundance-weighted sampling, but slightly worse with
-similarity-weighted sampling. In closed reference mode, The fraction of query
-sequences that can be fit to the reference OTUs decreases as the reference
-fraction increases.
+We also tested three methods for selecting the fraction of sequences to be used
+as the reference; a simple random sample, weighting sequences by relative
+abundance, and weighting by similarity to other sequences in the dataset. OTU
+quality was similar with the simple and abundance-weighted sampling (median MCCs
+**X** and **Y** respectively), but **Z%** worse with similarity-weighted
+sampling. In closed reference mode, the fraction of query sequences that can be
+fit to the reference OTUs decreases as the reference fraction increases from an
+MCC of **A** with **J** reference sequences to an MCC of **B** with **K**
+reference sequences.
 

### 2021-03-05 paper/paper.Rmd
moar wordz

@@
 sequences, it is discarded. While reference-based clustering is generally fast,
-it is limited by the diversity represented in the reference database. Rare or
-novel sequences in the sample will be lost if they are not represented by a similar sequence in the
-database. _De novo_ methods cluster sequences based on their distance to each
-other, without the use of an external reference. _De novo_ clustering overcomes
-the limitations of reference databases by considering only sequences in the
-dataset, but is more computationally intensive and generates different OTU
-assignments when new sequences are introduced. Unstable OTU assignments make it
-difficult to use _de novo_ clustering to compare taxonomic composition of
-communities between studies or to use machine learning models trained with _de
-novo_ OTUs to make predictions on new data. Open reference methods take a hybrid
-approach, first performing closed reference clustering, then any sequences that
-cannot be assigned to reference OTUs are clustered _de novo_ to create
-additional OTUs. Previous studies found that the OptiClust _de novo_ clustering
-algorithm created the highest quality OTU assignments of all clustering methods
-based on the Matthews correlation coefficient (MCC) [@schloss_application_2016;
-@westcott_novo_2015].
+it is limited by the diversity of the reference database. Rare or novel
+sequences in the sample will be lost if they are not represented by a similar
+sequence in the database. _De novo_ methods cluster sequences based on their
+distance to each other, without the use of an external reference. _De novo_
+clustering overcomes the limitations of reference databases by considering only
+sequences in the dataset, but is more computationally intensive and generates
+different OTU assignments when new sequences are introduced. Unstable OTU
+assignments make it difficult to use _de novo_ clustering to compare taxonomic
+composition of communities between studies or to use machine learning models
+trained with _de novo_ OTUs to make predictions on new data. Open reference
+methods take a hybrid approach, first performing closed reference clustering,
+then any sequences that cannot be assigned to reference OTUs are clustered _de
+novo_ to create additional OTUs. Previous studies found that the OptiClust _de
+novo_ clustering algorithm created the highest quality OTU assignments of all
+clustering methods based on the Matthews correlation coefficient (MCC)
+[@westcott_opticlust_2017]. As a result, we have recommended OptiClust as the
+preferred method for OTU clustering whenever OTU stability is not required.
 
@@
 to. **TODO: more words here?** 
-Here, we tested the OptiFit algorithm with the reference as databases or _de
+Here, we tested the OptiFit algorithm with the reference as a database or _de
 novo_ OTUs and compared the performance to existing tools.
@@
 ## Results
-
-**TODO: make a plan for the figures**
 
@@
 sequences are randomly seeded in reference OTUs, then for each sequence the
-algorithm calculates the quality metric based on whether the sequences stays in
+algorithm calculates the quality metric based on whether the sequence stays in
 its current OTU or moves to each of the other OTUs. This process is repeated
@@
 closed-reference mode, any query sequences that cannot be assigned are thrown
-out (**TODO: what determines whether a seq can't be assigned?**), and the
-results only contain OTUs that exist in the original reference. In
-open-reference mode, un-assigned query sequences are clustered _de novo_ using
+out (**TODO: exactly what determines whether a seq can't be assigned?**), and
+the results only contain OTUs that exist in the original reference. In
+open-reference mode, unassigned query sequences are clustered _de novo_ using
 OptiClust to generate additional OTUs. The final quality score is reported with
@@
 independent database, or 2) split the dataset into a reference and query
-fraction, then fit the query sequences to OTUs generated by clustering the
-reference sequences _de novo_. For each dataset repeated with 100 random seeds,
-we generated OTUs with OptiFit using both strategies, and also clustered _de
-novo_ OTUs with OptiClust for comparison. All clustering was performed at a
-sequence distance threshold of 0.03 and OTU quality was evaluated using the
-Matthews Correlation Coefficient (MCC) as described previously
-[@schloss_application_2016; @westcott_novo_2015]. We calculated the fraction of
-query sequences that were fit to existing OTUs in closed reference mode as an
-additional measure of quality for this mode.
+fraction, cluster the reference sequences _de novo_, then fit the query
+sequences to the reference OTUs. For each dataset repeated with 100 random
+seeds, we generated OTUs with OptiFit using both strategies. To compare to
+existing software, we also clustered OTUs _de novo_ using OptiClust and vsearch,
+and with vsearch in reference-based mode against the greengenes database. All
+clustering was performed at a distance threshold of 0.03 and OTU quality was
+evaluated using the Matthews Correlation Coefficient (MCC) as described
+previously [@westcott_opticlust_2017]. We calculated the fraction of query
+sequences that were fit to existing OTUs in closed reference mode as an
+additional measure of quality for this mode. _De novo_ clustering with OptiFit
+produced **X%** higher quality OTUs than vsearch, but performed **Y%** slower
+than vsearch.
 
@@
 dataset to reference OTUs generated by _de novo_ clustering the Greengenes
-database (v13_8_99), Silva non-redundant database (v132), and the Ribosomal
-Database Project (RDP; v16). As a comparison to existing software, vsearch was
-used to cluster OTUs _de novo_ or with reference-based clustering to the
-greengenes database. In open reference mode, OTU quality was similar between
-fitting the datasets to reference OTUs with OptiFit and clustering the datasets
-_de novo_ with OptiClust (**TODO: diff. in medians**). This held true for all
-datasets and reference databases. OptiFit produced higher quality OTUs than
-vsearch when open reference clustering against the greengenes database, with
-median MCC scores of **A** and **B%** (respectively).
-**TODO: open ref runtime of vsearch vs optifit vs opticlust**.
+database (v13_8_99), Silva non-redundant database (v132), or the Ribosomal
+Database Project (RDP; v16). **TODO: cite database papers?** In open reference
+mode, fitting the datasets to reference OTUs with OptiFit produced OTUs of
+similar quality as clustering the datasets _de novo_ with OptiClust (**TODO:
+diff. in medians**). This held true for all datasets and reference databases.
+OptiFit produced higher quality OTUs than vsearch when open reference clustering
+against the greengenes database, with median MCC scores of **A** and **B%**
+(respectively). However, open reference vsearch and OptiClust respectively ran
+**X%** and **Y%** faster than OptiFit in open reference mode.
 
@@
 able to fit **X** more query sequences to reference OTUs created with the
-Greengenes and Silva databases than with RDP. vsearch was able to map more query
-sequences to the greengenes reference than OptiFit in closed reference mode. In
-terms of run time, closed reference OptiFit outperformed OptiClust by **X%**,
-while OptiClust outperformed open reference OptiFit by **Y%**. **TODO vsearch
-closed ref run time**.
-
-For all datasets and clustering methods (_de novo_, open reference, and closed
-reference), mothur's clustering algorithms produced higher quality OTUs than
-vsearch.  In terms of runtime, OptiFit generally performed faster than vsearch
-when reference clustering, while vsearch _de novo_ clustering outperformed
-OptiClust.
+Greengenes and Silva databases than with RDP. vsearch was able to fit **W%**
+more query sequences to the greengenes reference than OptiFit in closed
+reference mode. In terms of run time, closed reference OptiFit outperformed
+OptiClust by **A%** and vsearch by **B%**.
 
@@
 
+A split dataset strategy was employed to assess how well OptiFit performs for
+tasks where new sequences are added to existing OTUs, such as when comparing OTUs
+across studies or making predictions with machine learning models.
 Datasets were randomly split into a reference fraction and a query fraction.
-Reference sizes from 10% to 80% of the sequences were created, with the
+Reference sizes from 10% to 90% of the sequences were created, with the
 remaining sequences used for the query. Reference sequences were clustered _de
@@
 mode (**TODO: diff in MCC medians**). OTU quality was remarkably stable across
-100 different random seeds. In terms of runtime, closed reference OptiFit
-performed faster than OptiClust on whole datasets by **Z%**. In open reference
-mode, OptiClust performed **X to Y%** faster than OptiFit only when the OptiFit
-reference fraction was 30% or less. The split dataset strategy performed just as
-well as the database strategy in open reference mode regardless of database
-used, and outperformed the database strategy in closed reference mode by **W%**.
+100 different random seeds (**TODO: variation or stdev?**). In terms of runtime,
+closed reference OptiFit performed faster than OptiClust on whole datasets by
+**Z%**. In open reference mode, OptiClust performed **X to Y%** faster than
+OptiFit only when the OptiFit reference fraction was 30% or less. The split
+dataset strategy performed just as well as the database strategy in open
+reference mode regardless of database used, and outperformed the database
+strategy in closed reference mode by **W%**.
 
@@
 ## Discussion
-
-- **TODO:** for these data, we don't see a compelling reason to use reference-based clustering over _de novo_.
-it was supposed to speed things up.
-the reason to do reference-based is if you like the ref OTUs -- e.g.
-for ML or downstream tools e.g.
-PiCrust?
-- **TODO:** highlight difference between what we're doing and what previously was done.
-  others use single ref seq to define an OTU, while we use all the ref & query seqs to define an otu. highlight why ours is better than previous methods.
 
@@
 databases, which are unlikely to contain sequences from rare and novel microbes.
-This drop in quality was most notable with RDP. We recommend users who require
-an independent reference database opt for Greengenes or Silva instead. Since
-OptiClust performs faster than open reference OptiFit and creates higher quality
-OTUs than closed reference OptiFit with the database strategy, we recommend
-using OptiClust rather than fitting to a database where possible.
-(**TODO: "if you don't have breadth, closed ref will suck." - Pat**)
+This drop in quality was most notable with RDP, which has only about 16,000 V4
+16S sequences compared to about 170,000 in Silva and greengenes. We recommend
+users who require an independent reference database opt for large databases with
+good coverage of microbial diversity. Since OptiClust performs faster than open
+reference OptiFit and creates higher quality OTUs than closed reference OptiFit
+with the database strategy, we recommend using OptiClust rather than fitting to
+a database whenever stable OTUs are not required for the study at hand.
 
-The mothur algorithms produced higher quality OTUs than vsearch in
-open reference, closed reference, or _de novo_ modes. However, vsearch was able
-to fit more sequence into OTUs than OptiFit in closed reference mode. While both
+The mothur algorithms produced higher quality OTUs than vsearch in open
+reference, closed reference, or _de novo_ modes. However, vsearch was able to
+fit more sequence into OTUs than OptiFit in closed reference mode. While both
 mothur and vsearch use a dissimilarity threshold for determining how to assign
-sequences into OTUs, vsearch is more permissive than mothur. Mothur requires
-that all pairs of sequences in an OTU are within the dissimilarity threshold
-without penalizing the MCC, while vsearch only requires sequences to be similar
-to one other sequence in the OTU. In this way, vsearch sacrifices OTU quality in
-order to allow more sequences to fit to OTUs. Users who require closed reference
-clustering may prefer to use vsearch if they wish to maximize the fraction of
-sequences that can be fit at the cost of OTU quality. However, mothur's
-OptiClust or OptiFit are preferred for _de novo_ or open reference clustering.
+sequences into OTUs, vsearch is more permissive than mothur. Mothur's OptiFit
+and OptiClust use all of the sequences to define an OTU, requiring that all
+pairs of sequences (including reference and query sequences) in an OTU are
+within the distance threshold without penalizing the MCC. In contrast,
+vsearch only requires each query sequence to be similar to the single sequence
+that seeded the OTU. In this way, vsearch sacrifices OTU quality in order to
+allow more sequences to fit to OTUs. Users who require closed reference
+clustering to the greengenes database may prefer to use vsearch if they wish to
+maximize the fraction of sequences that can be fit at the cost of OTU quality.
+However, mothur's OptiClust or OptiFit are recommended for _de novo_ or open
+reference clustering to produce OTUs of the highest possible quality.
 
@@
 such as when using already-trained machine learning models to make predictions
-on new data.
+on new data or comparing OTUs across studies. However, when stable OTUs are not
+required, we recommend using OptiClust for _de novo_ clustering over the split
+strategy with OptiFit since OptiClust simpler to execute but performs similarly.
 

### 2021-03-05 paper/paper.Rmd
minor polish

@@
 fit more sequence into OTUs than OptiFit in closed reference mode. While both
-mothur and vsearch use a dissimilarity threshold for determining how to assign
-sequences into OTUs, vsearch is more permissive than mothur. Mothur's OptiFit
-and OptiClust use all of the sequences to define an OTU, requiring that all
-pairs of sequences (including reference and query sequences) in an OTU are
-within the distance threshold without penalizing the MCC. In contrast,
-vsearch only requires each query sequence to be similar to the single sequence
-that seeded the OTU. In this way, vsearch sacrifices OTU quality in order to
-allow more sequences to fit to OTUs. Users who require closed reference
-clustering to the greengenes database may prefer to use vsearch if they wish to
-maximize the fraction of sequences that can be fit at the cost of OTU quality.
-However, mothur's OptiClust or OptiFit are recommended for _de novo_ or open
-reference clustering to produce OTUs of the highest possible quality.
+mothur and vsearch use a distance or similarity threshold for determining how to
+assign sequences to OTUs, vsearch is more permissive than mothur. Mothur's
+OptiFit and OptiClust use all of the sequences to define an OTU, requiring that
+all pairs of sequences (including reference and query sequences) in an OTU are
+within the distance threshold without penalizing the MCC. In contrast, vsearch
+only requires each query sequence to be similar to the single sequence that
+seeded the OTU. In this way, vsearch sacrifices OTU quality in order to allow
+more sequences to fit to OTUs. Users who require closed reference clustering to
+the greengenes database may prefer to use vsearch if they wish to maximize the
+fraction of sequences that can be fit at the cost of OTU quality. However,
+mothur's OptiClust or OptiFit are recommended for _de novo_ or open reference
+clustering to produce OTUs of the highest possible quality.
 
@@
 required, we recommend using OptiClust for _de novo_ clustering over the split
-strategy with OptiFit since OptiClust simpler to execute but performs similarly.
+strategy with OptiFit since OptiClust is simpler to execute but performs
+similarly in terms of both run time and OTU quality.
 

### 2021-03-05 paper/paper.Rmd
Describe data processing

@@
 dataset to reference OTUs generated by _de novo_ clustering the Greengenes
-database (v13_8_99), Silva non-redundant database (v132), or the Ribosomal
-Database Project (RDP; v16). **TODO: cite database papers?** In open reference
+database, Silva non-redundant database, or the Ribosomal
+Database Project (RDP). **TODO: cite database papers?** In open reference
 mode, fitting the datasets to reference OTUs with OptiFit produced OTUs of
@@
 
+We downloaded 16S rRNA gene amplicon sequences from four published datasets
+isolated from soil [@johnston_metagenomics_2016], marine
+[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
+gut [@baxter_microbiota-based_2016] samples.
+Raw sequences were processed using mothur according to the Schloss Lab MiSeq SOP
+as described in the mothur wiki and accompanying study by Kozich _et al._
+[@schloss_miseq_nodate; @kozich_development_2013].
+These steps included trimming and filtering for quality, aligning to the Silva
+reference alignment [**CITE**], discarding sequences that aligned outside the V4
+region, removing chimeric reads with UCHIME, and calculating distances between
+all pairs of sequences within each dataset prior to clustering.
+To generate reference OTUs from independent databases, we downloaded sequences
+from the Greengenes database (v13_8_99) [**CITE**], Silva non-redundant database
+(v132) [**CITE**], and the Ribosomal Database Project (RDP; v16) [**CITE**].
+These sequences were processed using the same steps outlined above followed by
+clustering sequences into _de novo_ OTUs with OptiClust.
+Processed reads from each of the four datasets were clustered with OptiFit to
+the reference OTUs generated form each of the three datasets.
+When reference clustering with vsearch, processed datasets were fit directly
+to the unprocessed greengenes reference alignment, since this is how vsearch
+is typically used by the QIIME2 software [@bolyen_reproducible_2019].
+
 ### Benchmarking
@@
 relied on Python [@van_rossum_python_2009], R [@r_core_team_r_2020], and GNU
-bash. Software used includes mothur v1.45.0 [@schloss_introducing_2009], vsearch
+bash.
+Software used includes mothur v1.45.0 [@schloss_introducing_2009], vsearch
 v2.13.3 [@rognes_vsearch_2016], numpy [@harris_array_2020], the Tidyverse
 metapackage [@wickham_welcome_2019], R Markdown [@xie_r_2018], and the conda
-environment manager [@noauthor_anaconda_2016]. The complete workflow,
-manuscript, and conda environment are available at **TODO: UPDATED REPO LINK**.
+environment manager [@noauthor_anaconda_2016].
+The complete workflow, manuscript, and conda environment are available at
+**TODO: UPDATED REPO LINK**.
 

### 2021-03-05 paper/paper.Rmd
Cite everybody

--- 2021-03-05 paper/paper.Rmd
@@
 
-- **TODO: summarize results & conclusion**
+**TODO: summarize results & conclusion**
 
@@
 
-- **TODO: ask Sarah Westcott to check the accuracy of this description**
+**TODO: ask Sarah Westcott to check the accuracy of this description**
 
@@
 existing software, we also clustered OTUs _de novo_ using OptiClust and vsearch,
-and with vsearch in reference-based mode against the greengenes database. All
+and with vsearch in reference-based mode against the Greengenes database. All
 clustering was performed at a distance threshold of 0.03 and OTU quality was
@@
 sequences that were fit to existing OTUs in closed reference mode as an
-additional measure of quality for this mode. _De novo_ clustering with OptiFit
+additional measure of quality for this mode. _De novo_ clustering with OptiClust
 produced **X%** higher quality OTUs than vsearch, but performed **Y%** slower
@@
 dataset to reference OTUs generated by _de novo_ clustering the Greengenes
-database (v13_8_99), Silva non-redundant database (v132), or the Ribosomal
-Database Project (RDP; v16). **TODO: cite database papers?** In open reference
-mode, fitting the datasets to reference OTUs with OptiFit produced OTUs of
-similar quality as clustering the datasets _de novo_ with OptiClust (**TODO:
-diff. in medians**). This held true for all datasets and reference databases.
+database, SILVA non-redundant database, or the Ribosomal Database Project (RDP)
+[@desantis_greengenes_2006; @quast_silva_2013; @cole_ribosomal_2014].
+
+In open reference mode, fitting the datasets to reference OTUs with OptiFit
+produced OTUs of similar quality as clustering the datasets _de novo_ with
+OptiClust (**diff. in medians**).
+This held true for all datasets and reference databases.
 OptiFit produced higher quality OTUs than vsearch when open reference clustering
-against the greengenes database, with median MCC scores of **A** and **B%**
-(respectively). However, open reference vsearch and OptiClust respectively ran
-**X%** and **Y%** faster than OptiFit in open reference mode.
+against the Greengenes database, with median MCC scores of **A** and **B%**
+(respectively).
+However, open reference vsearch and OptiClust respectively ran **X%** and **Y%**
+faster than OptiFit in open reference mode.
 
 In closed reference mode, OptiFit produced lower quality OTUs than OptiClust by
-**X%** when fitting sequences to Greengenes and Silva, and **Y%** worse when
+**X%** when fitting sequences to Greengenes and SILVA, and **Y%** worse when
 fitting to RDP. Only up to **Z%** of query sequences were fit to reference OTUs
@@
 able to fit **X** more query sequences to reference OTUs created with the
-Greengenes and Silva databases than with RDP. vsearch was able to fit **W%**
-more query sequences to the greengenes reference than OptiFit in closed
+Greengenes and SILVA databases than with RDP. vsearch was able to fit **W%**
+more query sequences to the Greengenes reference than OptiFit in closed
 reference mode. In terms of run time, closed reference OptiFit outperformed
@@
 Datasets were randomly split into a reference fraction and a query fraction.
-Reference sizes from 10% to 90% of the sequences were created, with the
+Reference sizes from 10% to 80% of the sequences were created, with the
 remaining sequences used for the query. Reference sequences were clustered _de
@@
 This drop in quality was most notable with RDP, which has only about 16,000 V4
-16S sequences compared to about 170,000 in Silva and greengenes. We recommend
+16S sequences compared to about 170,000 in SILVA and Greengenes. We recommend
 users who require an independent reference database opt for large databases with
@@
 more sequences to fit to OTUs. Users who require closed reference clustering to
-the greengenes database may prefer to use vsearch if they wish to maximize the
+the Greengenes database may prefer to use vsearch if they wish to maximize the
 fraction of sequences that can be fit at the cost of OTU quality. However,
@@
 
+We downloaded 16S rRNA gene amplicon sequences from four published datasets
+isolated from soil [@johnston_metagenomics_2016], marine
+[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
+gut [@baxter_microbiota-based_2016] samples.
+Raw sequences were processed using mothur according to the Schloss Lab MiSeq SOP
+as described in the mothur wiki and accompanying study by Kozich _et al._
+[@schloss_miseq_nodate; @kozich_development_2013].
+These steps included trimming and filtering for quality, aligning to the SILVA
+reference alignment [@quast_silva_2013], discarding sequences that aligned
+outside the V4 region, removing chimeric reads with UCHIME [@edgar_uchime_2011],
+and calculating distances between all pairs of sequences within each dataset
+prior to clustering.
+
+### Reference database clustering
+
+To generate reference OTUs from independent databases, we downloaded sequences
+from the Greengenes database (v13_8_99) [@desantis_greengenes_2006], SILVA
+non-redundant database (v132) [@quast_silva_2013], and the Ribosomal Database
+Project (v16) [@cole_ribosomal_2014].
+These sequences were processed using the same steps outlined above followed by
+clustering sequences into _de novo_ OTUs with OptiClust.
+Processed reads from each of the four datasets were clustered with OptiFit to
+the reference OTUs generated from each of the three databases.
+When reference clustering with vsearch, processed datasets were fit directly to
+the unprocessed Greengenes reference alignment, since this method is how vsearch
+is typically used by the QIIME2 software reference-based clustering
+[@bolyen_reproducible_2019; @noauthor_clustering_nodate].
+
+### Split dataset clustering
+
+**TODO: describe split dataset creation**
+
 ### Benchmarking
+
+**TODO**
 
@@
 We implemented the analysis workflow in Snakemake [@koster_snakemake_2012] and
-relied on Python [@van_rossum_python_2009], R [@r_core_team_r_2020], and GNU
-bash. Software used includes mothur v1.45.0 [@schloss_introducing_2009], vsearch
+wrote scripts in R [@r_core_team_r_2020], Python [@van_rossum_python_2009], and
+GNU bash [@noauthor_bash_nodate].
+Software used includes mothur v1.45.0 [@schloss_introducing_2009], vsearch
 v2.13.3 [@rognes_vsearch_2016], numpy [@harris_array_2020], the Tidyverse
-metapackage [@wickham_welcome_2019], R Markdown [@xie_r_2018], and the conda
-environment manager [@noauthor_anaconda_2016]. The complete workflow,
-manuscript, and conda environment are available at **TODO: UPDATED REPO LINK**.
+metapackage [@wickham_welcome_2019], R Markdown [@xie_r_2018], the SRA toolkit
+[@noauthor_sra-tools_nodate], and the conda environment manager
+[@noauthor_anaconda_2016].
+The complete workflow, manuscript, and conda environment are available at
+**TODO: UPDATED REPO LINK**.
 

### 2021-03-05 paper/paper.Rmd
Start new sentences on new lines

@@
 important step in characterizing the composition of microbial communities across
-large datasets. OptiClust, a _de novo_ OTU clustering method in the mothur
-program, has been shown to produce higher quality OTU assignments than other
-methods and at comparable or faster speeds [@westcott_opticlust_2017;
-@schloss_introducing_2009]. A notable difference between _de novo_ clustering
-and database-dependent methods is that OTU assignments clustered with _de novo_
-methods are not stable when new sequences are added to a dataset
-[@westcott_novo_2015]. However, in some cases one may wish to incorporate new
-samples into a previously clustered dataset without performing clustering again
-on all sequences, such as when deploying a machine learning model where OTUs are
-features [@topcuoglu_effective_2019]. To provide an efficient and robust method
-to fit amplicon sequence data to existing OTUs, we developed the OptiFit
-algorithm as a new component of the mothur program.
+large datasets.
+OptiClust, a _de novo_ OTU clustering method in the mothur program, has been
+shown to produce higher quality OTU assignments than other methods and at
+comparable or faster speeds [@westcott_opticlust_2017;
+@schloss_introducing_2009].
+A notable difference between _de novo_ clustering and database-dependent methods
+is that OTU assignments clustered with _de novo_ methods are not stable when new
+sequences are added to a dataset [@westcott_novo_2015].
+However, in some cases one may wish to incorporate new samples into a previously
+clustered dataset without performing clustering again on all sequences, such as
+when deploying a machine learning model where OTUs are features
+[@topcuoglu_effective_2019].
+To provide an efficient and robust method to fit amplicon sequence data to
+existing OTUs, we developed the OptiFit algorithm as a new component of the
+mothur program.
 
@@
 Amplicon sequencing has become a mainstay of microbial ecology and
-host-associated microbiome research. Researchers can affordably generate
-millions of sequences to characterize the composition of hundreds of samples
-from culture-independent microbial communities. In a typical analysis pipeline,
+host-associated microbiome research.
+Researchers can affordably generate millions of sequences to characterize the
+composition of hundreds of samples from culture-independent microbial
+communities. In a typical analysis pipeline,
 16S rRNA gene sequences are assigned to Operational Taxonomic Units (OTUs) to
-facilitate comparison of taxonomic composition between communities. A distance
-threshold of 3% (or sequence similarity of 97%) is commonly used to cluster
-sequences into OTUs based on either a reference database or pairwise comparisons
-of the sequences within the dataset. The method chosen for clustering affects
-the quality of OTU assignments and thus may impact downstream analyses of
-community composition [@westcott_opticlust_2017; @schloss_application_2016;
+facilitate comparison of taxonomic composition between communities.
+A distance threshold of 3% (or sequence similarity of 97%) is commonly used to
+cluster sequences into OTUs based on either a reference database or pairwise
+comparisons of the sequences within the dataset.
+The method chosen for clustering affects the quality of OTU assignments and thus
+may impact downstream analyses of community composition
+[@westcott_opticlust_2017; @schloss_application_2016;
 @westcott_novo_2015].
@@
 There are three main categories of OTU clustering algorithms: closed reference,
-open reference, and _de novo_ clustering. Closed reference methods assign
-sequences to a set of pre-made OTUs generated from reference sequences. If a
-query sequence is not within the distance threshold to any of the reference
-sequences, it is discarded. While reference-based clustering is generally fast,
-it is limited by the diversity of the reference database. Rare or novel
-sequences in the sample will be lost if they are not represented by a similar
-sequence in the database. _De novo_ methods cluster sequences based on their
-distance to each other, without the use of an external reference. _De novo_
-clustering overcomes the limitations of reference databases by considering only
-sequences in the dataset, but is more computationally intensive and generates
-different OTU assignments when new sequences are introduced. Unstable OTU
-assignments make it difficult to use _de novo_ clustering to compare taxonomic
-composition of communities between studies or to use machine learning models
-trained with _de novo_ OTUs to make predictions on new data. Open reference
-methods take a hybrid approach, first performing closed reference clustering,
-then any sequences that cannot be assigned to reference OTUs are clustered _de
-novo_ to create additional OTUs. Previous studies found that the OptiClust _de
-novo_ clustering algorithm created the highest quality OTU assignments of all
-clustering methods based on the Matthews correlation coefficient (MCC)
-[@westcott_opticlust_2017]. As a result, we have recommended OptiClust as the
-preferred method for OTU clustering whenever OTU stability is not required.
+open reference, and _de novo_ clustering. 
+Closed reference methods assign sequences to a set of pre-made OTUs generated
+from reference sequences.
+If a query sequence is not within the distance threshold to any of the reference
+sequences, it is discarded.
+While reference-based clustering is generally fast, it is limited by the
+diversity of the reference database.
+Rare or novel sequences in the sample will be lost if they are not represented
+by a similar sequence in the database.
+_De novo_ methods cluster sequences based on their distance to each other,
+without the use of an external reference.
+_De novo_ clustering overcomes the limitations of reference databases by
+considering only sequences in the dataset, but is more computationally intensive
+and generates different OTU assignments when new sequences are introduced.
+Unstable OTU assignments make it difficult to use _de novo_ clustering to
+compare taxonomic composition of communities between studies or to use machine
+learning models trained with _de novo_ OTUs to make predictions on new data.
+Open reference methods take a hybrid approach, first performing closed reference
+clustering, then any sequences that cannot be assigned to reference OTUs are
+clustered _de novo_ to create additional OTUs.
+Previous studies found that the OptiClust _de novo_ clustering algorithm created
+the highest quality OTU assignments of all clustering methods based on the
+Matthews correlation coefficient (MCC) [@westcott_opticlust_2017].
+As a result, we have recommended OptiClust as the preferred method for OTU
+clustering whenever OTU stability is not required.
 
@@
 mothur program which takes existing OTUs as the reference to fit new sequences
-to. **TODO: more words here?** 
+to. 
+**TODO: more words here?** 
 Here, we tested the OptiFit algorithm with the reference as a database or _de
@@
 sequences to OTUs to produce the highest quality OTUs possible, and extends this
-method for reference-based clustering. OptiFit takes as input a list of
-reference OTUs and their sequences, a list of query sequences to assign to the
-reference OTUs, the sequence pairs that are within the distance threshold (e.g.
-0.03), and the metric to assess clustering quality (default: MCC). Query
-sequences are randomly seeded in reference OTUs, then for each sequence the
-algorithm calculates the quality metric based on whether the sequence stays in
-its current OTU or moves to each of the other OTUs. This process is repeated
-until the quality metric stabilizes, changing by no more than 0.0001 by default,
-or until a maximum number of iterations is reached (default: 100). In
-closed-reference mode, any query sequences that cannot be assigned are thrown
+method for reference-based clustering.
+OptiFit takes as input a list of reference OTUs and their sequences, a list of
+query sequences to assign to the reference OTUs, the sequence pairs that are
+within the distance threshold (e.g. 0.03), and the metric to assess clustering
+quality (default: MCC).
+Query sequences are randomly seeded in reference OTUs, then for each sequence
+the algorithm calculates the quality metric based on whether the sequence stays
+in its current OTU or moves to each of the other OTUs.
+This process is repeated until the quality metric stabilizes, changing by no
+more than 0.0001 by default, or until a maximum number of iterations is reached
+(default: 100).
+In closed-reference mode, any query sequences that cannot be assigned are thrown
 out (**TODO: exactly what determines whether a seq can't be assigned?**), and
-the results only contain OTUs that exist in the original reference. In
-open-reference mode, unassigned query sequences are clustered _de novo_ using
+the results only contain OTUs that exist in the original reference.
+In open-reference mode, unassigned query sequences are clustered _de novo_ using
 OptiClust to generate additional OTUs. The final quality score is reported with
@@
 [@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
-gut [@baxter_microbiota-based_2016] samples. There are two strategies for
-generating OTUs with OptiFit: 1) fit sequences to reference OTUs of an
-independent database, or 2) split the dataset into a reference and query
-fraction, cluster the reference sequences _de novo_, then fit the query
-sequences to the reference OTUs. For each dataset repeated with 100 random
-seeds, we generated OTUs with OptiFit using both strategies. To compare to
-existing software, we also clustered OTUs _de novo_ using OptiClust and vsearch,
-and with vsearch in reference-based mode against the Greengenes database. All
-clustering was performed at a distance threshold of 0.03 and OTU quality was
+gut [@baxter_microbiota-based_2016] samples. 
+There are two strategies for generating OTUs with OptiFit: 1) fit sequences to
+reference OTUs of an independent database, or 2) split the dataset into a
+reference and query fraction, cluster the reference sequences _de novo_, then
+fit the query sequences to the reference OTUs.
+For each dataset repeated with 100 random seeds, we generated OTUs with OptiFit
+using both strategies.
+To compare to existing software, we also clustered OTUs _de novo_ using
+OptiClust and vsearch, and with vsearch in reference-based mode against the
+Greengenes database.
+All clustering was performed at a distance threshold of 0.03 and OTU quality was
 evaluated using the Matthews Correlation Coefficient (MCC) as described
-previously [@westcott_opticlust_2017]. We calculated the fraction of query
-sequences that were fit to existing OTUs in closed reference mode as an
-additional measure of quality for this mode. _De novo_ clustering with OptiClust
-produced **X%** higher quality OTUs than vsearch, but performed **Y%** slower
-than vsearch.
+previously [@westcott_opticlust_2017].
+We calculated the fraction of query sequences that were fit to existing OTUs in
+closed reference mode as an additional measure of quality for this mode.
+_De novo_ clustering with OptiClust produced **X%** higher quality OTUs than
+vsearch, but performed **Y%** slower than vsearch.
 
@@
 **X%** when fitting sequences to Greengenes and SILVA, and **Y%** worse when
-fitting to RDP. Only up to **Z%** of query sequences were fit to reference OTUs
-in closed reference mode across any dataset/database combination. OptiFit was
-able to fit **X** more query sequences to reference OTUs created with the
-Greengenes and SILVA databases than with RDP. vsearch was able to fit **W%**
-more query sequences to the Greengenes reference than OptiFit in closed
-reference mode. In terms of run time, closed reference OptiFit outperformed
+fitting to RDP.
+Only up to **Z%** of query sequences were fit to reference OTUs in closed
+reference mode across any dataset/database combination.
+OptiFit was able to fit **X** more query sequences to reference OTUs created
+with the Greengenes and SILVA databases than with RDP.
+Vsearch was able to fit **W%** more query sequences to the Greengenes reference
+than OptiFit in closed reference mode.
+In terms of run time, closed reference OptiFit outperformed
 OptiClust by **A%** and vsearch by **B%**.
@@
 A split dataset strategy was employed to assess how well OptiFit performs for
-tasks where new sequences are added to existing OTUs, such as when comparing OTUs
-across studies or making predictions with machine learning models.
+tasks where new sequences are added to existing OTUs, such as when comparing
+OTUs across studies or making predictions with machine learning models.
 Datasets were randomly split into a reference fraction and a query fraction.
 Reference sizes from 10% to 80% of the sequences were created, with the
-remaining sequences used for the query. Reference sequences were clustered _de
-novo_ with OptiClust, then query sequences were fit to the _de novo_ OTUs with
-OptiFit.
+remaining sequences used for the query.
+Reference sequences were clustered _de novo_ with OptiClust, then query
+sequences were fit to the _de novo_ OTUs with OptiFit.
 
@@
 that from _de novo_ clustering the whole dataset with OptiClust regardless of
-mode (**TODO: diff in MCC medians**). OTU quality was remarkably stable across
-100 different random seeds (**TODO: variation or stdev?**). In terms of runtime,
-closed reference OptiFit performed faster than OptiClust on whole datasets by
-**Z%**. In open reference mode, OptiClust performed **X to Y%** faster than
-OptiFit only when the OptiFit reference fraction was 30% or less. The split
-dataset strategy performed just as well as the database strategy in open
-reference mode regardless of database used, and outperformed the database
+mode (**TODO: diff in MCC medians**).
+OTU quality was remarkably stable across 100 different random seeds (**TODO:
+variation or stdev?**).
+In terms of runtime, closed reference OptiFit performed faster than OptiClust on
+whole datasets by **Z%**.
+In open reference mode, OptiClust performed **X to Y%** faster than OptiFit only
+when the OptiFit reference fraction was 30% or less.
+The split dataset strategy performed just as well as the database strategy in
+open reference mode regardless of database used, and outperformed the database
 strategy in closed reference mode by **W%**.
@@
 as the reference; a simple random sample, weighting sequences by relative
-abundance, and weighting by similarity to other sequences in the dataset. OTU
-quality was similar with the simple and abundance-weighted sampling (median MCCs
-**X** and **Y** respectively), but **Z%** worse with similarity-weighted
-sampling. In closed reference mode, the fraction of query sequences that can be
-fit to the reference OTUs decreases as the reference fraction increases from an
-MCC of **A** with **J** reference sequences to an MCC of **B** with **K**
-reference sequences.
+abundance, and weighting by similarity to other sequences in the dataset.
+OTU quality was similar with the simple and abundance-weighted sampling (median
+MCCs **X** and **Y** respectively), but **Z%** worse with similarity-weighted
+sampling.
+In closed reference mode, the fraction of query sequences that can be fit to the
+reference OTUs decreases as the reference fraction increases from an MCC of
+**A** with **J** reference sequences to an MCC of **B** with **K** reference
+sequences.
 
@@
 We developed a new algorithm for fitting sequences to existing OTUs and have
-demonstrated its suitability for reference-based clustering. OptiFit makes the
-iterative method employed by OptiClust available for tasks where
-reference-based clustering is required. We have shown that OTU quality is
-similar between OptiClust and OptiFit in open reference mode, regardless of
-strategy employed. open reference OptiFit does perform slower than OptiClust
-due to the additional _de novo_ clustering step, so users may prefer OptiClust
-for tasks that do not require reference OTUs.
+demonstrated its suitability for reference-based clustering.
+OptiFit makes the iterative method employed by OptiClust available for tasks
+where reference-based clustering is required.
+We have shown that OTU quality is similar between OptiClust and OptiFit in open
+reference mode, regardless of strategy employed.
+Open reference OptiFit performs slower than OptiClust due to the additional _de
+novo_ clustering step, so users may prefer OptiClust for tasks that do not
+require reference OTUs.
 
@@
 than half of query sequences were able to be fit to OTUs across any
-dataset/database combination. This may reflect limitations of reference
+dataset/database combination. 
+This may reflect limitations of reference
 databases, which are unlikely to contain sequences from rare and novel microbes.
 This drop in quality was most notable with RDP, which has only about 16,000 V4
-16S sequences compared to about 170,000 in SILVA and Greengenes. We recommend
-users who require an independent reference database opt for large databases with
-good coverage of microbial diversity. Since OptiClust performs faster than open
-reference OptiFit and creates higher quality OTUs than closed reference OptiFit
-with the database strategy, we recommend using OptiClust rather than fitting to
-a database whenever stable OTUs are not required for the study at hand.
+16S sequences compared to about 170,000 in SILVA and Greengenes (**TODO: insert
+R code to generate these numbers**).
+We recommend that users who require an independent reference database opt for
+large databases with good coverage of microbial diversity.
+Since OptiClust performs faster than open reference OptiFit and creates higher
+quality OTUs than closed reference OptiFit with the database strategy, we
+recommend using OptiClust rather than fitting to a database whenever stable OTUs
+are not required for the study at hand.
 
-The mothur algorithms produced higher quality OTUs than vsearch in open
-reference, closed reference, or _de novo_ modes. However, vsearch was able to
-fit more sequence into OTUs than OptiFit in closed reference mode. While both
-mothur and vsearch use a distance or similarity threshold for determining how to
-assign sequences to OTUs, vsearch is more permissive than mothur. Mothur's
-OptiFit and OptiClust use all of the sequences to define an OTU, requiring that
-all pairs of sequences (including reference and query sequences) in an OTU are
-within the distance threshold without penalizing the MCC. In contrast, vsearch
-only requires each query sequence to be similar to the single sequence that
-seeded the OTU. In this way, vsearch sacrifices OTU quality in order to allow
-more sequences to fit to OTUs. Users who require closed reference clustering to
-the Greengenes database may prefer to use vsearch if they wish to maximize the
-fraction of sequences that can be fit at the cost of OTU quality. However,
-mothur's OptiClust or OptiFit are recommended for _de novo_ or open reference
-clustering to produce OTUs of the highest possible quality.
+The OptiClust and OptiFit algorithms provided by mothur produced higher quality
+OTUs than vsearch in open reference, closed reference, or _de novo_ modes.
+However, vsearch was able to fit more sequence into OTUs than OptiFit in closed
+reference mode.
+While both mothur and vsearch use a distance or similarity threshold for
+determining how to assign sequences to OTUs, vsearch is more permissive than
+mothur.
+Mothur's OptiFit and OptiClust algorithms use all of the sequences to define an
+OTU, requiring that all pairs of sequences (including reference and query
+sequences) in an OTU are within the distance threshold without penalizing the
+MCC.
+In contrast, vsearch only requires each query sequence to be similar to the
+single sequence that seeded the OTU.
+In this way, vsearch sacrifices OTU quality in order to allow more sequences to
+fit to OTUs.
+Users who require closed reference clustering to the Greengenes database may
+prefer to use vsearch if they wish to maximize the fraction of sequences that
+can be fit at the cost of OTU quality.
+However, mothur's OptiClust or OptiFit are recommended for _de novo_ or open
+reference clustering to produce OTUs of the highest possible quality.
 
@@
 abundance, but quality was slightly worse when sequences were weighted by
-similarity. We recommend using a simple random sample since the more
-sophisticated reference selection methods do not offer any benefit. The
-similarity in OTU quality between OptiClust and OptiFit with this strategy
+similarity. 
+We recommend using a simple random sample since the more
+sophisticated reference selection methods do not offer any benefit. 
+The similarity in OTU quality between OptiClust and OptiFit with this strategy
 demonstrates the suitability of using OptiFit to fit sequences to existing OTUs,
 such as when using already-trained machine learning models to make predictions
-on new data or comparing OTUs across studies. However, when stable OTUs are not
-required, we recommend using OptiClust for _de novo_ clustering over the split
-strategy with OptiFit since OptiClust is simpler to execute but performs
-similarly in terms of both run time and OTU quality.
+on new data or comparing OTUs across studies.
+However, when stable OTUs are not required, we recommend using OptiClust for _de
+novo_ clustering over the split strategy with OptiFit since OptiClust is simpler
+to execute but performs similarly in terms of both run time and OTU quality.
 
-- **TODO: big picture concluding paragraph**
+**TODO: big picture concluding paragraph.**
+We have developed a new reference-based clustering algorithm that allows users
+to produce high quality OTUs when stability is a requirement for the task at hand.
 

### 2021-03-05 paper/paper.Rmd
Describe benchmarking methods

@@
 in its current OTU or moves to each of the other OTUs.
+If two or more OTU assignments are of equal quality, a random number generator
+is used to break the tie.
 This process is repeated until the quality metric stabilizes, changing by no
@@
 To compare to existing software, we also clustered OTUs _de novo_ using
-OptiClust and vsearch, and with vsearch in reference-based mode against the
+OptiClust and VSEARCH, and with VSEARCH in reference-based mode against the
 Greengenes database.
 All clustering was performed at a distance threshold of 0.03 and OTU quality was
-evaluated using the Matthews Correlation Coefficient (MCC) as described
-previously [@westcott_opticlust_2017].
+evaluated using the MCC as described previously [@westcott_opticlust_2017].
 We calculated the fraction of query sequences that were fit to existing OTUs in
@@
 _De novo_ clustering with OptiClust produced **X%** higher quality OTUs than
-vsearch, but performed **Y%** slower than vsearch.
+VSEARCH, but performed **Y%** slower than VSEARCH.
 
@@
 This held true for all datasets and reference databases.
-OptiFit produced higher quality OTUs than vsearch when open reference clustering
+OptiFit produced higher quality OTUs than VSEARCH when open reference clustering
 against the Greengenes database, with median MCC scores of **A** and **B%**
 (respectively).
-However, open reference vsearch and OptiClust respectively ran **X%** and **Y%**
+However, open reference VSEARCH and OptiClust respectively ran **X%** and **Y%**
 faster than OptiFit in open reference mode.
@@
 with the Greengenes and SILVA databases than with RDP.
-Vsearch was able to fit **W%** more query sequences to the Greengenes reference
+VSEARCH was able to fit **W%** more query sequences to the Greengenes reference
 than OptiFit in closed reference mode.
 In terms of run time, closed reference OptiFit outperformed
-OptiClust by **A%** and vsearch by **B%**.
+OptiClust by **A%** and VSEARCH by **B%**.
 
@@
 The OptiClust and OptiFit algorithms provided by mothur produced higher quality
-OTUs than vsearch in open reference, closed reference, or _de novo_ modes.
-However, vsearch was able to fit more sequence into OTUs than OptiFit in closed
+OTUs than VSEARCH in open reference, closed reference, or _de novo_ modes.
+However, VSEARCH was able to fit more sequence into OTUs than OptiFit in closed
 reference mode.
-While both mothur and vsearch use a distance or similarity threshold for
-determining how to assign sequences to OTUs, vsearch is more permissive than
+While both mothur and VSEARCH use a distance or similarity threshold for
+determining how to assign sequences to OTUs, VSEARCH is more permissive than
 mothur.
@@
 MCC.
-In contrast, vsearch only requires each query sequence to be similar to the
+In contrast, VSEARCH only requires each query sequence to be similar to the
 single sequence that seeded the OTU.
-In this way, vsearch sacrifices OTU quality in order to allow more sequences to
+In this way, VSEARCH sacrifices OTU quality in order to allow more sequences to
 fit to OTUs.
 Users who require closed reference clustering to the Greengenes database may
-prefer to use vsearch if they wish to maximize the fraction of sequences that
+prefer to use VSEARCH if they wish to maximize the fraction of sequences that
 can be fit at the cost of OTU quality.
@@
 **TODO: big picture concluding paragraph.**
-We have developed a new reference-based clustering algorithm that allows users
-to produce high quality OTUs when stability is a requirement for the task at hand.
+We have developed a new clustering algorithm that allows users to produce high
+quality OTUs using already existing OTUs as a reference.
 
@@
 the reference OTUs generated from each of the three databases.
-When reference clustering with vsearch, processed datasets were fit directly to
-the unprocessed Greengenes reference alignment, since this method is how vsearch
+When reference clustering with VSEARCH, processed datasets were fit directly to
+the unprocessed Greengenes reference alignment, since this method is how VSEARCH
 is typically used by the QIIME2 software reference-based clustering
@@
 
-**TODO: describe split dataset creation**
+For each dataset, a fraction of the sequences was selected to be clustered _de
+novo_ into reference OTUs with OptiClust.
+We used three methods for selecting the fraction of sequences to be used
+as the reference; a simple random sample, weighting sequences by relative
+abundance, and weighting by similarity to other sequences in the dataset.
+Dataset splitting was repeated with reference fractions ranging from 10% to 80%
+of the dataset and for 100 random seeds.
+For each dataset split, the remaining sequences were assigned to the reference
+OTUs with OptiFit.
 
@@
 
-**TODO**
+Since OptiClust and OptiFit employ a random number generator to break ties
+when OTU assignments are of equal quality, they produce slightly different OTU
+assignments when repeated with different random seeds.
+To capture any variation in OTU quality or execution time, clustering was
+repeated with 100 random seeds for each combination of parameters and
+input datasets.
+We used the benchmark feature provided by Snakemake to measure the run time of
+every clustering job.
+We calculated the MCC on each set of OTUs to quantify the quality of clustering,
+as described by Westcott _et al._ [@westcott_opticlust_2017].
 
@@
 GNU bash [@noauthor_bash_nodate].
-Software used includes mothur v1.45.0 [@schloss_introducing_2009], vsearch
+Software used includes mothur v1.45.0 [@schloss_introducing_2009], VSEARCH
 v2.13.3 [@rognes_vsearch_2016], numpy [@harris_array_2020], the Tidyverse

### 2021-03-05 code/R/summarize_results.R
Write rule to summarize results

@@ -1,0 +1,39 @@
+library(tidyverse)
+library(glue)
+
+mutate_perf <- function(dat) {
+  dat %>% 
+    mutate(mem_mb = max_rss,
+           mem_gb = mem_mb / 1024) %>% 
+    rename(sec = s,
+           num_otus = sobs)
+}
+
+opticlust <- read_tsv(snakemake@input[['opticlust']]) %>% 
+  mutate_perf() %>% 
+  mutate(strategy = method)
+optifit_db <- read_tsv(snakemake@input[['optifit_db']])%>% 
+  mutate_perf() %>% 
+  mutate(strategy = glue('database_{ref}'))
+optifit_split <- read_tsv(snakemake@input[['optifit_split']])%>% 
+  mutate_perf() %>% 
+  mutate(strategy = 'self-split')
+vsearch <- read_tsv(snakemake@input[['vsearch']])%>% 
+  mutate_perf() %>% 
+  mutate(strategy = case_when(
+    method == 'de_novo' ~ method,
+    TRUE ~ as.character(glue('database_{ref}'))))
+
+results_agg <- list(opticlust, optifit_db, optifit_split, vsearch) %>% 
+  reduce(full_join)
+
+results_sum <- results_agg %>% 
+  group_by(tool, strategy, method, dataset, database, ref_frac, ref_weight) %>% 
+  summarize(n = n(),
+            mcc_median = median(mcc),  # TODO: tidy way to avoid this repetitiveness?
+            sec_median = median(sec),
+            mem_gb_median = median(mem_gb),
+            frac_map_median = median(fraction_mapped))
+
+write_tsv(results_agg, snakemake@output[['agg']])
+write_tsv(results_sum, snakemake@output[['sum']])

### 2021-03-05 code/R/summarize_results.R
🎨 Style R code ⬡

@@ -2,38 +2,45 @@ library(tidyverse)
 library(glue)
 
 mutate_perf <- function(dat) {
-  dat %>% 
-    mutate(mem_mb = max_rss,
-           mem_gb = mem_mb / 1024) %>% 
-    rename(sec = s,
-           num_otus = sobs)
+  dat %>%
+    mutate(
+      mem_mb = max_rss,
+      mem_gb = mem_mb / 1024
+    ) %>%
+    rename(
+      sec = s,
+      num_otus = sobs
+    )
 }
 
-opticlust <- read_tsv(snakemake@input[['opticlust']]) %>% 
-  mutate_perf() %>% 
+opticlust <- read_tsv(snakemake@input[["opticlust"]]) %>%
+  mutate_perf() %>%
   mutate(strategy = method)
-optifit_db <- read_tsv(snakemake@input[['optifit_db']])%>% 
-  mutate_perf() %>% 
-  mutate(strategy = glue('database_{ref}'))
-optifit_split <- read_tsv(snakemake@input[['optifit_split']])%>% 
-  mutate_perf() %>% 
-  mutate(strategy = 'self-split')
-vsearch <- read_tsv(snakemake@input[['vsearch']])%>% 
-  mutate_perf() %>% 
+optifit_db <- read_tsv(snakemake@input[["optifit_db"]]) %>%
+  mutate_perf() %>%
+  mutate(strategy = glue("database_{ref}"))
+optifit_split <- read_tsv(snakemake@input[["optifit_split"]]) %>%
+  mutate_perf() %>%
+  mutate(strategy = "self-split")
+vsearch <- read_tsv(snakemake@input[["vsearch"]]) %>%
+  mutate_perf() %>%
   mutate(strategy = case_when(
-    method == 'de_novo' ~ method,
-    TRUE ~ as.character(glue('database_{ref}'))))
+    method == "de_novo" ~ method,
+    TRUE ~ as.character(glue("database_{ref}"))
+  ))
 
-results_agg <- list(opticlust, optifit_db, optifit_split, vsearch) %>% 
+results_agg <- list(opticlust, optifit_db, optifit_split, vsearch) %>%
   reduce(full_join)
 
-results_sum <- results_agg %>% 
-  group_by(tool, strategy, method, dataset, database, ref_frac, ref_weight) %>% 
-  summarize(n = n(),
-            mcc_median = median(mcc),  # TODO: tidy way to avoid this repetitiveness?
-            sec_median = median(sec),
[…]

### 2021-03-05 paper/paper.Rmd
Compare OptiClust vs de novo vsearch
Start inserting numbers in results w/ R code

@@
 mothuR::set_knitr_opts()
+```
+
+```{r deps, message = FALSE}
+library(here)
+library(mothuR)
+library(tidyverse)
+dat <- read_tsv(here('results', 'summarized.tsv'))
+rel_diff <- function(final, init, percent = TRUE) {
+  mult <- if (isTRUE(percent)) 100 else 1
+  return((final - init) / init * mult)
+}
 ```
@@
 closed reference mode as an additional measure of quality for this mode.
-_De novo_ clustering with OptiClust produced **X%** higher quality OTUs than
-VSEARCH, but performed **Y%** slower than VSEARCH.
 
 ### Reference clustering with public databases
+
+```{r ref_db}
+open_fit_mcc <- dat %>% filter(method == 'open', strategy == 'database', tool == 'mothur') %>% pull(mcc_median) %>% median()
+opticlust_mcc <- dat %>% filter(method == 'de_novo', tool == 'mothur') %>% pull(mcc_median) %>% median()
+open_fit_db_vs_clust <- rel_diff(opticlust_mcc, open_fit_mcc)
+
+open_fit_gg_mcc <- dat %>% filter(method == 'open', strategy == 'database', tool == 'mothur', ref == 'gg') %>% pull(mcc_median) %>% median()
+open_vsearch_mcc <- dat %>% filter(method == 'open', strategy == 'database', tool == 'vsearch', ref == 'gg') %>% pull(mcc_median) %>% median()
+open_fit_db_vs_vsearch <- rel_diff(open_fit_gg_mcc, open_vsearch_mcc)
+```
 
@@
 In open reference mode, fitting the datasets to reference OTUs with OptiFit
-produced OTUs of similar quality as clustering the datasets _de novo_ with
-OptiClust (**diff. in medians**).
-This held true for all datasets and reference databases.
+produced OTUs of similar quality (`r open_fit_db_vs_clust`% difference in median
+MCC) as clustering the datasets _de novo_ with OptiClust  across all datasets
+and reference databases.
 OptiFit produced higher quality OTUs than VSEARCH when open reference clustering
-against the Greengenes database, with median MCC scores of **A** and **B%**
-(respectively).
+against the Greengenes database, with median MCC scores of `r open_fit_gg_mcc`
+and `r open_vsearch_mcc` (respectively).
 However, open reference VSEARCH and OptiClust respectively ran **X%** and **Y%**
 faster than OptiFit in open reference mode.
+_De novo_ clustering with OptiClust produced **X%** higher quality OTUs than
+VSEARCH, but performed **Y%** slower than VSEARCH.
 

### 2021-03-05 paper/paper.Rmd
Update raw database seq counts

@@
 databases, which are unlikely to contain sequences from rare and novel microbes.
-This drop in quality was most notable with RDP, which has only about 16,000 V4
-16S sequences compared to about 170,000 in SILVA and Greengenes (**TODO: insert
-R code to generate these numbers**).
+This drop in quality was most notable with RDP, which contains only about 21,000
+sequences compared to over 200,000 sequences in SILVA and Greengenes each at the
+time of this writing.
 We recommend that users who require an independent reference database opt for

### 2021-03-05 paper/paper.Rmd
More numbers

@@
 opticlust_mcc <- dat %>% filter(method == 'de_novo', tool == 'mothur') %>% pull(mcc_median) %>% median()
-open_fit_db_vs_clust <- rel_diff(opticlust_mcc, open_fit_mcc)
+mcc_open_fit_db_vs_clust <- rel_diff(opticlust_mcc, open_fit_mcc)
 
@@
 open_vsearch_mcc <- dat %>% filter(method == 'open', strategy == 'database', tool == 'vsearch', ref == 'gg') %>% pull(mcc_median) %>% median()
-open_fit_db_vs_vsearch <- rel_diff(open_fit_gg_mcc, open_vsearch_mcc)
+mcc_open_fit_db_vs_vsearch <- rel_diff(open_fit_gg_mcc, open_vsearch_mcc)
+
+open_vsearch_sec <- dat %>% filter(method == 'open', strategy == 'database', tool == 'vsearch', ref == 'gg') %>% pull(sec_median) %>% median()
+open_fit_sec <- dat %>% filter(method == 'open', strategy == 'database', tool == 'mothur') %>% pull(sec_median) %>% median()
+opticlust_sec <- dat %>% filter(method == 'de_novo', tool == 'mothur') %>% pull(sec_median) %>% median()
+sec_vsearch_vs_open_fit_db <- rel_diff(open_vsearch_sec, open_fit_sec)
+sec_opticlust_vs_open_fit_db <- rel_diff(open_fit_sec, opticlust_sec)
+```
+
+```{r de_novo}
+dn_vsearch_mcc <- dat %>% filter(strategy == 'de_novo', tool == 'vsearch') %>% pull(mcc_median) %>% median()
+dn_vsearch_sec <- dat %>% filter(strategy == 'de_novo', tool == 'vsearch') %>% pull(sec_median) %>% median()
+mcc_opticlust_vs_vsearch <- rel_diff(opticlust_mcc, dn_vsearch_mcc)
+sec_opticlust_vs_vsearch <- rel_diff(opticlust_sec, dn_vsearch_sec)
 ```
@@
 In open reference mode, fitting the datasets to reference OTUs with OptiFit
-produced OTUs of similar quality (`r open_fit_db_vs_clust`% difference in median
+produced OTUs of similar quality (`r mcc_open_fit_db_vs_clust`% difference in median
 MCC) as clustering the datasets _de novo_ with OptiClust  across all datasets
@@
 and `r open_vsearch_mcc` (respectively).
-However, open reference VSEARCH and OptiClust respectively ran **X%** and **Y%**
-faster than OptiFit in open reference mode.
-_De novo_ clustering with OptiClust produced **X%** higher quality OTUs than
-VSEARCH, but performed **Y%** slower than VSEARCH.
+However, OptiClust and VSEARCH each ran `r sec_opticlust_vs_open_fit_db`% and 
+`r sec_vsearch_vs_open_fit_db`% faster than OptiFit in open reference mode.
+_De novo_ clustering with OptiClust produced `r mcc_opticlust_vs_vsearch`%
+higher quality OTUs than VSEARCH, but performed `r sec_opticlust_vs_vsearch`%
+slower than VSEARCH.
 

### 2021-03-05 paper/paper.Rmd
'as fast as' vs 'slower than'

@@
 mcc_opticlust_vs_vsearch <- rel_diff(opticlust_mcc, dn_vsearch_mcc)
-sec_opticlust_vs_vsearch <- rel_diff(opticlust_sec, dn_vsearch_sec)
+sec_opticlust_vs_vsearch <- abs(rel_diff(dn_vsearch_sec, opticlust_sec))
 ```
@@
 _De novo_ clustering with OptiClust produced `r mcc_opticlust_vs_vsearch`%
-higher quality OTUs than VSEARCH, but performed `r sec_opticlust_vs_vsearch`%
-slower than VSEARCH.
+higher quality OTUs than VSEARCH, but performed only 
+`r sec_opticlust_vs_vsearch`% as fast as VSEARCH.
 

### 2021-03-05 paper/paper.Rmd

@@
 
+- **TODO: current method for open/closed is vsearch against greengenes.**
+- **TODO**: 2 categories of clustering: _de novo_ and reference based.
+separate paragraphs.
+describe opticlust first in de novo paragraph.
+2nd paragraph: ref methods are good cause they're fast and don't use much ram.
+dependent on order of db.
+people use greengenes, which are rep seqs from 3% otus from full length.
+- reader should know what opticlust is, closed & open ref clustering is, strengths & weakness are of each. then we solve these problems.
+- **TODO:** note that greengenes is defunct now?!
+
 To overcome the limitations of _de novo_ clustering while maintaining OTU
@@
 novo_ OTUs and compared the performance to existing tools.
+To evaluate the OptiFit algorithm and compare to existing methods, we used four
+published datasets isolated from soil [@johnston_metagenomics_2016], marine
+[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
+gut [@baxter_microbiota-based_2016] samples.
 
@@
 
-**TODO: ask Sarah Westcott to check the accuracy of this description**
+- **TODO: ask Sarah Westcott to check the accuracy of this description**.
+- **TODO: toy example like opticlust with figure**.
 
@@
 method for reference-based clustering.
+**TODO: more narrative, less documentation-like.**
+**TODO: extension of opticlust except ref otus are fixed. just like opticlust only query seqs can move otus.**
 OptiFit takes as input a list of reference OTUs and their sequences, a list of
@@
 the algorithm calculates the quality metric based on whether the sequence stays
-in its current OTU or moves to each of the other OTUs.
+in its current OTU, moves to each of the other OTUs, or is discarded.
 If two or more OTU assignments are of equal quality, a random number generator
@@
 the best OTU assignments.
-
-To evaluate the OptiFit algorithm and compare to existing methods, we used four
-published datasets isolated from soil [@johnston_metagenomics_2016], marine
-[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
-gut [@baxter_microbiota-based_2016] samples. 
 There are two strategies for generating OTUs with OptiFit: 1) fit sequences to
@@
 fit the query sequences to the reference OTUs.
+
+**TODO: make sure this info is baked into relevant sections, then remove.**
 For each dataset repeated with 100 random seeds, we generated OTUs with OptiFit
@@
 Greengenes database.
-All clustering was performed at a distance threshold of 0.03 and OTU quality was
-evaluated using the MCC as described previously [@westcott_opticlust_2017].
-We calculated the fraction of query sequences that were fit to existing OTUs in
-closed reference mode as an additional measure of quality for this mode.
 
@@
 
-To evaluate reference-based clustering with independent databases, we fit each
-dataset to reference OTUs generated by _de novo_ clustering the Greengenes
+We assigned blah ref dbs to 3% otus with opticlust. _de novo_ clustering the Greengenes
 database, SILVA non-redundant database, or the Ribosomal Database Project (RDP)
 [@desantis_greengenes_2006; @quast_silva_2013; @cole_ribosomal_2014].
+All clustering was performed at a distance threshold of 0.03 and OTU quality was
+evaluated using the MCC as described previously [@westcott_opticlust_2017].
+To evaluate reference-based clustering with independent databases, we fit each
+dataset to reference OTUs generated by
+repeated 100 times.
+
+TODO: use word "map" for what vsearch does, "fit" for what optifit does.
 
@@
 OTU quality was remarkably stable across 100 different random seeds (**TODO:
-variation or stdev?**).
+coeff of variation?**).
 In terms of runtime, closed reference OptiFit performed faster than OptiClust on
@@
 mothur.
-Mothur's OptiFit and OptiClust algorithms use all of the sequences to define an
+The OptiFit and OptiClust algorithms use all of the sequences to define an
 OTU, requiring that all pairs of sequences (including reference and query
@@
 quality OTUs using already existing OTUs as a reference.
+**TODO: Point to courtney's paper metaphorically. wow what a cool application someone should do _wink wink_.**
 

### 2021-03-07 exploratory/2021/plots.Rmd
Fix typo

@@
   list(geom_hline(data = dat, aes(yintercept = {{yint}})),
-       labs(caption = "Black line: _de novo_ clustering0"),
+       labs(caption = "Black line: _de novo_ clustering"),
        theme(plot.caption = element_markdown())

### 2021-03-10 docs/slides/2021-03-11_bistro.Rmd
Make xaringan theme work

@@
   xaringan::moon_reader:
-    css: libs/xaringan-themer.css
+    css: xaringan-themer.css
     lib_dir: libs
@@
 
-# TODO: use http urls instead of here filepaths for figures
 ```
@@
 library(xaringanthemer)
-style_mono_light(
+style_mono_accent(
+  base_color = "#23395b",
   header_font_google = google_font("Helvetica"),
@@
 
+distance threshold: clumping vs splitting. some bacteria have multiple copies of rRNA gene.
+
 .footnote[

### 2021-03-11 docs/slides/2021-03-11_bistro.Rmd
Update slides for bistro

--- 2021-03-03 docs/slides/2021-03-11_bistro.Rmd
@@
 ---
-title: "OptiFit project update"
-subtitle: "+ a case study in debugging"
+title: "Comparing microbial communities with OptiFit: a fast method for fitting sequences to existing OTUs"
+subtitle: ""
 author: "Kelly Sovacool"
-date: "2021-03-11 (updated: `r Sys.Date()`)"
+date: "2021-03-11"
 output:
   xaringan::moon_reader:
-    css: libs/xaringan-themer.css
+    css: xaringan-themer.css
     lib_dir: libs
@@
 # prev slides:
+# - fall 2019 JC https://github.com/kelly-sovacool/bioinf603-fall2019
+# - Aug 2020 prelim https://docs.google.com/presentation/d/19kXLBr-lk_3Wva_raEzOGOYW2AH-GfA2ngcSSdOFJlY/edit
 # - Oct 2020 research talk https://docs.google.com/presentation/d/1GbbCYmlF4jkrO5BGyCTgtFBOL4pfzPDIQLMZBND0Hzg/edit#slide=id.p
 # - Dec 2020 research talk https://github.com/SchlossLab/OptiFitAnalysis/blob/master/docs/slides/2020-12-14_lab-mtg.Rmd
-# - fall 2019 JC https://github.com/kelly-sovacool/bioinf603-fall2019
 
-# TODO: use http urls instead of here filepaths for figures
 ```
@@
 )
-# style_mono_light(base_color = "#23395b")
-# style_mono_accent(
-#   base_color = "#1c5253",
-#   header_font_google = google_font("Josefin Sans"),
-#   text_font_google   = google_font("Montserrat", "300", "300i"),
-#   code_font_google   = google_font("Fira Mono")
-# )
 ```
@@
 
+## Role of the microbiome in human health
+
+.center[
+<img src="https://raw.githubusercontent.com/kelly-sovacool/bioinf603-fall2019/master/docs/figures/human_microbiome.gif" height="500"/>
+]
+---
+
+## The gut microbiome changes in colorectal cancer
+
+.pull-left[
+- Colorectal cancer patients have an altered gut microbiome compared to healthy individuals.
+
+- Fecal matter transplants from cancerous mice increase tumor formation in germ-free mice.
+
+- Machine learning models trained on microbiome data to predict CRC have modest performance (mean AUROC ~0.7) 
+]
+
+.pull-right[
+![](../../figures/zackular-3a.jpg)
+]
+
+.footnote[
+[Zackular et al. (2013) _mBio_](doi.org/10.1128/mBio.00692-13)
+]
+
+---
+
+## Characterizing microbial communities
+
+.pull-left[
+<img src="../../figures/omics-cascade.jpg" height=500/>
+]
+
+--
+
+.pull-right[
+
+- Genomics: Which microbes are there & what are they capable of doing
+
+- Transcriptomics: What are they trying to do
+
+- Proteomics: What are they actually doing
+
+- Metabolomics: How are they affecting the host & each other
+
+.footnote[
+[Haukaas et al. (2017) _Metabolites_](doi.org/10.3390/metabo7020018)
+]
+]
+
+---
+
+## Sequencing techniques
+
+|   | Metagenomics | Amplicon Sequencing |
+|------|--------------|----------|
+| **Technique** | Whole-metagenome shotgun sequencing (every gene of every microbe) | Sequence just 250bp of the 16S rRNA gene |
+| **Taxonomic resolution** | Species or strain | Genus |
+| **Organisms** | Bacteria, archaea, viruses, fungi | Bacteria & archaea; or fungi separately | 
+| **Purpose** | Identify which microbial genes are present; assemble microbial genomes | Identify which microbes are present and their abundances |
+| **Cost** | 💲💲💲 | 💲 |
+
+---
+
+## Operational Taxonomic Units (OTUs)
+
+- Cluster similar sequences together into OTUs.
+- OTUs clustered at 3% distance (97% sequence similarity) generally can be labelled at the genus level.
+
+.center[
+<img src="https://raw.githubusercontent.com/kelly-sovacool/bioinf603-fall2019/master/docs/figures/OTU_clustering.png" height="400" />
+]
+.pull-right[Image credit: M. Brodie Mumphrey]
+
+---
+
+## Why cluster sequences into OTUs?
+
+- Bacterial taxonomy is imperfect. Some species are overly broad, others are too narrow.
+  - e.g. some strains of _E. coli_ share only 20% of their genomes.
+
+- Many bacteria have multiple copies of the rRNA gene.
+  - Pick a sequence distance threshold to minimize splitting sequences from the same genome into different OTUs.
+
+- Bacteria evolve rapidly. Don't want to exclude novel microbes from an analysis.
+
+???
+
+Bacterial taxonomy is hard
+
+Goal of classification is to reflect evolutionary history.
+
+Challenges:
+- Asexual reproduction
+- Fast evolutionary change
+- Horizontal gene transfer
+
+Current classifications are far from perfect.
+
+---
+
+## Evaluating OTU Quality
+
+Pairs of sequences that are under the distance threshold should be assigned to the same OTU; sequences above the threshold should be assigned to different OTUs.
+
+Confusion Matrix:
+
+|   | Actual Positive | Actual Negative |
+|---|---|---|
+| Predicted Positive | TP | FP |
+| Predicted Negative | FN | TP |
+
+--
+
+Matthews Correlation Coefficient:
+
+$$MCC = \frac{TP \times TN - FP \times FN}{\sqrt{(TP + FP)(TP + FN)(TN + FP)(TN + FN)}}$$
+
+Range of MCC:
+
+- 1 - perfect prediction
+- 0 - random
+- -1 - completely wrong
+
+---
+
 ## OTU Clustering methods
 
-- _De novo_ (OptiClust)
-  - Cluster sequences based on their similarity.
-  - OTU's aren't stable; you'll get different OTU assignments with different random seeds.
-- Reference-based (OptiFit)
-  - First cluster reference sequences _de novo_, then fit query sequences to those OTUs.
+- _De novo_ 
+  - Cluster sequences based on their similarity to each other.
+  - Limitation: OTUs aren't stable; you'll get different OTU assignments when new sequences are added.
+- Reference-based 
+  - Cluster sequences based on their similarity to a reference database.
+  - Limitation: novel sequences won't be represented. OTUs are limited by the diversity represented in the database.
+
+.center[
+<img src="../../figures/schloss-2016-msystems_all-method-comparison.png"  height="350"/>
+]
+
+???
+
+_De novo_ methods out-perform reference-based methods.
 
 .footnote[
+
   [Westcott & Schloss (2015) _PeerJ_ ](https://doi.org/10.7717%2Fpeerj.1487);
@@
 
---
+---
 
-### Reference-based strategies
+## OptiClust: _de novo_ OTUs that optimize the MCC
+
+1. Calculate all pairwise sequence distance scores.
+
+1. Begin with each sequence assigned to its own OTU.
+
+1. For each sequence, consider whether to move it to a different OTU or not. Do whichever option results in the best MCC score. Repeat until the MCC doesn't change.
+
+.center[
+<img src="../../figures/opticlust-supp-ex.png" width="700"/>
+]
+<img src="https://raw.githubusercontent.com/mothur/logo/master/mothur_RGB.png" width="200"/>
+
+???
+if equally good, pick one at random
+
+---
+
+## OptiFit: assign new sequences to existing OTUs
+
+1. Start with OTUs that were previously clustered _de novo_ to use as a reference.
+
+1. Calculate pairwise sequence distance scores between all reference and query sequences.
+
+1. Begin with each query sequence randomly assigned to a reference OTU.
+
+1. For each query sequence, consider whether to move it to a different OTU or not. Do whichever option results in the best MCC score. Repeat until the MCC doesn't change.
+
+&nbsp;  
+&nbsp;
+&nbsp;
+&nbsp;
+
+.pull-left[
+<img src="https://raw.githubusercontent.com/mothur/logo/master/mothur_RGB.png" width="200"/>
+]
+
+---
+
+## Reference-based strategies
 
@@
   - Closed - sequences that can't be fit to the reference are thrown out.
+
 - Reference choice
-  - External database (Silva, greengenes, RDP).
+  - External, independent database (Silva, greengenes, RDP).
   - Split the dataset into a reference & query set.
@@
 
-![](../../exploratory/2020-11_Nov-Dec/figures/mcc_fit-db-1.png)
+## Reference clustering against a database
+
+Questions
+
+- Which database should we use?
+  - SILVA, Greengenes, Ribosomal database project
+
+- How does the performance of fitting datasets to independent databases compare to _de novo_ clustering?
+
+- Are the results consistent across datasets from a variety of sources? 
+  - Human gut, mouse gut, marine, soil
+
+- How does mothur compare to vsearch?
+
+---
+
+## Reference clustering against a database
+
+.pull-left[
+
+<img src='../../exploratory/2021/figures/mcc_fit-db-1.png' />
+]
+
+--
+
+.pull-right[
+
+<img src='../../exploratory/2021/figures/fraction-mapped_fit-db-1.png' />
+]
+
+???
+- _De novo_ and open-reference perform equally well.
+- Closed reference performs slightly worse.
+- RDP doesn't perform well.
+- Prior to switching to silva v132 & updating mothur, closed-reference artificially outperformed open-reference.
+
+---
+
+## Reference clustering against a database
+
+.pull-left[
+
+<img src='../../exploratory/2021/figures/runtime_fit-db-1.png' />
+]
+
+--
+
+.pull-right[
+
+<img src='../../exploratory/2021/figures/memory_fit-db-1.png' />
+]
+
+---
+
+## mothur vs. vsearch
+
+<img src='../../exploratory/2021/figures/metrics_all-human-1.png' />
+
+???
+- Just showing the human dataset here because results patterns were basically the same across all datasets.
+- Only used greengenes for reference because that's what
+QIIME does.
+
+---
+
+## Splitting datasets into reference & query sets
+
+Questions
+
+- What fraction of sequences should be in the reference?
+
+- How to select which sequences to be in the reference?
+  - Simple random sample
+  - Weight by sequence abundance
+  - Weight by connectivity (number of pairs within the distance threshold)
+
+- How does the performance of fitting datasets to reference fractions compare to de novo clustering?
+
+- Are the results consistent across datasets from a variety of sources?
+  - Human gut, mouse gut, marine, soil
+
+---
+
+## Splitting datasets into reference & query sets
+
+<img src='../../exploratory/2020-05_May-Oct/figures/mcc_complement-4.png' />
+
+---
+
+## Splitting datasets into reference & query sets
+
+<img src='../../exploratory/2021/figures/fraction-mapped_fit-split-1.png' />
+
+---
+
+## Splitting datasets into reference & query sets
+
+<img src='../../exploratory/2021/figures/runtime_fit-split-1.png' />
+
+---
+
+## Summarizing mothur results
+
+<img src='../../exploratory/2021/figures/mcc_all-opti-1.png' />
+
+???
+
+Whenever there's virtually no variation in the metric across random seeds.
+
+---
+## Conclusions
+
+- If you need to assign new sequences to existing OTUs, OptiFit results in OTUs of similar quality as OptiClust.
+
+- _De novo_ clustering still usually outperforms reference clustering against an external database.
+
+- VSEARCH runs faster but produces lower quality OTUs than OptiFit or OptiClust.
+
+- OptiFit runs slower than OptiClust in some cases; continue using OptiClust if stable OTUs aren't required.
+
+--
+
+## What's next
+
+- Demonstrate that OptiFit works well for machine learning.
+
+- Integrate OTUs with other microbial omics data.
+
+---
+
+## Reproducible Pipelines with Snakemake
+
+.center[
+<img src="https://snakemake2019.readthedocs.io/en/latest/_images/workflow_management.png" width="750"/>
+]
+
+---
+
+.pull-right[
+<img src="../../figures/dag_fit_db.png" height="600" />
+]
+
+---
+
+## Sequence similarity / distance
+
+1. Calculate all pairwise sequence identity scores to create a distance matrix.
+2. Apply cutoff of 97% sequence identity to create an adjacency matrix.
+.center[
+```
+seqA AGGGTACG
+seqB ACG-TACG
+```
+$$ \%\ sequence\ identity =  100 * \frac{number\ of\ matches}{sequence\  length} $$
+]
+.pull-left[
+1) Distance matrix:
+
+| | Seq1 | Seq2 | Seq3 |
+|-|------|------|------|
+| Seq1 | - | 98 | 87 |
+| Seq2 | 98 | - | 76 |
+| Seq3 | 87 | 76 | - |
+]
+.pull-right[
+2) Adjacency matrix:
+
+| | Seq1 | Seq2 | Seq3 |
+|-|------|------|------|
+| Seq1 | - | 1 | 0 |
+| Seq2 | 1 | - | 0 |
+| Seq3 | 0 | 0 | - |
+]

### 2021-03-11 docs/slides/2021-03-11_bistro.Rmd
Add lots of logos

@@
 style_mono_accent(
-  base_color = "#23395b",
+  #base_color = "#23395b",
+  base_color = "#000000",
   header_font_google = google_font("Helvetica"),
@@
 
+background-image: url("../../figures/SchlossLab.png") 
+background-size: contain
+background-color: "#000000"
+class: inverse
+
+---
+
 ## Role of the microbiome in human health
@@
 ]
+
 ---
@@
 |------|--------------|----------|
-| **Technique** | Whole-metagenome shotgun sequencing (every gene of every microbe) | Sequence just 250bp of the 16S rRNA gene |
+| **Technique** | Whole-metagenome shotgun sequencing: every gene of every microbe | Sequence just 250bp of the 16S rRNA gene |
+| **Cost** | 💲💲💲 | 💲 |
 | **Taxonomic resolution** | Species or strain | Genus |
@@
 | **Purpose** | Identify which microbial genes are present; assemble microbial genomes | Identify which microbes are present and their abundances |
-| **Cost** | 💲💲💲 | 💲 |
+
+???
+metagenomics is prohibitively expensive for very large datasets
 
@@
 
-- Cluster similar sequences together into OTUs.
+- Cluster similar sequences together into OTUs as a proxy for genus/species.
 - OTUs clustered at 3% distance (97% sequence similarity) generally can be labelled at the genus level.
@@
 Current classifications are far from perfect.
+
+Courtney's resolution figure?
 
@@
 ]
+.footnote[
 <img src="https://raw.githubusercontent.com/mothur/logo/master/mothur_RGB.png" width="200"/>
-
+]
 ???
@@
 
-## Reproducible Pipelines with Snakemake
+## Managing Bioinformatics Pipelines
 
@@
 
+.footnote[
+[snakemake.readthedocs.io](snakemake.readthedocs.io)
+]
+
+???
+I ran about 20,000 jobs on the GreatLakes HPC
+
 ---
 
-.pull-right[
-<img src="../../figures/dag_fit_db.png" height="600" />
+## Pipelines with <img src="https://raw.githubusercontent.com/snakemake/snakemake/f55801bfabfb916f77fa4ec8f9eb0667e5c8a92e/images/biglogo.png" height="35"/>
+
+.center[
+<img src="https://raw.githubusercontent.com/snakemake/snakemake/f55801bfabfb916f77fa4ec8f9eb0667e5c8a92e/docs/project_info/img/idea.png" width="900" />
+]
+
+.footnote[
+[snakemake.readthedocs.io](snakemake.readthedocs.io)
+]
+
+---
+
+## Tools for Reproducible Bioinformatics
+
+| package mgmt | scripts | reports | workflow | version control |
+|---|---|---|---|---|
+|   <img src="https://docs.conda.io/en/latest/_images/conda_logo.svg" width="175"/> | <img src="https://longhowlam.files.wordpress.com/2017/04/rp.png" height="175"/> | <img src="https://ulyngs.github.io/rmarkdown-workshop-2019/slides/figures/rmarkdown.png" height="165"/> | <img src="https://avatars.githubusercontent.com/u/33450111?s=400&v=4" height="175"> |<img src="https://miro.medium.com/max/2625/1*8fPMdk2Cd5iJQ7dI7jXCbA.jpeg" height="175"/> |
+
+--
+
+.center[
+### ...and headache prevention!
 ]

### 2021-03-11 docs/slides/2021-03-11_bistro.Rmd
Adjust snakemake links

@@
 .footnote[
-[snakemake.readthedocs.io](snakemake.readthedocs.io)
+[snakemake.readthedocs.io](https://snakemake.readthedocs.io)
 ]
@@
 
+&nbsp;
+
 .center[
-<img src="https://raw.githubusercontent.com/snakemake/snakemake/f55801bfabfb916f77fa4ec8f9eb0667e5c8a92e/docs/project_info/img/idea.png" width="900" />
+<img src="https://raw.githubusercontent.com/snakemake/snakemake/f55801bfabfb916f77fa4ec8f9eb0667e5c8a92e/docs/project_info/img/idea.png" width="1000" />
 ]

### 2021-03-11 docs/slides/2021-03-11_bistro.Rmd
Minor tweaks

@@
 class: inverse
+
+???
+Sarah Westcott, software engineer.
+implemented algorithms I'm discussing today.
 
@@
 
-1. Calculate all pairwise sequence distance scores.
+1. Calculate all pairwise distance scores.
+
+  .center[
+```
+seqA AGGGTACG
+seqB ACG-TACG
+```
+$$distance = 1 - sequence\ similarity = 1 - \frac{number\ of\ matches}{sequence\  length}$$
+]
 
@@
 ]
+
 .footnote[
-<img src="https://raw.githubusercontent.com/mothur/logo/master/mothur_RGB.png" width="200"/>
+<img src="https://raw.githubusercontent.com/mothur/logo/master/mothur_RGB.png" height="150"/>
 ]
+
 ???
-if equally good, pick one at random
+if equally good, pick one at random.
+
+opticlust is great, but still has limitations of _de novo_ clustering: OTU assignments aren't stable when new sequences are added.
 
@@
 - Modes
+  - Closed - sequences that can't be fit to the reference are thrown out.
   - Open - sequences that can't be fit to the reference are then clustered _de novo_ to create additional OTUs.
-  - Closed - sequences that can't be fit to the reference are thrown out.
 
 - Reference choice
-  - External, independent database (Silva, greengenes, RDP).
+  - External, independent database (SILVA, greengenes, RDP).
   - Split the dataset into a reference & query set.
@@
 ]
-
----
-
-## Sequence similarity / distance
-
-1. Calculate all pairwise sequence identity scores to create a distance matrix.
-2. Apply cutoff of 97% sequence identity to create an adjacency matrix.
-.center[
-```
-seqA AGGGTACG
-seqB ACG-TACG
-```
-$$ \%\ sequence\ identity =  100 * \frac{number\ of\ matches}{sequence\  length} $$
-]
-.pull-left[
-1) Distance matrix:
-
-| | Seq1 | Seq2 | Seq3 |
-|-|------|------|------|
-| Seq1 | - | 98 | 87 |
-| Seq2 | 98 | - | 76 |
-| Seq3 | 87 | 76 | - |
-]
-.pull-right[
-2) Adjacency matrix:
-
-| | Seq1 | Seq2 | Seq3 |
-|-|------|------|------|
-| Seq1 | - | 1 | 0 |
-| Seq2 | 1 | - | 0 |
-| Seq3 | 0 | 0 | - |
-]

### 2021-03-11 docs/slides/2021-03-11_bistro.Rmd
Minor updates

@@
 ---
-title: "Comparing microbial communities with OptiFit: a fast method for fitting sequences to existing OTUs"
+title: "OptiFit: a fast method for fitting sequences to existing OTUs for microbial ecology"
 subtitle: ""
@@
 
+???
+Human Microbes: The Power within (2018)
+
+[The Human Microbiome Project (2012) Nature](https://dx.doi.org/10.1038%2Fnature11234)
+
 ---
@@
 metagenomics is prohibitively expensive for very large datasets
+
+mothur: command line program for processing amplicon sequence data and comparing microbial communities
 

### 2021-03-16 paper/paper.Rmd
Calculate stats for paper in separate script

@@
 library(tidyverse)
-dat <- read_tsv(here('results', 'summarized.tsv'))
-rel_diff <- function(final, init, percent = TRUE) {
-  mult <- if (isTRUE(percent)) 100 else 1
-  return((final - init) / init * mult)
-}
+load(here('results', 'stats.RData'))
 ```
@@
 
-```{r ref_db}
-open_fit_mcc <- dat %>% filter(method == 'open', strategy == 'database', tool == 'mothur') %>% pull(mcc_median) %>% median()
-opticlust_mcc <- dat %>% filter(method == 'de_novo', tool == 'mothur') %>% pull(mcc_median) %>% median()
-mcc_open_fit_db_vs_clust <- rel_diff(opticlust_mcc, open_fit_mcc)
+**TODO: use word "map" for what vsearch does, "fit" for what optifit does.**
 
-open_fit_gg_mcc <- dat %>% filter(method == 'open', strategy == 'database', tool == 'mothur', ref == 'gg') %>% pull(mcc_median) %>% median()
-open_vsearch_mcc <- dat %>% filter(method == 'open', strategy == 'database', tool == 'vsearch', ref == 'gg') %>% pull(mcc_median) %>% median()
-mcc_open_fit_db_vs_vsearch <- rel_diff(open_fit_gg_mcc, open_vsearch_mcc)
-
-open_vsearch_sec <- dat %>% filter(method == 'open', strategy == 'database', tool == 'vsearch', ref == 'gg') %>% pull(sec_median) %>% median()
-open_fit_sec <- dat %>% filter(method == 'open', strategy == 'database', tool == 'mothur') %>% pull(sec_median) %>% median()
-opticlust_sec <- dat %>% filter(method == 'de_novo', tool == 'mothur') %>% pull(sec_median) %>% median()
-sec_vsearch_vs_open_fit_db <- rel_diff(open_vsearch_sec, open_fit_sec)
-sec_opticlust_vs_open_fit_db <- rel_diff(open_fit_sec, opticlust_sec)
-```
-
-```{r de_novo}
-dn_vsearch_mcc <- dat %>% filter(strategy == 'de_novo', tool == 'vsearch') %>% pull(mcc_median) %>% median()
-dn_vsearch_sec <- dat %>% filter(strategy == 'de_novo', tool == 'vsearch') %>% pull(sec_median) %>% median()
-mcc_opticlust_vs_vsearch <- rel_diff(opticlust_mcc, dn_vsearch_mcc)
-sec_opticlust_vs_vsearch <- abs(rel_diff(dn_vsearch_sec, opticlust_sec))
-```
-
+**TODO: ref clustering method paragraph here.**
 We assigned blah ref dbs to 3% otus with opticlust. _de novo_ clustering the Greengenes
@@
 repeated 100 times.
-
-TODO: use word "map" for what vsearch does, "fit" for what optifit does.
 
@@
 
-In closed reference mode, OptiFit produced lower quality OTUs than OptiClust by
-**X%** when fitting sequences to Greengenes and SILVA, and **Y%** worse when
-fitting to RDP.
-Only up to **Z%** of query sequences were fit to reference OTUs in closed
-reference mode across any dataset/database combination.
-OptiFit was able to fit **X** more query sequences to reference OTUs created
-with the Greengenes and SILVA databases than with RDP.
-VSEARCH was able to fit **W%** more query sequences to the Greengenes reference
-than OptiFit in closed reference mode.
-In terms of run time, closed reference OptiFit outperformed
-OptiClust by **A%** and VSEARCH by **B%**.
+In closed reference mode against databases, OptiFit produced lower quality OTUs
+than OptiClust by `r mcc_closed_fit_db_vs_clust` on average.
+Fitting sequences to Greengenes and SILVA in closed reference mode performed
+similarly with median MCC scores of `r closed_fit_gg_mcc` and 
+`r closed_fit_silva_mcc` respectively, while when fitting to RDP the median MCC 
+dropped to `r closed_fit_rdp_mcc`.
+An average of only `r frac_fit_db` of query sequences were fit to reference
+database OTUs in closed reference mode across all dataset/database combinations.
+VSEARCH was able to map `r frac_vsearch_vs_fit` more query sequences than
+OptiFit to the Greengenes reference database.
+In terms of run time, closed reference OptiFit outperformed OptiClust by 
+`r sec_closed_fit_db_vs_clust` and VSEARCH by `r sec_closed_fit_db_vs_vsearch`.
 

### 2021-03-16 code/R/calc_results_stats.R

@@ -1,0 +1,117 @@
+library(here)
+library(tidyverse)
+
+rel_diff <- function(final, init, percent = TRUE) {
+  mult <- if (isTRUE(percent)) 100 else 1
+  return((final - init) / init * mult)
+}
+
+
+dat <- read_tsv(here('results', 'summarized.tsv'))
+
+# de novo
+opticlust_mcc <- dat %>% filter(method == 'de_novo', 
+                                tool == 'mothur') %>% 
+  pull(mcc_median) %>% median()
+opticlust_sec <- dat %>% filter(method == 'de_novo', 
+                                tool == 'mothur') %>% 
+  pull(sec_median) %>% median()
+dn_vsearch_mcc <-
+  dat %>% filter(strategy == 'de_novo', tool == 'vsearch') %>% 
+  pull(mcc_median) %>% median()
+dn_vsearch_sec <-
+  dat %>% filter(strategy == 'de_novo', tool == 'vsearch') %>% 
+  pull(sec_median) %>% median()
+mcc_opticlust_vs_vsearch <- rel_diff(opticlust_mcc, dn_vsearch_mcc)
+sec_opticlust_vs_vsearch <- abs(rel_diff(dn_vsearch_sec, opticlust_sec))
+
+# ref db open 
+open_fit_mcc <- dat %>% filter(method == 'open', 
+                               strategy == 'database', 
+                               tool == 'mothur') %>% 
+  pull(mcc_median) %>% median()
+mcc_open_fit_db_vs_clust <- rel_diff(opticlust_mcc, open_fit_mcc)
+
+open_fit_gg_mcc <- dat %>% filter(method == 'open',
+                                  strategy == 'database',
+                                  tool == 'mothur',
+                                  ref == 'gg') %>% pull(mcc_median) %>% median()
+open_vsearch_mcc <- dat %>% filter(method == 'open',
+                                   strategy == 'database',
+                                   tool == 'vsearch',
+                                   ref == 'gg') %>% pull(mcc_median) %>% median()
+mcc_open_fit_db_vs_vsearch <- rel_diff(open_fit_gg_mcc, open_vsearch_mcc)
+
+open_vsearch_sec <- dat %>% filter(method == 'open',
+                                   strategy == 'database',
+                                   tool == 'vsearch',
+                                   ref == 'gg') %>% pull(sec_median) %>% median()
+open_fit_sec <- dat %>% filter(method == 'open', 
[…]

### 2021-03-16 code/R/calc_results_stats.R
🎨 Style R code ⬡

@@ -7,111 +7,190 @@ rel_diff <- function(final, init, percent = TRUE) {
 }
 
 
-dat <- read_tsv(here('results', 'summarized.tsv'))
+dat <- read_tsv(here("results", "summarized.tsv"))
 
 # de novo
-opticlust_mcc <- dat %>% filter(method == 'de_novo', 
-                                tool == 'mothur') %>% 
-  pull(mcc_median) %>% median()
-opticlust_sec <- dat %>% filter(method == 'de_novo', 
-                                tool == 'mothur') %>% 
-  pull(sec_median) %>% median()
+opticlust_mcc <- dat %>%
+  filter(
+    method == "de_novo",
+    tool == "mothur"
+  ) %>%
+  pull(mcc_median) %>%
+  median()
+opticlust_sec <- dat %>%
+  filter(
+    method == "de_novo",
+    tool == "mothur"
+  ) %>%
+  pull(sec_median) %>%
+  median()
 dn_vsearch_mcc <-
-  dat %>% filter(strategy == 'de_novo', tool == 'vsearch') %>% 
-  pull(mcc_median) %>% median()
+  dat %>%
+  filter(strategy == "de_novo", tool == "vsearch") %>%
+  pull(mcc_median) %>%
+  median()
 dn_vsearch_sec <-
-  dat %>% filter(strategy == 'de_novo', tool == 'vsearch') %>% 
-  pull(sec_median) %>% median()
+  dat %>%
+  filter(strategy == "de_novo", tool == "vsearch") %>%
+  pull(sec_median) %>%
+  median()
 mcc_opticlust_vs_vsearch <- rel_diff(opticlust_mcc, dn_vsearch_mcc)
 sec_opticlust_vs_vsearch <- abs(rel_diff(dn_vsearch_sec, opticlust_sec))
 
-# ref db open 
-open_fit_mcc <- dat %>% filter(method == 'open', 
-                               strategy == 'database', 
-                               tool == 'mothur') %>% 
-  pull(mcc_median) %>% median()
+# ref db open
+open_fit_mcc <- dat %>%
+  filter(
+    method == "open",
+    strategy == "database",
+    tool == "mothur"
+  ) %>%
+  pull(mcc_median) %>%
+  median()
 mcc_open_fit_db_vs_clust <- rel_diff(opticlust_mcc, open_fit_mcc)
 
-open_fit_gg_mcc <- dat %>% filter(method == 'open',
-                                  strategy == 'database',
[…]

### 2021-03-16 paper/paper.Rmd
Fill in rest of numbers

@@
 To evaluate reference-based clustering with independent databases, we fit each
-dataset to reference OTUs generated by
+dataset to reference OTUs generated by...
 repeated 100 times.
@@
 and `r open_vsearch_mcc` (respectively).
-However, OptiClust and VSEARCH each ran `r sec_opticlust_vs_open_fit_db`% and 
-`r sec_vsearch_vs_open_fit_db`% faster than OptiFit in open reference mode.
+OptiFit ran faster than VSEARCH by `r sec_vsearch_vs_open_fit_db`%  in open
+reference mode, but slower than OptiClust by `r sec_opticlust_vs_open_fit_db`%.
 _De novo_ clustering with OptiClust produced `r mcc_opticlust_vs_vsearch`%
-higher quality OTUs than VSEARCH, but performed only 
-`r sec_opticlust_vs_vsearch`% as fast as VSEARCH.
+higher quality OTUs than VSEARCH, but performed `r sec_opticlust_vs_vsearch`% 
+slower than VSEARCH.
 
@@
 dropped to `r closed_fit_rdp_mcc`.
-An average of only `r frac_fit_db` of query sequences were fit to reference
+An average of only `r frac_fit_db`% of query sequences were fit to reference
 database OTUs in closed reference mode across all dataset/database combinations.
-VSEARCH was able to map `r frac_vsearch_vs_fit` more query sequences than
+VSEARCH was able to map `r frac_vsearch_vs_fit`% more query sequences than
 OptiFit to the Greengenes reference database.
 In terms of run time, closed reference OptiFit outperformed OptiClust by 
-`r sec_closed_fit_db_vs_clust` and VSEARCH by `r sec_closed_fit_db_vs_vsearch`.
+`r sec_closed_fit_db_vs_clust`% and VSEARCH by `r sec_closed_fit_db_vs_vsearch`%.
 
@@
 that from _de novo_ clustering the whole dataset with OptiClust regardless of
-mode (**TODO: diff in MCC medians**).
-OTU quality was remarkably stable across 100 different random seeds (**TODO:
-coeff of variation?**).
-In terms of runtime, closed reference OptiFit performed faster than OptiClust on
-whole datasets by **Z%**.
-In open reference mode, OptiClust performed **X to Y%** faster than OptiFit only
-when the OptiFit reference fraction was 30% or less.
-The split dataset strategy performed just as well as the database strategy in
-open reference mode regardless of database used, and outperformed the database
-strategy in closed reference mode by **W%**.
+mode (`r mcc_opticlust_vs_fit_split`% difference in median MCC).
+OTU quality was remarkably stable across reference fraction sizes within
+parameter sets (Fig. **XX**).
+For example, splitting the human dataset 100 times yielded a coefficient of
+variation of `r cv_fit_split_mcc_human` for the MCC score.
+In terms of runtime, closed and open reference OptiFit performed faster than
+OptiClust on whole datasets by `r sec_closed_fit_split_vs_clust`% and 
+`r sec_open_fit_split_vs_clust` respectively.
+The split dataset strategy performed `r sec_closed_fit_split_vs_db`% faster than
+the database strategy in closed reference mode and `r sec_open_fit_split_vs_db`%
+in open reference mode.
 
@@
 OTU quality was similar with the simple and abundance-weighted sampling (median
-MCCs **X** and **Y** respectively), but **Z%** worse with similarity-weighted
-sampling.
+MCCs `r mcc_fit_split_simple` and `r mcc_fit_split_abun` respectively), but worse for similarity-weighted
+sampling with a median MCC of `r mcc_fit_split_dist`.
 In closed reference mode, the fraction of query sequences that can be fit to the
-reference OTUs decreases as the reference fraction increases from an MCC of
-**A** with **J** reference sequences to an MCC of **B** with **K** reference
-sequences.
+reference OTUs decreases as the reference fraction increases from 
+`r frac_fit_split_0.1`% of query sequences fit with `10`% of the dataset as the
+reference, to `r frac_fit_split_0.8`% of query sequences fit with `80`% of the
+dataset as the reference.
 
@@
 OTUs than VSEARCH in open reference, closed reference, or _de novo_ modes.
-However, VSEARCH was able to fit more sequence into OTUs than OptiFit in closed
+However, VSEARCH was able to map more sequences to OTUs than OptiFit in closed
 reference mode.

### 2021-03-16 code/R/calc_results_stats.R

@@ -5,6 +5,9 @@ rel_diff <- function(final, init, percent = TRUE) {
   mult <- if (isTRUE(percent)) 100 else 1
   return((final - init) / init * mult)
 }
+coeff_var <- function(x) {
+  return(sd(x) / mean(x))
+}
 
 
 dat <- read_tsv(here('results', 'summarized.tsv'))
@@ -26,11 +29,11 @@ mcc_opticlust_vs_vsearch <- rel_diff(opticlust_mcc, dn_vsearch_mcc)
 sec_opticlust_vs_vsearch <- abs(rel_diff(dn_vsearch_sec, opticlust_sec))
 
 # ref db open 
-open_fit_mcc <- dat %>% filter(method == 'open', 
+open_fit_db_mcc <- dat %>% filter(method == 'open', 
                                strategy == 'database', 
                                tool == 'mothur') %>% 
   pull(mcc_median) %>% median()
-mcc_open_fit_db_vs_clust <- rel_diff(opticlust_mcc, open_fit_mcc)
+mcc_open_fit_db_vs_clust <- rel_diff(opticlust_mcc, open_fit_db_mcc)
 
 open_fit_gg_mcc <- dat %>% filter(method == 'open',
                                   strategy == 'database',
@@ -46,12 +49,12 @@ open_vsearch_sec <- dat %>% filter(method == 'open',
                                    strategy == 'database',
                                    tool == 'vsearch',
                                    ref == 'gg') %>% pull(sec_median) %>% median()
-open_fit_sec <- dat %>% filter(method == 'open', 
+open_fit_db_sec <- dat %>% filter(method == 'open', 
                                strategy == 'database', 
                                tool == 'mothur') %>%
   pull(sec_median) %>% median()
-sec_vsearch_vs_open_fit_db <- rel_diff(open_vsearch_sec, open_fit_sec)
-sec_opticlust_vs_open_fit_db <- rel_diff(open_fit_sec, opticlust_sec)
+sec_vsearch_vs_open_fit_db <- rel_diff(open_vsearch_sec, open_fit_db_sec)
+sec_opticlust_vs_open_fit_db <- rel_diff(opticlust_sec, open_fit_db_sec) %>% abs()
 
 # ref db closed
 closed_fit_db_mcc <- dat %>% filter(method == 'closed',
@@ -79,39 +82,95 @@ closed_fit_rdp_mcc <- dat %>% filter(method == 'closed',
 frac_fit_db <-  dat %>% filter(method == 'closed',
[…]

### 2021-03-16 code/R/calc_results_stats.R
🎨 Style R code ⬡

@@ -10,167 +10,282 @@ coeff_var <- function(x) {
 }
 
 
-dat <- read_tsv(here('results', 'summarized.tsv'))
+dat <- read_tsv(here("results", "summarized.tsv"))
 
 # de novo
-opticlust_mcc <- dat %>% filter(method == 'de_novo', 
-                                tool == 'mothur') %>% 
-  pull(mcc_median) %>% median()
-opticlust_sec <- dat %>% filter(method == 'de_novo', 
-                                tool == 'mothur') %>% 
-  pull(sec_median) %>% median()
+opticlust_mcc <- dat %>%
+  filter(
+    method == "de_novo",
+    tool == "mothur"
+  ) %>%
+  pull(mcc_median) %>%
+  median()
+opticlust_sec <- dat %>%
+  filter(
+    method == "de_novo",
+    tool == "mothur"
+  ) %>%
+  pull(sec_median) %>%
+  median()
 dn_vsearch_mcc <-
-  dat %>% filter(strategy == 'de_novo', tool == 'vsearch') %>% 
-  pull(mcc_median) %>% median()
+  dat %>%
+  filter(strategy == "de_novo", tool == "vsearch") %>%
+  pull(mcc_median) %>%
+  median()
 dn_vsearch_sec <-
-  dat %>% filter(strategy == 'de_novo', tool == 'vsearch') %>% 
-  pull(sec_median) %>% median()
+  dat %>%
+  filter(strategy == "de_novo", tool == "vsearch") %>%
+  pull(sec_median) %>%
+  median()
 mcc_opticlust_vs_vsearch <- rel_diff(opticlust_mcc, dn_vsearch_mcc)
 sec_opticlust_vs_vsearch <- abs(rel_diff(dn_vsearch_sec, opticlust_sec))
 
-# ref db open 
-open_fit_db_mcc <- dat %>% filter(method == 'open', 
-                               strategy == 'database', 
-                               tool == 'mothur') %>% 
-  pull(mcc_median) %>% median()
+# ref db open
+open_fit_db_mcc <- dat %>%
+  filter(
+    method == "open",
+    strategy == "database",
+    tool == "mothur"
+  ) %>%
+  pull(mcc_median) %>%
+  median()
 mcc_open_fit_db_vs_clust <- rel_diff(opticlust_mcc, open_fit_db_mcc)
 
-open_fit_gg_mcc <- dat %>% filter(method == 'open',
-                                  strategy == 'database',
-                                  tool == 'mothur',
[…]

### 2021-03-16 paper/paper.Rmd
Fix ref-db methods paragraph in results

@@
 - **TODO: current method for open/closed is vsearch against greengenes.**
+- **TODO: use word "map" for what vsearch does, "fit" for what optifit does.**
 - **TODO**: 2 categories of clustering: _de novo_ and reference based.
@@
 
-**TODO: make sure this info is baked into relevant sections, then remove.**
-For each dataset repeated with 100 random seeds, we generated OTUs with OptiFit
-using both strategies.
-To compare to existing software, we also clustered OTUs _de novo_ using
-OptiClust and VSEARCH, and with VSEARCH in reference-based mode against the
-Greengenes database.
-
 ### Reference clustering with public databases
 
-**TODO: use word "map" for what vsearch does, "fit" for what optifit does.**
-
-**TODO: ref clustering method paragraph here.**
-We assigned blah ref dbs to 3% otus with opticlust. _de novo_ clustering the Greengenes
-database, SILVA non-redundant database, or the Ribosomal Database Project (RDP)
-[@desantis_greengenes_2006; @quast_silva_2013; @cole_ribosomal_2014].
-All clustering was performed at a distance threshold of 0.03 and OTU quality was
-evaluated using the MCC as described previously [@westcott_opticlust_2017].
-To evaluate reference-based clustering with independent databases, we fit each
-dataset to reference OTUs generated by...
-repeated 100 times.
+We performed _de novo_ clustering with OptiClust at distance threshold of 4%
+using the Greengenes database, SILVA non-redundant database, or the Ribosomal
+Database Project (RDP) [@desantis_greengenes_2006; @quast_silva_2013;
+@cole_ribosomal_2014].
+We then performed refernce clustering with OptiFit, fitting each dataset to each 
+set of database OTUs repeated 100 times.
+We measured the algorithm run time and evaluated OTU quality using the MCC as
+described previously [@westcott_opticlust_2017].
+To compare to existing software, we also clustered each dataset into _de novo_
+OTUs  using OptiClust and VSEARCH, and with VSEARCH in reference-based mode
+against the Greengenes database.
 
@@
 OTU quality was similar with the simple and abundance-weighted sampling (median
-MCCs `r mcc_fit_split_simple` and `r mcc_fit_split_abun` respectively), but worse for similarity-weighted
-sampling with a median MCC of `r mcc_fit_split_dist`.
+MCCs `r mcc_fit_split_simple` and `r mcc_fit_split_abun` respectively), but
+worse for similarity-weighted sampling with a median MCC of `r mcc_fit_split_dist`.
 In closed reference mode, the fraction of query sequences that can be fit to the
-reference OTUs decreases as the reference fraction increases from 
-`r frac_fit_split_0.1`% of query sequences fit with `10`% of the dataset as the
-reference, to `r frac_fit_split_0.8`% of query sequences fit with `80`% of the
-dataset as the reference.
+reference OTUs increases as the reference fraction increases; from 
+`r frac_fit_split_0.1`% of query sequences fit when using 10% of the dataset as 
+the reference, to `r frac_fit_split_0.8`% of query sequences fit when using 80% 
+of the dataset as the reference.
 

### 2021-03-16 code/R/calc_results_stats.R

@@ -162,14 +162,14 @@ frac_fit_split_0.1 <- dat %>% filter(strategy == 'self-split',
                                      tool == 'mothur',
                                     method == 'closed',
                                      ref_frac == 0.1) %>% 
-  pull(frac_map_median) %>% median()
+  pull(frac_map_median) %>% median() * 100
 
 
 frac_fit_split_0.8 <- dat %>% filter(strategy == 'self-split', 
                                     tool == 'mothur',
                                     method == 'closed',
                                     ref_frac == 0.8) %>% 
-  pull(frac_map_median) %>% median()
+  pull(frac_map_median) %>% median() * 100
 
 
 # save results

### 2021-03-16 paper/paper.Rmd
Add TO-DOs from mtg w/ Pat

@@
 
-We performed _de novo_ clustering with OptiClust at distance threshold of 4%
+- TODO: start paragraph with question, end with why we should care. transitions to move the story along.
+- start:"In order to compare Optifit.."
+- end: "optifit performs better"
+- AND, BUT, THEREFORE, structure
+- order of importance: mcc, mapping, sec
+
+**TODO: lead with comparison of optifit vs current state of the art. give context at beginning.**
+We performed _de novo_ clustering with OptiClust at distance threshold of 3% (TODO R CODE HERE)
 using the Greengenes database, SILVA non-redundant database, or the Ribosomal
@@
 against the Greengenes database.
+
+In closed reference mode against databases, OptiFit produced lower quality OTUs
+than OptiClust by `r mcc_closed_fit_db_vs_clust` on average.
+Fitting sequences to Greengenes and SILVA in closed reference mode performed
+similarly with median MCC scores of `r closed_fit_gg_mcc` and 
+`r closed_fit_silva_mcc` respectively, while when fitting to RDP the median MCC 
+dropped to `r closed_fit_rdp_mcc`.
+An average of only `r frac_fit_db`% of query sequences were fit to reference
+database OTUs in closed reference mode across all dataset/database combinations.
+VSEARCH was able to map `r frac_vsearch_vs_fit`% more query sequences than
+OptiFit to the Greengenes reference database.
+In terms of run time, closed reference OptiFit outperformed OptiClust by 
+`r sec_closed_fit_db_vs_clust`% and VSEARCH by `r sec_closed_fit_db_vs_vsearch`%.
+
+since closed ref threw out a lot of stuff, we performed open ref.
 
@@
 
-In closed reference mode against databases, OptiFit produced lower quality OTUs
-than OptiClust by `r mcc_closed_fit_db_vs_clust` on average.
-Fitting sequences to Greengenes and SILVA in closed reference mode performed
-similarly with median MCC scores of `r closed_fit_gg_mcc` and 
-`r closed_fit_silva_mcc` respectively, while when fitting to RDP the median MCC 
-dropped to `r closed_fit_rdp_mcc`.
-An average of only `r frac_fit_db`% of query sequences were fit to reference
-database OTUs in closed reference mode across all dataset/database combinations.
-VSEARCH was able to map `r frac_vsearch_vs_fit`% more query sequences than
-OptiFit to the Greengenes reference database.
-In terms of run time, closed reference OptiFit outperformed OptiClust by 
-`r sec_closed_fit_db_vs_clust`% and VSEARCH by `r sec_closed_fit_db_vs_vsearch`%.
+### Reference clustering with split datasets
 
-### Reference clustering with split datasets
+- TODO; start with problem with current state of the art. 
+- tied to reference dbs taht may be unrelated to samples studying. can we use our own data as a reference?
+- lay out the use case and how this is a benefit over ref-db clustering.
+- we wanted ot use x% of data as ref, weren't sure what size to use and strategy for selecting.
+- important result: it works!
 
@@
 
+TODO; ref weight simple only here.
 OTU quality from the split dataset strategy with OptiFit was highly similar to
@@
 in open reference mode.
+TODO: fraction mapped.
 
+TODO: pick a fraction (e.g. 50%). this comparison would be supplemental if there's a figure.
 We also tested three methods for selecting the fraction of sequences to be used

### 2021-03-16 code/R/calc_results_stats.R
🎨 Style R code ⬡

@@ -10,167 +10,282 @@ coeff_var <- function(x) {
 }
 
 
-dat <- read_tsv(here('results', 'summarized.tsv'))
+dat <- read_tsv(here("results", "summarized.tsv"))
 
 # de novo
-opticlust_mcc <- dat %>% filter(method == 'de_novo', 
-                                tool == 'mothur') %>% 
-  pull(mcc_median) %>% median()
-opticlust_sec <- dat %>% filter(method == 'de_novo', 
-                                tool == 'mothur') %>% 
-  pull(sec_median) %>% median()
+opticlust_mcc <- dat %>%
+  filter(
+    method == "de_novo",
+    tool == "mothur"
+  ) %>%
+  pull(mcc_median) %>%
+  median()
+opticlust_sec <- dat %>%
+  filter(
+    method == "de_novo",
+    tool == "mothur"
+  ) %>%
+  pull(sec_median) %>%
+  median()
 dn_vsearch_mcc <-
-  dat %>% filter(strategy == 'de_novo', tool == 'vsearch') %>% 
-  pull(mcc_median) %>% median()
+  dat %>%
+  filter(strategy == "de_novo", tool == "vsearch") %>%
+  pull(mcc_median) %>%
+  median()
 dn_vsearch_sec <-
-  dat %>% filter(strategy == 'de_novo', tool == 'vsearch') %>% 
-  pull(sec_median) %>% median()
+  dat %>%
+  filter(strategy == "de_novo", tool == "vsearch") %>%
+  pull(sec_median) %>%
+  median()
 mcc_opticlust_vs_vsearch <- rel_diff(opticlust_mcc, dn_vsearch_mcc)
 sec_opticlust_vs_vsearch <- abs(rel_diff(dn_vsearch_sec, opticlust_sec))
 
-# ref db open 
-open_fit_db_mcc <- dat %>% filter(method == 'open', 
-                               strategy == 'database', 
-                               tool == 'mothur') %>% 
-  pull(mcc_median) %>% median()
+# ref db open
+open_fit_db_mcc <- dat %>%
+  filter(
+    method == "open",
+    strategy == "database",
+    tool == "mothur"
+  ) %>%
+  pull(mcc_median) %>%
+  median()
 mcc_open_fit_db_vs_clust <- rel_diff(opticlust_mcc, open_fit_db_mcc)
 
-open_fit_gg_mcc <- dat %>% filter(method == 'open',
-                                  strategy == 'database',
-                                  tool == 'mothur',
[…]

### 2021-03-19 paper/paper.Rmd
Update algorithm description

@@
 - **TODO: ask Sarah Westcott to check the accuracy of this description**.
-- **TODO: toy example like opticlust with figure**.
 
@@
 method for reference-based clustering.
-**TODO: more narrative, less documentation-like.**
-**TODO: extension of opticlust except ref otus are fixed. just like opticlust only query seqs can move otus.**
-OptiFit takes as input a list of reference OTUs and their sequences, a list of
-query sequences to assign to the reference OTUs, the sequence pairs that are
-within the distance threshold (e.g. 0.03), and the metric to assess clustering
-quality (default: MCC).
+OptiFit begins where OptiClust ends, starting with a list of reference OTUs and
+their sequences, a list of query sequences to assign to the reference OTUs, and
+the sequence pairs that are within the distance threshold (e.g. 0.03).
 Query sequences are randomly seeded in reference OTUs, then for each sequence
-the algorithm calculates the quality metric based on whether the sequence stays
-in its current OTU, moves to each of the other OTUs, or is discarded.
+the algorithm calculates the quality metric (default: MCC) based on whether the
+sequence stays in its current OTU, moves to each of the other OTUs, or is
+discarded.
 If two or more OTU assignments are of equal quality, a random number generator
@@
 
-- TODO; start with problem with current state of the art. 
+- **TODO**; start with problem with current state of the art. 
 - tied to reference dbs taht may be unrelated to samples studying. can we use our own data as a reference?
@@
 
-TODO; ref weight simple only here.
+**TODO; ref weight simple only here.**
 OTU quality from the split dataset strategy with OptiFit was highly similar to
@@
 
-TODO: pick a fraction (e.g. 50%). this comparison would be supplemental if there's a figure.
+**TODO: pick a fraction (e.g. 50%). this part is less important. figure would be supplemental.**
 We also tested three methods for selecting the fraction of sequences to be used

### 2021-03-22 paper/paper.Rmd
Edit algorithm description w/ feedback from SLW (@mothur-westcott)

@@
 
-- **TODO: ask Sarah Westcott to check the accuracy of this description**.
-
 OptiFit leverages the method employed by OptiClust of iteratively assigning
@@
 the sequence pairs that are within the distance threshold (e.g. 0.03).
-Query sequences are randomly seeded in reference OTUs, then for each sequence
-the algorithm calculates the quality metric (default: MCC) based on whether the
-sequence stays in its current OTU, moves to each of the other OTUs, or is
-discarded.
-If two or more OTU assignments are of equal quality, a random number generator
-is used to break the tie.
-This process is repeated until the quality metric stabilizes, changing by no
-more than 0.0001 by default, or until a maximum number of iterations is reached
-(default: 100).
-In closed-reference mode, any query sequences that cannot be assigned are thrown
-out (**TODO: exactly what determines whether a seq can't be assigned?**), and
-the results only contain OTUs that exist in the original reference.
+Initially, query sequences are placed in singleton OTUs. 
+Then, the algorithm iteratively reassigns the query sequences to the reference 
+OTUs to optimize the Matthews correlation coefficient (MCC). 
+A sequence will remain unassigned if the MCC value is maximized when the
+sequence is a singleton rather than assigned to a reference OTU.
+This process is repeated until the MCC stabilizes, changing by no more than
+0.0001 by default, or until a maximum number of iterations is reached (default:
+100).
+In closed-reference mode, any query sequences that cannot be assigned to
+references OTUs are discarded, and the results will only contain OTUs that exist
+in the original reference.
 In open-reference mode, unassigned query sequences are clustered _de novo_ using
-OptiClust to generate additional OTUs. The final quality score is reported with
-the best OTU assignments.
+OptiClust to generate additional OTUs. 
+The final quality score is reported with the best OTU assignments.
 There are two strategies for generating OTUs with OptiFit: 1) fit sequences to
@@
 - AND, BUT, THEREFORE, structure
-- order of importance: mcc, mapping, sec
+- order of importance: mcc, mapping, sec (1 paragraph each, in this order)
 
 **TODO: lead with comparison of optifit vs current state of the art. give context at beginning.**
-We performed _de novo_ clustering with OptiClust at distance threshold of 3% (TODO R CODE HERE)
+We performed _de novo_ clustering with OptiClust at distance threshold of 3%
 using the Greengenes database, SILVA non-redundant database, or the Ribosomal

### 2021-03-22 paper/paper.Rmd
Give context for ref clustering against DBs

@@
 
-**TODO: lead with comparison of optifit vs current state of the art. give context at beginning.**
-We performed _de novo_ clustering with OptiClust at distance threshold of 3%
-using the Greengenes database, SILVA non-redundant database, or the Ribosomal
-Database Project (RDP) [@desantis_greengenes_2006; @quast_silva_2013;
-@cole_ribosomal_2014].
-We then performed refernce clustering with OptiFit, fitting each dataset to each 
-set of database OTUs repeated 100 times.
-We measured the algorithm run time and evaluated OTU quality using the MCC as
-described previously [@westcott_opticlust_2017].
+While _de novo_ clustering produces high quality OTUs, researchers may prefer to
+perform reference clustering to a public database because reference-based
+methods produce stable OTUs and are generally faster than _de novo_ methods.
+To test how OptiFit performs for this purpose, we fit each dataset to three
+databases of reference OTUs; the Greengenes database, the SILVA non-redundant
+database, and the Ribosomal Database Project (RDP) [@desantis_greengenes_2006;
+@quast_silva_2013; @cole_ribosomal_2014].
+Reference OTUs for each database were created by performing _de novo_ clustering
+with OptiClust at a distance threshold of 3%.
 To compare to existing software, we also clustered each dataset into _de novo_
 OTUs using OptiClust and VSEARCH, and with VSEARCH in reference-based mode
-against the Greengenes database.
+against the Greengenes database of OTUs clustered at 97% similarity.
 

### 2021-03-24 exploratory/2021/test_criteria.Rmd
Plot mcc & fraction mapped from criteria test

---
date: '`r Sys.Date()`'
output: 
  github_document:
    html_preview: false
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      fig.path = "figures/"
                      )
```

```{r lib}
set.seed(2018)
library(cowplot)
library(ggtext)
library(glue)
library(here)
library(knitr)
library(tidyverse)

theme_set(theme_bw())

mutate_perf <- function(dat) {
  dat %>% 
    mutate(mem_mb = max_rss,
           mem_gb = mem_mb / 1024) %>% 
    rename(sec = s)
}
```

```{r read}
dat <- read_tsv(here('subworkflows', '3_fit_sample_split', 'results', 
                     'optifit_crit_check.tsv')) %>% 
  mutate_perf() %>% 
  select(dataset, printref, criteria, mcc, fraction_mapped, sec, mem_gb)
```

## OTU Quality

```{r crit_mcc}
dat %>% 
  ggplot(aes(x = criteria, y = mcc, color = printref)) +
  geom_point(alpha = 0.5) +
  ylim(0, 1)
```
## Fraction mapped

```{r crit_map}
dat %>% 
  ggplot(aes(x = criteria, y = fraction_mapped, color = printref)) +
  geom_point(alpha = 0.5) +
  ylim(0, 1)
```

### 2021-03-24 exploratory/2021/test_criteria.Rmd
Plot criteria results w/ 20 seeds

@@
 
+Sarah W's description of the criteria parameter:
+
+"We have a criteria parameter that indicates when the reads should be moved. We
+calculated two mcc coefficients. One for the query reads (ignoring the
+references) and one for the query reads and reference OTUs. The screen output
+shows this as “combo” and “fit” lines. By default we move a read if either
+metric improves. The criteria parameter lets you indicate you only want to move
+if the one or the other improves. Options are “both” (default), “combo” or
+“fit”. In practice we found “both” gave the best overall results."
+
+I ran the 3 criteria options with the human dataset at 50/50 ref/query split
+chosen by simple random sample, method=closed, repeated for 20 seeds.
+
 ## OTU Quality
@@
   ggplot(aes(x = criteria, y = mcc, color = printref)) +
-  geom_point(alpha = 0.5) +
-  ylim(0, 1)
+  geom_jitter(alpha = 0.5, width = 0.2) 
 ```
+
 ## Fraction mapped
@@
   ggplot(aes(x = criteria, y = fraction_mapped, color = printref)) +
-  geom_point(alpha = 0.5) +
-  ylim(0, 1)
+  geom_jitter(alpha = 0.5, width = 0.2)
 ```
+
+## Runtime
+
+```{r crit_runtime}
+dat %>% 
+  ggplot(aes(x = criteria, y = sec, color = printref)) +
+  geom_jitter(alpha = 0.5, width = 0.2)
+```

### 2021-03-24 exploratory/2021/test_criteria.Rmd
Describe criteria options

@@
 
+My interpretation of the criteria parameter options:
+
+- `fit` = "query only mcc" - Move query sequence `X` to OTU `Y` if it improves the mcc calculated using **only the query seqs**.
+- `combo` = "query & ref mcc" - Move query sequence `X` to OTU `Y` if it improves the mcc calculated using **the query & reference seqs**.
+- `both` = "either" - Move query sequence `X` to OTU `Y` if **either** the `fit` mcc ("query only") or the `combo` mcc ("query & ref") score improves.
+
 I ran the 3 criteria options with the human dataset at 50/50 ref/query split

### 2021-03-24 exploratory/2021/test_criteria.Rmd
Formatting

--- 2021-03-24 exploratory/2021/test_criteria.Rmd
@@
 
-"We have a criteria parameter that indicates when the reads should be moved. We
+> We have a criteria parameter that indicates when the reads should be moved. We
 calculated two mcc coefficients. One for the query reads (ignoring the
@@
 if the one or the other improves. Options are “both” (default), “combo” or
-“fit”. In practice we found “both” gave the best overall results."
+“fit”. In practice we found “both” gave the best overall results.
+
+My interpretation of the criteria parameter options:
+
+- `fit` = "query only mcc" - Move query sequence X to OTU Y if it improves the mcc calculated using **only the query seqs**.
+- `combo` = "query & ref mcc" - Move query sequence X to OTU Y if it improves the mcc calculated using **the query & reference seqs**.
+- `both` = "either" - Move query sequence X to OTU Y if **either** the `fit` mcc ("query only") or the `combo` mcc ("query & ref") score improves.
 

### 2021-03-24 exploratory/2021/test_criteria.Rmd
Select seed col

@@
   mutate_perf() %>% 
-  select(dataset, printref, criteria, mcc, fraction_mapped, sec, mem_gb) 
+  select(dataset, printref, criteria, mcc, fraction_mapped, sec, mem_gb, seed) 
 ```

### 2021-03-30 exploratory/2021/plots.Rmd
Update w/ new opticlust; virtually the same

@@
            mem_gb = mem_mb / 1024) %>% 
-    rename(sec = s,
-           num_otus = sobs)
+    rename(sec = s)
 }

### 2021-04-06 paper/paper.Rmd
Restructure ref-db section

@@
 ```
+
+- **start paragraph with question, end with why we should care. transitions to move the story along.**
+- **AND, BUT, THEREFORE, structure**
 
@@
 
-- TODO: start paragraph with question, end with why we should care. transitions to move the story along.
-- start:"In order to compare Optifit.."
-- end: "optifit performs better"
-- AND, BUT, THEREFORE, structure
-- order of importance: mcc, mapping, sec (1 paragraph each, in this order)
-
 While _de novo_ clustering produces high quality OTUs, researchers may prefer to
@@
 methods produce stable OTUs and are generally faster than _de novo_ methods.
+In closed reference mode, sequences that cannot be assigned to reference OTUs
+are thrown out, so that the final clustering contains only OTUs that exist in
+the reference.
 To test how OptiFit performs for this purpose, we fit each dataset to three
@@
 with OptiClust at a distance threshold of 3%.
-To compare to existing software, we also clustered each dataset into _de novo_
-OTUs using OptiClust and VSEARCH, and with VSEARCH in reference-based mode
-against the Greengenes database of OTUs clustered at 97% similarity.
-
-In closed reference mode against databases, OptiFit produced lower quality OTUs
-than OptiClust by `r mcc_closed_fit_db_vs_clust` on average.
 Fitting sequences to Greengenes and SILVA in closed reference mode performed
-similarly with median MCC scores of `r closed_fit_gg_mcc` and 
+similarly, with median MCC scores of `r closed_fit_gg_mcc` and 
 `r closed_fit_silva_mcc` respectively, while when fitting to RDP the median MCC 
 dropped to `r closed_fit_rdp_mcc`.
-An average of only `r frac_fit_db`% of query sequences were fit to reference
-database OTUs in closed reference mode across all dataset/database combinations.
+For comparison to the state of the art, clustering datasets _de novo_ with
+OptiClust produced an average MCC score of `r opticlust_mcc`.
+This gap in OTU quality mostly disappears when clustering in open reference
+mode, which produced median MCCs of `r open_fit_gg_mcc` with greengenes, 
+`r open_fit_silva_mcc` with Silva, and `r open_fit_rdp_mcc` with RDP.
+Thus, open reference OptiFit produces OTUs of very similar quality as _de novo_
+clustering, and closed reference OptiFit follows closely behind as long as a
+suitable reference database is chosen.
+
+Since closed reference clustering throws out query sequences that cannot be
+assigned to reference OTUs, an additional measure of clustering performance to
+consider is the fraction of query sequences that were able to be assigned.
+On average, more sequences were assigned with Greengenes as the reference 
+(`r frac_fit_gg`%) than with Silva (`r frac_fit_silva`%) or RDP (`r frac_fit_rdp`%).
+This mirrors the result reported above that Greengenes produced better OTUs in
+terms of MCC score than either Silva or RDP.
+Note that _De novo_ and open reference clustering methods always assign 100% of
+sequences to OTUs.
+The database chosen affects the final OTU assignments considerably in terms of 
+both MCC score and fraction of query sequences fit.
+
+Despite the drawbacks, closed reference methods have been used when fast
+execution speed is required such as when using very large datasets.
+To compare performance in terms of speed, we repeated each OptiFit and OptiClust
+run 100 times and measured the execution time.
+Closed reference OptiFit outperformed OptiClust by 
+`r sec_closed_fit_db_vs_clust`% and VSEARCH by `r sec_closed_fit_db_vs_vsearch`%.
+In open reference mode, OptiFit ran faster than VSEARCH by 
+`r sec_vsearch_vs_open_fit_db`%, but slower than OptiClust by 
+`r sec_opticlust_vs_open_fit_db`%.
+
+To compare to the reference clustering method used by QIIME2, we clustered each
+dataset with VSEARCH against the Greengenes database of OTUs previously 
+clustered at 97% sequence similarity.
+**TODO note vsearch otu definition vs. ours.**
+_De novo_ clustering with OptiClust produced `r mcc_opticlust_vs_vsearch`%
+higher quality OTUs than VSEARCH, but performed `r sec_opticlust_vs_vsearch`% 
+slower than VSEARCH.
 VSEARCH was able to map `r frac_vsearch_vs_fit`% more query sequences than
-OptiFit to the Greengenes reference database.
-In terms of run time, closed reference OptiFit outperformed OptiClust by 
-`r sec_closed_fit_db_vs_clust`% and VSEARCH by `r sec_closed_fit_db_vs_vsearch`%.
-
-since closed ref threw out a lot of stuff, we performed open ref.
-
-In open reference mode, fitting the datasets to reference OTUs with OptiFit
-produced OTUs of similar quality (`r mcc_open_fit_db_vs_clust`% difference in median
-MCC) as clustering the datasets _de novo_ with OptiClust  across all datasets
-and reference databases.
+OptiFit to the Greengenes reference database in closed reference mode.
 OptiFit produced higher quality OTUs than VSEARCH when open reference clustering
@@
 and `r open_vsearch_mcc` (respectively).
-OptiFit ran faster than VSEARCH by `r sec_vsearch_vs_open_fit_db`%  in open
-reference mode, but slower than OptiClust by `r sec_opticlust_vs_open_fit_db`%.
-_De novo_ clustering with OptiClust produced `r mcc_opticlust_vs_vsearch`%
-higher quality OTUs than VSEARCH, but performed `r sec_opticlust_vs_vsearch`% 
-slower than VSEARCH.
 

### 2021-04-06 code/R/calc_results_stats.R

@@ -63,6 +63,27 @@ open_fit_gg_mcc <- dat %>%
   ) %>%
   pull(mcc_median) %>%
   median()
+
+open_fit_silva_mcc <- dat %>%
+  filter(
+    method == "open",
+    strategy == "database",
+    tool == "mothur",
+    ref == "silva"
+  ) %>%
+  pull(mcc_median) %>%
+  median()
+
+open_fit_rdp_mcc <- dat %>%
+  filter(
+    method == "open",
+    strategy == "database",
+    tool == "mothur",
+    ref == "rdp"
+  ) %>%
+  pull(mcc_median) %>%
+  median()
+
 open_vsearch_mcc <- dat %>%
   filter(
     method == "open",

### 2021-04-06 paper/paper.Rmd
Edit vsearch paragraph

@@
 clustered at 97% sequence similarity.
-**TODO note vsearch otu definition vs. ours.**
+Each reference OTU from the Greengenes 97% database contains one reference
+sequence, and VSEARCH maps sequences to the reference based on each individual
+sequence's similarity to the single reference OTU.
+In contrast, OptiFit accepts reference OTUs which each contain multiple
+sequences, and the sequence similarity between all query and reference sequences
+is considered when assign sequences to OTUs.
 _De novo_ clustering with OptiClust produced `r mcc_opticlust_vs_vsearch`%
@@
 slower than VSEARCH.
-VSEARCH was able to map `r frac_vsearch_vs_fit`% more query sequences than
-OptiFit to the Greengenes reference database in closed reference mode.
-OptiFit produced higher quality OTUs than VSEARCH when open reference clustering
-against the Greengenes database, with median MCC scores of `r open_fit_gg_mcc`
-and `r open_vsearch_mcc` (respectively).
+In closed reference mode, VSEARCH was able to map `r frac_vsearch_vs_fit`% more
+query sequences than OptiFit to the Greengenes reference database.
+When open reference clustering, OptiFit produced higher quality OTUs than
+VSEARCH against the Greengenes database, with median MCC scores of 
+`r open_fit_gg_mcc` and `r open_vsearch_mcc` (respectively).
+In terms of run time, OptiFit outperformed VSEARCH in both closed and open
+reference mode by `r sec_closed_fit_db_vs_vsearch`% and 
+`r sec_vsearch_vs_open_fit_db`% on average respectively.
+**TODO:** The stark difference in OTU definitions between mothur and VSEARCH resulted in...
 
@@
 
-- **TODO**; start with problem with current state of the art. 
-- tied to reference dbs taht may be unrelated to samples studying. can we use our own data as a reference?
-- lay out the use case and how this is a benefit over ref-db clustering.
 - we wanted ot use x% of data as ref, weren't sure what size to use and strategy for selecting.
 - important result: it works!
+
+When performing reference clustering against public databases, the database 
+chosen greatly affects the quality of OTUs produced.
+OTUs may be of poor quality when the reference database is too unrelated to the 
+samples of interest, such as when samples contain new and rapidly evolving microbes.
+For this reason, we were interested in whether we could use the samples themselves
+as the reference to cluster against.
 

### 2021-04-06 paper/paper.Rmd
Conclude ref-db runtime paragraph

@@
 run 100 times and measured the execution time.
-Closed reference OptiFit outperformed OptiClust by 
-`r sec_closed_fit_db_vs_clust`% and VSEARCH by `r sec_closed_fit_db_vs_vsearch`%.
-In open reference mode, OptiFit ran faster than VSEARCH by 
-`r sec_vsearch_vs_open_fit_db`%, but slower than OptiClust by 
-`r sec_opticlust_vs_open_fit_db`%.
+Closed reference OptiFit outperformed both OptiClust and open reference OptiFit,
+with respective average run times of `r closed_fit_db_sec`, `r opticlust_sec`,
+and `r open_fit_db_sec` seconds.
+Thus, the OptiFit algorithm continues the precedent that closed reference
+clustering sacrifices OTU quality for execution speed.
 

### 2021-04-07 paper/paper.Rmd
Better structure for split results

@@
 `r sec_vsearch_vs_open_fit_db`% on average respectively.
-**TODO:** The stark difference in OTU definitions between mothur and VSEARCH resulted in...
+**TODO: conclude:** The stark difference in OTU definitions between mothur and VSEARCH resulted in...
 
@@
 
-- we wanted ot use x% of data as ref, weren't sure what size to use and strategy for selecting.
+- we wanted to use x% of data as ref, weren't sure what size to use and strategy for selecting.
 - important result: it works!
@@
 chosen greatly affects the quality of OTUs produced.
-OTUs may be of poor quality when the reference database is too unrelated to the 
-samples of interest, such as when samples contain new and rapidly evolving microbes.
-For this reason, we were interested in whether we could use the samples themselves
-as the reference to cluster against.
-
-A split dataset strategy was employed to assess how well OptiFit performs for
-tasks where new sequences are added to existing OTUs, such as when comparing
-OTUs across studies or making predictions with machine learning models.
-Datasets were randomly split into a reference fraction and a query fraction.
-Reference sizes from 10% to 80% of the sequences were created, with the
-remaining sequences used for the query.
+OTUs quality may be poor when the reference database is too unrelated to the
+samples of interest, such as when samples contain low abundant or rapidly
+evolving microbes.
+While _de novo_ clustering overcomes the quality limitations of reference
+clustering to databases, OTU assignments are not stable when new sequences are
+added.
+Researchers may wish to fit new sequences to existing OTUs when comparing OTUs
+across studies or when making predictions with machine learning models.
+To determine how well OptiFit performs for fitting new sequences to existing
+OTUs, we employed a split dataset strategy, where each dataset was randomly
+split into a reference fraction and a query fraction.
 Reference sequences were clustered _de novo_ with OptiClust, then query
@@
 
-**TODO; ref weight simple only here.**
+**TODO; ref weight simple only here. First we wanted to know; does this strategy perform as well as de novo with 50% split?**
 OTU quality from the split dataset strategy with OptiFit was highly similar to
 that from _de novo_ clustering the whole dataset with OptiClust regardless of
-mode (`r mcc_opticlust_vs_fit_split`% difference in median MCC).
+mode (`r mcc_opticlust_vs_fit_split_simple`% difference in median MCC).
 OTU quality was remarkably stable across reference fraction sizes within
-parameter sets (Fig. **XX**).
+parameter sets.
 For example, splitting the human dataset 100 times yielded a coefficient of
@@
 the database strategy in closed reference mode and `r sec_open_fit_split_vs_db`%
-in open reference mode.
-TODO: fraction mapped.
+faster in open reference mode.
+**TODO: fraction mapped.**
 
+**Then we wanted to know; what fraction of sequences should be in the reference?**
+To test the best reference size, reference sizes from 10% to 80% of the sequences were created, with the remaining sequences used for the query.
+
+**Finally, we wanted to know the best way to select the reference sequences.**
 **TODO: pick a fraction (e.g. 50%). this part is less important. figure would be supplemental.**

### 2021-04-07 code/R/calc_results_stats.R

@@ -223,31 +223,28 @@ sec_closed_fit_db_vs_vsearch <- rel_diff(closed_fit_db_sec, closed_vsearch_sec) 
 
 ################################################################################
 # fit split
-fit_split_mcc <- dat %>%
-  filter(
-    strategy == "self-split",
-    tool == "mothur"
-  ) %>%
-  pull(mcc_median) %>%
-  median()
-mcc_opticlust_vs_fit_split <- rel_diff(opticlust_mcc, fit_split_mcc)
-
 cv_fit_split_mcc <- coeff_var(dat %>% filter(
   strategy == "self-split",
-  tool == "mothur"
+  tool == "mothur",
+  ref_weight == 'simple',
+  ref_frac == 0.5
 ) %>%
   pull(mcc_median))
 cv_fit_split_mcc_human <- coeff_var(dat %>% filter(
   strategy == "self-split",
   tool == "mothur",
-  dataset == "human"
+  dataset == "human",
+  ref_weight == 'simple',
+  ref_frac == 0.5
 ) %>%
   pull(mcc_median))
 closed_fit_split_sec <- dat %>%
   filter(
     method == "closed",
     strategy == "self-split",
-    tool == "mothur"
+    tool == "mothur",
+    ref_weight == 'simple',
+    ref_frac == 0.5
   ) %>%
   pull(sec_median) %>%
   median()
@@ -256,7 +253,9 @@ open_fit_split_sec <- dat %>%
   filter(
     method == "open",
     strategy == "self-split",
-    tool == "mothur"
+    tool == "mothur",
+    ref_weight == 'simple',
+    ref_frac == 0.5
   ) %>%
   pull(sec_median) %>%
   median()
@@ -268,16 +267,19 @@ mcc_fit_split_simple <- dat %>%
   filter(
     strategy == "self-split",
     tool == "mothur",
-    ref_weight == "simple"
+    ref_weight == "simple",
+    ref_frac == 0.5
   ) %>%
   pull(mcc_median) %>%
   median()
+mcc_opticlust_vs_fit_split_simple <- rel_diff(opticlust_mcc, mcc_fit_split_simple)
 
 mcc_fit_split_abun <- dat %>%
   filter(
     strategy == "self-split",
     tool == "mothur",
-    ref_weight == "abundance"
+    ref_weight == "abundance",
+    ref_frac == 0.5
   ) %>%
   pull(mcc_median) %>%
   median()
[…]

### 2021-04-07 code/R/calc_results_stats.R
🎨  Style R code

@@ -226,7 +226,7 @@ sec_closed_fit_db_vs_vsearch <- rel_diff(closed_fit_db_sec, closed_vsearch_sec) 
 cv_fit_split_mcc <- coeff_var(dat %>% filter(
   strategy == "self-split",
   tool == "mothur",
-  ref_weight == 'simple',
+  ref_weight == "simple",
   ref_frac == 0.5
 ) %>%
   pull(mcc_median))
@@ -234,7 +234,7 @@ cv_fit_split_mcc_human <- coeff_var(dat %>% filter(
   strategy == "self-split",
   tool == "mothur",
   dataset == "human",
-  ref_weight == 'simple',
+  ref_weight == "simple",
   ref_frac == 0.5
 ) %>%
   pull(mcc_median))
@@ -243,7 +243,7 @@ closed_fit_split_sec <- dat %>%
     method == "closed",
     strategy == "self-split",
     tool == "mothur",
-    ref_weight == 'simple',
+    ref_weight == "simple",
     ref_frac == 0.5
   ) %>%
   pull(sec_median) %>%
@@ -254,7 +254,7 @@ open_fit_split_sec <- dat %>%
     method == "open",
     strategy == "self-split",
     tool == "mothur",
-    ref_weight == 'simple',
+    ref_weight == "simple",
     ref_frac == 0.5
   ) %>%
   pull(sec_median) %>%

### 2021-04-08 paper/paper.Rmd
Improve split dataset results

@@
 
-- we wanted to use x% of data as ref, weren't sure what size to use and strategy for selecting.
-- important result: it works!
-
 When performing reference clustering against public databases, the database 
@@
 
-**TODO; ref weight simple only here. First we wanted to know; does this strategy perform as well as de novo with 50% split?**
-OTU quality from the split dataset strategy with OptiFit was highly similar to
-that from _de novo_ clustering the whole dataset with OptiClust regardless of
+First, we tested whether OptiFit performed as well as _de novo_ clustering when
+using the split dataset strategy with half of the sequences selected for
+the reference by a simple random sample.
+OTU quality was highly similar to that from OptiClust regardless of
 mode (`r mcc_opticlust_vs_fit_split_simple`% difference in median MCC).
-OTU quality was remarkably stable across reference fraction sizes within
-parameter sets.
-For example, splitting the human dataset 100 times yielded a coefficient of
-variation of `r cv_fit_split_mcc_human` for the MCC score.
+In closed reference mode, OptiFit was able to fit `r frac_fit_split`% of query
+sequences to reference OTUs with the split strategy, a great improvement over
+the average `r frac_fit_gg`% of sequences fit to the greengenes database.
 In terms of runtime, closed and open reference OptiFit performed faster than
@@
 `r sec_open_fit_split_vs_clust` respectively.
-The split dataset strategy performed `r sec_closed_fit_split_vs_db`% faster than
-the database strategy in closed reference mode and `r sec_open_fit_split_vs_db`%
-faster in open reference mode.
-**TODO: fraction mapped.**
+The split dataset strategy also performed `r sec_closed_fit_split_vs_db`% faster
+than the database strategy in closed reference mode and 
+`r sec_open_fit_split_vs_db`% faster in open reference mode.
+Thus, reference clustering with the split dataset strategy creates as high
+quality OTUs as _de novo_ clustering yet at a faster run time, and fits far more
+query sequences than the database strategy.
 
 **Then we wanted to know; what fraction of sequences should be in the reference?**
-To test the best reference size, reference sizes from 10% to 80% of the sequences were created, with the remaining sequences used for the query.
+To test the best reference size, reference sizes from 10% to 80% of the
+sequences were created, with the remaining sequences used for the query.
+OTU quality was remarkably stable across reference fraction sizes.
+For example, splitting the human dataset 100 times yielded a coefficient of
+variation of `r cv_fit_split_mcc_human` for the MCC score across all fractions. **TODO: revisit how to report this**
 

### 2021-04-08 code/R/calc_results_stats.R

@@ -230,14 +230,18 @@ cv_fit_split_mcc <- coeff_var(dat %>% filter(
   ref_frac == 0.5
 ) %>%
   pull(mcc_median))
-cv_fit_split_mcc_human <- coeff_var(dat %>% filter(
-  strategy == "self-split",
-  tool == "mothur",
-  dataset == "human",
-  ref_weight == 'simple',
-  ref_frac == 0.5
-) %>%
-  pull(mcc_median))
+
+frac_fit_split <- dat %>%
+  filter(
+    strategy == "self-split",
+    tool == "mothur",
+    ref_weight == 'simple',
+    method == "closed",
+    ref_frac == 0.5
+  ) %>%
+  pull(frac_map_median) %>%
+  median() * 100
+
 closed_fit_split_sec <- dat %>%
   filter(
     method == "closed",
@@ -248,7 +252,6 @@ closed_fit_split_sec <- dat %>%
   ) %>%
   pull(sec_median) %>%
   median()
-sec_closed_fit_split_vs_clust <- rel_diff(closed_fit_split_sec, opticlust_sec) %>% abs()
 open_fit_split_sec <- dat %>%
   filter(
     method == "open",
@@ -259,9 +262,19 @@ open_fit_split_sec <- dat %>%
   ) %>%
   pull(sec_median) %>%
   median()
+
+sec_closed_fit_split_vs_clust <- rel_diff(closed_fit_split_sec, opticlust_sec) %>% abs()
 sec_open_fit_split_vs_clust <- rel_diff(open_fit_split_sec, opticlust_sec) %>% abs()
 sec_open_fit_split_vs_db <- rel_diff(open_fit_split_sec, open_fit_db_sec) %>% abs()
 sec_closed_fit_split_vs_db <- rel_diff(closed_fit_split_sec, closed_fit_db_sec) %>% abs()
+
+cv_fit_split_mcc_human <- coeff_var(dat %>% filter(
+  strategy == "self-split",
+  tool == "mothur",
+  dataset == "human",
+  ref_weight == 'simple'
+) %>%
+  pull(mcc_median))
 
 mcc_fit_split_simple <- dat %>%
   filter(
@@ -303,7 +316,6 @@ frac_fit_split_0.1 <- dat %>%
   ) %>%
   pull(frac_map_median) %>%
   median() * 100
-
 
 frac_fit_split_0.8 <- dat %>%
   filter(

### 2021-04-08 code/R/calc_results_stats.R
🎨  Style R code

@@ -226,7 +226,7 @@ sec_closed_fit_db_vs_vsearch <- rel_diff(closed_fit_db_sec, closed_vsearch_sec) 
 cv_fit_split_mcc <- coeff_var(dat %>% filter(
   strategy == "self-split",
   tool == "mothur",
-  ref_weight == 'simple',
+  ref_weight == "simple",
   ref_frac == 0.5
 ) %>%
   pull(mcc_median))
@@ -235,7 +235,7 @@ frac_fit_split <- dat %>%
   filter(
     strategy == "self-split",
     tool == "mothur",
-    ref_weight == 'simple',
+    ref_weight == "simple",
     method == "closed",
     ref_frac == 0.5
   ) %>%
@@ -247,7 +247,7 @@ closed_fit_split_sec <- dat %>%
     method == "closed",
     strategy == "self-split",
     tool == "mothur",
-    ref_weight == 'simple',
+    ref_weight == "simple",
     ref_frac == 0.5
   ) %>%
   pull(sec_median) %>%
@@ -257,7 +257,7 @@ open_fit_split_sec <- dat %>%
     method == "open",
     strategy == "self-split",
     tool == "mothur",
-    ref_weight == 'simple',
+    ref_weight == "simple",
     ref_frac == 0.5
   ) %>%
   pull(sec_median) %>%
@@ -272,7 +272,7 @@ cv_fit_split_mcc_human <- coeff_var(dat %>% filter(
   strategy == "self-split",
   tool == "mothur",
   dataset == "human",
-  ref_weight == 'simple'
+  ref_weight == "simple"
 ) %>%
   pull(mcc_median))
 

### 2021-04-14 paper/paper.Rmd
Incorporate Pat's minor edits to results

@@
 
-- **start paragraph with question, end with why we should care. transitions to move the story along.**
-- **AND, BUT, THEREFORE, structure**
+- **"AND, BUT, THEREFORE" structure. start paragraph with question, end with why we should care. transitions to move the story along.**
+- From Pat: "briefly looking through the Discussion and Intro, one point that we may have forgotten is that a benefit of our approach is that it is much easier to customize to a specific region.
+The greengenes reference OTUs are based on full length sequences.
+This causes problems when considering shorter (e.g.
+V4) sequences since reference OTUs may be more similar  than the full length and even identical to each other in the subregion.
+Because we can easily spin up a subregion specific set of reference OTUs from a public database or the reference fraction this isn't a problem.
+This is described in one of my earlier papers looking at open/closed reference clustering and was part of the reason that the order of the database was important."
 
@@
 method for reference-based clustering.
+**TODO: brief description of the opticlust algorithm.**
 OptiFit begins where OptiClust ends, starting with a list of reference OTUs and
@@
 OTUs to optimize the Matthews correlation coefficient (MCC). 
-A sequence will remain unassigned if the MCC value is maximized when the
-sequence is a singleton rather than assigned to a reference OTU.
-This process is repeated until the MCC stabilizes, changing by no more than
-0.0001 by default, or until a maximum number of iterations is reached (default:
-100).
-In closed-reference mode, any query sequences that cannot be assigned to
+Alternatively, a sequence will remain unassigned if the MCC value is maximized
+when the sequence is a singleton rather than assigned to a reference OTU.
+This process is repeated until the MCC changes by no more than 0.0001 (default)
+or until a maximum number of iterations is reached (default: 100).
+In the closed reference mode, any query sequences that cannot be assigned to
 references OTUs are discarded, and the results will only contain OTUs that exist
 in the original reference.
-In open-reference mode, unassigned query sequences are clustered _de novo_ using
-OptiClust to generate additional OTUs. 
-The final quality score is reported with the best OTU assignments.
-There are two strategies for generating OTUs with OptiFit: 1) fit sequences to
-reference OTUs of an independent database, or 2) split the dataset into a
-reference and query fraction, cluster the reference sequences _de novo_, then
-fit the query sequences to the reference OTUs.
+In the open reference mode, unassigned query sequences are clustered _de novo_
+using OptiClust to generate new OTUs.
+The final MCC is reported with the best OTU assignments.
+There are two strategies for generating OTUs with OptiFit: 
+1) fit query sequences to reference OTUs generated by _de novo_ clustering an
+independent database, or
+2) split the dataset into a reference and query fraction, cluster the reference
+sequences _de novo_, then fit the query sequences to the reference OTUs.
+**TODO: describe data sets here**
 
@@
 perform reference clustering to a public database because reference-based
-methods produce stable OTUs and are generally faster than _de novo_ methods.
+methods produce consistent OTUs and are generally faster than _de novo_ methods.
 In closed reference mode, sequences that cannot be assigned to reference OTUs
@@
 To test how OptiFit performs for this purpose, we fit each dataset to three
-databases of reference OTUs; the Greengenes database, the SILVA non-redundant
+databases of reference OTUs: the Greengenes database, the SILVA non-redundant
 database, and the Ribosomal Database Project (RDP) [@desantis_greengenes_2006;
@@
 with OptiClust at a distance threshold of 3%.
+The _de novo_ MCC scores for the three databases were **TODO**.
 Fitting sequences to Greengenes and SILVA in closed reference mode performed
 similarly, with median MCC scores of `r closed_fit_gg_mcc` and 
-`r closed_fit_silva_mcc` respectively, while when fitting to RDP the median MCC 
-dropped to `r closed_fit_rdp_mcc`.
-For comparison to the state of the art, clustering datasets _de novo_ with
+`r closed_fit_silva_mcc` respectively, while the median MCC 
+dropped to `r closed_fit_rdp_mcc` when fitting to RDP.
+For comparison, clustering datasets _de novo_ with
 OptiClust produced an average MCC score of `r opticlust_mcc`.
-This gap in OTU quality mostly disappears when clustering in open reference
+This gap in OTU quality mostly disappeared when clustering in open reference
 mode, which produced median MCCs of `r open_fit_gg_mcc` with greengenes, 
-`r open_fit_silva_mcc` with Silva, and `r open_fit_rdp_mcc` with RDP.
-Thus, open reference OptiFit produces OTUs of very similar quality as _de novo_
-clustering, and closed reference OptiFit follows closely behind as long as a
-suitable reference database is chosen.
+`r open_fit_silva_mcc` with SILVA, and `r open_fit_rdp_mcc` with RDP.
+Thus, open reference OptiFit produced OTUs of very similar quality as _de novo_
+clustering, and closed reference OptiFit followed closely behind as long as a
+suitable reference database was chosen.
 
-Since closed reference clustering throws out query sequences that cannot be
-assigned to reference OTUs, an additional measure of clustering performance to
-consider is the fraction of query sequences that were able to be assigned.
+Since closed reference clustering does not cluster query sequences that could
+not be assigned to reference OTUs, an additional measure of clustering
+performance to consider is the fraction of query sequences that were able to be
+assigned.
 On average, more sequences were assigned with Greengenes as the reference 
-(`r frac_fit_gg`%) than with Silva (`r frac_fit_silva`%) or RDP (`r frac_fit_rdp`%).
-This mirrors the result reported above that Greengenes produced better OTUs in
-terms of MCC score than either Silva or RDP.
-Note that _De novo_ and open reference clustering methods always assign 100% of
+(`r frac_fit_gg`%) than with SILVA (`r frac_fit_silva`%) or RDP (`r frac_fit_rdp`%).
+This mirrored the result reported above that Greengenes produced better OTUs in
+terms of MCC score than either SILVA or RDP.
+Note that _de novo_ and open reference clustering methods always assign 100% of
 sequences to OTUs.
 The database chosen affects the final OTU assignments considerably in terms of
-both MCC score and fraction of query sequences fit.
+both MCC score and fraction of query sequences that could be fit to the
+reference OTUs.
 
@@
 Closed reference OptiFit outperformed both OptiClust and open reference OptiFit,
-with respective average run times of `r closed_fit_db_sec`, `r opticlust_sec`,
-and `r open_fit_db_sec` seconds.
+with average run times of `r closed_fit_db_sec`, `r opticlust_sec`,
+and `r open_fit_db_sec` seconds, respectively.
+**TODO: don't average by all datasets & datasets?.**
 Thus, the OptiFit algorithm continues the precedent that closed reference
@@
 sequence's similarity to the single reference OTU.
-In contrast, OptiFit accepts reference OTUs which each contain multiple
+In contrast, OptiFit accepts reference OTUs, which each may contain multiple
 sequences, and the sequence similarity between all query and reference sequences
@@
 query sequences than OptiFit to the Greengenes reference database.
+This is because VSEARCH only considers the distances between each query sequence
+to the single reference sequence, while OptiFit considers the distances between
+all pairs of sequences in an OTU. <!-- VSEARCH "diameter" vs OptiFit "radius"
+-->
 When open reference clustering, OptiFit produced higher quality OTUs than
@@
 chosen greatly affects the quality of OTUs produced.
-OTUs quality may be poor when the reference database is too unrelated to the
-samples of interest, such as when samples contain low abundant or rapidly
-evolving microbes.
+OTU quality may be poor when the reference database is too unrelated to the
+samples of interest, such as when samples contain low abundant or novel
+populations.
 While _de novo_ clustering overcomes the quality limitations of reference
-clustering to databases, OTU assignments are not stable when new sequences are
+clustering to databases, OTU assignments are not consistent when new sequences are
 added.

### 2021-04-21 paper/paper.Rmd
Describe OptiClust algorithm

@@
 method for reference-based clustering.
-**TODO: brief description of the opticlust algorithm.**
+OptiClust first seeds each sequence into its own OTU as a singleton. 
+Then for each sequence, OptiClust considers whether the sequence should move to a
+different OTU or remain in its current OTU, choosing the option that results in
+a better Matthews correlation coefficient (MCC).
+Iterations continue until the MCC stabilizes or until a maximum number of
+iterations is reached.
+This process produces _de novo_ OTU assignments with the most optimal MCC given
+the input sequences.
 OptiFit begins where OptiClust ends, starting with a list of reference OTUs and
@@
 Then, the algorithm iteratively reassigns the query sequences to the reference 
-OTUs to optimize the Matthews correlation coefficient (MCC). 
+OTUs to optimize the MCC. 
 Alternatively, a sequence will remain unassigned if the MCC value is maximized
@@
 sequences _de novo_, then fit the query sequences to the reference OTUs.
-**TODO: describe data sets here**
+We clustered sequences from four datasets isolated from soil
+[@johnston_metagenomics_2016], marine [@henson_artificial_2016], mouse gut
+[@schloss_stabilization_2012], and human gut [@baxter_microbiota-based_2016]
+samples to test the performance of OptiFit with both of these strategies.
 

### 2021-05-03 paper/paper.Rmd
Report mcc for de novo clustering ref dbs

@@
 mothuR::set_knitr_opts()
+knitr::knit_hooks$set(inline = mothuR::paste_oxford_list)
 ```
@@
 with OptiClust at a distance threshold of 3%.
-The _de novo_ MCC scores for the three databases were **TODO**.
+The _de novo_ MCC scores were `r dn_dbs_mcc` for `r names(dn_dbs_mcc)`, 
+respectively.
 Fitting sequences to Greengenes and SILVA in closed reference mode performed

### 2021-05-03 code/R/calc_results_stats.R

@@ -13,7 +13,7 @@ coeff_var <- function(x) {
 dat <- read_tsv(here("results", "summarized.tsv"))
 
 ################################################################################
-# de novo
+# de novo datasets
 opticlust_mcc <- dat %>%
   filter(
     method == "de_novo",
@@ -40,6 +40,14 @@ dn_vsearch_sec <-
   median()
 mcc_opticlust_vs_vsearch <- rel_diff(opticlust_mcc, dn_vsearch_mcc)
 sec_opticlust_vs_vsearch <- abs(rel_diff(dn_vsearch_sec, opticlust_sec))
+
+################################################################################
+# de novo ref dbs
+
+dn_dbs_mcc <- read_tsv('subworkflows/2_fit_reference_db/results/denovo_dbs.tsv') %>% 
+  group_by(ref) %>% 
+  summarize(med_mcc = median(mcc)) %>% 
+  deframe()
 
 ################################################################################
 # ref db open

### 2021-05-03 code/R/calc_results_stats.R
🎨  Style R code

@@ -44,9 +44,9 @@ sec_opticlust_vs_vsearch <- abs(rel_diff(dn_vsearch_sec, opticlust_sec))
 ################################################################################
 # de novo ref dbs
 
-dn_dbs_mcc <- read_tsv('subworkflows/2_fit_reference_db/results/denovo_dbs.tsv') %>% 
-  group_by(ref) %>% 
-  summarize(med_mcc = median(mcc)) %>% 
+dn_dbs_mcc <- read_tsv("subworkflows/2_fit_reference_db/results/denovo_dbs.tsv") %>%
+  group_by(ref) %>%
+  summarize(med_mcc = median(mcc)) %>%
   deframe()
 
 ################################################################################

### 2021-05-05 paper/paper.Rmd
Report human x silva optifit runtime

@@
 run 100 times and measured the execution time.
-Closed reference OptiFit outperformed both OptiClust and open reference OptiFit,
-with average run times of `r closed_fit_db_sec`, `r opticlust_sec`,
-and `r open_fit_db_sec` seconds, respectively.
-**TODO: don't average by all datasets & datasets?.**
+Across all dataset and database combinations, closed reference OptiFit
+outperformed both OptiClust and open reference OptiFit.
+For example, with the human dataset fit to SILVA reference OTUs, the average run
+times in seconds were `r closed_fit_silva_human_sec` for closed reference
+OptiFit, `r opticlust_human_sec` for _de novo_ clustering the dataset, and 
+`r open_fit_silva_human_sec` for open reference OptiFit. 
 Thus, the OptiFit algorithm continues the precedent that closed reference
@@
 sequences, and the sequence similarity between all query and reference sequences
-is considered when assign sequences to OTUs.
+is considered when assigning sequences to OTUs.
 _De novo_ clustering with OptiClust produced `r mcc_opticlust_vs_vsearch`%

### 2021-05-05 code/R/calc_results_stats.R

@@ -123,6 +123,36 @@ open_fit_db_sec <- dat %>%
 sec_vsearch_vs_open_fit_db <- rel_diff(open_vsearch_sec, open_fit_db_sec)
 sec_opticlust_vs_open_fit_db <- rel_diff(opticlust_sec, open_fit_db_sec) %>% abs()
 
+# human dataset to silva 
+open_fit_silva_human_sec <- dat %>%
+  filter(
+    method == "open",
+    strategy == "database",
+    tool == "mothur",
+    ref == "silva",
+    dataset == 'human'
+  ) %>%
+  pull(sec_median) %>%
+  median()
+closed_fit_silva_human_sec <- dat %>%
+  filter(
+    method == "closed",
+    strategy == "database",
+    tool == "mothur",
+    ref == "silva",
+    dataset == 'human'
+  ) %>%
+  pull(sec_median) %>%
+  median()
+opticlust_human_sec <- dat %>%
+  filter(
+    method == "de_novo",
+    tool == "mothur",
+    dataset == 'human'
+  ) %>%
+  pull(sec_median) %>%
+  median()
+
 ################################################################################
 # ref db closed
 closed_fit_db_mcc <- dat %>%

### 2021-05-05 code/R/calc_results_stats.R
🎨  Style R code

@@ -123,14 +123,14 @@ open_fit_db_sec <- dat %>%
 sec_vsearch_vs_open_fit_db <- rel_diff(open_vsearch_sec, open_fit_db_sec)
 sec_opticlust_vs_open_fit_db <- rel_diff(opticlust_sec, open_fit_db_sec) %>% abs()
 
-# human dataset to silva 
+# human dataset to silva
 open_fit_silva_human_sec <- dat %>%
   filter(
     method == "open",
     strategy == "database",
     tool == "mothur",
     ref == "silva",
-    dataset == 'human'
+    dataset == "human"
   ) %>%
   pull(sec_median) %>%
   median()
@@ -140,7 +140,7 @@ closed_fit_silva_human_sec <- dat %>%
     strategy == "database",
     tool == "mothur",
     ref == "silva",
-    dataset == 'human'
+    dataset == "human"
   ) %>%
   pull(sec_median) %>%
   median()
@@ -148,7 +148,7 @@ opticlust_human_sec <- dat %>%
   filter(
     method == "de_novo",
     tool == "mothur",
-    dataset == 'human'
+    dataset == "human"
   ) %>%
   pull(sec_median) %>%
   median()

### 2021-05-05 paper/paper.Rmd
Refine comparison to VSEARCH paragraph

@@
 Despite the drawbacks, closed reference methods have been used when fast
-execution speed is required such as when using very large datasets.
+execution speed is required, such as when using very large datasets.
 To compare performance in terms of speed, we repeated each OptiFit and OptiClust
@@
 sequence, and VSEARCH maps sequences to the reference based on each individual
-sequence's similarity to the single reference OTU.
-In contrast, OptiFit accepts reference OTUs, which each may contain multiple
+query sequence's similarity to the single reference sequence.
+In contrast, OptiFit accepts reference OTUs which each may contain multiple
 sequences, and the sequence similarity between all query and reference sequences
@@
 _De novo_ clustering with OptiClust produced `r mcc_opticlust_vs_vsearch`%
-higher quality OTUs than VSEARCH, but performed `r sec_opticlust_vs_vsearch`% 
-slower than VSEARCH.
-In closed reference mode, VSEARCH was able to map `r frac_vsearch_vs_fit`% more
-query sequences than OptiFit to the Greengenes reference database.
+higher quality OTUs than VSEARCH in terms of MCC, but performed 
+`r sec_opticlust_vs_vsearch`% slower than VSEARCH.
+In closed reference mode, OptiFit produced `r mcc_closed_fit_db_vs_vsearch`%
+higher quality OTUs than VSEARCH, but VSEARCH was able to map 
+`r frac_vsearch_vs_fit`% more query sequences than OptiFit to the Greengenes
+reference database.
 This is because VSEARCH only considers the distances between each query sequence
@@
 `r sec_vsearch_vs_open_fit_db`% on average respectively.
-**TODO: conclude:** The stark difference in OTU definitions between mothur and VSEARCH resulted in...
+Thus, the more stringent OTU definition employed by OptiFit resulted in fewer
+sequences being fit to reference OTUs than when using VSEARCH, but caused
+OptiFit to outperform VSEARCH in terms of both OTU quality and execution time.
 

### 2021-05-05 code/R/calc_results_stats.R

@@ -258,6 +258,15 @@ closed_vsearch_sec <- dat %>%
   median()
 sec_closed_fit_db_vs_vsearch <- rel_diff(closed_fit_db_sec, closed_vsearch_sec) %>% abs()
 
+closed_vsearch_mcc <- dat %>%
+  filter(
+    method == "closed",
+    strategy == "database",
+    tool == "vsearch"
+  ) %>%
+  pull(mcc_median) %>%
+  median()
+mcc_closed_fit_db_vs_vsearch <- rel_diff(closed_fit_db_mcc, closed_vsearch_mcc) %>% abs()
 
 ################################################################################
 # fit split

### 2021-05-10 compare-strategies_2021-05.Rmd
Prep for lab mtg plot critiquing

---
title: "Comparing OTU quality across clustering strategies"
date: '`r Sys.Date()`'
output: 
  github_document:
    html_preview: false
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      fig.path = "figures/"
                      )
```

```{r lib}
set.seed(2018)
library(cowplot)
library(ggtext)
library(glue)
library(here)
library(knitr)
library(tidyverse)

theme_set(theme_classic())
color_palette <- RColorBrewer::brewer.pal(4, "Dark2")
dataset_colors <- c(
  human = color_palette[[3]],
  marine = color_palette[[1]],
  mouse = color_palette[[4]],
  soil = color_palette[[2]]
)
tri_colors <-  c("darkorange","darkorchid","cyan4") # https://allisonhorst.github.io/palmerpenguins/articles/intro.html
mutate_perf <- function(dat) {
  dat %>% 
    mutate(mem_mb = max_rss,
           mem_gb = mem_mb / 1024) %>% 
    rename(sec = s)
}
plot_denovo_hline <- function(yint, dat = sum_opticlust) {
  list(geom_hline(data = dat, aes(yintercept = {{yint}})),
       labs(caption = "Black line: _de novo_ clustering"),
       theme(plot.caption = element_markdown())
  )
}
```

```{r load-data}
opticlust <- read_tsv(here('subworkflows/1_prep_samples/results/opticlust_results.tsv')) %>% 
  full_join(read_tsv(here('subworkflows/1_prep_samples/results/dataset_sizes.tsv'))) %>% 
  mutate_perf()
sum_opticlust <- opticlust %>% 
  group_by(dataset) %>% 
  summarize(mcc_median = median(mcc),
            sec_median = median(sec),
            mem_gb_median = median(mem_gb)) %>% 
  mutate(frac_map_median = 1)

optifit_dbs <- read_tsv(here('subworkflows/2_fit_reference_db/results/optifit_dbs_results.tsv')) %>% 
  mutate_perf()
optifit_split <- read_tsv(here('subworkflows/3_fit_sample_split/results/optifit_split_results.tsv')) %>% 
  mutate_perf()
optifit_all <- list(optifit_dbs %>% 
                   mutate(strategy = glue('database_{ref}')),
                 optifit_split %>% 
                   mutate(strategy = 'self-split')) %>% 
  reduce(full_join)

sum_optifit <- optifit_all %>% 
  group_by(dataset, strategy, method) %>% 
  summarize(n = n(),
            mcc_median = median(mcc),  # TODO: tidy way to avoid this repetitiveness?
            sec_median = median(sec),
            mem_gb_median = median(mem_gb),
            frac_map_median = median(fraction_mapped))
head(sum_optifit)
```

# with facet_grid()

```{r otu-quality_facet}
sum_optifit %>% 
  pivot_longer(c(frac_map_median, mcc_median), names_to = 'quality_metric') %>% 
  ggplot(aes(strategy, value, color = method)) + 
  plot_denovo_hline(yint = value, dat = sum_opticlust %>% 
                      pivot_longer(c(frac_map_median, mcc_median), 
                                   names_to = 'quality_metric')) +
  geom_point() +
  facet_grid(quality_metric ~ dataset) +
  scale_color_manual(values = tri_colors) +
  coord_flip() +
  labs(x = '', y = '')  + 
  guides(color = guide_legend(nrow = 1)) +
  theme(legend.position = "bottom")
```

Problems:

- The sideways grid labels are hard to read.
- There's too much whitespace around the legend.

## with cowplot

```{r otu-quality_cowplot, fig.width=4, fig.height=2}
plot_quality <- function(dat, y_val, title = '') {
  dat %>% 
    ggplot(aes(strategy, {{ y_val }}, color = method)) + 
    geom_hline(data = sum_opticlust, aes(yintercept = {{ y_val }})) +
    geom_point() +
    facet_wrap(dataset ~ ., nrow=1) +
    scale_color_manual(values = tri_colors) +
    ylim(0, 1) +
    coord_flip() +
    labs(x = '', y = '', title = title) + 
    theme(legend.position="none")
}
mcc_plot <- sum_optifit %>% 
  plot_quality(mcc_median, title = 'Median MCC')
frac_plot <- sum_optifit %>% 
  plot_quality(frac_map_median, title = 'Median fraction mapped')  + 
  labs(caption = "Black line: _de novo_ clustering") +
  theme(plot.caption = element_markdown(hjust = 0),
        plot.caption.position = 'plot')

shared_legend <- get_legend(mcc_plot + 
                              guides(color = guide_legend(nrow = 1)) +
                              theme(legend.position = "bottom")
                            )

main_plot <- plot_grid(mcc_plot, frac_plot, 
                       ncol = 1, align = 'v', labels = 'AUTO')

plot_grid(main_plot, shared_legend, 
          ncol = 1, rel_heights = c(1, 0.1))
```

TODO:

- Get caption & legend side-by-side. Maybe use `ggannotate` instead of `labs(caption)` to accomplish this?
- Remove excess whitespace around the legend. https://wilkelab.org/cowplot/articles/shared_legends.html

### 2021-05-10 exploratory/2021/compare-strategies_2021-05.Rmd
auto figure size

--- 2021-05-10 compare-strategies_2021-05.Rmd
@@
 ```
+
+### data prep
 
@@
 
-# with facet_grid()
+# using facet_grid()
 
@@
 
-## with cowplot
+## using cowplot
 
-```{r otu-quality_cowplot, fig.width=4, fig.height=2}
+```{r otu-quality_cowplot}
 plot_quality <- function(dat, y_val, title = '') {

### 2021-05-10 exploratory/2021/compare-strategies_2021-05.Rmd
Adjust fig dim

@@
 
-```{r otu-quality_cowplot}
+```{r otu-quality_cowplot, fig.width = 9, fig.height = 5}
 plot_quality <- function(dat, y_val, title = '') {
@@
 - Remove excess whitespace around the legend. https://wilkelab.org/cowplot/articles/shared_legends.html
+- Adjust figure dimensions.

### 2021-05-10 exploratory/2021/compare-strategies_2021-05.Rmd
Edit cowplot from lab mtg feedback

@@
 
-```{r otu-quality_cowplot, fig.width = 9, fig.height = 5}
+```{r otu-quality_cowplot, fig.width = 9, fig.height = 4.5}
 plot_quality <- function(dat, y_val, title = '') {
@@
 - Adjust figure dimensions.
+
+### Feedback from lab meeting
+
+- [x] vertical jitter
+- [x] frac map only makes sense for closed
+- [x] dotted line for _de novo_
+- [x] filled/unfilled for closed/open
+- [x] box around facets
+- [x] bigger points
+- [ ] set theme elements to add space between facets: `+ theme(panel.spacing = unit(c(3,3),'lines'))`
+- generally: build for the dimensions required by the journal
+
+```{r otu-quality_cowplot-revised, fig.width = 6, fig.height = 4.5}
+plot_quality <- function(dat, y_val, title = '') {
+  dat %>% 
+    ggplot(aes(strategy, {{ y_val }}, 
+               #color = method, 
+               shape = method)) + 
+    geom_point(size = 3,
+               position = position_jitter(width = 0.3, height = 0)) +
+    facet_wrap(dataset ~ ., nrow=1) +
+    scale_shape_manual(values = list(open = 1, closed = 19)) +
+    #scale_color_manual(values = tri_colors) +
+    scale_y_continuous(labels = c('0', '0.5', '1'), 
+                       breaks = c(0, 0.5, 1),
+                       limits = c(0, 1)) +
+    coord_flip() +
+    labs(x = '', y = '', title = title) +
+    theme_bw()
+}
+mcc_plot <- sum_optifit %>% 
+  plot_quality(mcc_median, title = 'Median MCC') +
+    geom_hline(data = sum_opticlust, 
+               aes(yintercept = mcc_median), 
+               linetype = 'dashed', color = 'red')
+frac_plot <- sum_optifit %>% filter(method == 'closed') %>% 
+  plot_quality(frac_map_median, title = 'Median fraction mapped')  + 
+  labs(caption = "dashed line: _de novo_ clustering") +
+  theme(plot.caption = element_markdown(hjust = 0),
+        plot.caption.position = 'plot')
+
+shared_legend <- get_legend(mcc_plot + 
+                              guides(color = guide_legend(nrow = 1)) +
+                              theme(legend.position = "bottom")
+                            )
+
+main_plot <- plot_grid(mcc_plot + 
+    theme(legend.position="none"), frac_plot + 
+    theme(legend.position="none"), 
+                       ncol = 1, align = 'v', labels = 'AUTO')
+
+plot_grid(main_plot, shared_legend, 
+          ncol = 1, rel_heights = c(1, 0.1))
+```
+
+more TODOs:
+
+- [ ] get height of A & B to be equal. B decreased since I removed the de novo & open data.

### 2021-05-10 exploratory/2021/compare-strategies_2021-05.Rmd
No jitter for fraction mapped plot

@@
                shape = method)) + 
-    geom_point(size = 3,
-               position = position_jitter(width = 0.3, height = 0)) +
     facet_wrap(dataset ~ ., nrow=1) +
@@
   plot_quality(mcc_median, title = 'Median MCC') +
+    geom_point(size = 3,
+               position = position_jitter(width = 0.3, height = 0)) +
     geom_hline(data = sum_opticlust, 
@@
   plot_quality(frac_map_median, title = 'Median fraction mapped')  + 
+    geom_point(size = 3) +
   labs(caption = "dashed line: _de novo_ clustering") +
@@
 ```
-
-more TODOs:
-
-- [ ] get height of A & B to be equal. B decreased since I removed the de novo & open data.

### 2021-05-11 exploratory/2021/compare-strategies_2021-05.Rmd
Use position_dodge() instead of jitter

@@
 ```{r lib}
-set.seed(2018)
+set.seed(20210509)
 library(cowplot)
@@
   plot_quality(mcc_median, title = 'Median MCC') +
-    geom_point(size = 3,
-               position = position_jitter(width = 0.3, height = 0)) +
     geom_hline(data = sum_opticlust, 
                aes(yintercept = mcc_median), 
-               linetype = 'dashed', color = 'red')
+               linetype = 'dashed', color = 'red') +
+    geom_point(size = 3,
+               position = position_dodge(width = 0.3))
 frac_plot <- sum_optifit %>% filter(method == 'closed') %>% 

### 2021-05-11 exploratory/2021/compare-strategies_2021-05.Rmd
Increase dodge width

@@
     geom_point(size = 3,
-               position = position_dodge(width = 0.3))
+               position = position_dodge(width = 0.4))
 frac_plot <- sum_optifit %>% filter(method == 'closed') %>% 

### 2021-05-11 exploratory/2021/compare-strategies_2021-05.Rmd
Move caption to make plots equal height

@@
     labs(x = '', y = '', title = title) +
-    theme_bw()
+    theme_bw() + 
+    theme(legend.position="none")
 }
+
 mcc_plot <- sum_optifit %>%
   plot_quality(mcc_median, title = 'Median MCC') +
-    geom_hline(data = sum_opticlust, 
-               aes(yintercept = mcc_median), 
+  geom_hline(data = sum_opticlust, aes(yintercept = mcc_median),
              linetype = 'dashed', color = 'red') +
-    geom_point(size = 3,
-               position = position_dodge(width = 0.4))
+  geom_point(size = 3, position = position_dodge(width = 0.4))
+
 frac_plot <- sum_optifit %>% filter(method == 'closed') %>% 
   plot_quality(frac_map_median, title = 'Median fraction mapped')  + 
-    geom_point(size = 3) +
-  labs(caption = "dashed line: _de novo_ clustering") +
-  theme(plot.caption = element_markdown(hjust = 0),
-        plot.caption.position = 'plot')
+    geom_point(size = 3)
 
@@
 
-main_plot <- plot_grid(mcc_plot + 
-    theme(legend.position="none"), frac_plot + 
-    theme(legend.position="none"), 
-                       ncol = 1, align = 'v', labels = 'AUTO')
+main_plot <- plot_grid(mcc_plot, frac_plot,
+                       ncol = 1, align = 'v', labels = 'AUTO'
+                       ) +
+  labs(caption = "dashed line: _de novo_ clustering") +
+  theme(plot.caption = element_markdown(size = rel(0.8)))
 

### 2021-05-12 exploratory/2021/compare-strategies_2021-05.Rmd
Create version of otu quality plot with patchwork

@@
 
-# using facet_grid()
+## facet_grid()
 
@@
 
-## using cowplot
+## cowplot
 
@@
 ```
+
+Idea: add vsearch _de novo_ & gg to the plot?
+
+## patchwork
+
+```{r otu-quality_patchwork}
+library(patchwork)
+
+(mcc_plot  + 
+  guides(color = guide_legend(nrow = 1)) +
+  theme(legend.position = "bottom")) / 
+frac_plot / 
+  guide_area() + 
+  plot_layout(guides = 'collect') +
+  plot_annotation(caption = "dashed line: _de novo_ clustering",
+                  tag_levels = 'A',
+                  theme = theme(plot.caption = element_markdown()))
+```
+
+TODO:
+
+- [ ] reduce the whitespace
+- [ ] figure out how to get the caption & legend side-by-side

### 2021-05-16 paper/paper.Rmd
Updates from Wednesday writing group
Added transition statements for split datasets section.

@@
 While _de novo_ clustering overcomes the quality limitations of reference
-clustering to databases, OTU assignments are not consistent when new sequences are
-added.
+clustering to databases, OTU assignments are not consistent when new sequences
+are added.
 Researchers may wish to fit new sequences to existing OTUs when comparing OTUs
@@
 
-**Then we wanted to know; what fraction of sequences should be in the reference?**
+While we initially tested this strategy using an even split of the data into
+reference and query fractions, we then wanted to investigate whether there was
+an optimal reference fraction size.
 To test the best reference size, reference sizes from 10% to 80% of the
@@
 variation of `r cv_fit_split_mcc_human` for the MCC score across all fractions. **TODO: revisit how to report this**
+**TODO: runtime & fraction mapped across sizes.**
 
-**Finally, we wanted to know the best way to select the reference sequences.**
 **TODO: pick a fraction (e.g. 50%). this part is less important. figure would be supplemental.**
-We also tested three methods for selecting the fraction of sequences to be used
+After testing the split strategy using a simple random sample to select the reference sequences, we then investigated other methods of splitting the data.
+We tested three methods for selecting the fraction of sequences to be used
 as the reference; a simple random sample, weighting sequences by relative
@@
 the reference, to `r frac_fit_split_0.8`% of query sequences fit when using 80% 
-of the dataset as the reference.
+of the dataset as the reference. **TODO: report up to 90% fraction**
 

### 2021-05-16 exploratory/2021/compare-strategies_2021-05.Rmd
Compare mothur vs vsearch in one plot

@@
 
-theme_set(theme_classic())
 color_palette <- RColorBrewer::brewer.pal(4, "Dark2")
@@
 }
+select_cols <- function(dat) {
+  dat %>% 
+    select(dataset, strategy, method, tool, mcc, sec, mem_gb, fraction_mapped)
+}
+group_sum <- function(dat) {
+  dat %>% 
+  group_by(dataset, tool, method, strategy) %>% 
+  summarize(mcc_median = median(mcc),  # TODO: tidy way to avoid this repetitiveness?
+            sec_median = median(sec),
+            mem_gb_median = median(mem_gb),
+            frac_map_median = median(fraction_mapped))
+}
 ```
@@
 ```{r load-data}
+
 opticlust <- read_tsv(here('subworkflows/1_prep_samples/results/opticlust_results.tsv')) %>% 
   full_join(read_tsv(here('subworkflows/1_prep_samples/results/dataset_sizes.tsv'))) %>% 
-  mutate_perf()
-sum_opticlust <- opticlust %>% 
-  group_by(dataset) %>% 
-  summarize(mcc_median = median(mcc),
-            sec_median = median(sec),
-            mem_gb_median = median(mem_gb)) %>% 
-  mutate(frac_map_median = 1)
-
+  mutate_perf() %>% mutate(fraction_mapped = 1, strategy = method)
+sum_opticlust <- opticlust %>% group_sum()
 optifit_dbs <- read_tsv(here('subworkflows/2_fit_reference_db/results/optifit_dbs_results.tsv')) %>% 
@@
   reduce(full_join)
-
-sum_optifit <- optifit_all %>% 
-  group_by(dataset, strategy, method) %>% 
-  summarize(n = n(),
-            mcc_median = median(mcc),  # TODO: tidy way to avoid this repetitiveness?
-            sec_median = median(sec),
-            mem_gb_median = median(mem_gb),
-            frac_map_median = median(fraction_mapped))
-head(sum_optifit)
+sum_optifit <- optifit_all %>% group_sum()
+vsearch <- read_tsv(here('subworkflows/4_vsearch/results/vsearch_results.tsv')) %>% 
+  mutate_perf() %>% 
+  mutate(strategy = case_when(
+    method == 'de_novo' ~ method,
+    TRUE ~ as.character(glue('database_{ref}')))) 
+mothur_vsearch <- list(optifit_all, opticlust, vsearch) %>% 
+  lapply(select_cols) %>% 
+  reduce(bind_rows) %>% 
+  mutate(strategy = factor(strategy, levels = c('database_rdp', 'database_silva', 'database_gg', 'self-split', 'de_novo')),
+         method = factor(method, levels = c('open', 'closed', 'de_novo')))
+sum_all <- mothur_vsearch %>% group_sum()
+head(sum_all)
 ```
@@
   guides(color = guide_legend(nrow = 1)) +
+  theme_classic() +
   theme(legend.position = "bottom")
@@
     labs(x = '', y = '', title = title) + 
+    theme_classic() +
     theme(legend.position="none")
@@
 - [ ] figure out how to get the caption & legend side-by-side
+
+## mothur vs vsearch
+
+- color: mothur or vsearch
+- shape: open, closed, or _de novo_
+- sort by mcc?
+
+```{r otu-quality_mothur-vs-vsearch, fig.width = 6, fig.height = 4.5}
+plot_quality <- function(dat, y_val, title = '') {
+  dat %>% 
+    ggplot(aes(strategy, {{ y_val }}, 
+               color = tool, 
+               shape = method)) + 
+    geom_point(size = 3, position = position_dodge(width = 0.4)) +
+    facet_wrap(dataset ~ ., nrow=1) +
+    scale_shape_manual(values = list(open = 1, closed = 19, de_novo = 17)) +
+    #scale_color_manual(values = tri_colors) +
+    scale_y_continuous(labels = c('0', '0.5', '1'), 
+                       breaks = c(0, 0.5, 1),
+                       limits = c(0, 1)) +
+    coord_flip() +
+    labs(x = '', y = '', title = title) +
+    theme_bw() +
+    theme(legend.position="none")
+}
+
+mcc_plot <- sum_all %>% 
+  plot_quality(mcc_median, title = "MCC")
+frac_plot <- sum_all %>% filter(method == 'closed') %>% 
+  plot_quality(frac_map_median, title = "Fraction Mapped")
+
+shared_legend <- get_legend(mcc_plot + 
+                              guides(color = guide_legend(nrow = 1)) +
+                              theme(legend.position = "bottom",
+                                    legend.title = element_blank())
+                            )
+
+main_plot <- plot_grid(mcc_plot, frac_plot,
+                       ncol = 1, align = 'v', labels = 'AUTO'
+                       )
+
+plot_grid(main_plot, shared_legend, 
+          ncol = 1, rel_heights = c(1, 0.1))
+```
+
+## Runtime
+
+```{r runtime_mothur-vs-vsearch}
+plot_runtime <- function(dat, yval, title = '') {
+  dat %>% 
+    ggplot(aes(strategy, {{ yval }}, 
+               color = tool, 
+               shape = method
+               )) + 
+    geom_point(size = 3, position = position_dodge(width = 0.4)) +
+    facet_wrap(dataset ~ ., nrow = 1, scales = 'free_x') +
+    scale_shape_manual(values = list(open = 1, closed = 19, de_novo = 17)) +
+    scale_y_log10() +
+    #scale_color_manual(values = tri_colors) +
+    coord_flip() +
+    labs(x = '', y = '', title = title) +
+    theme_bw() +
+    guides(color = guide_legend(nrow = 1)) +
+    theme(legend.position = "none")
+}
+runtime_plot <- sum_all %>% plot_runtime(sec_median) + labs(title = 'Runtime (sec)')
+runtime_plot
+```
+
+## Performance combined
+
+```{r combined_mothur-vs-vsearch, fig.width = 6, fig.height = 6}
+main_plot <- plot_grid(mcc_plot, frac_plot, runtime_plot,
+                       ncol = 1, align = 'v', labels = 'AUTO'
+                       )
+
+plot_grid(main_plot, shared_legend, 
+          ncol = 1, rel_heights = c(1, 0.1))
+```

### 2021-05-16 code/R/summarize_results.R
Tweak aggregated results strategy

@@ -9,8 +9,7 @@ mutate_perf <- function(dat) {
       mem_gb = mem_mb / 1024
     ) %>%
     rename(
-      sec = s,
-      num_otus = sobs
+      sec = s
     )
 }
 #
@@ -24,7 +23,7 @@ opticlust <- read_tsv(here("subworkflows/1_prep_samples/results/opticlust_result
   mutate(strategy = method)
 optifit_db <- read_tsv(here("subworkflows/2_fit_reference_db/results/optifit_dbs_results.tsv")) %>%
   mutate_perf() %>%
-  mutate(strategy = glue("database"))
+  mutate(strategy = glue("database_{ref}"))
 optifit_split <- read_tsv(here("subworkflows/3_fit_sample_split/results/optifit_split_results.tsv")) %>%
   mutate_perf() %>%
   mutate(strategy = "self-split")
@@ -32,7 +31,7 @@ vsearch <- read_tsv(here("subworkflows/4_vsearch/results/vsearch_results.tsv")) 
   mutate_perf() %>%
   mutate(strategy = case_when(
     method == "de_novo" ~ method,
-    TRUE ~ as.character(glue("database"))
+    TRUE ~ as.character(glue("database_{ref}"))
   ))
 results_agg <- list(opticlust, optifit_db, optifit_split, vsearch) %>%
   reduce(full_join)

### 2021-05-20 paper/paper.Rmd
More details for split datasets section

@@
 ```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
-mothuR::set_knitr_opts()
-knitr::knit_hooks$set(inline = mothuR::paste_oxford_list)
+schtools::set_knitr_opts()
 ```
@@
 library(here)
-library(mothuR)
+library(schtools)
 library(tidyverse)
@@
 ```
-
+<!--
 - **"AND, BUT, THEREFORE" structure. start paragraph with question, end with why we should care. transitions to move the story along.**
@@
 This is described in one of my earlier papers looking at open/closed reference clustering and was part of the reason that the order of the database was important."
+-->
 
@@
 the average `r frac_fit_gg`% of sequences fit to the greengenes database.
-In terms of runtime, closed and open reference OptiFit performed faster than
+In terms of run time, closed and open reference OptiFit performed faster than
 OptiClust on whole datasets by `r sec_closed_fit_split_vs_clust`% and 
@@
 For example, splitting the human dataset 100 times yielded a coefficient of
-variation of `r cv_fit_split_mcc_human` for the MCC score across all fractions. **TODO: revisit how to report this**
-**TODO: runtime & fraction mapped across sizes.**
+variation of `r cv_fit_split_mcc_human_simple` for the MCC score across all fractions.
+Run time generally decreased as the reference fraction increased; for the human
+dataset, the median run time was `r sec_fit_split_human_simple_1` with 10% of
+sequences in the reference and `r sec_fit_split_human_simple_9` with 90% of
+sequences in the reference.
+In closed reference mode, the fraction of sequences that mapped increased as the
+reference size increased; for the human dataset, the median fraction mapped was
+`r frac_fit_split_human_simple_1` with 10% of sequences in the reference and 
+`r frac_fit_split_human_simple_9` with 90% of sequences in the reference.
+These trends hold for the other datasets as well (see supplemental figure **TODO**).
+Thus, the reference fraction does not affect OTU quality in terms of MCC score,
+but does affect the run time and the fraction of sequences that map during closed reference mode.
 
@@
 reference OTUs increases as the reference fraction increases; from 
-`r frac_fit_split_0.1`% of query sequences fit when using 10% of the dataset as 
-the reference, to `r frac_fit_split_0.8`% of query sequences fit when using 80% 
+`r frac_fit_split_1`% of query sequences fit when using 10% of the dataset as 
+the reference, to `r frac_fit_split_9`% of query sequences fit when using 90% 
 of the dataset as the reference. **TODO: report up to 90% fraction**

### 2021-05-20 code/R/summarize_results.R
Don't include the refname for database strategy

@@ -23,7 +23,7 @@ opticlust <- read_tsv(here("subworkflows/1_prep_samples/results/opticlust_result
   mutate(strategy = method)
 optifit_db <- read_tsv(here("subworkflows/2_fit_reference_db/results/optifit_dbs_results.tsv")) %>%
   mutate_perf() %>%
-  mutate(strategy = glue("database_{ref}"))
+  mutate(strategy = "database")
 optifit_split <- read_tsv(here("subworkflows/3_fit_sample_split/results/optifit_split_results.tsv")) %>%
   mutate_perf() %>%
   mutate(strategy = "self-split")
@@ -31,8 +31,8 @@ vsearch <- read_tsv(here("subworkflows/4_vsearch/results/vsearch_results.tsv")) 
   mutate_perf() %>%
   mutate(strategy = case_when(
     method == "de_novo" ~ method,
-    TRUE ~ as.character(glue("database_{ref}"))
-  ))
+    TRUE ~ "database")
+    )
 results_agg <- list(opticlust, optifit_db, optifit_split, vsearch) %>%
   reduce(full_join)
 

### 2021-06-02 paper/paper.Rmd
Report sampling methods results at reffrac 0.5

@@
 OptiClust first seeds each sequence into its own OTU as a singleton. 
-Then for each sequence, OptiClust considers whether the sequence should move to a
-different OTU or remain in its current OTU, choosing the option that results in
-a better Matthews correlation coefficient (MCC).
+Then for each sequence, OptiClust considers whether the sequence should move to
+a different OTU or remain in its current OTU, choosing the option that results
+in a better Matthews correlation coefficient (MCC).
 Iterations continue until the MCC stabilizes or until a maximum number of
@@
 For example, splitting the human dataset 100 times yielded a coefficient of
-variation of `r cv_fit_split_mcc_human_simple` for the MCC score across all fractions.
+variation of `r cv_fit_split_mcc_human_simple` for the MCC score across all
+fractions.
 Run time generally decreased as the reference fraction increased; for the human
@@
 Thus, the reference fraction does not affect OTU quality in terms of MCC score,
-but does affect the run time and the fraction of sequences that map during closed reference mode.
+but does affect the run time and the fraction of sequences that map during
+closed reference mode.
 
-**TODO: pick a fraction (e.g. 50%). this part is less important. figure would be supplemental.**
-After testing the split strategy using a simple random sample to select the reference sequences, we then investigated other methods of splitting the data.
-We tested three methods for selecting the fraction of sequences to be used
-as the reference; a simple random sample, weighting sequences by relative
-abundance, and weighting by similarity to other sequences in the dataset.
-OTU quality was similar with the simple and abundance-weighted sampling (median
-MCCs `r mcc_fit_split_simple` and `r mcc_fit_split_abun` respectively), but
-worse for similarity-weighted sampling with a median MCC of `r mcc_fit_split_dist`.
-In closed reference mode, the fraction of query sequences that can be fit to the
-reference OTUs increases as the reference fraction increases; from 
-`r frac_fit_split_1`% of query sequences fit when using 10% of the dataset as 
-the reference, to `r frac_fit_split_9`% of query sequences fit when using 90% 
-of the dataset as the reference. **TODO: report up to 90% fraction**
+After testing the split strategy using a simple random sample to select the
+reference sequences, we then investigated other methods of splitting the data.
+We tested three methods for selecting the fraction of sequences to be used as
+the reference at a size of 50%; a simple random sample, weighting sequences by
+relative abundance, and weighting by similarity to other sequences in the
+dataset.
+OTU quality in terms of MCC was similar with the simple and abundance-weighted
+sampling (median MCCs of `r mcc_fit_split_simple` and `r mcc_fit_split_abun`
+respectively), but worse for similarity-weighted sampling (median MCC of
+`r mcc_fit_split_dist`).
+In closed-reference clustering mode, the fraction of sequences that mapped were
+similar for simple and abundance-weighted sampling (median fraction mapped of 
+`r frac_fit_split_simple` and `r frac_fit_split_abun` respectively), but worse 
+for similarity-weighted sampling (median fraction mapped of `r frac_fit_split_dist`).
+While simple and abundance-weighted sampling produced better quality OTUs than
+similarity-weighted sampling, OptiFit performed faster on
+similarity-weighted samples with a median runtime of `r sec_fit_split_dist`
+seconds compared to `r sec_fit_split_simple` and `r sec_fit_split_abun` seconds
+for simple and abundance-weighted sampling (respectively).
+Thus, employing more complicated sampling strategies such as abundance-weighted
+and similarity-weighted sampling did not confer any advantages over selecting
+the reference via a simple random sample, and in fact decreased OTU quality in
+the case of similarity-weighted sampling.
 

### 2021-06-02 code/R/calc_results_stats.R

@@ -378,6 +378,9 @@ mcc_fit_split_simple <- dat %>%
   median()
 mcc_opticlust_vs_fit_split_simple <- rel_diff(opticlust_mcc, mcc_fit_split_simple)
 
+#####
+# fit split at ref_frac 0.5
+
 mcc_fit_split_abun <- dat %>%
   filter(
     strategy == "self-split",
@@ -397,6 +400,68 @@ mcc_fit_split_dist <- dat %>%
   ) %>%
   pull(mcc_median) %>%
   median()
+
+frac_fit_split_simple <- dat %>%
+  filter(
+    strategy == "self-split",
+    tool == "mothur",
+    ref_weight == "simple",
+    ref_frac == 0.5
+  ) %>%
+  pull(frac_map_median) %>%
+  median()
+
+frac_fit_split_abun <- dat %>%
+  filter(
+    strategy == "self-split",
+    tool == "mothur",
+    ref_weight == "abundance",
+    ref_frac == 0.5
+  ) %>%
+  pull(frac_map_median) %>%
+  median()
+
+frac_fit_split_dist <- dat %>%
+  filter(
+    strategy == "self-split",
+    tool == "mothur",
+    ref_weight == "distance",
+    ref_frac == 0.5
+  ) %>%
+  pull(frac_map_median) %>%
+  median()
+
+sec_fit_split_simple <- dat %>%
+  filter(
+    strategy == "self-split",
+    tool == "mothur",
+    ref_weight == "simple",
+    ref_frac == 0.5
+  ) %>%
+  pull(sec_median) %>%
+  median()
+
+sec_fit_split_abun <- dat %>%
+  filter(
+    strategy == "self-split",
+    tool == "mothur",
+    ref_weight == "abundance",
+    ref_frac == 0.5
+  ) %>%
+  pull(sec_median) %>%
+  median()
+
+sec_fit_split_dist <- dat %>%
+  filter(
+    strategy == "self-split",
+    tool == "mothur",
+    ref_weight == "distance",
+    ref_frac == 0.5
+  ) %>%
+  pull(sec_median) %>%
+  median()
+
+##########
 
 frac_fit_split_1 <- dat %>%
   filter(

### 2021-06-08 exploratory/2021/compare-strategies_2021-05.Rmd
Reformat strategy & method labels gotta have 'de novo' italicized

@@
   reduce(bind_rows) %>% 
-  mutate(strategy = factor(strategy, levels = c('database_rdp', 'database_silva', 'database_gg', 'self-split', 'de_novo')),
-         method = factor(method, levels = c('open', 'closed', 'de_novo')))
+  mutate(method = as.character(method),
+         strategy = as.character(strategy))
 sum_all <- mothur_vsearch %>% group_sum()
+head(sum_all)
+```
+
+### make the strategy & method labels prettier
+
+```{r format-labels}
+sum_all <- sum_all %>%
+  mutate(
+    strategy = factor(
+      case_when(
+        strategy == "de_novo"        ~ "_de novo_",
+        strategy == 'database_rdp'   ~ "db: RDP",
+        strategy == 'database_silva' ~ "db: SILVA",
+        strategy == 'database_gg'    ~ "db: Greengenes",
+        TRUE                         ~ strategy
+      ),
+      levels = c('db: RDP', 'db: SILVA', 'db: Greengenes', 
+                 'self-split',  '_de novo_')
+    ),
+    method = factor(
+      case_when(method == "de_novo" ~ "_de novo_",
+                TRUE                ~ method),
+      levels = c('open', 'closed', '_de novo_')
+    )
+  )
 head(sum_all)
@@
 - generally: build for the dimensions required by the journal
+- [ ] fix strategy labels (remove underscores, italicize _de novo_)
 
@@
     facet_wrap(dataset ~ ., nrow=1) +
-    scale_shape_manual(values = list(open = 1, closed = 19, de_novo = 17)) +
+    scale_shape_manual(values = list(open = 1, closed = 19, `_de novo_` = 17)) +
     #scale_color_manual(values = tri_colors) +
@@
     theme_bw() +
-    theme(legend.position="none")
+    theme(legend.position="none",
+          axis.text.y = element_markdown())
 }
@@
                               theme(legend.position = "bottom",
-                                    legend.title = element_blank())
+                                    legend.title = element_blank(),
+                                    legend.text = element_markdown())
                             )
@@
     facet_wrap(dataset ~ ., nrow = 1, scales = 'free_x') +
-    scale_shape_manual(values = list(open = 1, closed = 19, de_novo = 17)) +
+    scale_shape_manual(values = list(open = 1, closed = 19, `_de novo_` = 17)) +
     scale_y_log10() +
@@
     guides(color = guide_legend(nrow = 1)) +
-    theme(legend.position = "none")
+    theme(legend.position = "none",
+          axis.text.y = element_markdown())
 }
-runtime_plot <- sum_all %>% plot_runtime(sec_median) + labs(title = 'Runtime (sec)')
+runtime_plot <- sum_all %>% plot_runtime(sec_median) + 
+  labs(title = 'Runtime (sec)')
 runtime_plot

### 2021-06-09 paper/paper.Rmd
Separate intro paragraphs for _de novo_ and reference clustering

@@
 composition of hundreds of samples from culture-independent microbial
-communities. In a typical analysis pipeline,
-16S rRNA gene sequences are assigned to Operational Taxonomic Units (OTUs) to
-facilitate comparison of taxonomic composition between communities.
+communities. 
+In a typical analysis pipeline, 16S rRNA gene sequences are assigned to
+Operational Taxonomic Units (OTUs) to facilitate comparison of taxonomic
+composition between communities.
 A distance threshold of 3% (or sequence similarity of 97%) is commonly used to
@@
 
-There are three main categories of OTU clustering algorithms: closed reference,
-open reference, and _de novo_ clustering. 
-Closed reference methods assign sequences to a set of pre-made OTUs generated
-from reference sequences.
-If a query sequence is not within the distance threshold to any of the reference
-sequences, it is discarded.
+There are two main categories of OTU clustering algorithms: _de novo_ and
+reference-based.
+OptiClust, the _de novo_ clustering method provided by the mothur software
+program, uses the distance score between all pairs of sequences in the dataset
+to cluster them into OTUs by maximizing the Matthews correlation coefficient
+(MCC) [@westcott_opticlust_2017].
+This approach takes into account the distances between all pairs of sequences
+when assigning query sequences to OTUs, in contrast to other tools such as
+VSEARCH which only considers the distance between the query sequence and a
+representative centroid sequence in the OTU.
+A limitation of _de novo_ clustering is that different OTU assignments will be
+produced when new sequences are added to a dataset, making it difficult to use
+_de novo_ clustering to compare OTUs between different studies.
+Furthermore, since _de novo_ clustering requires distance calculations between all
+sequences in a dataset, the execution time can be slow for very large datasets.
+
+Reference clustering attempts to overcome the limitations of _de novo_
+clustering methods by using a representative set of sequences from a database,
+with each reference sequence seeding an OTU.
+Commonly, the Greengenes set of representative full length sequences clustered
+at 97% similarity is used as the reference with VSEARCH
+[@desantis_greengenes_2006].
+Query sequences are then assigned to OTUs based on their similarity to the
+reference sequences.
+Any query sequences that are not within the distance threshold to any of the
+reference sequences are either thrown out (closed reference mode) or clustered
+_de novo_ to create additional OTUs (open reference mode).
 While reference-based clustering is generally fast, it is limited by the
 diversity of the reference database.
-Rare or novel sequences in the sample will be lost if they are not represented
-by a similar sequence in the database.
-_De novo_ methods cluster sequences based on their distance to each other,
-without the use of an external reference.
-_De novo_ clustering overcomes the limitations of reference databases by
-considering only sequences in the dataset, but is more computationally intensive
-and generates different OTU assignments when new sequences are introduced.
-Unstable OTU assignments make it difficult to use _de novo_ clustering to
-compare taxonomic composition of communities between studies or to use machine
-learning models trained with _de novo_ OTUs to make predictions on new data.
-Open reference methods take a hybrid approach, first performing closed reference
-clustering, then any sequences that cannot be assigned to reference OTUs are
-clustered _de novo_ to create additional OTUs.
+Rare or novel sequences in the sample will be lost in closed reference mode if
+they are not represented by a similar sequence in the database.
 Previous studies found that the OptiClust _de novo_ clustering algorithm created
@@
 
-- **TODO: current method for open/closed is vsearch against greengenes.**
-- **TODO: use word "map" for what vsearch does, "fit" for what optifit does.**
-- **TODO**: 2 categories of clustering: _de novo_ and reference based.
-separate paragraphs.
-describe opticlust first in de novo paragraph.
-2nd paragraph: ref methods are good cause they're fast and don't use much ram.
-dependent on order of db.
-people use greengenes, which are rep seqs from 3% otus from full length.
-- reader should know what opticlust is, closed & open ref clustering is, strengths & weakness are of each. then we solve these problems.
-- **TODO:** note that greengenes is defunct now?!
-
 To overcome the limitations of _de novo_ clustering while maintaining OTU
@@
 to. 
-**TODO: more words here?** 
+In contrast to other tools, OptiFit considers all pairwise distance scores
+between reference and query sequences when assigning sequences to OTUs in order
+to produce OTUs of the highest possible quality.
 Here, we tested the OptiFit algorithm with the reference as a database or _de
@@
 time of this writing.
+**TODO:** note that greengenes is defunct now?!
 We recommend that users who require an independent reference database opt for

### 2021-06-15 paper/paper.Rmd
Note gg hasn't been updated since 2013

@@
 This drop in quality was most notable with RDP, which contains only about 21,000
-sequences compared to over 200,000 sequences in SILVA and Greengenes each at the
-time of this writing.
-**TODO:** note that greengenes is defunct now?!
+sequences compared to over 200,000 sequences in SILVA and Greengenes each.
+Note that Greengenes has not been updated since 2013 at the time of this
+writing, while SILVA is updated every one to two years with the most recent
+release in 2019.
 We recommend that users who require an independent reference database opt for
-large databases with good coverage of microbial diversity.
+large databases with regular updates and good coverage of microbial diversity.
 Since OptiClust performs faster than open reference OptiFit and creates higher

### 2021-06-15 paper/paper.Rmd
Concluding discussion paragraph

@@
 demonstrates the suitability of using OptiFit to fit sequences to existing OTUs,
-such as when using already-trained machine learning models to make predictions
-on new data or comparing OTUs across studies.
+such as when comparing OTUs across studies.
 However, when stable OTUs are not required, we recommend using OptiClust for _de
@@
 
-**TODO: big picture concluding paragraph.**
 We have developed a new clustering algorithm that allows users to produce high
 quality OTUs using already existing OTUs as a reference.
-**TODO: Point to courtney's paper metaphorically. wow what a cool application someone should do _wink wink_.**
+Unlike existing reference-based methods which map query sequences to a single
+centroid sequence in each reference OTU, OptiFit considers all sequences in each
+reference OTU when fitting query sequences, resulting in OTUs of a similar high
+quality as those produced by the _de novo_ OptiClust algorithm.
+Potential applications include fitting sequences to reference databases,
+comparing taxonomic composition of microbiomes across different studies, 
+or using OTU-based machine learning models to make predictions on new data.
+OptiFit fills the missing option for fitting query sequences to existing OTUs
+that does not sacrifice OTU quality for stability of OTU assignments.
 

### 2021-06-15 paper/paper.Rmd
Summarize results in abstract

@@
 mothur program.
-
-**TODO: summarize results & conclusion**
+We tested OptiFit using four microbiome datasets with two different strategies;
+by fitting to an external reference database, or by splitting the dataset into a 
+reference and query set and fitting the query sequences to the reference set.
+OptiFit produces OTUs of similar quality as OptiClust and at faster speeds when
+using the split dataset strategy, while the OTU quality and speed depends on the
+database chosen when using the external database strategy.
+OptiFit provides a suitable option for users who require stable reference-based
+OTU assignments at the same quality afforded by _de novo_ clustering methods.
 

### 2021-06-17 paper/paper.Rmd
Create figure for analysis workflow

@@
 schtools::set_knitr_opts()
+knitr::opts_chunk$set(fig.path = here::here('figures/'))
 ```
@@
 ```{r deps, message = FALSE}
+library(cowplot)
 library(here)
+library(knitr)
 library(schtools)
@@
 Reference OTUs for each database were created by performing _de novo_ clustering
-with OptiClust at a distance threshold of 3%.
+with OptiClust at a distance threshold of 3% (see Figure \@ref(fig:workflow)).
 The _de novo_ MCC scores were `r dn_dbs_mcc` for `r names(dn_dbs_mcc)`, 
@@
 suitable reference database was chosen.
+
+```{r workflow, dpi=300, fig.width=6, fig.cap="The Analysis Workflow"}
+tiff::readTIFF(here('figures', 'workflow.tiff')) %>% 
+  grid::grid.raster() 
+```
 

### 2021-06-17 paper/head.tex
Use bookdown pdf format to fix cross referencing of figures

@@
 
-$1$ Department of Computational Medicine and Bioinformatics,
-University of Michigan, Ann Arbor, MI 48109
+$1$ Department of Computational Medicine and Bioinformatics, University of Michigan
 
-$2$ Department of Microbiology and Immunology,
-University of Michigan, Ann Arbor, MI 48109
+$2$ Department of Microbiology and Immunology, University of Michigan
 

### 2021-06-17 paper/paper.Rmd
Use bookdown pdf format to fix cross referencing of figures

@@
     url: https://github.com/kelly-sovacool
-    affiliation: Department of Computational Medicine and Bioinformatics, University of Michigan, Ann Arbor, MI
+    affiliation: Department of Computational Medicine and Bioinformatics, University of Michigan
     orcid_id: 0000-0003-3283-829X
@@
     url: https://github.com/mothur-westcott
-    affiliation: Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI
+    affiliation: Department of Microbiology and Immunology, University of Michigan
   - name: "M. Brodie Mumphrey"
     url: https://github.com/MBMumphrey
-    affiliation: Department of Computational Medicine and Bioinformatics, University of Michigan, Ann Arbor, MI
+    affiliation: Department of Computational Medicine and Bioinformatics, University of Michigan
   - name: "Gabrielle A. Dotson"
     url: https://github.com/dotsonga
-    affiliation: Department of Computational Medicine and Bioinformatics, University of Michigan, Ann Arbor, MI
+    affiliation: Department of Computational Medicine and Bioinformatics, University of Michigan
   - name: "Patrick D. Schloss"
     url: https://github.com/pschloss
-    affiliation: Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI
+    affiliation: Department of Microbiology and Immunology, University of Michigan
 bibliography: references.bib
@@
     toc_depth: 3
-  pdf_document:
+  bookdown::pdf_document2:
     keep_tex: false
@@
       after_body: tail.tex
+    number_sections: false
+    toc: false
 knit: (function(inputFile, encoding) {

### 2021-06-17 code/R/algorithm_diagram.R
Reflow

@@ -9,8 +9,9 @@ set.seed(20200308)
 # Pat's MCC function
 # source: https://github.com/SchlossLab/Westcott_OptiClust_mSphere_2017/blob/a8bc26855423bba85acc0b8e7cca075e5c94f533/submission/supplemental_text.Rmd#L26-L28
 mcc <- function(tp, tn, fp, fn) {
-  format(round((tp * tn - fp * fn) / sqrt((tp + fp) * (tp + fn) * (tn + fp) *
-    (tn + fn)), digits = 2),
+  format(round((tp * tn - fp * fn) / 
+                 sqrt((tp + fp) * (tp + fn) * (tn + fp) * (tn + fn)), 
+               digits = 2),
   digits = 2,
   nsmall = 2L
   )

### 2021-06-17 paper/paper.Rmd
Make sure fig dimensions in Rmd chunk match actual tiff file
Add draft figure of summarized results

@@
 `r closed_fit_silva_mcc` respectively, while the median MCC 
-dropped to `r closed_fit_rdp_mcc` when fitting to RDP.
+dropped to `r closed_fit_rdp_mcc` when fitting to RDP 
+(see Figure \@ref(fig:workflow)).
 For comparison, clustering datasets _de novo_ with
@@
 
-```{r workflow, dpi=300, fig.width=6, fig.cap="The Analysis Workflow"}
-tiff::readTIFF(here('figures', 'workflow.tiff')) %>% 
+```{r workflow, dpi=300, fig.dim=c(4,3), fig.cap="The Analysis Workflow"}
+# TODO: write caption
+here('figures', 'workflow.tiff') %>% 
+  tiff::readTIFF() %>% 
   grid::grid.raster() 
@@
 clustering sacrifices OTU quality for execution speed.
+
+```{r results, dpi=100, fig.dim=c(6,6), fig.cap="Benchmarking results"}
+# TODO: write caption
+here('figures', 'results.tiff') %>% 
+  tiff::readTIFF() %>% 
+  grid::grid.raster() 
+```
 

### 2021-06-17 code/R/plot_results.R

@@ -1,0 +1,161 @@
+
+### data prep
+
+set.seed(20210509)
+library(cowplot)
+library(ggtext)
+library(glue)
+library(here)
+library(knitr)
+library(tidyverse)
+
+color_palette <- RColorBrewer::brewer.pal(4, "Dark2")
+dataset_colors <- c(
+  human = color_palette[[3]],
+  marine = color_palette[[1]],
+  mouse = color_palette[[4]],
+  soil = color_palette[[2]]
+)
+tri_colors <-  c("darkorange","darkorchid","cyan4") # https://allisonhorst.github.io/palmerpenguins/articles/intro.html
+mutate_perf <- function(dat) {
+  dat %>% 
+    mutate(mem_mb = max_rss,
+           mem_gb = mem_mb / 1024) %>% 
+    rename(sec = s)
+}
+plot_denovo_hline <- function(yint, dat = sum_opticlust) {
+  list(geom_hline(data = dat, aes(yintercept = {{yint}})),
+       labs(caption = "Black line: _de novo_ clustering"),
+       theme(plot.caption = element_markdown())
+  )
+}
+select_cols <- function(dat) {
+  dat %>% 
+    select(dataset, strategy, method, tool, mcc, sec, mem_gb, fraction_mapped)
+}
+group_sum <- function(dat) {
+  dat %>% 
+    group_by(dataset, tool, method, strategy) %>% 
+    summarize(mcc_median = median(mcc),  # TODO: tidy way to avoid this repetitiveness?
+              sec_median = median(sec),
+              mem_gb_median = median(mem_gb),
+              frac_map_median = median(fraction_mapped))
+}
+
+opticlust <- read_tsv(here('subworkflows/1_prep_samples/results/opticlust_results.tsv')) %>% 
+  full_join(read_tsv(here('subworkflows/1_prep_samples/results/dataset_sizes.tsv'))) %>% 
+  mutate_perf() %>% mutate(fraction_mapped = 1, strategy = method)
+sum_opticlust <- opticlust %>% group_sum()
+optifit_dbs <- read_tsv(here('subworkflows/2_fit_reference_db/results/optifit_dbs_results.tsv')) %>% 
+  mutate_perf()
+optifit_split <- read_tsv(here('subworkflows/3_fit_sample_split/results/optifit_split_results.tsv')) %>% 
+  mutate_perf()
+optifit_all <- list(optifit_dbs %>% 
+                      mutate(strategy = glue('database_{ref}')),
+                    optifit_split %>% 
[…]

### 2021-06-21 docs/slides/2021-03-11_bistro.Rmd
Reorganize figures used in slides to subdir

@@
 
-background-image: url("../../figures/SchlossLab.png") 
+background-image: url("../../figures/for_slides/SchlossLab.png") 
 background-size: contain
@@
 .pull-right[
-![](../../figures/zackular-3a.jpg)
+![](../../figures/for_slides/zackular-3a.jpg)
 ]
@@
 .center[
-<img src="../../figures/schloss-2016-msystems_all-method-comparison.png"  height="350"/>
+<img src="../../figures/for_slides/schloss-2016-msystems_all-method-comparison.png"  height="350"/>
 ]
@@
 .center[
-<img src="../../figures/opticlust-supp-ex.png" width="700"/>
+<img src="../../figures/for_slides/opticlust-supp-ex.png" width="700"/>
 ]

### 2021-06-21 paper/paper.Rmd
Re-knit

@@
 - **"AND, BUT, THEREFORE" structure. start paragraph with question, end with why we should care. transitions to move the story along.**
-- From Pat: "briefly looking through the Discussion and Intro, one point that we may have forgotten is that a benefit of our approach is that it is much easier to customize to a specific region.
-The greengenes reference OTUs are based on full length sequences.
-This causes problems when considering shorter (e.g.
-V4) sequences since reference OTUs may be more similar  than the full length and even identical to each other in the subregion.
-Because we can easily spin up a subregion specific set of reference OTUs from a public database or the reference fraction this isn't a problem.
-This is described in one of my earlier papers looking at open/closed reference clustering and was part of the reason that the order of the database was important."
+
+- Go back through https://peerj.com/articles/1487/ to get all the arguments for why the qiime based methods are problematic…
+1. The threshold with closed reference is a radius and with de novo its a diameter
+2. The gg reference was defined for full length sequences, within a subregion there are duplicate and nearly duplicate sequences
+3.
+Because VSEARCH/USEARCH is a greedy heuristic, its looking for a reference that is close.
+Although the ordering of the database is not relevant (it was with USEARCH) it gets the wrong number of its (see Figure 5)
 -->

### 2021-06-22 paper/paper.Rmd
Just use pdf_document output format

@@
 date: '`r Sys.Date()`'
-authors:
-  - name: "Kelly L. Sovacool"
-    url: https://github.com/kelly-sovacool
-    affiliation: Department of Computational Medicine and Bioinformatics, University of Michigan
-    orcid_id: 0000-0003-3283-829X
-  - name: "Sarah L. Westcott"
-    url: https://github.com/mothur-westcott
-    affiliation: Department of Microbiology and Immunology, University of Michigan
-  - name: "M. Brodie Mumphrey"
-    url: https://github.com/MBMumphrey
-    affiliation: Department of Computational Medicine and Bioinformatics, University of Michigan
-  - name: "Gabrielle A. Dotson"
-    url: https://github.com/dotsonga
-    affiliation: Department of Computational Medicine and Bioinformatics, University of Michigan
-  - name: "Patrick D. Schloss"
-    url: https://github.com/pschloss
-    affiliation: Department of Microbiology and Immunology, University of Michigan
 bibliography: references.bib
 output:
-  distill::distill_article:
-    toc: true
-    toc_float: true
-    toc_depth: 3
-  bookdown::pdf_document2:
+  pdf_document:
     keep_tex: false
@@
       before_body: head.tex
-      after_body: tail.tex
     number_sections: false
@@
 repository_url: https://github.com/SchlossLab/OptiFitAnalysis
-creative_commons: CC BY-SA
-twitter:
-  creator: "@kelly_sovacool"
-  site: "@PatSchloss"
 ---
@@
 
-## Acknowledgements {.appendix}
+## Acknowledgements
 
@@
 
-## Author Contributions {.appendix}
+## Author Contributions
 
@@
 All authors reviewed and edited the manuscript.
+
+## References

### 2021-06-23 code/R/plot_results.R
Cleanup unneeded code

@@ -22,8 +22,7 @@ select_cols <- function(dat) {
 opticlust <- read_tsv(here('subworkflows/1_prep_samples/results/opticlust_results.tsv')) %>% 
   full_join(read_tsv(here('subworkflows/1_prep_samples/results/dataset_sizes.tsv'))) %>% 
   mutate_perf() %>% 
-  mutate(strategy = method)
-sum_opticlust <- opticlust %>% group_sum()
+  mutate(strategy = method, fraction_mapped = NA)
 optifit_dbs <- read_tsv(here('subworkflows/2_fit_reference_db/results/optifit_dbs_results.tsv')) %>% 
   mutate_perf()
 optifit_split <- read_tsv(here('subworkflows/3_fit_sample_split/results/optifit_split_results.tsv')) %>% 
@@ -106,6 +105,6 @@ mothur_vsearch  %>%
         plot.margin=unit(x=c(3,3,3,0),units="pt")
   )
 
-ggsave(here('figures', 'results.tiff'), 
+ggsave(here('figures', 'results-sum.tiff'), 
        device = 'tiff', dpi=300,
        width=6, height=6, units='in')

### 2021-06-23 paper/paper.Rmd
Rename results figure slug

@@
 dropped to `r closed_fit_rdp_mcc` when fitting to RDP 
-(see Figure \@ref(fig:workflow)).
+(see Figure \@ref(fig:results-sum)).
 For comparison, clustering datasets _de novo_ with
@@
 
-```{r workflow, dpi=300, fig.dim=c(4,3), fig.cap="The Analysis Workflow"}
+```{r workflow, dpi=300, fig.dim=c(4.5,3.5), fig.cap="**The Analysis Workflow.**"}
 # TODO: write caption
@@
 
-```{r results, dpi=100, fig.dim=c(6,6), fig.cap="Benchmarking results"}
+```{r results-sum, dpi=100, fig.dim=c(6,6), fig.cap="**Benchmarking results**"}
 # TODO: write caption
-here('figures', 'results.tiff') %>% 
+here('figures', 'results-sum.tiff') %>% 
   tiff::readTIFF() %>% 

### 2021-06-23 paper/paper.Rmd
Write figure metadata in separate yaml file

--- 2021-06-22 paper/paper.Rmd
@@
 repository_url: https://github.com/SchlossLab/OptiFitAnalysis
+params:
+    smk: "none"
 ---
@@
 schtools::set_knitr_opts()
-knitr::opts_chunk$set(fig.path = here::here('figures/'))
+knitr::opts_chunk$set(fig.path = here::here('figures/'),
+                      dpi = 300)
 ```
@@
 library(tidyverse)
-load(here('results', 'stats.RData'))
+snakemake <- params$smk
+load(here(snakemake@input[['rda']]))
 ```
+```{r figs-meta}
+figures_meta <- yaml::read_yaml(here('config', 'figures.yaml'))
+figs <- function(fig_name, attribute) {
+  value <- figures_meta[[fig_name]][[attribute]]
+  if (attribute != 'cap') {
+    value <- eval(parse(text = value))
+  }
+  return(value)
+}
+```
+
 <!--
@@
 dropped to `r closed_fit_rdp_mcc` when fitting to RDP 
-(see Figure \@ref(fig:workflow)).
+(see Figure \@ref(fig:results_sum)).
 For comparison, clustering datasets _de novo_ with
@@
 
-```{r workflow, dpi=300, fig.dim=c(4,3), fig.cap="The Analysis Workflow"}
-# TODO: write caption
-here('figures', 'workflow.tiff') %>% 
+```{r workflow, fig.dim=figs('workflow','dim'), fig.cap=figs('workflow','cap')}
+figs('workflow', 'path') %>% 
   tiff::readTIFF() %>% 
@@
 
-```{r results, dpi=100, fig.dim=c(6,6), fig.cap="Benchmarking results"}
-# TODO: write caption
-here('figures', 'results.tiff') %>% 
+```{r results_sum, fig.dim=figs('results_sum', 'dim'), fig.cap=figs('results_sum', 'cap')}
+figs('results_sum', 'path') %>% 
   tiff::readTIFF() %>% 

### 2021-06-23 code/R/plot_results.R

@@ -9,46 +9,46 @@ library(knitr)
 library(tidyverse)
 dual_colors <- RColorBrewer::brewer.pal(3, 'Set1')[1:2]
 mutate_perf <- function(dat) {
-  dat %>% 
+  dat %>%
     mutate(mem_mb = max_rss,
-           mem_gb = mem_mb / 1024) %>% 
+           mem_gb = mem_mb / 1024) %>%
     rename(sec = s)
 }
 select_cols <- function(dat) {
-  dat %>% 
+  dat %>%
     select(dataset, strategy, method, tool, mcc, sec, mem_gb, fraction_mapped)
 }
 
-opticlust <- read_tsv(here('subworkflows/1_prep_samples/results/opticlust_results.tsv')) %>% 
-  full_join(read_tsv(here('subworkflows/1_prep_samples/results/dataset_sizes.tsv'))) %>% 
-  mutate_perf() %>% 
+opticlust <- read_tsv(here('subworkflows/1_prep_samples/results/opticlust_results.tsv')) %>%
+  full_join(read_tsv(here('subworkflows/1_prep_samples/results/dataset_sizes.tsv'))) %>%
+  mutate_perf() %>%
   mutate(strategy = method, fraction_mapped = NA)
-optifit_dbs <- read_tsv(here('subworkflows/2_fit_reference_db/results/optifit_dbs_results.tsv')) %>% 
+optifit_dbs <- read_tsv(here('subworkflows/2_fit_reference_db/results/optifit_dbs_results.tsv')) %>%
   mutate_perf()
-optifit_split <- read_tsv(here('subworkflows/3_fit_sample_split/results/optifit_split_results.tsv')) %>% 
-  filter(ref_frac == 0.5, ref_weight == 'simple') %>% 
+optifit_split <- read_tsv(here('subworkflows/3_fit_sample_split/results/optifit_split_results.tsv')) %>%
+  filter(ref_frac == 0.5, ref_weight == 'simple') %>%
   mutate_perf()
-optifit_all <- list(optifit_dbs %>% 
+optifit_all <- list(optifit_dbs %>%
                       mutate(strategy = glue('database_{ref}')),
-                    optifit_split %>% 
-                      mutate(strategy = 'self-split')) %>% 
+                    optifit_split %>%
+                      mutate(strategy = 'self-split')) %>%
   reduce(full_join)
-vsearch <- read_tsv(here('subworkflows/4_vsearch/results/vsearch_results.tsv')) %>% 
-  mutate_perf() %>% 
[…]

### 2021-06-23 code/R/render.R

@@ -1,6 +1,8 @@
 source(snakemake@input[["fcns"]])
 log_smk()
-rmarkdown::render(here::here(snakemake@input[["Rmd"]]),
+rmarkdown::render(
+  here::here(snakemake@input[["Rmd"]]),
+  params = list(smk = snakemake),
   output_format = snakemake@params[["format"]],
   output_file = here::here(snakemake@output[["file"]])
 )

### 2021-06-23 paper/paper.Rmd
Spell out 'snakemake' for Rmd param

@@
 params:
-    smk: "none"
+    snakemake: "none"
 ---
@@
 library(tidyverse)
-snakemake <- params$smk
+snakemake <- params$snakemake
 load(here(snakemake@input[['rda']]))
-```
-```{r figs-meta}
 figures_meta <- yaml::read_yaml(here('config', 'figures.yaml'))

### 2021-06-23 code/R/render.R

@@ -2,7 +2,7 @@ source(snakemake@input[["fcns"]])
 log_smk()
 rmarkdown::render(
   here::here(snakemake@input[["Rmd"]]),
-  params = list(smk = snakemake),
+  params = list(snakemake = snakemake),
   output_format = snakemake@params[["format"]],
   output_file = here::here(snakemake@output[["file"]])
 )

### 2021-06-23 paper/paper.Rmd
Fix figure cross-referencing with pdf_document format

@@
   pdf_document:
-    keep_tex: false
     includes:
@@
       before_body: head.tex
-    number_sections: false
-    toc: false
+    fig_caption: yes
 knit: (function(inputFile, encoding) {
@@
 library(cowplot)
+library(glue)
 library(here)
@@
   value <- figures_meta[[fig_name]][[attribute]]
-  if (attribute != 'cap') {
+  if (attribute == 'cap') {
+    value <- glue("\\label{fig:[fig_name]} [value]", .open='[', .close = ']')
+  } else {
     value <- eval(parse(text = value))
@@
 Reference OTUs for each database were created by performing _de novo_ clustering
-with OptiClust at a distance threshold of 3% (see Figure \@ref(fig:workflow)).
+with OptiClust at a distance threshold of 3% (see Figure \ref{fig:workflow}).
 The _de novo_ MCC scores were `r dn_dbs_mcc` for `r names(dn_dbs_mcc)`, 
@@
 dropped to `r closed_fit_rdp_mcc` when fitting to RDP 
-(see Figure \@ref(fig:results_sum)).
+(see Figure \ref{fig:results_sum}).
 For comparison, clustering datasets _de novo_ with

### 2021-06-23 paper/paper.Rmd
Make figure labels bold

@@
   if (attribute == 'cap') {
-    value <- glue("\\label{fig:[fig_name]} [value]", .open='[', .close = ']')
+    value <- glue("\\label{fig:[fig_name]}[value]", .open='[', .close = ']')
   } else {

### 2021-06-23 paper/paper.Rmd
link citations & remove knit override

@@
 bibliography: references.bib
+link-citations: true
 output:
@@
     fig_caption: yes
-knit: (function(inputFile, encoding) {
-  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs", output_format = "all") })
 csl: msystems.csl

### 2021-06-24 paper/paper.Rmd
Roughhhh draft of figure for split strategy params

@@
 
+```{r results_split, fig.dim=figs('results_split', 'dim'), fig.cap=figs('results_split', 'cap')}
+figs('results_split', 'path') %>% 
+  tiff::readTIFF() %>% 
+  grid::grid.raster() 
+```
+
 While we initially tested this strategy using an even split of the data into
@@
 sequences in the reference and `r sec_fit_split_human_simple_9` with 90% of
-sequences in the reference.
+sequences in the reference (Figure \ref{fig:results_split}).
 In closed reference mode, the fraction of sequences that mapped increased as the

### 2021-06-24 code/R/plot_results_split.R

@@ -1,0 +1,158 @@
+# plot a summary of the results for all datasets & clustering strategies
+
+set.seed(20210509)
+library(cowplot)
+library(ggtext)
+library(glue)
+library(here)
+library(knitr)
+library(tidyverse)
+tri_colors <-  c("darkorange","darkorchid","cyan4")
+color_palette <- RColorBrewer::brewer.pal(4, "Dark2")
+dataset_colors <- c(
+  human = color_palette[[3]],
+  marine = color_palette[[1]],
+  mouse = color_palette[[4]],
+  soil = color_palette[[2]]
+)
+mutate_perf <- function(dat) {
+  dat %>%
+    mutate(mem_mb = max_rss,
+           mem_gb = mem_mb / 1024) %>%
+    rename(sec = s)
+}
+select_cols <- function(dat) {
+  dat %>%
+    select(dataset, strategy, method, tool, mcc, sec, mem_gb, fraction_mapped, 
+           ref_frac, ref_weight)
+}
+
+opticlust <- read_tsv(here('subworkflows/1_prep_samples/results/opticlust_results.tsv')) %>%
+  full_join(read_tsv(here('subworkflows/1_prep_samples/results/dataset_sizes.tsv'))) %>%
+  mutate_perf() %>%
+  mutate(strategy = method, fraction_mapped = NA)
+optifit_split <- read_tsv(here('subworkflows/3_fit_sample_split/results/optifit_split_results.tsv')) %>%
+  mutate_perf() %>%
+  mutate(strategy = 'self-split')
+
+dat <- list(optifit_split) %>% #, opticlust) %>%
+  lapply(select_cols) %>%
+  reduce(bind_rows) %>%
+  mutate(method = as.character(method),
+         strategy = as.character(strategy)) %>%
+  mutate(fraction_mapped = case_when(
+    method %>% as.character() != 'closed'  ~ NA_real_,
+    TRUE                                   ~ fraction_mapped
+  )) %>%
+  pivot_longer(c(mcc, fraction_mapped, sec),
+               names_to = 'metric') %>%
+  mutate(
+    metric = factor(
+      case_when(
+        metric == "mcc"        ~ "MCC",
+        metric == 'fraction_mapped'   ~ "Fraction Mapped",
+        metric == 'sec'        ~ "Runtime (sec)",
+        TRUE                          ~ metric
+      ),
+      levels = c('MCC', 'Fraction Mapped', 'Runtime (sec)')
+    ),
+    strategy = factor(
+      case_when(
[…]

### 2021-06-24 code/R/plot_results_split.R
Use ggtext to color legend text & ditch symbols

@@ -79,7 +79,13 @@ med_iqr <- function(x) {
                     ymin = quantile(x)[2],
                     ymax = quantile(x)[4]))
 }
-
+color_breaks <- list(simple="#FF8C00",
+                     abundance="#9932CC",
+                     similarity="#008B8B")
+color_labels <- lapply(names(color_breaks), 
+                       function(name) { glue("<span style = 'color:{color_breaks[[name]]};'>{name}</span>")
+                         }
+                       ) %>% unlist()
 dat %>% 
   filter((ref_weight == 'simple' | ref_frac == 0.5) | method == "_de novo_") %>% 
   ggplot(aes(ref_frac, value, color = ref_weight, shape = method)) +
@@ -90,11 +96,12 @@ dat %>%
                position = position_dodge(width = 0.05)) +
   facet_grid(dataset ~ metric, scales = 'free', switch = 'x') +
   scale_shape_manual(values = list(open = 1, closed = 19, `_de novo_` = 17)) +
-  scale_color_manual(values = list(simple="darkorange",
-                                   abundance="darkorchid",
-                                   similarity="cyan4",
-                                   `NA`="black"),
-                     breaks = c('simple', 'abundance', 'similarity')) +
+  scale_color_manual(values = list(simple="#FF8C00",
+                                   abundance="#9932CC",
+                                   similarity="#008B8B",
+                                   `NA`="#000000"),
+                     breaks = names(color_breaks),
+                     labels = color_labels) +
   scale_x_continuous(breaks = seq(0, 0.9, 0.1), 
                      limits = c(0, 0.9),
                      labels = c('NA', seq(0.1, 0.9, 0.1))) +
@@ -105,7 +112,11 @@ dat %>%
         legend.position="top",
         legend.margin=margin(t=0, r=0, b=0, l=0, unit='pt'),
         plot.margin=unit(x=c(0,3,0,3),units="pt")
-  )
+  ) + 
+  guides(shape = guide_legend(order = 1),
+         colour = guide_legend(override.aes = list(size = -1),
+                               order = 2)
+         )
 
 dim <- eval(parse(text=snakemake@params[['dim']]))
 ggsave(snakemake@output[['tiff']],

### 2021-06-24 code/R/plot_results_sum.R

@@ -7,7 +7,6 @@ library(glue)
 library(here)
 library(knitr)
 library(tidyverse)
-dual_colors <- RColorBrewer::brewer.pal(3, 'Set1')[1:2]
 mutate_perf <- function(dat) {
   dat %>%
     mutate(mem_mb = max_rss,
@@ -82,6 +81,13 @@ med_iqr <- function(x) {
                     ymax = quantile(x)[4]))
 }
 
+color_list <- list(mothur = RColorBrewer::brewer.pal(3, 'Set1')[1],
+                     vsearch = RColorBrewer::brewer.pal(3, 'Set1')[2])
+color_labels <- lapply(names(color_list), 
+                       function(name) { glue("<span style = 'color:{color_list[[name]]};'>{name}</span>")
+                       }
+) %>% unlist()
+
 mothur_vsearch  %>%
   ggplot(aes(value, strategy, color = tool, shape = method)) +
   # stat_summary(geom = "linerange",
@@ -93,7 +99,8 @@ mothur_vsearch  %>%
                position = position_dodge(width = 0.4)) +
   facet_grid(dataset ~ metric, scales = 'free', switch = 'x') +
   scale_shape_manual(values = list(open = 1, closed = 19, `_de novo_` = 17)) +
-  scale_color_manual(values = dual_colors) +
+  scale_color_manual(values = color_list,
+                     labels = color_labels) +
   labs(x = '', y = '') +
   theme_bw() +
   theme(strip.placement = "outside",
@@ -103,6 +110,10 @@ mothur_vsearch  %>%
         legend.position="top",
         legend.margin=margin(t=0, r=0, b=0, l=0, unit='pt'),
         plot.margin=unit(x=c(3,3,3,0),units="pt")
+  ) + 
+  guides(shape = guide_legend(order = 1),
+         colour = guide_legend(override.aes = list(size = -1),
+                               order = 2)
   )
 
 dim <- eval(parse(text=snakemake@params[['dim']]))

### 2021-06-24 paper/paper.Rmd
Try setting seed in paper to make figure pdfs stop changing

@@
 }
+set.seed(20210624)
 ```

### 2021-06-24 paper/paper.Rmd
Cite ggtext

@@
 Software used includes mothur v1.45.0 [@schloss_introducing_2009], VSEARCH
-v2.13.3 [@rognes_vsearch_2016], numpy [@harris_array_2020], the Tidyverse
-metapackage [@wickham_welcome_2019], R Markdown [@xie_r_2018], the SRA toolkit
-[@noauthor_sra-tools_nodate], and the conda environment manager
-[@noauthor_anaconda_2016].
+v2.13.3 [@rognes_vsearch_2016], numpy [@harris_array_2020], the tidyverse
+metapackage [@wickham_welcome_2019], R Markdown [@xie_r_2018], ggtext
+[@wilke_ggtext_2020], the SRA toolkit [@noauthor_sra-tools_nodate], and the
+conda environment manager [@noauthor_anaconda_2016].
 The complete workflow, manuscript, and conda environment are available at

### 2021-06-24 code/R/plot_results_split.R
Tweak jitter width

@@ -83,8 +83,9 @@ color_breaks <- list(simple="#FF8C00",
                      abundance="#9932CC",
                      similarity="#008B8B")
 color_labels <- lapply(names(color_breaks), 
-                       function(name) { glue("<span style = 'color:{color_breaks[[name]]};'>{name}</span>")
-                         }
+                       function(name) { 
+                         glue("<span style = 'color:{color_breaks[[name]]};'>{name}</span>")
+                       }
                        ) %>% unlist()
 dat %>% 
   filter((ref_weight == 'simple' | ref_frac == 0.5) | method == "_de novo_") %>% 
@@ -93,7 +94,7 @@ dat %>%
   stat_summary(geom = 'point',
                fun = median,
                size = 3,
-               position = position_dodge(width = 0.05)) +
+               position = position_dodge(width = 0.07)) +
   facet_grid(dataset ~ metric, scales = 'free', switch = 'x') +
   scale_shape_manual(values = list(open = 1, closed = 19, `_de novo_` = 17)) +
   scale_color_manual(values = list(simple="#FF8C00",

### 2021-06-25 paper/paper.Rmd
Incorporate most of Pat's edits
Co-authored-by: Pat Schloss <pschloss@users.noreply.github.com>

@@
 
-Assigning amplicon sequences to Operational Taxonomic Units (OTUs) is an
+Assigning amplicon sequences to operational taxonomic units (OTUs) is often an
 important step in characterizing the composition of microbial communities across
@@
 shown to produce higher quality OTU assignments than other methods and at
-comparable or faster speeds [@westcott_opticlust_2017;
-@schloss_introducing_2009].
+comparable or faster speeds.
 A notable difference between _de novo_ clustering and database-dependent methods
-is that OTU assignments clustered with _de novo_ methods are not stable when new
-sequences are added to a dataset [@westcott_novo_2015].
+is that OTU assignments clustered with _de novo_ methods may change when new
+sequences are added to a dataset.
 However, in some cases one may wish to incorporate new samples into a previously
 clustered dataset without performing clustering again on all sequences, such as
-when deploying a machine learning model where OTUs are features
-[@topcuoglu_effective_2019].
+when comparing across datasets or deploying machine learning models where OTUs
+are features.
 To provide an efficient and robust method to fit amplicon sequence data to
-existing OTUs, we developed the OptiFit algorithm as a new component of the
-mothur program.
-We tested OptiFit using four microbiome datasets with two different strategies;
+existing OTUs, we developed the OptiFit algorithm.
+We tested OptiFit using four microbiome datasets with two different strategies:
 by fitting to an external reference database, or by splitting the dataset into a 
 reference and query set and fitting the query sequences to the reference set.
+The result is a robust implementation of closed and open-refernce clustering.
 OptiFit produces OTUs of similar quality as OptiClust and at faster speeds when
-using the split dataset strategy, while the OTU quality and speed depends on the
-database chosen when using the external database strategy.
-OptiFit provides a suitable option for users who require stable reference-based
-OTU assignments at the same quality afforded by _de novo_ clustering methods.
+using the split dataset strategy, although the OTU quality and processing speed
+depends on the database chosen when using the external database strategy.
+OptiFit provides a suitable option for users who require consistent OTU
+assignments at the same quality afforded by _de novo_ clustering methods.
 
@@
 
-Amplicon sequencing has become a mainstay of microbial ecology and
-host-associated microbiome research.
+Amplicon sequencing has become a mainstay of microbial ecology.
 Researchers can affordably generate millions of sequences to characterize the
-composition of hundreds of samples from culture-independent microbial
-communities. 
-In a typical analysis pipeline, 16S rRNA gene sequences are assigned to
-Operational Taxonomic Units (OTUs) to facilitate comparison of taxonomic
-composition between communities.
+composition of hundreds of samples from microbial communities without the need
+for culturing.
+In many analysis pipelines, 16S rRNA gene sequences are assigned to
+operational taxonomic units (OTUs) to facilitate comparison of taxonomic
+composition between communities to avoid the need for classification.
 A distance threshold of 3% (or sequence similarity of 97%) is commonly used to
-cluster sequences into OTUs based on either a reference database or pairwise
-comparisons of the sequences within the dataset.
+cluster sequences into OTUs based on pairwise comparisons of the sequences
+within the dataset.
 The method chosen for clustering affects the quality of OTU assignments and thus
@@
 reference-based.
-OptiClust, the _de novo_ clustering method provided by the mothur software
-program, uses the distance score between all pairs of sequences in the dataset
-to cluster them into OTUs by maximizing the Matthews correlation coefficient
-(MCC) [@westcott_opticlust_2017].
+OptiClust is a _de novo_ clustering method which uses the distance score between
+all pairs of sequences in the dataset to cluster them into OTUs by maximizing
+the Matthews Correlation Coefficient (MCC) [@westcott_opticlust_2017].
 This approach takes into account the distances between all pairs of sequences
-when assigning query sequences to OTUs, in contrast to other tools such as
-VSEARCH which only considers the distance between the query sequence and a
-representative centroid sequence in the OTU.
+when assigning query sequences to OTUs, in contrast to other _de novo_
+algorithms such as VSEARCH which only considers the distance between the query
+sequence and a representative centroid sequence in the OTU [**TODO cite vsearch and schloss papers, also maybe qiime**].
 A limitation of _de novo_ clustering is that different OTU assignments will be
@@
 _de novo_ clustering to compare OTUs between different studies.
-Furthermore, since _de novo_ clustering requires distance calculations between all
-sequences in a dataset, the execution time can be slow for very large datasets.
-
+Furthermore, since _de novo_ clustering requires calculating and comparing
+distances between all sequences in a dataset, the execution time can be slow for
+very large datasets.
 Reference clustering attempts to overcome the limitations of _de novo_
@@
 at 97% similarity is used as the reference with VSEARCH
-[@desantis_greengenes_2006].
+[@desantis_greengenes_2006; **TODO cite vsearch**].
 Query sequences are then assigned to OTUs based on their similarity to the
@@
 Any query sequences that are not within the distance threshold to any of the
-reference sequences are either thrown out (closed reference mode) or clustered
-_de novo_ to create additional OTUs (open reference mode).
+reference sequences are either thrown out (closed reference clustering) or
+clustered _de novo_ to create additional OTUs (open reference clustering).
 While reference-based clustering is generally fast, it is limited by the
@@
 they are not represented by a similar sequence in the database.
+**TODO: closed uses radius of 3%. de novo uses diameter of 3%. open uses both. vsearch is order dependent.**
 Previous studies found that the OptiClust _de novo_ clustering algorithm created
-the highest quality OTU assignments of all clustering methods based on the
-Matthews correlation coefficient (MCC) [@westcott_opticlust_2017].
-As a result, we have recommended OptiClust as the preferred method for OTU
-clustering whenever OTU stability is not required.
+the highest quality OTU assignments of all clustering methods [@westcott_opticlust_2017].
 
-To overcome the limitations of _de novo_ clustering while maintaining OTU
-quality, we developed OptiFit, a reference-based clustering algorithm in the
-mothur program which takes existing OTUs as the reference to fit new sequences
-to. 
+To overcome the limitations of current reference-based and _de novo_ clustering
+algorithms while maintaining OTU quality, we developed OptiFit, a
+reference-based clustering algorithm which uses existing OTUs as the reference
+and fits new sequences those reference OTUs.
 In contrast to other tools, OptiFit considers all pairwise distance scores
@@
 gut [@baxter_microbiota-based_2016] samples. 
+OptiFit is available within the mothur software program.
 
@@
 in a better Matthews correlation coefficient (MCC).
+**TODO: describe how it generates a confusion matrix and calculates the MCC.**
 Iterations continue until the MCC stabilizes or until a maximum number of
@@
 the sequence pairs that are within the distance threshold (e.g. 0.03).
-Initially, query sequences are placed in singleton OTUs. 
+Initially, all query sequences are placed into separate OTUs. 
 Then, the algorithm iteratively reassigns the query sequences to the reference 
@@
 There are two strategies for generating OTUs with OptiFit: 
-1) fit query sequences to reference OTUs generated by _de novo_ clustering an
+1) fit the query sequences to reference OTUs generated by _de novo_ clustering an
 independent database, or
@@
 ### Reference clustering with public databases
+
+```{r db_names}
+db_names <- names(dn_dbs_mcc)
+db_names_upper <- case_when(db_names == 'gg' ~ 'Greengenes', 
+                           TRUE ~ toupper(db_names))
+```
 
@@
 with OptiClust at a distance threshold of 3% (see Figure \ref{fig:workflow}).
-The _de novo_ MCC scores were `r dn_dbs_mcc` for `r names(dn_dbs_mcc)`, 
+The _de novo_ MCC scores were `r dn_dbs_mcc` for `r db_names_upper`,
 respectively.
 Fitting sequences to Greengenes and SILVA in closed reference mode performed
-similarly, with median MCC scores of `r closed_fit_gg_mcc` and 
+similarly, with median MCC scores of `r closed_fit_gg_mcc` (**TODO: should be 0.80**) and 
 `r closed_fit_silva_mcc` respectively, while the median MCC 
-dropped to `r closed_fit_rdp_mcc` when fitting to RDP 
+was `r closed_fit_rdp_mcc` when fitting to RDP 
 (see Figure \ref{fig:results_sum}).
-For comparison, clustering datasets _de novo_ with
-OptiClust produced an average MCC score of `r opticlust_mcc`.
+For comparison, clustering datasets with OptiClust produced an average MCC score
+of `r opticlust_mcc`.
 This gap in OTU quality mostly disappeared when clustering in open reference
-mode, which produced median MCCs of `r open_fit_gg_mcc` with greengenes, 
-`r open_fit_silva_mcc` with SILVA, and `r open_fit_rdp_mcc` with RDP.
+mode, which produced median MCCs of `r open_fit_gg_mcc` with Greengenes, 
+`r open_fit_silva_mcc` with SILVA, and `r open_fit_rdp_mcc` with the RDP.
 Thus, open reference OptiFit produced OTUs of very similar quality as _de novo_
@@
 On average, more sequences were assigned with Greengenes as the reference 
-(`r frac_fit_gg`%) than with SILVA (`r frac_fit_silva`%) or RDP (`r frac_fit_rdp`%).
+(`r frac_fit_gg`%) than with SILVA (`r frac_fit_silva`%) or with the RDP 
+(`r frac_fit_rdp`%).
 This mirrored the result reported above that Greengenes produced better OTUs in
@@
 Despite the drawbacks, closed reference methods have been used when fast
-execution speed is required, such as when using very large datasets.
+execution speed is required, such as when using very large datasets [**TODO: REF**].
 To compare performance in terms of speed, we repeated each OptiFit and OptiClust
@@
 
-To compare to the reference clustering method used by QIIME2, we clustered each
+To compare to the reference clustering methods used by QIIME2, we clustered each
 dataset with VSEARCH against the Greengenes database of OTUs previously 
@@
 VSEARCH against the Greengenes database, with median MCC scores of 
-`r open_fit_gg_mcc` and `r open_vsearch_mcc` (respectively).
+`r open_fit_gg_mcc` and `r open_vsearch_mcc`, respectively).
 In terms of run time, OptiFit outperformed VSEARCH in both closed and open
@@
 `r sec_vsearch_vs_open_fit_db`% on average respectively.
-Thus, the more stringent OTU definition employed by OptiFit resulted in fewer
-sequences being fit to reference OTUs than when using VSEARCH, but caused
-OptiFit to outperform VSEARCH in terms of both OTU quality and execution time.
+Thus, the more stringent OTU definition employed by OptiFit, which requires the
+query to be similar to all other sequences in the OTU rather than to one
+sequence, resulted in fewer sequences being fit to reference OTUs than when
+using VSEARCH, but caused OptiFit to outperform VSEARCH in terms of both OTU
+quality and execution time.
 
@@
 using the split dataset strategy with half of the sequences selected for
-the reference by a simple random sample.
+the reference by a simple random sample (a 50% split).
 OTU quality was highly similar to that from OptiClust regardless of
@@
 
-While we initially tested this strategy using an even split of the data into
-reference and query fractions, we then wanted to investigate whether there was
+While we initially tested this strategy using a 50% split of the data into
+reference and query fractions, we next investigated whether there was
 an optimal reference fraction size.
-To test the best reference size, reference sizes from 10% to 80% of the
+To test the best reference size, reference sets with 10% to 80% of the
 sequences were created, with the remaining sequences used for the query.
@@
 `r frac_fit_split_human_simple_9` with 90% of sequences in the reference. 
-These trends hold for the other datasets as well (see supplemental figure **TODO**).
-Thus, the reference fraction does not affect OTU quality in terms of MCC score,
-but does affect the run time and the fraction of sequences that map during
-closed reference mode.
+These trends held for the other datasets as well (Figure 
+\ref{fig:results_split}).
+Thus, the reference fraction doid not affect OTU quality in terms of MCC score,
+but did affect the run time and the fraction of sequences that mapped during the
+closed reference clustering.
 
@@
 We tested three methods for selecting the fraction of sequences to be used as
-the reference at a size of 50%; a simple random sample, weighting sequences by
+the reference at a size of 50%: a simple random sample, weighting sequences by
 relative abundance, and weighting by similarity to other sequences in the
@@
 seconds compared to `r sec_fit_split_simple` and `r sec_fit_split_abun` seconds
-for simple and abundance-weighted sampling (respectively).
+for simple and abundance-weighted sampling, respectively.
 Thus, employing more complicated sampling strategies such as abundance-weighted
@@
 databases, which are unlikely to contain sequences from rare and novel microbes.
-This drop in quality was most notable with RDP, which contains only about 21,000
-sequences compared to over 200,000 sequences in SILVA and Greengenes each.
+This drop in quality was most notable with RDP, which contains only 21,000 (**TODO insert exact number**)
+sequences compared to over 200,000 (**TODO insert exact number**) sequences in SILVA and Greengenes each.
 Note that Greengenes has not been updated since 2013 at the time of this
@@
 large databases with regular updates and good coverage of microbial diversity.
-Since OptiClust performs faster than open reference OptiFit and creates higher
-quality OTUs than closed reference OptiFit with the database strategy, we
+Since OptiClust still performs faster than open reference OptiFit and creates
+higher quality OTUs than closed reference OptiFit with the database strategy, we
 recommend using OptiClust rather than fitting to a database whenever stable OTUs
@@
 
-The OptiClust and OptiFit algorithms provided by mothur produced higher quality
+The OptiClust and OptiFit algorithms produced higher quality
 OTUs than VSEARCH in open reference, closed reference, or _de novo_ modes.
@@
 fit to OTUs.
-Users who require closed reference clustering to the Greengenes database may
-prefer to use VSEARCH if they wish to maximize the fraction of sequences that
-can be fit at the cost of OTU quality.
-However, mothur's OptiClust or OptiFit are recommended for _de novo_ or open
-reference clustering to produce OTUs of the highest possible quality.
+**TODO: another problem with vsearch is dependence on order of the ref sequences.**
 
@@
 quality OTUs using already existing OTUs as a reference.
-Unlike existing reference-based methods which map query sequences to a single
+Unlike existing reference-based methods that map query sequences to a single
 centroid sequence in each reference OTU, OptiFit considers all sequences in each
@@
 gut [@baxter_microbiota-based_2016] samples.
+These were selected because **TODO**.
 Raw sequences were processed using mothur according to the Schloss Lab MiSeq SOP
@@
 When reference clustering with VSEARCH, processed datasets were fit directly to
-the unprocessed Greengenes reference alignment, since this method is how VSEARCH
-is typically used by the QIIME2 software reference-based clustering
+the unprocessed Greengenes 97% OTU reference alignment, since this method is how
+VSEARCH is typically used by the QIIME2 software reference-based clustering
 [@bolyen_reproducible_2019; @noauthor_clustering_nodate].
@@
 We used three methods for selecting the fraction of sequences to be used
-as the reference; a simple random sample, weighting sequences by relative
+as the reference: a simple random sample, weighting sequences by relative
 abundance, and weighting by similarity to other sequences in the dataset.
@@
 KLS received support from the NIH Training Program in Bioinformatics (T32 GM070449).
-
-PDS received support from **TODO: Pat's grant(s)**.
+Salary support for PDS came from NIH grants R01CA215574 and U01AI124255.
 

### 2021-06-25 paper/paper.Rmd
Cite USEARCH

@@
 <!--
-- **"AND, BUT, THEREFORE" structure. start paragraph with question, end with why we should care. transitions to move the story along.**
-
 - Go back through https://peerj.com/articles/1487/ to get all the arguments for why the qiime based methods are problematic…
@@
 
-**TODO**
+**TODO.**
+**"AND, BUT, THEREFORE" structure. start paragraph with question, end with why we should care. transitions to move the story along.**
 
@@
 when assigning query sequences to OTUs, in contrast to other _de novo_
-algorithms such as VSEARCH which only considers the distance between the query
-sequence and a representative centroid sequence in the OTU [**TODO cite vsearch and schloss papers, also maybe qiime**].
+methods such as the greedy clustering algorithms implemented by USEARCH and
+VSEARCH, which only considers the distance between the query sequence and a
+representative centroid sequence in the OTU [@edgar_search_2010;
+@rognes_vsearch_2016].
 A limitation of _de novo_ clustering is that different OTU assignments will be
@@
 at 97% similarity is used as the reference with VSEARCH
-[@desantis_greengenes_2006; **TODO cite vsearch**].
+[@desantis_greengenes_2006; @rognes_vsearch_2016; @noauthor_clustering_nodate].
 Query sequences are then assigned to OTUs based on their similarity to the
@@
 The complete workflow, manuscript, and conda environment are available at
-**TODO: UPDATED REPO LINK**.
+https://github.com/SchlossLab/Sovacool_OptiFit_2021.
+<!-- TODO: update repo link with journal title -->
 

### 2021-06-25 paper/paper.Rmd
Minor edits

@@
 reference and query set and fitting the query sequences to the reference set.
-The result is a robust implementation of closed and open-refernce clustering.
+The result is a robust implementation of closed and open-reference clustering.
 OptiFit produces OTUs of similar quality as OptiClust and at faster speeds when
@@
 **TODO.**
+<!--
 **"AND, BUT, THEREFORE" structure. start paragraph with question, end with why we should care. transitions to move the story along.**
-
+Advancements in sequencing technology have allowed researchers to affordably generate millions of reads of microorganisms from diverse natural communities,
+allowing us to characterize communities.
+-->
 \newpage
@@
 sequences were created, with the remaining sequences used for the query.
-OTU quality was remarkably stable across reference fraction sizes.
+OTU quality was remarkably consistent across reference fraction sizes.
 For example, splitting the human dataset 100 times yielded a coefficient of
@@
 higher quality OTUs than closed reference OptiFit with the database strategy, we
-recommend using OptiClust rather than fitting to a database whenever stable OTUs
-are not required for the study at hand.
+recommend using OptiClust rather than fitting to a database whenever consistent
+OTUs are not required for the study at hand.
 
@@
 such as when comparing OTUs across studies.
-However, when stable OTUs are not required, we recommend using OptiClust for _de
-novo_ clustering over the split strategy with OptiFit since OptiClust is simpler
-to execute but performs similarly in terms of both run time and OTU quality.
+However, when consistent OTUs are not required, we recommend using OptiClust for
+_de novo_ clustering over the split strategy with OptiFit since OptiClust is
+simpler to execute but performs similarly in terms of both run time and OTU
+quality.
 
@@
 OptiFit fills the missing option for fitting query sequences to existing OTUs
-that does not sacrifice OTU quality for stability of OTU assignments.
+that does not sacrifice OTU quality for consistency of OTU assignments.
 
@@
 gut [@baxter_microbiota-based_2016] samples.
-These were selected because **TODO**.
-Raw sequences were processed using mothur according to the Schloss Lab MiSeq SOP
-as described in the mothur wiki and accompanying study by Kozich _et al._
+These datasets represent a selection of the broad types of natural communities
+that microbial ecologists study.
+We processed the raw sequences using mothur according to the Schloss Lab MiSeq
+SOP as described in the mothur wiki and accompanying study by Kozich _et al._
 [@schloss_miseq_nodate; @kozich_development_2013].

### 2021-06-25 code/R/plot_results_split.R
Reduce margins to zero

@@ -110,7 +110,7 @@ dat %>%
         legend.title = element_blank(),
         legend.position="top",
         legend.margin=margin(t=0, r=0, b=0, l=0, unit='pt'),
-        plot.margin=unit(x=c(0,3,0,3),units="pt")
+        plot.margin=unit(x=c(0,0,0,0),units="pt")
   ) + 
   guides(shape = guide_legend(order = 1),
          colour = guide_legend(override.aes = list(size = -1),

### 2021-06-25 code/R/plot_results_sum.R

@@ -110,7 +110,7 @@ mothur_vsearch  %>%
         legend.text = element_markdown(),
         legend.position="top",
         legend.margin=margin(t=0, r=0, b=0, l=0, unit='pt'),
-        plot.margin=unit(x=c(3,3,3,0),units="pt")
+        plot.margin=unit(x=c(0,0,0,0),units="pt")
   ) + 
   guides(shape = guide_legend(order = 1),
          colour = guide_legend(override.aes = list(size = -1),

### 2021-06-25 paper/paper.Rmd
Expand on vsearch limitations & cite more stuff

@@
 reference-based.
-OptiClust is a _de novo_ clustering method which uses the distance score between
-all pairs of sequences in the dataset to cluster them into OTUs by maximizing
-the Matthews Correlation Coefficient (MCC) [@westcott_opticlust_2017].
+OptiClust is a _de novo_ clustering algorithm which uses the distance score
+between all pairs of sequences in the dataset to cluster them into OTUs by
+maximizing the Matthews Correlation Coefficient (MCC)
+[@westcott_opticlust_2017].
 This approach takes into account the distances between all pairs of sequences
 when assigning query sequences to OTUs, in contrast to other _de novo_
-methods such as the greedy clustering algorithms implemented by USEARCH and
-VSEARCH, which only considers the distance between the query sequence and a
+methods such as the greedy clustering algorithms implemented in USEARCH and
+VSEARCH, which only consider the distance between the query sequence and a
 representative centroid sequence in the OTU [@edgar_search_2010;
@@
 _de novo_ clustering to compare OTUs between different studies.
+Additionally, the greedy clustering algorithms implemented in VSEARCH are
+sensitive to the order of the input sequences: different OTU assignments are
+produced when the same sequences are randomly shuffled [@westcott_novo_2015;
+@he_stability_2015].
 Furthermore, since _de novo_ clustering requires calculating and comparing
@@
 they are not represented by a similar sequence in the database.
-**TODO: closed uses radius of 3%. de novo uses diameter of 3%. open uses both. vsearch is order dependent.**
+**TODO: closed uses radius of 3%. de novo uses diameter of 3%. open uses both.**
 Previous studies found that the OptiClust _de novo_ clustering algorithm created
@@
 Despite the drawbacks, closed reference methods have been used when fast
-execution speed is required, such as when using very large datasets [**TODO: REF**].
+execution speed is required, such as when using very large datasets
+[@navas-molina_chapter_2013].
 To compare performance in terms of speed, we repeated each OptiFit and OptiClust
@@
 an optimal reference fraction size.
-To test the best reference size, reference sets with 10% to 80% of the
+To test the best reference size, reference sets with 10% to 90% of the
 sequences were created, with the remaining sequences used for the query.
@@
 In contrast, VSEARCH only requires each query sequence to be similar to the
-single sequence that seeded the OTU.
+single centroid sequence that seeded the OTU.
+Additionally, OTU assignments clustered by VSEARCH are dependent on the order of
+the input sequences, because each query is assigned to the OTU of the first
+centroid sequence that is found within the distance threshold.
 In this way, VSEARCH sacrifices OTU quality in order to allow more sequences to
 fit to OTUs.
-**TODO: another problem with vsearch is dependence on order of the ref sequences.**
 

### 2021-06-28 paper/paper.Rmd
Insert actual numbers for ref db sizes

@@
 library(tidyverse)
+set.seed(20210624)
 snakemake <- params$snakemake
 load(here(snakemake@input[['rda']]))
+ref_seqs <- read_tsv(here('subworkflows', '0_prep_db', 'data', 
+                          'seq_counts.tsv')) %>% 
+  pivot_wider(names_from = ref, values_from = nseqs) %>% as.list()
+
 figures_meta <- yaml::read_yaml(here('config', 'figures.yaml'))
@@
 }
-set.seed(20210624)
 ```
@@
 _de novo_ clustering to compare OTUs between different studies.
-Additionally, the greedy clustering algorithms implemented in VSEARCH are
-sensitive to the order of the input sequences: different OTU assignments are
-produced when the same sequences are randomly shuffled [@westcott_novo_2015;
-@he_stability_2015].
+Additionally, the greedy clustering algorithms are sensitive to the order of the
+input sequences: different OTU assignments are produced when the same sequences
+are randomly shuffled [@westcott_novo_2015; @he_stability_2015].
 Furthermore, since _de novo_ clustering requires calculating and comparing
@@
 Fitting sequences to Greengenes and SILVA in closed reference mode performed
-similarly, with median MCC scores of `r closed_fit_gg_mcc` (**TODO: should be 0.80**) and 
+similarly, with median MCC scores of `r round(closed_fit_gg_mcc, 3)` (**TODO: Pat marked this as 0.80, but the actual number is 0.795414**) and 
 `r closed_fit_silva_mcc` respectively, while the median MCC 
@@
 databases, which are unlikely to contain sequences from rare and novel microbes.
-This drop in quality was most notable with RDP, which contains only 21,000 (**TODO insert exact number**)
-sequences compared to over 200,000 (**TODO insert exact number**) sequences in SILVA and Greengenes each.
+This drop in quality was most notable with the RDP reference, which contained only 
+`r ref_seqs[['rdp']]` sequences compared to `r ref_seqs[['silva']]` sequences in
+SILVA and `r ref_seqs[['gg']]` in Greengenes after trimming to the V4 region.
 Note that Greengenes has not been updated since 2013 at the time of this

### 2021-06-28 paper/paper.Rmd
Describe MCC calculation

@@
 a different OTU or remain in its current OTU, choosing the option that results
-in a better Matthews correlation coefficient (MCC).
-**TODO: describe how it generates a confusion matrix and calculates the MCC.**
-Iterations continue until the MCC stabilizes or until a maximum number of
-iterations is reached.
+in a better Matthews correlation coefficient (MCC) [@westcott_opticlust_2017].
+The MCC uses all values from a confusion matrix and ranges from zero to one,
+with a score of one occurring when all sequence pairs are true positives and
+true negatives, and a score of zero when all pairs are false positives and false
+negatives.
+Sequence pairs that are similar to each other (i.e. within the distance
+threshold) are counted as true positives if they are assigned to the same OTU,
+and false negatives if they are not in the the same OTU.
+Sequence pairs that are not similar to each other are true negatives if they are
+not assigned to the same OTU, and false positives if they are not in the same
+OTU.
+OptiClust iterations continue until the MCC stabilizes or until a maximum number
+of iterations is reached.
 This process produces _de novo_ OTU assignments with the most optimal MCC given
 the input sequences.
+
 OptiFit begins where OptiClust ends, starting with a list of reference OTUs and
@@
 when the sequence is a singleton rather than assigned to a reference OTU.
+All query and reference sequence pairs are considered when calculating the MCC.
 This process is repeated until the MCC changes by no more than 0.0001 (default)

### 2021-06-28 code/R/plot_algorithm_diagram.R
Insert placeholder for algorithm diagram

@@ -7,7 +7,9 @@ library(tidyverse)
 set.seed(20200308)
 
 # blank placeholder plot for now
-ggplot() + theme_void() + annotate('text', x=1, y=1, label='insert algorithm diagram here')
+ggplot() + theme_void() + annotate('text', x=1, y=1, 
+                                   label='insert algorithm diagram here', 
+                                   size=8, angle=45)
 dims <- eval(parse(text=snakemake@params[['dim']]))
 ggsave(snakemake@output[['tiff']],
        device = 'tiff', dpi=300,

### 2021-06-29 exploratory/2021/algorithm_diagram.Rmd
Experimenting with ggraph for algorithm diagram

---
date: '`r Sys.Date()`'
output: 
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      fig.path = "figures/"
                      )
```

```{r lib}
library(tidyverse)
library(diagram)
```

```{r functions}
# Pat's MCC function
# source: https://github.com/SchlossLab/Westcott_OptiClust_mSphere_2017/blob/a8bc26855423bba85acc0b8e7cca075e5c94f533/submission/supplemental_text.Rmd#L26-L28
mcc <- function(tp, tn, fp, fn) {
  format(round((tp * tn - fp * fn) / 
                 sqrt((tp + fp) * (tp + fn) * (tn + fp) * (tn + fn)), 
               digits = 2),
  digits = 2,
  nsmall = 2L
  )
}

calc_mcc_from_conf_mat <- function(conf_mat) {
  tp <- conf_mat[1, 1]
  tn <- conf_mat[2, 2]
  fp <- conf_mat[1, 2]
  fn <- conf_mat[2, 1]
  return(mcc(tp, tn, fp, fn))
}

get_seqs_to_otus <- function(otu_list) {
  seqs_to_otus <- list()
  for (i in 1:length(otu_list)) {
    otu <- otu_list[[i]]
    for (seq in otu) {
      seqs_to_otus[seq] <- i
    }
  }
  return(seqs_to_otus)
}

get_otu_label <- function(otus, idx) {
  x <- otus[[idx]]
  paste0(x, collapse = ",")
}

# assignments of sequences in OTUs as matrix
get_otu_mat_from_list <- function(otu_list, seqs) {
  otu_mat <- matrix(
    nrow = length(seqs), ncol = length(seqs),
    dimnames = list(seqs, seqs), data = 0
  )
  for (i in 1:length(otu_list)) {
    otu <- otu_list[[i]]
    for (seq1 in otu) {
      for (seq2 in otu) {
        otu_mat[seq1, seq2] <- 1
        otu_mat[seq2, seq1] <- 1
      }
    }
  }
  return(otu_mat)
}

get_condition <- function(pair) {
  seq1 <- pair[1]
  seq2 <- pair[2]
  in_dist <- dist_mat[seq1, seq2]
  in_otu <- otu_mat[seq1, seq2]
  return(c(actual = in_dist, pred = in_otu))
}

# compare otu_mat to dist_mat to get tp/tn/fp/fn values
get_conf_mat <- function(pairs, dist_mat, otu_mat) {
  conditions <- apply(seq_pairs, 2, get_condition)
  conf_mat <- caret::confusionMatrix(table(
    conditions["pred", ],
    conditions["actual", ]
  )) %>%
    as.matrix()
  return(conf_mat)
}
```

```{r opticlust}
# example data
seqs <- LETTERS[1:17]
seq_pairs <- combn(seqs, 2)
# distances within threshold
dat <- data.frame(
  Seq1 = c("D", "F", "G", "H", "I", "I", "J", "J", "N", "O", "P", "P", "P", "Q", "Q"),
  Seq2 = c("B", "E", "C", "A", "B", "D", "A", "H", "M", "L", "K", "L", "O", "E", "F"),
  Distance = c(0.024, 0.028, 0.028, 0.027, 0.016, 0.024, 0.028, 0.024, 0.024, 0.024, 0.016, 0.027, 0.027, 0.024, 0.028)
)

# matrix of pairwise sequences within distance threshold
dist_mat <- matrix(
  nrow = length(seqs), ncol = length(seqs),
  dimnames = list(seqs, seqs), data = 0
)
diag(dist_mat) <- 1
for (i in 1:nrow(dat)) {
  r <- dat[i, ]
  seq1 <- r[["Seq1"]]
  seq2 <- r[["Seq2"]]
  dist_mat[seq1, seq2] <- 1
  dist_mat[seq2, seq1] <- 1
}

# example otu assignment
otu_list <- list(
  c("B", "D", "I"),
  c("E", "F"),
  c("C", "G"),
  c("A", "H", "J"),
  c("M", "N"),
  c("L", "O", "P"),
  c("K"),
  c("Q")
)
seqs_to_otus <- get_seqs_to_otus(otu_list)
otu_mat <- get_otu_mat_from_list(otu_list, seqs)
current_mcc <- calc_mcc_from_conf_mat(get_conf_mat(seq_pairs, dist_mat, otu_mat))

# Pat's method for plotting the network graph
# last example from OptiClust supplemental text
tp <- 12
tn <- 1210
fp <- 0
fn <- 3
names <- c(	"B,D,I", "E,F,Q", "C,G", "A,H,J", "M,N", "L,O,P", "K", "...")
n_seqs <- length(names)
M <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(M) <- colnames(M) <- names
C <- matrix(nrow = n_seqs, ncol = n_seqs, byrow = TRUE, data = 0)
rownames(C) <- colnames(C) <- names

M["K","K"] <- mcc(tp=tp, tn=tn, fp=fp, fn=fn)
M["L,O,P", "K"] <- mcc(tp=tp+1, tn=tn-2, fp=fp+2, fn=fn-1)
C["L,O,P", "K"] <- 0.4

par(mar=c(1,2,1,2), xpd=T)

diagram::plotmat(M, pos=n_seqs, name = names, lwd = 1, curve=C, box.lwd = 2, 
        cex.txt = 0.6, box.size = 0.03, box.type = "circle", shadow.size=0, 
        arr.type="triangle", dtext=0.5, self.shiftx =-0.02, self.shifty=-0.04,
        xpd=T, box.cex=0.6, arr.length=0.3)

```

## ggraph

can I make part of a node label bold?
I modified `ggraph::geom_node_label` to use `ggtext::geom_richtext` by default.

```{r ggraph_mod}
# devtools::install_github('kelly-sovacool/ggraph', ref = 'iss-297_ggtext')
#library(ggraph) # something is wrong & this fails, but load_all() works!
devtools::load_all('../ggraph')
library(ggtext)
library(tidygraph)
```
```{r ggraph_exp}
graph <-
  tbl_graph(nodes = data.frame(name = c("A B C", "D E F", "G **H** I", "J K L")),
            edges = data.frame(from = c(2, 2, 2), 
                               to = c(3, 4, 2), 
                               mcc = c(0.97, 0.84, 0.80)))
ggraph(graph, layout = 'linear') +
  geom_edge_arc(aes(label = mcc,
                    start_cap = label_rect(node1.name),
                    end_cap = label_rect(node2.name)), 
                arrow = arrow(length = unit(4, 'mm')),
                edge_colour = 'gray',
                angle_calc = 'along',
                label_dodge = unit(2, 'mm')
                ) +
  geom_edge_loop(aes(label = mcc,
                     span=120,
                     direction=270,
                     strength=0.5
                     ),
                 angle_calc = 'along',
                 label_dodge = unit(-2, 'mm'),
                 edge_colour = 'gray') +
  geom_node_label(aes(label = name), richtext = TRUE) +
  theme(panel.background = element_rect(fill='white'))
```

### 2021-06-30 exploratory/2021/algorithm_diagram.Rmd
Use patchwork to stack graphs... need to enforce node order

@@
 library(tidygraph)
-```
-```{r ggraph_exp}
-graph <-
+library(patchwork)
+g1 <-
   tbl_graph(nodes = data.frame(name = c("A B C", "D E F", "G **H** I", "J K L")),
@@
                                mcc = c(0.97, 0.84, 0.80)))
-ggraph(graph, layout = 'linear') +
+g2 <-
+  tbl_graph(nodes = data.frame(name = c("A B C", "D **E** F", "G H I", "J K L")),
+            edges = data.frame(from = c(2, 2, 2), 
+                               to = c(1, 3, 2), 
+                               mcc = c(0.94, 0.81, 0.86)))
+g2_ranks <-
+  tbl_graph(nodes = data.frame(name = c("A B C", "D **E** F", "G H I", "J K L")),
+            edges = data.frame(from = c(1, 2, 3), 
+                               to = c(2, 3, 4)))
+```
+
+```{r ggraph_exp, fig.dim=c(6.8,1)}
+plot_graph <- function(graph, ranks) {
+  ggraph(graph) +
   geom_edge_arc(aes(label = mcc,
@@
                     end_cap = label_rect(node2.name)), 
-                arrow = arrow(length = unit(4, 'mm')),
+                arrow = arrow(length = unit(4, 'mm'),
+                              type = 'closed'),
                 edge_colour = 'gray',
                 angle_calc = 'along',
-                label_dodge = unit(2, 'mm')
+                label_dodge = unit(-2, 'mm')
                 ) +
-  geom_edge_loop(aes(label = mcc,
-                     span=120,
-                     direction=270,
-                     strength=0.5
-                     ),
-                 angle_calc = 'along',
-                 label_dodge = unit(-2, 'mm'),
-                 edge_colour = 'gray') +
-  geom_node_label(aes(label = name), richtext = TRUE) +
-  theme(panel.background = element_rect(fill='white'))
+    geom_edge_link(data = ranks) +
+  geom_node_label(aes(label = name)) +
+  theme_void() +
+  theme(plot.margin=unit(x=c(0,0,0,0),units="pt"))
+}
+
+pg1 <- plot_graph(g1, g1_ranks) 
+pg2 <- plot_graph(g2, g2_ranks)
+
+pg1 / pg2
 ```

### 2021-07-01 exploratory/2021/algorithm_diagram.Rmd
Enforce node order

--- 2021-06-29 exploratory/2021/algorithm_diagram.Rmd
@@
 library(tidygraph)
-```
-```{r ggraph_exp}
-graph <-
-  tbl_graph(nodes = data.frame(name = c("A B C", "D E F", "G **H** I", "J K L")),
+library(patchwork)
+g1 <-
+  tbl_graph(nodes = data.frame(name = c("A B C", "D E F", "G **H** I", "J K L"),
+                               i = 1:4),
             edges = data.frame(from = c(2, 2, 2), 
                                to = c(3, 4, 2), 
-                               mcc = c(0.97, 0.84, 0.80)))
-ggraph(graph, layout = 'linear') +
+                               mcc = c(0.97, 0.84, 0.80)) %>% 
+              mutate(is_link = from == to))
+g2 <-
+  tbl_graph(nodes = data.frame(name = c("A B C", "D **E** F", "G H I", "J K L"),
+                               i = 1:4),
+            edges = data.frame(from = c(2, 2, 2), 
+                               to = c(1, 3, 2), 
+                               mcc = c(0.94, 0.81, 0.86)) %>% 
+              mutate(is_link = from == to))
+g2_ranks <-
+  tbl_graph(nodes = data.frame(name = c("A B C", "D **E** F", "G H I", "J K L")),
+            edges = data.frame(from = c(1, 2, 3), 
+                               to = c(2, 3, 4)))
+```
+
+```{r ggraph_exp, fig.dim=c(6.8,1)}
+plot_graph <- function(graph) {
+  create_layout(graph, 'linear', sort.by = i) %>% 
+  ggraph() +
   geom_edge_arc(aes(label = mcc,
@@
                     end_cap = label_rect(node2.name)), 
-                arrow = arrow(length = unit(4, 'mm')),
+                arrow = arrow(length = unit(4, 'mm'),
+                              type = 'closed'),
                 edge_colour = 'gray',
                 angle_calc = 'along',
-                label_dodge = unit(2, 'mm')
+                label_dodge = unit(-2, 'mm')
                 ) +
-  geom_edge_loop(aes(label = mcc,
-                     span=120,
-                     direction=270,
-                     strength=0.5
-                     ),
-                 angle_calc = 'along',
-                 label_dodge = unit(-2, 'mm'),
-                 edge_colour = 'gray') +
-  geom_node_label(aes(label = name), richtext = TRUE) +
-  theme(panel.background = element_rect(fill='white'))
+  geom_edge_loop(aes(span = 1, direction = 270, strength = 0.1, color = is_link)) +
+  geom_node_label(aes(label = name)) +
+    scale_color_manual(c('black', 'white')) +
+  theme_void() +
+  theme(plot.margin=unit(x=c(0,0,0,0), units="pt"))
+}
+
+pg1 <- plot_graph(g1) 
+pg2 <- plot_graph(g2)
+
+pg1 / pg2
 ```

### 2021-07-02 exploratory/2021/algorithm_diagram.Rmd
playing around with layouts for algorithm diagram

--- 2021-06-29 exploratory/2021/algorithm_diagram.Rmd
@@
 # distances within threshold
-dat <- data.frame(
+dist_dat <- data.frame(
   Seq1 = c("D", "F", "G", "H", "I", "I", "J", "J", "N", "O", "P", "P", "P", "Q", "Q"),
@@
 diag(dist_mat) <- 1
-for (i in 1:nrow(dat)) {
-  r <- dat[i, ]
+for (i in 1:nrow(dist_dat)) {
+  r <- dist_dat[i, ]
   seq1 <- r[["Seq1"]]
@@
 library(tidygraph)
-```
-```{r ggraph_exp}
-graph <-
-  tbl_graph(nodes = data.frame(name = c("A B C", "D E F", "G **H** I", "J K L")),
+library(patchwork)
+library(gridExtra)
+g1 <-
+  tbl_graph(nodes = data.frame(name = c("A B C", "D E F", "G **H** I", "J K L"),
+                               i = 1:4),
             edges = data.frame(from = c(2, 2, 2), 
                                to = c(3, 4, 2), 
-                               mcc = c(0.97, 0.84, 0.80)))
-ggraph(graph, layout = 'linear') +
+                               mcc = c(0.97, 0.84, 0.80)) %>% 
+              mutate(is_loop = from == to))
+g2 <-
+  tbl_graph(nodes = data.frame(name = c("A B C", "D **E** F", "G H I", "J K L"),
+                               i = 1:4),
+            edges = data.frame(from = c(2, 2, 2), 
+                               to = c(1, 3, 2), 
+                               mcc = c(0.94, 0.81, 0.86)) %>% 
+              mutate(is_loop = from == to))
+```
+
+```{r ggraph_exp, fig.dim=c(6,5), dpi=300}
+plot_graph <- function(graph) {
+  create_layout(graph, 'linear', sort.by = i) %>% 
+  ggraph() +
   geom_edge_arc(aes(label = mcc,
@@
                     end_cap = label_rect(node2.name)), 
-                arrow = arrow(length = unit(4, 'mm')),
+                arrow = arrow(length = unit(4, 'mm'),
+                              type = 'closed'),
                 edge_colour = 'gray',
                 angle_calc = 'along',
-                label_dodge = unit(2, 'mm')
+                label_dodge = unit(-2, 'mm')
                 ) +
-  geom_edge_loop(aes(label = mcc,
-                     span=120,
-                     direction=270,
-                     strength=0.5
-                     ),
-                 angle_calc = 'along',
-                 label_dodge = unit(-2, 'mm'),
-                 edge_colour = 'gray') +
-  geom_node_label(aes(label = name), richtext = TRUE) +
-  theme(panel.background = element_rect(fill='white'))
+  geom_edge_loop(aes(span = 1, direction = 270, strength = 0.2, color = is_loop)) +
+  geom_node_label(aes(label = name)) +
+    scale_edge_color_manual(values = c('white')) +
+  theme_void() +
+  theme(plot.margin=unit(x=c(0,0,10,0), units="pt"),
+        legend.position = 'none')
+}
+
+pg1 <- plot_graph(g1) 
+pg2 <- plot_graph(g2)
+
+wrap_elements(tableGrob(t(dist_dat),
+                        theme = ttheme_default(base_size = 9, 
+                                               padding = unit(c(4,8), 'pt')))
+              ) /
+((pg1 / pg2 / pg1 / pg2) + 
+  plot_annotation(tag_levels = '1')) +
+plot_layout(heights = c(1,4))
+
 ```

### 2021-07-13 code/py/algorithm_diagram.py
Calculate conf mat & mcc

@@ -25,16 +25,35 @@ class otuMap(dict):
             otu_dict_set[otu_id].add(seq)
         return otu_dict_set
 
-    def conf_mat(self, dist_mat): # TODO
+    def conf_mat(self, dist_mat):
         """
-        :param dist_mat: dict of sets of the form {seqA: {seqB, seqC}, seqB: {seqA}}
+        :param dist_mat: dict of sets, e.g. {seqA: {seqB, seqC}, seqB: {seqA}}
         """
-        for seq1, seq2 in combinations(self.keys(), 2)
+        conf_mat = {'tp': 0, 'tn': 0, 'fp': 0, 'fn': 0}
+        for seq1, seq2 in combinations(self.keys(), 2):
+            # are they actually similar (true assignment)
+            true_otu = (seq2 in dist_mat[seq1]) or (seq1 in dist_mat[seq2])
+            # were they assigned to the same OTU (predicted assignment)
+            pred_otu = self[seq1] == self[seq2]
 
-    def mcc(self, dist_mat): # TODO
+            if true_otu and pred_otu:
+                result = 'tp'
+            elif true_otu and not pred_otu:
+                result = 'fn'
+            elif not true_otu and pred_otu:
+                result = 'fp'
+            elif not true_otu and not pred_otu:
+                result = 'tn'
+            else:
+                raise ValueError("this should never happen")
+            conf_mat[result] += 1
+        return conf_mat
+
+    def mcc(self, conf_mat):
         """
-        Calculate the Matthews Correlation coefficient given a distance matrix.
+        Calculate the Matthews Correlation coefficient given a confusion matrix.
         """
+        return round((tp * tn - fp * fn) / sqrt((tp + fp) * (tp + fn) * (tn + fp) * (tn + fn)), 2)
 
 
 class OptiFit: # TODO

### 2021-07-14 code/py/algorithm_diagram.py
Work on toy example

@@ -3,12 +3,23 @@
 from collections import defaultdict
 from itertools import combinations
 
+
+def mcc(conf_mat):
+    """
+    Calculate the Matthews Correlation coefficient given a confusion matrix.
+    """
+    return round((tp * tn - fp * fn) / sqrt((tp + fp) * (tp + fn) * (tn + fp) * (tn + fn)), 2)
+
+
 class otuMap(dict):
     def __init__(self):
         """
         Maps sequences to OTU assignments. The dict should be of the form {seqID: otuIndex}, e.g. {'seqA': 1, 'seqB': 2, 'seqC': 3}
         """
         super().__init__()
+
+    def __repr__(self):
+        return "%s(%r)" % (self.__class__, self.__dict__.items())
 
     @classmethod
     def from_list(cls, otu_list):
@@ -20,16 +31,23 @@ class otuMap(dict):
 
     @property
     def otu_to_seq(self):
+        """
+        :return: dictionary of sets, with keys as OTU IDs and values as sets containing sequence IDs.
+        """
         otu_dict_set = defaultdict(set)
         for seq, otu_id in self.items():
             otu_dict_set[otu_id].add(seq)
         return otu_dict_set
 
-    def conf_mat(self, dist_mat):
+    def conf_mat(self, dist_mat, baseline_conf_mat = False):
         """
         :param dist_mat: dict of sets, e.g. {seqA: {seqB, seqC}, seqB: {seqA}}
+        :param baseline_conf_mat: provide a confusion matrix (as a dictionary) with starting values other than zero.
         """
-        conf_mat = {'tp': 0, 'tn': 0, 'fp': 0, 'fn': 0}
+        if not baseline_conf_mat:
+            conf_mat = {'tp': 0, 'tn': 0, 'fp': 0, 'fn': 0}
+        else:
+            conf_mat = baseline_conf_mat
         for seq1, seq2 in combinations(self.keys(), 2):
             # are they actually similar (true assignment)
             true_otu = (seq2 in dist_mat[seq1]) or (seq1 in dist_mat[seq2])
@@ -49,15 +67,22 @@ class otuMap(dict):
             conf_mat[result] += 1
         return conf_mat
 
-    def mcc(self, conf_mat):
-        """
-        Calculate the Matthews Correlation coefficient given a confusion matrix.
-        """
[…]

### 2021-07-16 exploratory/2021/algorithm_diagram.Rmd
May try using reticulate
Specify python path
correct conf mat

@@
 library(diagram)
+library(reticulate)
+use_python('~/miniconda3/bin/python')
 ```
@@
 # last example from OptiClust supplemental text
-tp <- 12
+tp <- 14
 tn <- 1210
 fp <- 0
-fn <- 3
+fn <- 1
 names <- c(	"B,D,I", "E,F,Q", "C,G", "A,H,J", "M,N", "L,O,P", "K", "...")
@@
 
+## plot several optifit iterations
+
 ```{r ggraph_exp, fig.dim=c(6,5), dpi=300}
@@
 ```
+
+## maybe I'll do everything before plotting in python
+
+#```{python}
+#r.g1
+#```

### 2021-07-16 code/py/algorithm_diagram.py
Handle singleton seqs not in dist mat ('ghost pairs')

@@ -2,6 +2,63 @@
 """
 from collections import defaultdict, Counter
 from itertools import combinations
+from math import comb, sqrt
+          
+          
+def main():
+    ref_otus = opticlust_example()
+    print(ref_otus)
+
+
+def opticlust_example():
+    dist_frame = {
+        "seq1": [
+            "D",
+            "F",
+            "G",
+            "H",
+            "I",
+            "I",
+            "J",
+            "J",
+            "N",
+            "O",
+            "P",
+            "P",
+            "P",
+            "Q",
+            "Q",
+        ],
+        "seq2": [
+            "B",
+            "E",
+            "C",
+            "A",
+            "B",
+            "D",
+            "A",
+            "H",
+            "M",
+            "L",
+            "K",
+            "L",
+            "O",
+            "E",
+            "F",
+        ],
+    }
+    assert len(dist_frame['seq1']) == len(dist_frame['seq2'])
+    dist_mat = dist_pairs_to_sets(dist_frame)
+    otu_list = [{}, {'I', 'D', 'B'}, {'F', 'E', 'Q'}, {'C', 'G'},
+                {'H', 'J', 'A'}, {'M', 'N'}, {'P', 'L', 'O'}, {'K'}]
+    otus = otuMap.from_list(otu_list, dist_mat = dist_mat, n_seqs = 50)
+    #print(otus)
+    #conf_mat = otus.conf_mat(dist_mat)
+    #print('mcc current:', mcc(conf_mat),
+    #      '\nmcc from correct conf mat:', 
+    #      mcc({"tp": 14, "tn": 1210, "fp": 0, "fn": 1}),
+    #      '\nmcc correct: 0.97')
+    return otus
 
 
 def mcc(conf_mat):
@@ -48,30 +105,41 @@ def dist_pairs_to_sets(dframe):
         dist_set[seq1].add(seq2)
         dist_set[seq2].add(seq1)
     return dist_set
+            
 
-
-class otuMap():
+class otuMap:
     """
-    Maps sequences to OTU assignments. The dict should be of the form
-        {seqID: otuIndex}, e.g. {'seqA': 1, 'seqB': 2, 'seqC': 3}
+    Maps sequences to OTU assignments.
     """
 
-    def __init__(self, seqs_to_otus = None):
+    def __init__(self, seqs_to_otus = None, dist_mat = None, n_seqs = 0):
+        """
[…]

### 2021-07-16 exploratory/2021/algorithm_diagram.Rmd
Create ggraph from python dataframes

@@
 ```{r lib}
+library(here)
 library(tidyverse)
@@
   tbl_graph(nodes = data.frame(name = c("A B C", "D E F", "G **H** I", "J K L"),
-                               i = 1:4),
+                               id = 1:4),
             edges = data.frame(from = c(2, 2, 2), 
@@
   tbl_graph(nodes = data.frame(name = c("A B C", "D **E** F", "G H I", "J K L"),
-                               i = 1:4),
+                               id = 1:4),
             edges = data.frame(from = c(2, 2, 2), 
@@
 plot_graph <- function(graph) {
-  create_layout(graph, 'linear', sort.by = i) %>% 
+  create_layout(graph, 'linear', sort.by = id) %>% 
   ggraph() +
@@
 
-## maybe I'll do everything before plotting in python
+## maybe I'll do everything in python until the actual plotting step
 
-#```{python}
-#r.g1
-#```
+```{r pseudo_optifit}
+reticulate::source_python(here('code', 'py', 'algorithm_diagram.py'))
+optifit_iters <- run_optifit() %>% 
+  sapply(function(x) {
+    return(list(nodes = x[['nodes']] %>% py_to_r(),
+                edges = x[['edges']] %>% py_to_r()))
+    })
+```
+
+```{r graph_optifit, fig.dim=c(6,5), dpi=300}
+g <- tbl_graph(nodes = optifit_iters[,1]$nodes,
+               edges = optifit_iters[,1]$edges %>% 
+                 mutate(is_loop = from == to))
+plot_graph(g)
+```

### 2021-07-16 code/py/algorithm_diagram.py

@@ -4,7 +4,16 @@ from collections import defaultdict, Counter
 from copy import deepcopy
 from itertools import combinations
 from math import comb, sqrt
-import pandas as pd
+import pandas
+
+
+def run_optifit():
+    ref_otus = opticlust_example()
+    query_seqs = {'X', 'Y', 'Z'}
+    query_dist_mat = dist_pairs_to_sets({'seq1': ['X', 'X', 'X', 'X', 'Y'],
+                                         'seq2': ['Y', 'C', 'G', 'K', 'C']})
+    optifit = OptiFit(ref_otus, query_seqs, query_dist_mat, n_seqs = 53)
+    return optifit.iterate
 
 
 def opticlust_example():
@@ -202,6 +211,7 @@ class OptiIter:
         Store the nodes and edges as dictionaries for tidygraph.
         """
         sim_seqs = fitmap.dist_mat[curr_seq]
+        sim_seqs.add(curr_seq)
         options = list()
         edges = {'from': [], 'to': [], 'mcc': []}
         for sim_seq in sim_seqs:
@@ -209,10 +219,11 @@ class OptiIter:
             edges['from'].append(option.from_otu)
             edges['to'].append(option.to_otu)
             edges['mcc'].append(option.mcc)
-        self.edges = pd.DataFrame.from_dict(edges)
-        self.nodes = pd.DataFrame.from_dict(
+
+        self.edges = pandas.DataFrame.from_dict(edges)
+        self.nodes = pandas.DataFrame.from_dict(
           {'name': [' '.join([s if s != curr_seq else f"**{s}**"
-                              for s in otu])
+                              for s in sorted(otu)])
                     for otu in fitmap.otus_to_seqs.values()],
            'id': list(fitmap.otus_to_seqs.keys())
            })
@@ -243,16 +254,8 @@ class OptiOption:
 
 
 def main():
-    ref_otus = opticlust_example()
-    query_seqs = {'X', 'Y', 'Z'}
-    query_dist_mat = dist_pairs_to_sets({'seq1': ['X', 'X', 'X', 'X', 'Y'],
-                                         'seq2': ['Y', 'C', 'G', 'K', 'C']})
-    optifit = OptiFit(ref_otus, query_seqs, query_dist_mat, n_seqs = 53)
-    iters = optifit.iterate
-    print(iters)
+    run_optifit()
 
 
 if __name__ == "__main__":
     main()
-
-

### 2021-07-16 exploratory/2021/algorithm_diagram.Rmd
Format seq colors for ggtext

@@
 optifit_iters <- run_optifit() %>% 
-  sapply(function(x) {
+  lapply(function(x) {
     return(list(nodes = x[['nodes']] %>% py_to_r(),
-                edges = x[['edges']] %>% py_to_r()))
+                edges = x[['edges']] %>% py_to_r() %>% 
+                 mutate(is_loop = from == to)))
     })
@@
 
-```{r graph_optifit, fig.dim=c(6,5), dpi=300}
-g <- tbl_graph(nodes = optifit_iters[,1]$nodes,
-               edges = optifit_iters[,1]$edges %>% 
-                 mutate(is_loop = from == to))
+```{r optifit_example, fig.dim=c(6,5), dpi=300}
+g <- tbl_graph(nodes = optifit_iters[[1]]$nodes,
+               edges = optifit_iters[[1]]$edges)
 plot_graph(g)
 ```
+
+```{r plot_optifit_iters, fig.dim=c(6,5), dpi=300}
+lapply(optifit_iters, function(x) {
+  tbl_graph(nodes = x$nodes, edges = x$edges) %>% 
+    plot_graph()
+}) %>% 
+  wrap_plots(ncol = 1)
+```

### 2021-07-16 code/py/algorithm_diagram.py

@@ -81,6 +81,22 @@ def dist_pairs_to_sets(dframe):
     return dist_set
 
 
+def format_seq(s, optifit, curr_seq, color_curr_seq = False,
+               base_color = "#000000",
+               ref_color = "#D95F02",
+               query_color = "#1B9E77"):
+    """
+    format sequences with bold for the current iteration seq and
+    color-code reference and query sequences.
+    """
+    if s == curr_seq:
+        s = f"**{s}**"
+        color = query_color if color_curr_seq else base_color
+    else:
+        color = query_color if s in optifit.query_seqs else ref_color
+    return f"<span style = 'color:{color};'>{s}</span>"
+
+
 class otuMap:
     """
     Maps sequences to OTU assignments.
@@ -199,17 +215,19 @@ class OptiFit:
 
     @property
     def iterate(self):
-        return [OptiIter(self.fitmap, seq).to_dict
+        # TODO: after each iteration, modify fitmap with best option
+        return [OptiIter(self, seq).to_dict
                 for seq in self.query_seqs
                 if seq in self.fitmap.dist_mat]
 
 
 class OptiIter:
-    def __init__(self, fitmap, curr_seq):
+    def __init__(self, optifit, curr_seq):
         """
         Calculate possible MCCs if the current seq is moved to different OTUs.
         Store the nodes and edges as dictionaries for tidygraph.
         """
+        fitmap = optifit.fitmap
         sim_seqs = fitmap.dist_mat[curr_seq]
         sim_seqs.add(curr_seq)
         options = list()
@@ -222,11 +240,12 @@ class OptiIter:
 
         self.edges = pandas.DataFrame.from_dict(edges)
         self.nodes = pandas.DataFrame.from_dict(
-          {'name': [' '.join([s if s != curr_seq else f"**{s}**"
+          {'name': [' '.join([format_seq(s, optifit, curr_seq)
                               for s in sorted(otu)])
                     for otu in fitmap.otus_to_seqs.values()],
            'id': list(fitmap.otus_to_seqs.keys())
            })
+        # TODO: best_option, modify optifit object
 
     def __repr__(self):
         return "%s(%r)" % (self.__class__, self.__dict__.items())
@@ -245,12 +264,14 @@ class OptiOption:
         self.sim_seq = sim_seq
         self.from_otu = curr_otu_map.seqs_to_otus[curr_seq]
[…]

### 2021-07-16 code/py/algorithm_diagram.py
Attempt to fix otu renumbering

@@ -80,8 +80,16 @@ def dist_pairs_to_sets(dframe):
         dist_set[seq2].add(seq1)
     return dist_set
 
+def otu_list_to_dict(otu_list):
+    """
+    :param otu_list: should be of the form
+        [{seqA, seqB}, {seqC, secD}, {secE}]
+        where the index in the list is the OTU ID.
+    :return: an otuMap
+    """
+    return {seq: idx for idx, otu in enumerate(otu_list) for seq in otu}
 
-def format_seq(s, optifit, curr_seq, color_curr_seq = False,
+def format_seq(s, optifit, curr_seq, color_curr_seq = False, do_color = False,
                base_color = "#000000",
                ref_color = "#D95F02",
                query_color = "#1B9E77"):
@@ -94,7 +102,8 @@ def format_seq(s, optifit, curr_seq, color_curr_seq = False,
         color = query_color if color_curr_seq else base_color
     else:
         color = query_color if s in optifit.query_seqs else ref_color
-    return f"<span style = 'color:{color};'>{s}</span>"
+
+    return f"<span style = 'color:{color};'>{s}</span>" if do_color else s
 
 
 class otuMap:
@@ -116,6 +125,15 @@ class otuMap:
     def __repr__(self):
         return "%s(%r)" % (self.__class__, self.seqs_to_otus)
 
+    def renumber_otus(self):
+        """
+        re-number OTU ids so they're continuous
+        """
+        old_otus = self.otus_to_seqs
+        if not all(old_otus.values()): # there exists an empty set
+            print('renumbering otus')
+            self.seqs_to_otus = otu_list_to_dict([otu for otu in old_otus if otu])
+
     @classmethod
     def from_list(cls, otu_list, **kwargs):
         """
@@ -124,8 +142,7 @@ class otuMap:
             where the index in the list is the OTU ID.
         :return: an otuMap
         """
-        return cls(seqs_to_otus = {seq: idx for idx, otu in enumerate(otu_list)
-                                            for seq in otu}, **kwargs)
+        return cls(seqs_to_otus = otu_list_to_dict(otu_list), **kwargs)
 
     @property
     def seqs(self):
@@ -215,10 +232,13 @@ class OptiFit:
 
     @property
     def iterate(self):
-        # TODO: after each iteration, modify fitmap with best option
-        return [OptiIter(self, seq).to_dict
-                for seq in self.query_seqs
-                if seq in self.fitmap.dist_mat]
[…]

### 2021-07-16 code/py/algorithm_diagram.py
Fix bugs to renumber OTUs

@@ -9,7 +9,7 @@ import pandas
 
 def run_optifit():
     ref_otus = opticlust_example()
-    query_seqs = {'X', 'Y', 'Z'}
+    query_seqs = ['X', 'Y', 'Z']
     query_dist_mat = dist_pairs_to_sets({'seq1': ['X', 'X', 'X', 'X', 'Y'],
                                          'seq2': ['Y', 'C', 'G', 'K', 'C']})
     optifit = OptiFit(ref_otus, query_seqs, query_dist_mat, n_seqs = 53)
@@ -130,9 +130,8 @@ class otuMap:
         re-number OTU ids so they're continuous
         """
         old_otus = self.otus_to_seqs
-        if not all(old_otus.values()): # there exists an empty set
-            print('renumbering otus')
-            self.seqs_to_otus = otu_list_to_dict([otu for otu in old_otus if otu])
+        if max(old_otus) != len(old_otus): # then OTU IDs are not continuous
+            self.seqs_to_otus = otu_list_to_dict([{}]+[otu for idx, otu in old_otus.items() if otu])
 
     @classmethod
     def from_list(cls, otu_list, **kwargs):
@@ -221,7 +220,6 @@ class OptiFit:
         self.fitmap = otuMap(seqs_to_otus = seqs_to_otus,
                              dist_mat = dist_mat,
                              n_seqs = n_seqs)
-        #self.iterations = list()  # list of DataFrames for tidygraph/ggraph
 
     def __repr__(self):
         return "%s(%r)" % (self.__class__, self.fitmap)
@@ -233,27 +231,27 @@ class OptiFit:
     @property
     def iterate(self):
         iterations = list()
+        curr_fitmap = self.fitmap
         for seq in self.query_seqs:
             if seq in self.fitmap.dist_mat:
-                iter = OptiIter(self, seq)
-                iterations.append(iter.to_dict)
-                self.fitmap = iter.best_map
+                iteration = OptiIter(curr_fitmap, seq)
+                iterations.append(iteration.to_dict)
+                curr_fitmap = iteration.best_map
         return iterations
 
 
 class OptiIter:
-    def __init__(self, optifit, curr_seq):
+    def __init__(self, curr_fitmap, curr_seq):
         """
         Calculate possible MCCs if the current seq is moved to different OTUs.
         Store the nodes and edges as dictionaries for tidygraph.
         """
[…]

### 2021-07-17 exploratory/2021/algorithm_diagram.Rmd
Attempt to fix otu renumbering
Fix bugs to renumber OTUs
Fix bugs after refactor... i think

@@
                 ) +
-  geom_edge_loop(aes(span = 1, direction = 270, strength = 0.2, color = is_loop)) +
+  geom_edge_loop(aes(#span = 1, strength = 0.2, direction = 270, 
+                     color = is_loop)) +
   geom_node_label(aes(label = name)) +
-    scale_edge_color_manual(values = c('white')) +
+    scale_edge_color_manual(values = c('black')#c('white')
+                            ) +
   theme_void() +
@@
 
-## maybe I'll do everything in python until the actual plotting step
+## do backend in python until the actual plotting step
 
@@
 ```{r optifit_example, fig.dim=c(6,5), dpi=300}
-g <- tbl_graph(nodes = optifit_iters[[1]]$nodes,
-               edges = optifit_iters[[1]]$edges)
-plot_graph(g)
+tbl_graph(nodes = optifit_iters[[1]]$nodes,
+          edges = optifit_iters[[1]]$edges) %>% 
+  plot_graph()
 ```

### 2021-07-17 code/py/algorithm_diagram.py

@@ -90,7 +90,7 @@ def otu_list_to_dict(otu_list):
     """
     return {seq: idx for idx, otu in enumerate(otu_list) for seq in otu}
 
-def format_seq(s, optifit, curr_seq, color_curr_seq = False, do_color = False,
+def format_seq(s, curr_seq, query_seqs, color_curr_seq = False, do_color = False,
                base_color = "#000000",
                ref_color = "#D95F02",
                query_color = "#1B9E77"):
@@ -102,7 +102,7 @@ def format_seq(s, optifit, curr_seq, color_curr_seq = False, do_color = False,
         s = f"**{s}**"
         color = query_color if color_curr_seq else base_color
     else:
-        color = query_color if s in optifit.query_seqs else ref_color
+        color = query_color if s in query_seqs else ref_color
 
     return f"<span style = 'color:{color};'>{s}</span>" if do_color else s
 
@@ -238,14 +238,14 @@ class OptiFit:
             prev_mcc = self.mcc
             for seq in self.query_seqs:
                 if seq in self.fitmap.dist_mat:
-                    iteration = OptiIter(curr_fitmap, seq)
+                    iteration = OptiIter(curr_fitmap, seq, self.query_seqs)
                     iterations.append(iteration.to_dict)
                     curr_fitmap = iteration.best_map
         return iterations
 
 
 class OptiIter:
-    def __init__(self, curr_fitmap, curr_seq):
+    def __init__(self, curr_fitmap, curr_seq, query_seqs):
         """
         Calculate possible MCCs if the current seq is moved to different OTUs.
         Store the nodes and edges as dictionaries for tidygraph.
@@ -263,7 +263,7 @@ class OptiIter:
 
         self.edges = pandas.DataFrame.from_dict(edges)
         self.nodes = pandas.DataFrame.from_dict(
-          {'name': [' '.join([format_seq(s, optifit, curr_seq)
+          {'name': [' '.join([format_seq(s, curr_seq, query_seqs)
                               for s in sorted(otu)])
                     for otu in curr_fitmap.otus_to_seqs.values()],
            'id': list(curr_fitmap.otus_to_seqs.keys())

### 2021-07-17 code/py/algorithm_diagram.py
Slightly bigger example

@@ -10,9 +10,10 @@ import pandas
 
 def run_optifit():
     ref_otus = opticlust_example()
-    query_seqs = ['X', 'Y', 'Z']
-    query_dist_mat = dist_pairs_to_sets({'seq1': ['X', 'X', 'X', 'X', 'Y'],
-                                         'seq2': ['Y', 'C', 'G', 'K', 'C']})
+    query_seqs = ['W','X','Y','Z']
+    query_dist_mat = dist_pairs_to_sets(
+    {'seq1': ['X', 'X', 'X', 'X', 'Y', 'W', 'W', 'W'],
+     'seq2': ['Y', 'C', 'G', 'N', 'C', 'M', 'N', 'F']})
     optifit = OptiFit(ref_otus, query_seqs, query_dist_mat, n_seqs = 53)
     return optifit.iterate
 
@@ -90,7 +91,7 @@ def otu_list_to_dict(otu_list):
     """
     return {seq: idx for idx, otu in enumerate(otu_list) for seq in otu}
 
-def format_seq(s, curr_seq, query_seqs, color_curr_seq = False, do_color = False,
+def format_seq(s, curr_seq, query_seqs, color_curr_seq = False, do_color = True,
                base_color = "#000000",
                ref_color = "#D95F02",
                query_color = "#1B9E77"):

### 2021-07-17 exploratory/2021/algorithm_diagram.Rmd
Tweak plot annotations

@@
 library(diagram)
+library(glue)
 library(reticulate)
@@
 ```{r ggraph_exp, fig.dim=c(6,5), dpi=300}
-plot_graph <- function(graph) {
+plot_graph <- function(graph, title = '') {
   create_layout(graph, 'linear', sort.by = id) %>% 
@@
                 ) +
-  geom_edge_loop(aes(#span = 1, strength = 0.2, direction = 270, 
+  geom_edge_loop(aes(span = 1, direction = 90, strength = 0.2,# direction = 270, 
                      color = is_loop)) +
@@
                             ) +
+  labs(title = title) +
   theme_void() +
@@
 
-```{r plot_optifit_iters, fig.dim=c(6,5), dpi=300}
+## plot optifit iterations with ggraph & patchwork
+
+```{r plot_optifit_iters, fig.dim=c(6.875,9.0625), dpi=300}
+# TODO custom modifications for loop span & direction
 lapply(optifit_iters, function(x) {
   tbl_graph(nodes = x$nodes, edges = x$edges) %>% 
-    plot_graph()
+    plot_graph(title = glue('mcc = {x$edges %>% filter(is_loop) %>% pull(mcc)}'))
 }) %>% 
-  wrap_plots(ncol = 1)
+  wrap_plots(ncol = 1) +
+  plot_annotation(tag_levels = 'A')
 ```

### 2021-07-17 exploratory/2021/algorithm_diagram.Rmd
DIY plot label so it's inline with mcc

@@
 # TODO custom modifications for loop span & direction
+i <- 0
 lapply(optifit_iters, function(x) {
+  i <<- i + 1
   tbl_graph(nodes = x$nodes, edges = x$edges) %>% 
-    plot_graph(title = glue('mcc = {x$edges %>% filter(is_loop) %>% pull(mcc)}'))
+    plot_graph(title = glue('{i}) mcc = {x$edges %>% filter(is_loop) %>% pull(mcc)}'))
 }) %>% 
-  wrap_plots(ncol = 1) +
-  plot_annotation(tag_levels = 'A')
+  wrap_plots(ncol = 1)
 ```

### 2021-07-21 exploratory/2021/algorithm_diagram.Rmd

@@
 ```{r ggraph_exp, fig.dim=c(6,5), dpi=300}
-plot_graph <- function(graph, title = '') {
+plot_graph <- function(graph, title = '', 
+                       is_last_plot = FALSE,
+                       hide_loops = FALSE) {
+  loop_dir <- ifelse(is_last_plot, 270, 90)
+  loop_color <- ifelse(hide_loops, 'white', 'black')
   create_layout(graph, 'linear', sort.by = id) %>% 
@@
                 ) +
-  geom_edge_loop(aes(span = 1, direction = 90, strength = 0.2,# direction = 270, 
+  geom_edge_loop(aes(span = 1, 
+                     direction = loop_dir, 
+                     strength = 0.2,
                      color = is_loop)) +
   geom_node_label(aes(label = name)) +
-    scale_edge_color_manual(values = c('black')#c('white')
+    scale_edge_color_manual(values = c(loop_color)
                             ) +
@@
 
-```{r plot_optifit_iters, fig.dim=c(6.875,9.0625), dpi=300}
-# TODO custom modifications for loop span & direction
+```{r plot_optifit_iters, fig.dim=c(6.875,6.875), dpi=300}
 i <- 0
@@
   tbl_graph(nodes = x$nodes, edges = x$edges) %>% 
-    plot_graph(title = glue('{i}) mcc = {x$edges %>% filter(is_loop) %>% pull(mcc)}'))
+    plot_graph(title = glue('{i}) mcc = {x$edges %>% filter(is_loop) %>% pull(mcc)}'),
+               is_last_plot = i == length(optifit_iters),
+               hide_loops = FALSE)
 }) %>% 

### 2021-07-21 paper/paper.Rmd
Insert placeholder for algorithm diagram

@@
 
+```{r algorithm, fig.dim=figs('algorithm','dim'), fig.cap=figs('algorithm','cap')}
+figs('algorithm', 'path') %>% 
+  tiff::readTIFF() %>% 
+  grid::grid.raster() 
+```
+
 OptiFit begins where OptiClust ends, starting with a list of reference OTUs and
 their sequences, a list of query sequences to assign to the reference OTUs, and
-the sequence pairs that are within the distance threshold (e.g. 0.03).
+the sequence pairs that are within the distance threshold (e.g. 0.03) 
+(Figure \ref{fig:algorithm}).
 Initially, all query sequences are placed into separate OTUs. 
@@
 Fitting sequences to Greengenes and SILVA in closed reference mode performed
-similarly, with median MCC scores of `r round(closed_fit_gg_mcc, 3)` (**TODO: Pat marked this as 0.80, but the actual number is 0.795414**) and 
+similarly, with median MCC scores of `r closed_fit_gg_mcc` and 
 `r closed_fit_silva_mcc` respectively, while the median MCC 

### 2021-07-23 exploratory/2021/algorithm_diagram.Rmd
Trying to get dist_mat out of Python into R

@@
 ```
+
+## Pat's OptiClust diagram
 
@@
 ## do backend in python until the actual plotting step
-
-```{r pseudo_optifit}
+```{python optifit}
+import code.py.algorithm_diagram
+optifit = create_optifit()
+iters = optifit.iterate
+dists = dist_sets_to_array(optifit.fitmap.dist_mat)
+```
+```{r pytor}
 reticulate::source_python(here('code', 'py', 'algorithm_diagram.py'))
-optifit_iters <- run_optifit() %>% 
+optifit <- create_optifit() 
+dist_array <- optifit$fitmap$dists_to_array %>% py_to_r()
+optifit_iters <- optifit$iterate %>% 
   lapply(function(x) {

### 2021-07-23 code/py/algorithm_diagram.py

@@ -8,14 +8,14 @@ import numpy
 import pandas
 
 
-def run_optifit():
+def create_optifit():
     ref_otus = opticlust_example()
     query_seqs = ['W', 'X', 'Y', 'Z']
     query_dist_mat = dist_pairs_to_sets(
         {'seq1': ['X', 'X', 'X', 'X', 'Y', 'W', 'W', 'W'],
          'seq2': ['Y', 'C', 'G', 'N', 'C', 'M', 'N', 'F']})
     optifit = OptiFit(ref_otus, query_seqs, query_dist_mat, n_seqs=53)
-    return optifit.iterate
+    return optifit
 
 
 def opticlust_example():
@@ -85,6 +85,17 @@ def dist_pairs_to_sets(dframe):
         dist_set[seq2].add(seq1)
     return dist_set
 
+
+def dist_array_to_dframe(dist_array):
+    """
+    :param dist_array: pd.DataFrame(dist_array,
+                          columns=['A', 'B', 'C', 'D'],
+                          index=['A', 'B', 'C', 'D'])
+    :return: pd.DataFrame.from_dict({'seq1': ['A', 'A', 'B'],
+                                     'seq2': ['B', 'C', 'D']})
+
+    """
+    return # TODO
 
 def otu_list_to_dict(otu_list):
     """
@@ -205,6 +216,26 @@ class otuMap:
     def mcc(self):
         return mcc(self.conf_mat)
 
+    @property
+    def dists_to_array(self):
+        """
+        :param dist_mat: {'A': {'B','C'}, 'B': {'A','C','D'}, 'C': {'A','B'}, 'D': {'B'}}
+        :return: pd.DataFrame(dist_array,
+                              columns=['A', 'B', 'C', 'D'],
+                              index=['A', 'B', 'C', 'D'])
+        """
+        dist_sets = self.dist_mat
+        seqs = {seq2 for seq1 in dist_sets for seq2 in dist_sets[seq1]}
+        seqs.update({seq for seq in dist_sets})
+        seqs = list(sorted(seqs))
+        len_seqs = len(seqs)
+        dist_array = numpy.zeros(len_seqs, len_seqs)
+        for i, seqi in enumerate(seqs):
+            for j, seqj in enumerate(seqs):
+                if i == j or seqi in dist_sets[seqj] or seqj in dist_sets[seqi]:
+                    dist_array [i][j] = 1
+                    dist_array [j][i] = 1
[…]

### 2021-08-10 paper/paper.Rmd
Explain 'radius' vs 'diameter' result of clustering algorithms

@@
 @rognes_vsearch_2016].
+As a result of greedy clustering, some pairs of sequences in the same OTU may
+have a greater distance than the specified threshold since only the distance
+between each sequence and the centroid sequence is considered while clustering.
+In contrast, the OptiClust algorithm enforces that all pairs of sequences must
+be within the distance threshold. <!-- i.e. radius vs. diameter -->
 A limitation of _de novo_ clustering is that different OTU assignments will be
@@
 they are not represented by a similar sequence in the database.
-**TODO: closed uses radius of 3%. de novo uses diameter of 3%. open uses both.**
 Previous studies found that the OptiClust _de novo_ clustering algorithm created

### 2021-08-10 paper/paper.Rmd
Start writing importance section...

@@
 **TODO.**
-<!--
-**"AND, BUT, THEREFORE" structure. start paragraph with question, end with why we should care. transitions to move the story along.**
-Advancements in sequencing technology have allowed researchers to affordably generate millions of reads of microorganisms from diverse natural communities,
+Advancements in DNA sequencing technology have allowed researchers to affordably
+generate millions of reads from microorganisms in diverse natural communities.
 allowing us to characterize communities.
--->
 \newpage

### 2021-08-10 code/R/plot_algorithm_diagram.R
Tweak distance table headers

@@ -73,8 +73,8 @@ dist_dat <- get_dists()  %>%
     color2 = ifelse(seq2 %in% ref_seqs, ref_color, query_color),
   )
 dist_dat[["color3"]] <- rep.int('black', nrow(dist_dat))
-dist_dat[['distance']] <- runif(nrow(dist_dat), 0.1, 0.3) %>% 
-  as.character() %>% substring(2, 4)
+dist_dat[['dist']] <- runif(nrow(dist_dat), 1.0, 2.9) %>% 
+  format(digits = 2) %>% as.character() 
 table_colors <- dist_dat %>% select(color1, color2, color3) %>% as.matrix() %>% t()
 
 table_plot <- plot_grid(ggdraw() + 
@@ -84,7 +84,10 @@ table_plot <- plot_grid(ggdraw() +
                                      ) + 
                           theme(plot.margin = margin(5, 0, 5, 0)),
                    tableGrob(dist_dat %>% 
-                               select(seq1, seq2, distance) %>% 
+                               select(seq1, seq2, dist) %>% 
+                               rename(`% distance` = dist,
+                                      ` ` = seq1,
+                                      `  ` = seq2) %>% 
                                t(),
                              theme = ttheme_default(base_size = 10,
                                                     padding = unit(c(4, 4), 'pt'),

### 2021-08-13 code/R/plot_results_split.R
Fix points at 0.9 not showing up

@@ -101,9 +101,9 @@ dat %>%
   scale_color_manual(values = color_values,
                      breaks = names(color_breaks),
                      labels = color_labels) +
-  scale_x_continuous(breaks = seq(0, 0.9, 0.1), 
-                     limits = c(0, 0.9),
-                     labels = c('NA', seq(0.1, 0.9, 0.1))) +
+  scale_x_continuous(breaks = seq(0, 1, 0.1), 
+                     #limits = c(0, 0.9),
+                     labels = c('NA', seq(0.1, 1, 0.1))) +
   labs(x = 'reference fraction', y = '') +
   theme_bw() +
   theme(legend.text = element_markdown(),

### 2021-08-13 paper/paper.Rmd
Split datasets up to 90% reference fraction

@@
 abundance, and weighting by similarity to other sequences in the dataset.
-Dataset splitting was repeated with reference fractions ranging from 10% to 80%
+Dataset splitting was repeated with reference fractions ranging from 10% to 90%
 of the dataset and for 100 random seeds.
-For each dataset split, the remaining sequences were assigned to the reference
-OTUs with OptiFit.
+For each dataset split, the remaining query sequences were assigned to the
+reference OTUs with OptiFit.
 

### 2021-08-24 paper/paper.Rmd
Write figure caption for algorithm diagram
Finish importance section

@@
 
-<!--
+<!-- TODO
 - Go back through https://peerj.com/articles/1487/ to get all the arguments for why the qiime based methods are problematic…
@@
 
-**TODO.**
 Advancements in DNA sequencing technology have allowed researchers to affordably
 generate millions of reads from microorganisms in diverse natural communities.
-allowing us to characterize communities.
+Efficient software tools are needed to assign microbial sequences into
+taxonomic groups for characterization and comparison of communities.
+The OptiClust algorithm produces high quality groups by comparing sequences to
+each other, but the taxonomic assignments can change when new sequences are
+added to a dataset, making it difficult to compare different studies.
+Other approaches assign sequences to taxonomic groups by comparing them to
+sequences in a reference database to produce consistent assignments, but the
+quality of the groups produced is reduced compared to OptiClust.
+We developed OptiFit, a new reference-based algorithm that produces consistent
+yet high quality taxonomic assignments like OptiClust.
+OptiFit allows researchers to compare microbial communities across different
+studies or add new data to existing studies without sacrificing the quality of
+the taxonomic assignments.
+
 \newpage
@@
 
+We thank members of the Schloss Lab for their feedback on the figures.
+
 KLS received support from the NIH Training Program in Bioinformatics (T32 GM070449).
 Salary support for PDS came from NIH grants R01CA215574 and U01AI124255.
-
 The funders had no role in study design, data collection and interpretation, 

### 2021-09-01 paper/paper.Rmd
Incorporate (most of) Pat's revisions
Co-authored-by: Pat Schloss <pschloss@users.noreply.github.com>

--- 2021-08-13 paper/paper.Rmd
@@
 
+<!-- TODO
+- "require" and "enforce" for opti -- wrong word choice.
+- be consistent with cluster, assign, fit, and map.
+-->
+
 ## Abstract
@@
 large datasets.
-OptiClust, a _de novo_ OTU clustering method in the mothur program, has been
+OptiClust, a _de novo_ OTU clustering method, has been
 shown to produce higher quality OTU assignments than other methods and at
@@
 A notable difference between _de novo_ clustering and database-dependent methods
-is that OTU assignments clustered with _de novo_ methods may change when new
+is that OTU assignments from _de novo_ methods may change when new
 sequences are added to a dataset.
@@
 are features.
+**TODO**: sentence mentioning open/closed reference clustering and the inherent problems with the current methods.
 To provide an efficient and robust method to fit amplicon sequence data to
@@
 We tested OptiFit using four microbiome datasets with two different strategies:
-by fitting to an external reference database, or by splitting the dataset into a 
-reference and query set and fitting the query sequences to the reference set.
+by fitting to an external reference database or by splitting the dataset into a 
+reference and query set and fitting the query sequences to the reference set
+after clustering it using OptiClust.
 The result is a robust implementation of closed and open-reference clustering.
@@
 
-**TODO.**
 Advancements in DNA sequencing technology have allowed researchers to affordably
 generate millions of reads from microorganisms in diverse natural communities.
-allowing us to characterize communities.
+Efficient and robust software tools are needed to assign microbial sequences
+into taxonomic groups for characterization and comparison of communities.
+The OptiClust algorithm produces high quality groups by comparing sequences to
+each other, but the assignments can change when new sequences are
+added to a dataset, making it difficult to compare different studies.
+Other approaches assign sequences to groups by comparing them to
+sequences in a reference database to produce consistent assignments, but the
+quality of the groups produced is reduced compared to OptiClust.
+We developed OptiFit, a new reference-based algorithm that produces consistent
+yet high quality assignments like OptiClust.
+OptiFit allows researchers to compare microbial communities across different
+studies or add new data to existing studies without sacrificing the quality of
+the group assignments.
+
 \newpage
@@
 
-Amplicon sequencing has become a mainstay of microbial ecology.
+Amplicon sequencing is a mainstay of microbial ecology.
 Researchers can affordably generate millions of sequences to characterize the
@@
 operational taxonomic units (OTUs) to facilitate comparison of taxonomic
-composition between communities to avoid the need for classification.
+composition between communities to avoid the need for taxonomic classification.
 A distance threshold of 3% (or sequence similarity of 97%) is commonly used to
@@
 @rognes_vsearch_2016].
-As a result of greedy clustering, some pairs of sequences in the same OTU may
-have a greater distance than the specified threshold since only the distance
-between each sequence and the centroid sequence is considered while clustering.
+In methods employing greedy clustering algorithms, some pairs of sequences in
+the same OTU may have a greater distance than the specified threshold since only
+the distance between each sequence and the centroid sequence is considered while
+clustering.
 In contrast, the OptiClust algorithm enforces that all pairs of sequences must
@@
 algorithms while maintaining OTU quality, we developed OptiFit, a
-reference-based clustering algorithm which uses existing OTUs as the reference
-and fits new sequences those reference OTUs.
+reference-based clustering algorithm.
+While other tools represent reference OTUs with a single sequence, OptiFit uses
+multiple sequences in existing OTUs as the reference and fits new sequences
+those reference OTUs.
 In contrast to other tools, OptiFit considers all pairwise distance scores
@@
 to produce OTUs of the highest possible quality.
-Here, we tested the OptiFit algorithm with the reference as a database or _de
-novo_ OTUs and compared the performance to existing tools.
+Here, we tested the OptiFit algorithm with the reference as a public database
+(e.g. Greengenes) or _de novo_ OTUs and compared the performance to existing
+tools.
 To evaluate the OptiFit algorithm and compare to existing methods, we used four
@@
 In the closed reference mode, any query sequences that cannot be assigned to
-references OTUs are discarded, and the results will only contain OTUs that exist
+reference OTUs are discarded, and the results only contain OTUs that exist
 in the original reference.
@@
 sequences _de novo_, then fit the query sequences to the reference OTUs.
-We clustered sequences from four datasets isolated from soil
-[@johnston_metagenomics_2016], marine [@henson_artificial_2016], mouse gut
-[@schloss_stabilization_2012], and human gut [@baxter_microbiota-based_2016]
-samples to test the performance of OptiFit with both of these strategies.
 
@@
 
-While _de novo_ clustering produces high quality OTUs, researchers may prefer to
-perform reference clustering to a public database because reference-based
-methods produce consistent OTUs and are generally faster than _de novo_ methods.
+While _de novo_ clustering produces the highest quality OTUs, researchers may
+prefer to perform reference clustering to a public database because
+reference-based methods produce consistent OTUs and are generally faster than
+_de novo_ methods.
 In closed reference mode, sequences that cannot be assigned to reference OTUs
@@
 Reference OTUs for each database were created by performing _de novo_ clustering
-with OptiClust at a distance threshold of 3% (see Figure \ref{fig:workflow}).
+with OptiClust at a distance threshold of 3% using the V4 region of each
+sequence (see Figure \ref{fig:workflow}).
 The _de novo_ MCC scores were `r dn_dbs_mcc` for `r db_names_upper`,
@@
 chosen greatly affects the quality of OTUs produced.
-OTU quality may be poor when the reference database is too unrelated to the
-samples of interest, such as when samples contain low abundant or novel
+OTU quality may be poor when the reference database consists of sequences that
+are too unrelated to the samples of interest, such as when samples contain novel
 populations.
@@
 are added.
-Researchers may wish to fit new sequences to existing OTUs when comparing OTUs
-across studies or when making predictions with machine learning models.
+Researchers may wish to fit new sequences to existing OTUs or to compare OTUs
+across studies.
 To determine how well OptiFit performs for fitting new sequences to existing
@@
 sequences to reference OTUs with the split strategy, a great improvement over
-the average `r frac_fit_gg`% of sequences fit to the greengenes database.
+the average `r frac_fit_gg`% of sequences fit to the Greengenes database.
 In terms of run time, closed and open reference OptiFit performed faster than
@@
 an optimal reference fraction size.
-To test the best reference size, reference sets with 10% to 90% of the
+To identify the best reference size, reference sets with 10% to 90% of the
 sequences were created, with the remaining sequences used for the query.
@@
 \ref{fig:results_split}).
-Thus, the reference fraction doid not affect OTU quality in terms of MCC score,
+Thus, the reference fraction did not affect OTU quality in terms of MCC score,
 but did affect the run time and the fraction of sequences that mapped during the
@@
 OTU quality in terms of MCC was similar with the simple and abundance-weighted
-sampling (median MCCs of `r mcc_fit_split_simple` and `r mcc_fit_split_abun`
+sampling (median MCCs of `r mcc_fit_split_simple` and `r mcc_fit_split_abun`,
 respectively), but worse for similarity-weighted sampling (median MCC of
@@
 dataset/database combination. 
-This may reflect limitations of reference
-databases, which are unlikely to contain sequences from rare and novel microbes.
+This may reflect limitations of reference databases, which are unlikely to
+contain sequences from novel microbes.
 This drop in quality was most notable with the RDP reference, which contained only 
 `r ref_seqs[['rdp']]` sequences compared to `r ref_seqs[['silva']]` sequences in
-SILVA and `r ref_seqs[['gg']]` in Greengenes after trimming to the V4 region.
+SILVA and `r ref_seqs[['gg']]` in Greengenes.
 Note that Greengenes has not been updated since 2013 at the time of this
-writing, while SILVA is updated every one to two years with the most recent
-release in 2019.
+writing, while SILVA and the RDP are updated regularly.
 We recommend that users who require an independent reference database opt for
-large databases with regular updates and good coverage of microbial diversity.
+large databases with regular updates and good coverage of microbial diversity
+for their environment.
 Since OptiClust still performs faster than open reference OptiFit and creates
@@
 recommend using OptiClust rather than fitting to a database whenever consistent
-OTUs are not required for the study at hand.
+OTUs are not required.
 
@@
 reference mode.
-While both mothur and VSEARCH use a distance or similarity threshold for
+While both OptiFit and VSEARCH use a distance or similarity threshold for
 determining how to assign sequences to OTUs, VSEARCH is more permissive than
-mothur.
+OptiFit.
 The OptiFit and OptiClust algorithms use all of the sequences to define an
@@
 the input sequences, because each query is assigned to the OTU of the first
-centroid sequence that is found within the distance threshold.
-In this way, VSEARCH sacrifices OTU quality in order to allow more sequences to
+centroid sequence that is found within the distance threshold. TODO: is this true of VSEARCH or only USEARCH?
+Because of this, VSEARCH sacrifices OTU quality by allowing more sequences to
 fit to OTUs.
@@
 gut [@baxter_microbiota-based_2016] samples.
-These datasets represent a selection of the broad types of natural communities
-that microbial ecologists study.
+These datasets contain sequences from the V4 region of the 16S rRNA gene and
+represent a selection of the broad types of natural communities that microbial
+ecologists study.
 We processed the raw sequences using mothur according to the Schloss Lab MiSeq
-SOP as described in the mothur wiki and accompanying study by Kozich _et al._
+SOP and accompanying study by Kozich _et al._
 [@schloss_miseq_nodate; @kozich_development_2013].

### 2021-09-01 paper/paper.Rmd
Can we just not with 'robust'? ISO a better word.

@@
 ---
-title: "OptiFit: a fast method for fitting amplicon sequences to existing OTUs"
+title: "OptiFit: a new method for fitting amplicon sequences to existing OTUs"
 date: '`r Sys.Date()`'
@@
 - be consistent with cluster, assign, fit, and map.
+- can we replace "robust" with something better / more specific?
 -->

### 2021-09-02 paper/paper.Rmd
Use 'improved' instead of 'robust' in some places

@@
 ---
-title: "OptiFit: a new method for fitting amplicon sequences to existing OTUs"
+title: "OptiFit: an improved method for fitting amplicon sequences to existing OTUs"
 date: '`r Sys.Date()`'
@@
 after clustering it using OptiClust.
-The result is a robust implementation of closed and open-reference clustering.
+The result is an improved implementation of closed and open-reference clustering.
 OptiFit produces OTUs of similar quality as OptiClust and at faster speeds when

### 2021-09-03 paper/paper.Rmd
Note problems with existing reference methods in the abstract

@@
 comparable or faster speeds.
-A notable difference between _de novo_ clustering and database-dependent methods
-is that OTU assignments from _de novo_ methods may change when new
-sequences are added to a dataset.
+A notable difference between _de novo_ clustering and database-dependent
+reference clustering methods is that OTU assignments from _de novo_ methods may
+change when new sequences are added to a dataset.
 However, in some cases one may wish to incorporate new samples into a previously
@@
 are features.
-**TODO**: sentence mentioning open/closed reference clustering and the inherent problems with the current methods.
+Existing reference-based clustering methods produce consistent OTUs, but they
+only consider the similarity of each query sequence to a single reference
+sequence in an OTU, thus resulting in some dissimilar sequences being assigned
+to the same OTU.
 To provide an efficient and robust method to fit amplicon sequence data to
 existing OTUs, we developed the OptiFit algorithm.
+Like OptiFit, OptiClust considers the similarity of all pairs of reference and
+query sequences in an OTU to produce OTUs of the best possible quality.
 We tested OptiFit using four microbiome datasets with two different strategies:

### 2021-09-07 paper/paper.Rmd
Incorporate the rest of @pschloss' suggestions
    - Fixed incorrect assertion about Opti[Fit/Clust] requiring/enforcing the distance threshold.
    - Edited to use "cluster", "assign", "map", & "fit" in a consistent way. Mainly just use cluster.
    - Made the effects of greedy clustering vs opticlust more obvious without explicitly using the words "radius" or "diameter", since those haven't been used elsewhere.

@@
 
-<!-- TODO
-- Go back through https://peerj.com/articles/1487/ to get all the arguments for why the qiime based methods are problematic…
-1. The threshold with closed reference is a radius and with de novo its a diameter
-2. The gg reference was defined for full length sequences, within a subregion there are duplicate and nearly duplicate sequences
-3.
-Because VSEARCH/USEARCH is a greedy heuristic, its looking for a reference that is close.
-Although the ordering of the database is not relevant (it was with USEARCH) it gets the wrong number of its (see Figure 5)
--->
-
-<!-- TODO
-- "require" and "enforce" for opti -- wrong word choice.
-- be consistent with cluster, assign, fit, and map.
-- can we replace "robust" with something better / more specific?
--->
-
 ## Abstract
@@
 only consider the similarity of each query sequence to a single reference
-sequence in an OTU, thus resulting in some dissimilar sequences being assigned
+sequence in an OTU, thus resulting in some dissimilar sequences being clustered
 to the same OTU.
@@
 We tested OptiFit using four microbiome datasets with two different strategies:
-by fitting to an external reference database or by splitting the dataset into a 
-reference and query set and fitting the query sequences to the reference set
+by clustering to an external reference database or by splitting the dataset into a 
+reference and query set and clustering the query sequences to the reference set
 after clustering it using OptiClust.
@@
 @rognes_vsearch_2016].
-In methods employing greedy clustering algorithms, some pairs of sequences in
-the same OTU may have a greater distance than the specified threshold since only
-the distance between each sequence and the centroid sequence is considered while
-clustering.
-In contrast, the OptiClust algorithm enforces that all pairs of sequences must
-be within the distance threshold. <!-- i.e. radius vs. diameter -->
+In methods employing greedy clustering algorithms, only the distance between
+each sequence and the centroid sequence is considered while clustering.
+As a result, pairs of sequences in the same OTU may have a greater distance than
+the specified threshold, i.e. they are false positives.
+In contrast, the OptiClust algorithm takes into account the distance between all
+pairs of sequences when considering how to cluster sequences into OTUs and is
+thus less willing to take on false positives.
 A limitation of _de novo_ clustering is that different OTU assignments will be
@@
 _de novo_ clustering to compare OTUs between different studies.
-Additionally, the greedy clustering algorithms are sensitive to the order of the
-input sequences: different OTU assignments are produced when the same sequences
-are randomly shuffled [@westcott_novo_2015; @he_stability_2015].
 Furthermore, since _de novo_ clustering requires calculating and comparing
@@
 [@desantis_greengenes_2006; @rognes_vsearch_2016; @noauthor_clustering_nodate].
-Query sequences are then assigned to OTUs based on their similarity to the
+Query sequences are then clustered into OTUs based on their similarity to the
 reference sequences.
@@
 they are not represented by a similar sequence in the database.
+Just like in _de novo_ clustering, reference-based methods that use greedy
+clustering algorithms only consider the distance between each query and a single
+reference sequence, causing some sequences in the same OTU to be more dissimilar
+to each other than the specified threshold.
 Previous studies found that the OptiClust _de novo_ clustering algorithm created
@@
 Sequence pairs that are similar to each other (i.e. within the distance
-threshold) are counted as true positives if they are assigned to the same OTU,
+threshold) are counted as true positives if they are clustered into the same OTU,
 and false negatives if they are not in the the same OTU.
 Sequence pairs that are not similar to each other are true negatives if they are
-not assigned to the same OTU, and false positives if they are not in the same
+not clustered into the same OTU, and false positives if they are not in the same
 OTU.
@@
 OptiFit begins where OptiClust ends, starting with a list of reference OTUs and
-their sequences, a list of query sequences to assign to the reference OTUs, and
+their sequences, a list of query sequences to cluster to the reference OTUs, and
 the sequence pairs that are within the distance threshold (e.g. 0.03) 
@@
 Alternatively, a sequence will remain unassigned if the MCC value is maximized
-when the sequence is a singleton rather than assigned to a reference OTU.
+when the sequence is a singleton rather than clustered into a reference OTU.
 All query and reference sequence pairs are considered when calculating the MCC.
@@
 or until a maximum number of iterations is reached (default: 100).
-In the closed reference mode, any query sequences that cannot be assigned to
+In the closed reference mode, any query sequences that cannot be clustered into
 reference OTUs are discarded, and the results only contain OTUs that exist
@@
 There are two strategies for generating OTUs with OptiFit: 
-1) fit the query sequences to reference OTUs generated by _de novo_ clustering an
+1) cluster the query sequences to reference OTUs generated by _de novo_ clustering an
 independent database, or
 2) split the dataset into a reference and query fraction, cluster the reference
-sequences _de novo_, then fit the query sequences to the reference OTUs.
+sequences _de novo_, then cluster the query sequences to the reference OTUs.
 
@@
 _de novo_ methods.
-In closed reference mode, sequences that cannot be assigned to reference OTUs
+In closed reference mode, sequences that cannot be clustered into reference OTUs
 are thrown out, so that the final clustering contains only OTUs that exist in
 the reference.
-To test how OptiFit performs for this purpose, we fit each dataset to three
+To test how OptiFit performs for this purpose, we clustered each dataset to three
 databases of reference OTUs: the Greengenes database, the SILVA non-redundant
@@
 respectively.
-Fitting sequences to Greengenes and SILVA in closed reference mode performed
+clustering sequences to Greengenes and SILVA in closed reference mode performed
 similarly, with median MCC scores of `r closed_fit_gg_mcc` and 
 `r closed_fit_silva_mcc` respectively, while the median MCC 
-was `r closed_fit_rdp_mcc` when fitting to RDP 
+was `r closed_fit_rdp_mcc` when clustering to RDP 
 (see Figure \ref{fig:results_sum}).
@@
 Since closed reference clustering does not cluster query sequences that could
-not be assigned to reference OTUs, an additional measure of clustering
+not be clustered into reference OTUs, an additional measure of clustering
 performance to consider is the fraction of query sequences that were able to be
-assigned.
-On average, more sequences were assigned with Greengenes as the reference 
+clustered.
+On average, more sequences were clustered with Greengenes as the reference 
 (`r frac_fit_gg`%) than with SILVA (`r frac_fit_silva`%) or with the RDP 
@@
 terms of MCC score than either SILVA or RDP.
-Note that _de novo_ and open reference clustering methods always assign 100% of
-sequences to OTUs.
+Note that _de novo_ and open reference clustering methods always cluster 100% of
+sequences into OTUs.
 The database chosen affects the final OTU assignments considerably in terms of
-both MCC score and fraction of query sequences that could be fit to the
+both MCC score and fraction of query sequences that could be clustered into the
 reference OTUs.
@@
 In closed reference mode, OptiFit produced `r mcc_closed_fit_db_vs_vsearch`%
-higher quality OTUs than VSEARCH, but VSEARCH was able to map 
+higher quality OTUs than VSEARCH, but VSEARCH was able to cluster
 `r frac_vsearch_vs_fit`% more query sequences than OptiFit to the Greengenes
@@
 `r sec_vsearch_vs_open_fit_db`% on average respectively.
-Thus, the more stringent OTU definition employed by OptiFit, which requires the
-query to be similar to all other sequences in the OTU rather than to one
-sequence, resulted in fewer sequences being fit to reference OTUs than when
+Thus, the more stringent OTU definition employed by OptiFit, which prefers the
+query sequence to be similar to all other sequences in the OTU rather than to only one
+sequence, resulted in fewer sequences being clustered to reference OTUs than when
 using VSEARCH, but caused OptiFit to outperform VSEARCH in terms of both OTU
@@
 are added.
-Researchers may wish to fit new sequences to existing OTUs or to compare OTUs
+Researchers may wish to cluster new sequences to existing OTUs or to compare OTUs
 across studies.
-To determine how well OptiFit performs for fitting new sequences to existing
+To determine how well OptiFit performs for clustering new sequences to existing
 OTUs, we employed a split dataset strategy, where each dataset was randomly
@@
 Reference sequences were clustered _de novo_ with OptiClust, then query
-sequences were fit to the _de novo_ OTUs with OptiFit.
+sequences were clustered to the _de novo_ OTUs with OptiFit.
 
@@
 mode (`r mcc_opticlust_vs_fit_split_simple`% difference in median MCC).
-In closed reference mode, OptiFit was able to fit `r frac_fit_split`% of query
+In closed reference mode, OptiFit was able to cluster `r frac_fit_split`% of query
 sequences to reference OTUs with the split strategy, a great improvement over
-the average `r frac_fit_gg`% of sequences fit to the Greengenes database.
+the average `r frac_fit_gg`% of sequences clustered to the Greengenes database.
 In terms of run time, closed and open reference OptiFit performed faster than
@@
 
-We developed a new algorithm for fitting sequences to existing OTUs and have
+We developed a new algorithm for clustering sequences to existing OTUs and have
 demonstrated its suitability for reference-based clustering.
@@
 
-When fitting to public databases, OTU quality dropped in closed reference mode
+When clustering to public databases, OTU quality dropped in closed reference mode
 to different degrees depending on the database and dataset source, and no more
-than half of query sequences were able to be fit to OTUs across any
+than half of query sequences were able to be clustered into OTUs across any
 dataset/database combination. 
@@
 higher quality OTUs than closed reference OptiFit with the database strategy, we
-recommend using OptiClust rather than fitting to a database whenever consistent
+recommend using OptiClust rather than clustering to a database whenever consistent
 OTUs are not required.
@@
 OTUs than VSEARCH in open reference, closed reference, or _de novo_ modes.
-However, VSEARCH was able to map more sequences to OTUs than OptiFit in closed
+However, VSEARCH was able to cluster more sequences to OTUs than OptiFit in closed
 reference mode.
 While both OptiFit and VSEARCH use a distance or similarity threshold for
-determining how to assign sequences to OTUs, VSEARCH is more permissive than
+determining how to cluster sequences into OTUs, VSEARCH is more permissive than
 OptiFit.
-The OptiFit and OptiClust algorithms use all of the sequences to define an
-OTU, requiring that all pairs of sequences (including reference and query
-sequences) in an OTU are within the distance threshold without penalizing the
-MCC.
+The OptiFit and OptiClust algorithms use all of the sequences to define an OTU,
+preferring that all pairs of sequences (including reference and query sequences)
+in an OTU are within the distance threshold in order to maximize the MCC.
 In contrast, VSEARCH only requires each query sequence to be similar to the
 single centroid sequence that seeded the OTU.
-Additionally, OTU assignments clustered by VSEARCH are dependent on the order of
-the input sequences, because each query is assigned to the OTU of the first
-centroid sequence that is found within the distance threshold. TODO: is this true of VSEARCH or only USEARCH?
 Because of this, VSEARCH sacrifices OTU quality by allowing more sequences to
-fit to OTUs.
+be clustered into OTUs.
 
-When fitting with the split dataset strategy, OTU quality was remarkably similar
+When clustering with the split dataset strategy, OTU quality was remarkably similar
 when reference sequences were selected by a simple random sample or weighted by
@@
 The similarity in OTU quality between OptiClust and OptiFit with this strategy
-demonstrates the suitability of using OptiFit to fit sequences to existing OTUs,
+demonstrates the suitability of using OptiFit to cluster sequences to existing OTUs,
 such as when comparing OTUs across studies.
@@
 quality OTUs using already existing OTUs as a reference.
-Unlike existing reference-based methods that map query sequences to a single
+Unlike existing reference-based methods that cluster query sequences to a single
 centroid sequence in each reference OTU, OptiFit considers all sequences in each
-reference OTU when fitting query sequences, resulting in OTUs of a similar high
+reference OTU when clustering query sequences, resulting in OTUs of a similar high
 quality as those produced by the _de novo_ OptiClust algorithm.
-Potential applications include fitting sequences to reference databases,
+Potential applications include clustering sequences to reference databases,
 comparing taxonomic composition of microbiomes across different studies, 
 or using OTU-based machine learning models to make predictions on new data.
-OptiFit fills the missing option for fitting query sequences to existing OTUs
+OptiFit fills the missing option for clustering query sequences to existing OTUs
 that does not sacrifice OTU quality for consistency of OTU assignments.
@@
 We processed the raw sequences using mothur according to the Schloss Lab MiSeq
-SOP and accompanying study by Kozich _et al._
-[@schloss_miseq_nodate; @kozich_development_2013].
+SOP [@schloss_miseq_nodate] and accompanying study by Kozich _et al._
+[@kozich_development_2013].
 These steps included trimming and filtering for quality, aligning to the SILVA
@@
 the reference OTUs generated from each of the three databases.
-When reference clustering with VSEARCH, processed datasets were fit directly to
+When reference clustering with VSEARCH, processed datasets were clustered directly to
 the unprocessed Greengenes 97% OTU reference alignment, since this method is how
@@
 of the dataset and for 100 random seeds.
-For each dataset split, the remaining query sequences were assigned to the
+For each dataset split, the remaining query sequences were clustered into the
 reference OTUs with OptiFit.

### 2021-09-21 paper/paper.Rmd

@@
 snakemake <- params$snakemake
+if (is_character(snakemake)) {
+  stop("You need to knit this R Markdown document from the Snakemake workflow")
+}
 load(here(snakemake@input[['rda']]))
@@
 only consider the similarity of each query sequence to a single reference
-sequence in an OTU, thus resulting in some dissimilar sequences being clustered
-to the same OTU.
+sequence in an OTU, thus resulting in OTU assignments that are significantly 
+worse than those generated by _de novo_ methods.
 To provide an efficient and robust method to fit amplicon sequence data to
 existing OTUs, we developed the OptiFit algorithm.
-Like OptiFit, OptiClust considers the similarity of all pairs of reference and
+Inspired by OptiClust, OptiFit considers the similarity of all pairs of reference and
 query sequences in an OTU to produce OTUs of the best possible quality.
@@
 Advancements in DNA sequencing technology have allowed researchers to affordably
-generate millions of reads from microorganisms in diverse natural communities.
+generate millions of sequence reads from microorganisms in diverse environments.
 Efficient and robust software tools are needed to assign microbial sequences
@@
 each sequence and the centroid sequence is considered while clustering.
-As a result, pairs of sequences in the same OTU may have a greater distance than
-the specified threshold, i.e. they are false positives.
+As a result, distances between pairs of sequences in the same OTU are frequenty
+larger than the specified threshold, i.e. they are false positives.
 In contrast, the OptiClust algorithm takes into account the distance between all
@@
 Furthermore, since _de novo_ clustering requires calculating and comparing
-distances between all sequences in a dataset, the execution time can be slow for
-very large datasets.
+distances between all sequences in a dataset, the execution time can be slow and
+memory requirements can be prohibitive for very large datasets.
 Reference clustering attempts to overcome the limitations of _de novo_
@@
 diversity of the reference database.
-Rare or novel sequences in the sample will be lost in closed reference mode if
+Novel sequences in the sample will be lost in closed reference mode if
 they are not represented by a similar sequence in the database.
-Just like in _de novo_ clustering, reference-based methods that use greedy
-clustering algorithms only consider the distance between each query and a single
-reference sequence, causing some sequences in the same OTU to be more dissimilar
-to each other than the specified threshold.
 Previous studies found that the OptiClust _de novo_ clustering algorithm created
@@
 Here, we tested the OptiFit algorithm with the reference as a public database
-(e.g. Greengenes) or _de novo_ OTUs and compared the performance to existing
-tools.
+(e.g. Greengenes) or _de novo_ OTUs generated using a reference set from the
+full dataset and compared the performance to existing tools.
 To evaluate the OptiFit algorithm and compare to existing methods, we used four
@@
 
-While _de novo_ clustering produces the highest quality OTUs, researchers may
-prefer to perform reference clustering to a public database because
-reference-based methods produce consistent OTUs and are generally faster than
-_de novo_ methods.
-In closed reference mode, sequences that cannot be clustered into reference OTUs
-are thrown out, so that the final clustering contains only OTUs that exist in
-the reference.
-To test how OptiFit performs for this purpose, we clustered each dataset to three
-databases of reference OTUs: the Greengenes database, the SILVA non-redundant
-database, and the Ribosomal Database Project (RDP) [@desantis_greengenes_2006;
-@quast_silva_2013; @cole_ribosomal_2014].
+To test how OptiFit performs for reference-based clustering, we clustered each
+dataset to three databases of reference OTUs: the Greengenes database, the SILVA
+non-redundant database, and the Ribosomal Database Project (RDP)
+[@desantis_greengenes_2006; @quast_silva_2013; @cole_ribosomal_2014].
 Reference OTUs for each database were created by performing _de novo_ clustering
@@
 respectively.
-clustering sequences to Greengenes and SILVA in closed reference mode performed
+Clustering sequences to Greengenes and SILVA in closed reference mode performed
 similarly, with median MCC scores of `r closed_fit_gg_mcc` and 
@@
 was `r closed_fit_rdp_mcc` when clustering to RDP 
-(see Figure \ref{fig:results_sum}).
+(Figure \ref{fig:results_sum}).
 For comparison, clustering datasets with OptiClust produced an average MCC score
@@
 (`r frac_fit_gg`%) than with SILVA (`r frac_fit_silva`%) or with the RDP 
-(`r frac_fit_rdp`%).
+(`r frac_fit_rdp`%) (Figure \ref{fig:results_sum}).
 This mirrored the result reported above that Greengenes produced better OTUs in
@@
 sequences into OTUs.
-The database chosen affects the final OTU assignments considerably in terms of
-both MCC score and fraction of query sequences that could be clustered into the
-reference OTUs.
+The database chosen affects the final closed reference OTU assignments
+considerably in terms of both MCC score and fraction of query sequences that
+could be clustered into the reference OTUs.
 
@@
 Across all dataset and database combinations, closed reference OptiFit
-outperformed both OptiClust and open reference OptiFit.
+outperformed both OptiClust and open reference OptiFit (Figure \ref{fig:results_sum}).
 For example, with the human dataset fit to SILVA reference OTUs, the average run
@@
 is considered when assigning sequences to OTUs.
-_De novo_ clustering with OptiClust produced `r mcc_opticlust_vs_vsearch`%
+<!--**TODO: delete or move this sentence:** _De novo_ clustering with OptiClust produced `r mcc_opticlust_vs_vsearch`%
 higher quality OTUs than VSEARCH in terms of MCC, but performed 
-`r sec_opticlust_vs_vsearch`% slower than VSEARCH.
+`r sec_opticlust_vs_vsearch`% slower than VSEARCH.-->
 In closed reference mode, OptiFit produced `r mcc_closed_fit_db_vs_vsearch`%
@@
 `r frac_vsearch_vs_fit`% more query sequences than OptiFit to the Greengenes
-reference database.
+reference database (Figure \ref{fig:results_sum}).
 This is because VSEARCH only considers the distances between each query sequence
 to the single reference sequence, while OptiFit considers the distances between
-all pairs of sequences in an OTU. <!-- VSEARCH "diameter" vs OptiFit "radius"
--->
+all pairs of reference and query sequences in an OTU. 
 When open reference clustering, OptiFit produced higher quality OTUs than
@@
 using the split dataset strategy with half of the sequences selected for
-the reference by a simple random sample (a 50% split).
-OTU quality was highly similar to that from OptiClust regardless of
-mode (`r mcc_opticlust_vs_fit_split_simple`% difference in median MCC).
+the reference by a simple random sample (a 50% split) 
+(Figure \ref{fig:results_sum}; self-split).
+OTU quality was similar to that from OptiClust regardless of
+mode (`r mcc_opticlust_vs_fit_split_simple`% difference in median MCC [**TODO: why is opticlust worse than optifit now?!?!**]).
 In closed reference mode, OptiFit was able to cluster `r frac_fit_split`% of query
@@
 To identify the best reference size, reference sets with 10% to 90% of the
-sequences were created, with the remaining sequences used for the query.
+sequences were created, with the remaining sequences used for the query 
+(Figure \ref{fig:results_split}).
 OTU quality was remarkably consistent across reference fraction sizes.
 For example, splitting the human dataset 100 times yielded a coefficient of
-variation of `r cv_fit_split_mcc_human_simple` for the MCC score across all
-fractions.
+variation (i.e. the standard deviation divided by the mean) of 
+`r cv_fit_split_mcc_human_simple` for the MCC score across all fractions.
 Run time generally decreased as the reference fraction increased; for the human
-dataset, the median run time was `r sec_fit_split_human_simple_1` with 10% of
-sequences in the reference and `r sec_fit_split_human_simple_9` with 90% of
-sequences in the reference (Figure \ref{fig:results_split}).
+dataset, the median run time was `r sec_fit_split_human_simple_1` seconds with
+10% of sequences in the reference and `r sec_fit_split_human_simple_9` seconds
+with 90% of sequences in the reference.
 In closed reference mode, the fraction of sequences that mapped increased as the
@@
 `r frac_fit_split_human_simple_9` with 90% of sequences in the reference. 
-These trends held for the other datasets as well (Figure 
-\ref{fig:results_split}).
+These trends held for the other datasets as well.
 Thus, the reference fraction did not affect OTU quality in terms of MCC score,
@@
 relative abundance, and weighting by similarity to other sequences in the
-dataset.
+dataset (Figure \ref{fig:results_split}).
 OTU quality in terms of MCC was similar with the simple and abundance-weighted
@@
 similar for simple and abundance-weighted sampling (median fraction mapped of 
-`r frac_fit_split_simple` and `r frac_fit_split_abun` respectively), but worse 
+`r frac_fit_split_simple` and `r frac_fit_split_abun`, respectively), but worse 
 for similarity-weighted sampling (median fraction mapped of `r frac_fit_split_dist`).
@@
 determining how to cluster sequences into OTUs, VSEARCH is more permissive than
-OptiFit.
+OptiFit regardless of mode.
 The OptiFit and OptiClust algorithms use all of the sequences to define an OTU,
@@
 single centroid sequence that seeded the OTU.
-Because of this, VSEARCH sacrifices OTU quality by allowing more sequences to
-be clustered into OTUs.
+Because of this, VSEARCH sacrifices OTU quality by allowing more dissimilar
+sequences to be clustered into OTUs.
 
@@
 
-We have developed a new clustering algorithm that allows users to produce high
-quality OTUs using already existing OTUs as a reference.
 Unlike existing reference-based methods that cluster query sequences to a single

### 2021-09-21 code/slurm/submit.sh
May need to cd into subworkflows when using slurm profile

@@ -6,7 +6,7 @@
 #### #### ####  These are the most frequently changing options
 
 ####  Job name
-#SBATCH --job-name=ofa
+#SBATCH --job-name=OptiFit
 
 ####  Request resources here
 ####    These are typically, number of processors, amount of memory,
@@ -35,5 +35,11 @@
 #SBATCH --mail-user=YOUR_EMAIL
 #SBATCH --mail-type=BEGIN,END
 
+alias snakemake_cmd="time snakemake --profile config/slurm_KLS --latency-wait 90"
 source /etc/profile.d/http_proxy.sh  # required for internet on the Great Lakes cluster
-time snakemake --profile config/slurm --latency-wait 90
+for dir in $(ls subworkflows); do
+    pushd subworkflows/${dir}
+    snakemake_cmd
+    popd
+done
+snakemake_cmd

### 2021-09-22 paper/paper.Rmd

@@
     fig_caption: yes
+  github_document:
+    includes:
+      in_header: preamble.tex
+      before_body: head.tex
+    html_preview: false
 csl: msystems.csl

### 2021-10-21 code/tests/testthat/test-functions.R
🎨  Style R code

@@ -1,1 +1,1 @@
-source(here('code', 'R', 'functions.R'))
+source(here("code", "R", "functions.R"))

### 2021-10-21 code/py/combine_open_lists.py
🎨  Style Python code 🐍

@@ -14,22 +14,28 @@ class mothurList:
         self.otu_assigns = otu_assigns
 
     def __eq__(self, other):
-        return all([self.__class__ ==  other.__class__,
-                    self.label == other.label,
-                    self.otu_assigns == other.otu_assigns])
+        return all(
+            [
+                self.__class__ == other.__class__,
+                self.label == other.label,
+                self.otu_assigns == other.otu_assigns,
+            ]
+        )
 
     @classmethod
     def from_list_file(cls, list_filename):
         with open(list_filename, "r") as listfile:
-            lines = [line.strip().split('\t') for line in listfile]
+            lines = [line.strip().split("\t") for line in listfile]
             len_lines = len(lines)
             dat_line = []
-            if len_lines == 1: # for when there's no header line
+            if len_lines == 1:  # for when there's no header line
                 dat_line = lines[0]
-            elif len_lines == 2: # ditch the header line
+            elif len_lines == 2:  # ditch the header line
                 dat_line == lines[1]
-            else: # this shouldn't happen
-                raise ValueError(f"List file contains {len_lines} lines, but only 1 or 2 were expected.")
+            else:  # this shouldn't happen
+                raise ValueError(
+                    f"List file contains {len_lines} lines, but only 1 or 2 were expected."
+                )
             list_label = dat_line[0]
             otu_assigns = dat_line[2:]
         return cls(list_label, otu_assigns)
@@ -56,11 +62,15 @@ class mothurList:
             listfile.write(f"{self.label}\t{str(self.num_otus)}")
             for otu in self.otu_assigns:
                 listfile.write(f"\t{otu}")
-            listfile.write('\n')
+            listfile.write("\n")
 
 
 if __name__ == "__main__":
     if "snakemake" in locals() or "snakemake" in globals():
-        main(snakemake.input.list_closed, snakemake.input.list_denovo, snakemake.output.list)
+        main(
+            snakemake.input.list_closed,
+            snakemake.input.list_denovo,
+            snakemake.output.list,
+        )
     else:
         main(sys.argv[1], sys.argv[2], sys.argv[3])

### 2021-10-21 code/py/create_test_uc_files.py

@@ -1,25 +1,24 @@
 def main(uc1_filename, uc2_filename, list_filename):
-    otus1 = {1: ['a', 'b'],
-             2: ['c', 'd'],
-             3: ['e']
-             }
-    otus2 = {1: ['f', 'g'],
-             2: ['h']
-             }
-    for filename, otus in ((uc1_filename, otus1),
-                           (uc2_filename, otus2)):
-        with open(filename, 'w') as uc_file:
+    otus1 = {1: ["a", "b"], 2: ["c", "d"], 3: ["e"]}
+    otus2 = {1: ["f", "g"], 2: ["h"]}
+    for filename, otus in ((uc1_filename, otus1), (uc2_filename, otus2)):
+        with open(filename, "w") as uc_file:
             for otu_id, seqs in otus.items():
                 for seq_id in seqs:
-                    uc_file.write(f"H\t{otu_id}\t1\t100\t+\t-\t-\t=\t{seq_id}\t{otu_id}\n")
-    combined = [','.join(seqs) for otus in [otus1, otus2] for seqs in otus.values()]
-    with open(list_filename, 'w') as listfile:
+                    uc_file.write(
+                        f"H\t{otu_id}\t1\t100\t+\t-\t-\t=\t{seq_id}\t{otu_id}\n"
+                    )
+    combined = [",".join(seqs) for otus in [otus1, otus2] for seqs in otus.values()]
+    with open(list_filename, "w") as listfile:
         listfile.write(f"userLabel\t{str(len(combined))}")
         for otu in combined:
             listfile.write(f"\t{otu}")
-        listfile.write('\n')
+        listfile.write("\n")
+
 
 if __name__ == "__main__":
-    main('code/tests/data/closed.uc',
-         'code/tests/data/denovo.uc',
-         'code/tests/data/oracle_open.list')
+    main(
+        "code/tests/data/closed.uc",
+        "code/tests/data/denovo.uc",
+        "code/tests/data/oracle_open.list",
+    )

### 2021-10-21 code/py/fraction_mapped_sanity_check.py

@@ -4,10 +4,12 @@ calculate the fraction of reads that mapped during closed-reference clustering
 """
 import numpy as np
 
+
 def count_seqs(infilename):
     with open(infilename, "r") as seq_file:
-        num_seqs = sum(1 for line in seq_file if line.startswith('>'))
+        num_seqs = sum(1 for line in seq_file if line.startswith(">"))
     return num_seqs
+
 
 def get_fraction_mapped(num_query_seqs, num_unmapped_seqs):
     fraction_mapped = 1 - round(num_unmapped_seqs / num_query_seqs, 3)
@@ -20,22 +22,23 @@ def main():
     num_unmapped = count_seqs(snakemake.input.unmapped)
     frac_mapped_fasta = get_fraction_mapped(num_query, num_unmapped)
 
-    with open(snakemake.input.tsv, 'r') as tsv_file:
-        header_line = next(tsv_file).split('\t')
+    with open(snakemake.input.tsv, "r") as tsv_file:
+        header_line = next(tsv_file).split("\t")
         print(header_line)
-        assert header_line[4].strip() == 'fraction_mapped'
-        data_line = next(tsv_file).split('\t')
+        assert header_line[4].strip() == "fraction_mapped"
+        data_line = next(tsv_file).split("\t")
         frac_mapped_list = float(data_line[4].strip())
 
-    print('frac_mapped_fasta: ', frac_mapped_fasta)
-    print('frac_mapped_list:  ', frac_mapped_list)
+    print("frac_mapped_fasta: ", frac_mapped_fasta)
+    print("frac_mapped_list:  ", frac_mapped_list)
 
     is_close = np.isclose(frac_mapped_fasta, frac_mapped_list)
-    print('is_close: ', is_close)
+    print("is_close: ", is_close)
     assert is_close
 
-    with open(snakemake.output.txt, 'w') as outfile:
-        outfile.write('frac_mapped_fasta\tfrac_mapped_list\tis_close\n')
+    with open(snakemake.output.txt, "w") as outfile:
+        outfile.write("frac_mapped_fasta\tfrac_mapped_list\tis_close\n")
         outfile.write(f"{frac_mapped_fasta}\t{frac_mapped_list}\t{is_close}\n")
+
 
 main()

### 2021-10-21 code/py/uc_to_list.py

@@ -1,25 +1,27 @@
 import collections
 import sys
 
+
 def uc_to_list(clustered_filename, list_filename):
     clusters = collections.defaultdict(set)
-    with open(clustered_filename, 'r') as clustered_file:
+    with open(clustered_filename, "r") as clustered_file:
         for line in clustered_file:
-            line = line.split('\t')
+            line = line.split("\t")
             # https://drive5.com/usearch/manual/opt_uc.html
             record_type = line[0]
             cluster_num = line[1]
-            query_label = line[8].split(';size=')[0]
-            if record_type in {'H','S'}:
+            query_label = line[8].split(";size=")[0]
+            if record_type in {"H", "S"}:
                 clusters[cluster_num].add(query_label)
-    with open(list_filename, 'w') as list_file:
-        list_file.write('userLabel\t' + str(len(clusters)))
+    with open(list_filename, "w") as list_file:
+        list_file.write("userLabel\t" + str(len(clusters)))
         for cluster_id, seqs in clusters.items():
-            list_file.write('\t' + ','.join(sorted(seqs)))
-        list_file.write('\n')
+            list_file.write("\t" + ",".join(sorted(seqs)))
+        list_file.write("\n")
+
 
 if __name__ == "__main__":
-    if 'snakemake' in globals() or 'snakemake' in locals():
+    if "snakemake" in globals() or "snakemake" in locals():
         uc_to_list(snakemake.input.uc, snakemake.output.list)
     else:
         uc_to_list(sys.argv[1], sys.argv[2])

### 2021-10-21 code/tests/test_python.py

@@ -4,35 +4,40 @@ from ..py.uc_to_list import uc_to_list
 
 
 class testMothurList(unittest.TestCase):
-
     def test_combine(self):
-        ml1 = mothurList('userLabel', ['a,b', 'c,d', 'e'])
-        ml2 = mothurList('userLabel', ['f,g', 'h'])
+        ml1 = mothurList("userLabel", ["a,b", "c,d", "e"])
+        ml2 = mothurList("userLabel", ["f,g", "h"])
         ml1.combine(ml2)
-        self.assertEqual(ml1, mothurList('userLabel', ['a,b', 'c,d', 'e', 'f,g', 'h']))
+        self.assertEqual(ml1, mothurList("userLabel", ["a,b", "c,d", "e", "f,g", "h"]))
 
-    def test_write(self, list_filename_test = 'code/tests/data/test_open.list',
-                         list_filename_oracle = 'code/tests/data/oracle_open.list'):
-        ml1 = mothurList('userLabel', ['a,b', 'c,d', 'e', 'f,g', 'h'])
+    def test_write(
+        self,
+        list_filename_test="code/tests/data/test_open.list",
+        list_filename_oracle="code/tests/data/oracle_open.list",
+    ):
+        ml1 = mothurList("userLabel", ["a,b", "c,d", "e", "f,g", "h"])
         ml1.write(list_filename_test)
 
-        with open(list_filename_test, 'r') as infile:
+        with open(list_filename_test, "r") as infile:
             test_list = infile.readlines()
-        with open(list_filename_oracle, 'r') as infile:
+        with open(list_filename_oracle, "r") as infile:
             oracle_list = infile.readlines()
 
         self.assertEqual(test_list, oracle_list)
 
 
 class testUcList(unittest.TestCase):
-    def test_convert(self, uc_filename = 'code/tests/data/closed.uc',
-                           list_filename_test = 'code/tests/data/test_closed.list',
-                           list_filename_oracle = 'code/tests/data/oracle_closed.list'):
+    def test_convert(
+        self,
+        uc_filename="code/tests/data/closed.uc",
+        list_filename_test="code/tests/data/test_closed.list",
+        list_filename_oracle="code/tests/data/oracle_closed.list",
+    ):
         uc_to_list(uc_filename, list_filename_test)
 
-        with open(list_filename_test, 'r') as infile:
+        with open(list_filename_test, "r") as infile:
             test_list = infile.readlines()
[…]

### 2021-10-21 paper/paper.Rmd

@@
 reference mode by `r sec_closed_fit_db_vs_vsearch`% and 
-`r sec_vsearch_vs_open_fit_db`% on average respectively.
+`r sec_vsearch_vs_open_fit_db`% on average, respectively.
 Thus, the more stringent OTU definition employed by OptiFit, which prefers the
@@
 OTU quality was similar to that from OptiClust regardless of
-mode (`r mcc_opticlust_vs_fit_split_simple`% difference in median MCC [**TODO: why is opticlust worse than optifit now?!?!**]).
+mode (`r mcc_opticlust_vs_fit_split_simple`% difference in median MCC).
 In closed reference mode, OptiFit was able to cluster `r frac_fit_split`% of query
@@
 OptiClust on whole datasets by `r sec_closed_fit_split_vs_clust`% and 
-`r sec_open_fit_split_vs_clust` respectively.
+`r sec_open_fit_split_vs_clust`%, respectively.
 The split dataset strategy also performed `r sec_closed_fit_split_vs_db`% faster

### 2021-10-21 exploratory/2021/seq_stats.Rmd
Plot similarity & abundance of seqs

---
date: '`r Sys.Date()`'
output: 
  github_document:
    html_preview: false
---

# Similarity & Abundance of Sequences

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      fig.path = "figures/",
                      dpi = 300
                      )
```
```{r deps, warning=FALSE, message=FALSE}
set.seed(20211021)
library(glue)
library(here)
library(tidyverse)
theme_set(theme_bw())
```

```{r dat}
seq_stats <- c("human","marine","mouse","soil") %>% map_dfr(function(x) {
  read_tsv(glue('subworkflows/1_prep_samples/data/{x}/seq_stats.tsv')) %>% 
    mutate(dataset = x)
}) %>% pivot_longer(c(similarity, abundance), names_to = 'statistic')
```

```{r hist_sim_abun}
seq_stats %>% 
  ggplot(aes(x=value, fill=statistic)) +
  geom_histogram() +
  facet_grid(dataset ~ statistic, scales='free')
```

## log10

```{r hist_sim_abun_log10}
seq_stats %>% 
  ggplot(aes(x=value, fill=statistic)) +
  geom_histogram() +
  scale_x_log10() +
  facet_wrap("dataset")
```

## abundance vs similarity

```{r point_abun_sim}
seq_stats %>% 
  pivot_wider(names_from = statistic) %>% 
  ggplot(aes(x = abundance, y = similarity, color = dataset)) +
  geom_point(alpha=0.7) +
  scale_x_log10() +
  facet_wrap('dataset')
```

### 2021-10-22 exploratory/2021/seq_stats.Rmd
Make a boxplot too
Calculate relative values & summary stats

@@
 # Similarity & Abundance of Sequences
+
+Each sequence has an absolute abundance and similarity value, where similarity
+is the count of other sequences within the dataset that it is similar to per the
+0.03 distance threshold. I was wondering if the distributions of those values
+would have some pattern that corresponded to what we see with the MCC scores for
+weighting by those methods. There isn't anything obviously weird or wrong. For
+all datasets most sequences are very low abundance (expected). Both marine and
+soil have a much lower median similarity value than human and mouse, and the
+marine and soil datasets also have a huge spread in MCC between open & closed
+with similarity weighting.
 
@@
 
+### histograms
+
 ```{r hist_sim_abun}
@@
 ```
-
-## log10
 
@@
 
-## abundance vs similarity
+### boxplots
+
+```{r box_sim_abun}
+seq_stats %>% 
+  ggplot(aes(x=value, y=dataset, fill=statistic)) +
+  geom_boxplot() +
+  facet_wrap('statistic', scales = 'free')
+``` 
+```{r box_sim_abun_log10}
+seq_stats %>% 
+  ggplot(aes(x=value, y=dataset, fill=statistic)) +
+  geom_boxplot() +
+  scale_x_log10() +
+  facet_wrap('statistic', scales = 'free')
+```
+
+### abundance vs similarity
 
@@
 ```
+
+## relative abundance & simlarity
+
+```{r calc_relative_stats}
+totals <- seq_stats %>%
+  pivot_wider(names_from = statistic) %>% 
+  group_by(dataset) %>% 
+  summarize(tot_simi = sum(similarity),
+            tot_abun = sum(abundance))
+
+rel_stats <- c("human", "marine", "mouse", "soil") %>% map_dfr(function(x) {
+  read_tsv(glue('subworkflows/1_prep_samples/data/{x}/seq_stats.tsv')) %>% 
+    mutate(dataset = x,
+           rel_sim = similarity / (totals %>% 
+                                     filter(dataset == x) %>% 
+                                     pull(tot_simi)),
+           rel_abun = abundance / (totals %>% 
+                                     filter(dataset == x) %>% 
+                                     pull(tot_abun))
+           )
+})
+```
+
+### rel abun vs rel sim
+
+```{r point_rel_abun_sim}
+rel_stats %>% 
+  ggplot(aes(x = rel_abun, y = rel_sim, color = dataset)) +
+  geom_point(alpha=0.7) +
+  facet_wrap('dataset')
+```
+```{r point_rel_abun_sim_log10}
+rel_stats %>% 
+  ggplot(aes(x = rel_abun, y = rel_sim, color = dataset)) +
+  geom_point(alpha=0.7) +
+  scale_x_log10() +
+  facet_wrap('dataset')
+```
+
+### relative boxplots
+
+```{r box_rel_sim_abun}
+rel_stats %>% pivot_longer(c(rel_sim, rel_abun), names_to = 'statistic') %>% 
+  ggplot(aes(x=value, y=dataset, fill=statistic)) +
+  geom_boxplot() +
+  facet_wrap('statistic', scales = 'free')
+```
+
+```{r box_rel_sim_abun_log10}
+rel_stats %>% pivot_longer(c(rel_sim, rel_abun), names_to = 'statistic') %>% 
+  ggplot(aes(x=value, y=dataset, fill=statistic)) +
+  geom_boxplot() +
+  scale_x_log10() +
+  facet_wrap('statistic', scales = 'free')
+```
+
+### relative histograms
+
+```{r hist_rel_sim_abun}
+rel_stats %>%  pivot_longer(c(rel_sim, rel_abun), names_to = 'statistic') %>% 
+  ggplot(aes(x=value, fill=statistic)) +
+  geom_histogram() +
+  facet_wrap("dataset")
+```
+
+```{r hist_rel_sim_abun_log10}
+rel_stats %>%  pivot_longer(c(rel_sim, rel_abun), names_to = 'statistic') %>% 
+  ggplot(aes(x=value, fill=statistic)) +
+  geom_histogram() +
+  scale_x_log10() +
+  facet_wrap("dataset")
+```
+
+## summary stats
+
+```{r summary_abundance}
+rel_stats %>% 
+  group_by(dataset) %>% 
+  summarize_at(vars(abundance), 
+               list(min=min, median=median, mean=mean, max=max)) %>% 
+  knitr::kable()
+```
+
+```{r summary_similarity}
+rel_stats %>% 
+  group_by(dataset) %>% 
+  summarize_at(vars(similarity), 
+               list(min=min, median=median, mean=mean, max=max)) %>% 
+  knitr::kable()
+```

### 2021-10-22 exploratory/2021/seq_stats.Rmd
Make a table of quantiles

@@
 ```{r dat}
-seq_stats <- c("human","marine","mouse","soil") %>% map_dfr(function(x) {
+dataset_names <- c("human","marine","mouse","soil")
+seq_stats <- dataset_names %>% map_dfr(function(x) {
   read_tsv(glue('subworkflows/1_prep_samples/data/{x}/seq_stats.tsv')) %>% 
@@
   facet_wrap("dataset")
+```
+```{r hist_sim_abun_log10bins}
+seq_stats %>% 
+  ggplot(aes(x=value, fill=statistic)) +
+  geom_histogram(bins=10) +
+  scale_x_log10() +
+  facet_grid(dataset ~ statistic, scales='free')
 ```
@@
 
-rel_stats <- c("human", "marine", "mouse", "soil") %>% map_dfr(function(x) {
+rel_stats <- dataset_names %>% map_dfr(function(x) {
   read_tsv(glue('subworkflows/1_prep_samples/data/{x}/seq_stats.tsv')) %>% 
@@
 ```
+
+```{r quantiles}
+quantile_custom <- function(x, by = 0.05) {
+  x %>% 
+    quantile(., seq.int(from = 0.1, to = 1, by = by))
+}
+get_quantile <- function(dat, dataset_name, var = abundance) {
+  quantiles <- dat %>% 
+    filter(dataset == dataset_name) %>% 
+    pull({{ var }}) %>%  
+    quantile_custom()
+  data.frame(quant = names(quantiles), 
+             value = quantiles, 
+             row.names = NULL) %>% 
+    mutate(dataset = dataset_name)
+}
+get_dataset_quantiles <- function(dat, column = abundance) {
+  dataset_names %>% 
+    map_dfr(function(x) {get_quantile(dat = dat, 
+                                    dataset_name = x,
+                                    var = {{column}})
+                         }) %>% 
+    pivot_wider(names_from = dataset, values_from = value) 
+}
+```
+
+### abundance quantiles
+
+```{r abun_quant}
+rel_stats %>% 
+  get_dataset_quantiles(column = abundance) %>% 
+  knitr::kable()
+```
+
+### similarity quantiles
+
+```{r simi_quant}
+rel_stats %>% 
+  get_dataset_quantiles(column = similarity) %>% 
+  knitr::kable()
+```

### 2021-10-22 exploratory/2021/printref-test_2021-10/README.Rmd
Tested printref=t for split dataset strategy

---
date: '`r Sys.Date()`'
output: 
  github_document:
    html_preview: false
---

# printref for split dataset strategy

With `printref=f` the reference sequences are not included in the final MCC
score, which makes sense when fitting to an independent database like
greengenes. but **for the split dataset strategy, wouldn't it make more sense to
include ref seqs in the MCC to be a more comparable comparison to de novo?**

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      fig.path = "figures/",
                      dpi = 300
                      )
```
```{r deps, warning=FALSE, message=FALSE}
set.seed(20211022)
library(cowplot)
library(ggtext)
library(glue)
library(here)
library(knitr)
library(tidyverse)

mutate_perf <- function(dat) {
  dat %>%
    mutate(
      mem_mb = max_rss,
      mem_gb = mem_mb / 1024
    ) %>%
    rename(sec = s)
}
select_cols <- function(dat) {
  dat %>%
    select(
      dataset, strategy, method, tool, mcc, sec, mem_gb, fraction_mapped,
      ref_frac, ref_weight, printref
    )
}
color_breaks <- list(
  simple = "#FF8C00",
  abundance = "#9932CC",
  similarity = "#008B8B"
)
color_labels <- lapply(
  names(color_breaks),
  function(name) {
    glue("<span style = 'color:{color_breaks[[name]]};'>{name}</span>")
  }
) %>% unlist()
color_values <- append(color_breaks, list(`NA` = "#000000"))
```

```{r dat}

opticlust <- read_tsv(here("subworkflows/1_prep_samples/results/opticlust_results.tsv")) %>%
  full_join(read_tsv(here("subworkflows/1_prep_samples/results/dataset_sizes.tsv"))) %>%
  mutate_perf() %>%
  mutate(
    strategy = method,
    fraction_mapped = NA,
    ref_frac = 0,
    ref_weight = "NA",
    printref = TRUE
  )
optifit_split <- read_tsv(here("exploratory", "2021", "printref-test_2021-10", 
                               "data", "optifit_split_results.tsv")) %>%
  mutate_perf() %>%
  mutate(strategy = "self-split")

dat <- list(optifit_split, opticlust) %>%
  lapply(select_cols) %>%
  reduce(bind_rows) %>%
  mutate(
    method = as.character(method),
    strategy = as.character(strategy)
  ) %>%
  mutate(fraction_mapped = case_when(
    method %>% as.character() != "closed" ~ NA_real_,
    TRUE ~ fraction_mapped
  )) %>%
  pivot_longer(c(mcc, fraction_mapped, sec),
    names_to = "metric"
  ) %>%
  mutate(
    metric = factor(
      case_when(
        metric == "mcc" ~ "MCC",
        metric == "fraction_mapped" ~ "Fraction Mapped",
        metric == "sec" ~ "Runtime (sec)",
        TRUE ~ metric
      ),
      levels = c("MCC", "Fraction Mapped", "Runtime (sec)")
    ),
    strategy = factor(
      case_when(
        strategy == "de_novo" ~ "_de novo_",
        strategy == "database_rdp" ~ "db: RDP",
        strategy == "database_silva" ~ "db: SILVA",
        strategy == "database_gg" ~ "db: Greengenes",
        TRUE ~ strategy
      ),
      levels = c(
        "db: RDP", "db: SILVA", "db: Greengenes",
        "self-split", "_de novo_"
      )
    ),
    method = factor(
      case_when(
        method == "de_novo" ~ "_de novo_",
        TRUE ~ method
      ),
      levels = c("open", "closed", "_de novo_")
    ),
    ref_weight = factor(
      case_when(
        ref_weight == "distance" ~ "similarity",
        TRUE ~ ref_weight
      ),
      levels = c("simple", "abundance", "similarity", "NA")
    )
  )
```

## facet by printref

```{r strategy_printref-facet}
dat %>%
  filter(dataset == 'human') %>% 
  #filter(((ref_weight == "simple" | ref_frac == 0.5) | method == "_de novo_") & !is.na(value)) %>%
  ggplot(aes(ref_frac, value, color = ref_weight, shape = method)) +
  coord_flip() +
  stat_summary(
    geom = "point",
    fun = median,
    size = 2,
    position = position_dodge(width = 0.07)
  ) +
  facet_grid(printref ~ metric, scales = "free", switch = "x") +
  scale_shape_manual(values = list(open = 1, closed = 19, `_de novo_` = 17)) +
  scale_color_manual(
    values = color_values,
    breaks = names(color_breaks),
    labels = color_labels
  ) +
  scale_x_continuous(
    breaks = c(0, 0.4, 0.5, 0.6),
    labels = c("NA", 0.4, 0.5, 0.6)
  ) +
  labs(x = "reference fraction", y = "") +
  theme_bw() +
  theme(
    legend.text = element_markdown(),
    legend.title = element_blank(),
    legend.position = "top",
    legend.margin = margin(t = 0, r = 0, b = 0, l = 0, unit = "pt"),
    plot.margin = unit(x = c(0, 0, 0, 0), units = "pt"),
    panel.grid.minor.y = element_blank(),
    axis.title.x = element_blank(),
    strip.placement = "outside",
    strip.background = element_blank()
  ) +
  guides(
    shape = guide_legend(order = 1),
    colour = guide_legend(
      override.aes = list(size = -1),
      order = 2
    )
  )
```

## printref=TRUE

```{r strategy_printref-true}
dat %>%
  filter(printref & dataset == 'human') %>% 
  #filter(((ref_weight == "simple" | ref_frac == 0.5) | method == "_de novo_") & !is.na(value)) %>%
  ggplot(aes(ref_frac, value, color = ref_weight, shape = method)) +
  coord_flip() +
  stat_summary(
    geom = "point",
    fun = median,
    size = 2,
    position = position_dodge(width = 0.07)
  ) +
  facet_grid(dataset ~ metric, scales = "free", switch = "x") +
  scale_shape_manual(values = list(open = 1, closed = 19, `_de novo_` = 17)) +
  scale_color_manual(
    values = color_values,
    breaks = names(color_breaks),
    labels = color_labels
  ) +
  scale_x_continuous(
    breaks = c(0, 0.4, 0.5, 0.6),
    labels = c("NA", 0.4, 0.5, 0.6)
  ) +
  labs(x = "reference fraction", y = "") +
  theme_bw() +
  theme(
    legend.text = element_markdown(),
    legend.title = element_blank(),
    legend.position = "top",
    legend.margin = margin(t = 0, r = 0, b = 0, l = 0, unit = "pt"),
    plot.margin = unit(x = c(0, 0, 0, 0), units = "pt"),
    panel.grid.minor.y = element_blank(),
    axis.title.x = element_blank(),
    strip.placement = "outside",
    strip.background = element_blank()
  ) +
  guides(
    shape = guide_legend(order = 1),
    colour = guide_legend(
      override.aes = list(size = -1),
      order = 2
    )
  )
```

### 2021-10-22 paper/paper.Rmd
Use assert_that() to check assumptions

@@
 ```{r deps, message = FALSE}
+library(assertthat)
 library(cowplot)
@@
 }
-load(here(snakemake@input[['rda']]))
-ref_seqs <- read_tsv(here('subworkflows', '0_prep_db', 'data', 
-                          'seq_counts.tsv')) %>% 
-  pivot_wider(names_from = ref, values_from = nseqs) %>% as.list()
+load(here("results", "stats.RData"))
 
@@
   return(value)
+}
+
+close_enough <- function(x, y, tol = 10^-2) {
+  assert_that(all(near(x, y, tol = tol)))
 }
@@
                            TRUE ~ toupper(db_names))
+close_enough(dn_dbs_mcc['gg'], dn_dbs_mcc['silva'])
+close_enough(dn_dbs_mcc['gg'], dn_dbs_mcc['rdp'], 10^-1)
+mcc_decreasing <- c(closed_fit_rdp_mcc, closed_fit_silva_mcc, closed_fit_gg_mcc,
+                    open_fit_silva_mcc, open_fit_gg_mcc, open_fit_rdp_mcc, 
+                    opticlust_mcc)
+assert_that(!is.unsorted(mcc_decreasing))
 ```
@@
 
+```{r}
+assert_that(!is.unsorted(c(frac_fit_rdp, frac_fit_silva, frac_fit_gg)))
+assert_that((frac_fit_gg - frac_fit_silva) < 10)
+assert_that((frac_fit_gg - frac_fit_rdp) > 40)
+```
+
 Since closed reference clustering does not cluster query sequences that could
@@
 could be clustered into the reference OTUs.
+
+```{r}
+assert_that(!is.unsorted(c(closed_fit_silva_human_sec, 
+                           opticlust_human_sec,
+                           open_fit_silva_human_sec)))
+```
 
@@
   grid::grid.raster() 
+```
+
+```{r}
+assert_that(mcc_closed_fit_db_vs_vsearch > 10)
+assert_that(frac_vsearch_vs_fit > 10)
+assert_that((open_fit_gg_mcc - open_vsearch_mcc) > 0.20)
+assert_that((sec_vsearch_vs_open_fit_db - sec_closed_fit_db_vs_vsearch) < 5)
 ```
@@
 
+```{r}
+close_enough(mcc_opticlust_vs_fit_split_simple, 0.02)
+assert_that((frac_fit_split - frac_fit_gg) > 20)
+assert_that(!is.unsorted(c(sec_closed_fit_split_vs_db, sec_closed_fit_db_vs_clust, sec_open_fit_split_vs_clust, sec_open_fit_split_vs_db)))
+```
+
 First, we tested whether OptiFit performed as well as _de novo_ clustering when
@@
   grid::grid.raster() 
+```
+```{r}
+close_enough(cv_fit_split_mcc_human_simple, 0, tol = 10^-3)
+assert_that(sec_fit_split_human_simple_1 > sec_fit_split_human_simple_9)
+assert_that(frac_fit_split_human_simple_1 < frac_fit_split_human_simple_9)
 ```
@@
 
+```{r}
+close_enough(c(mcc_fit_split_simple, mcc_fit_split_abun), mcc_fit_split_dist)
+assert_that(sec_fit_split_dist < sec_fit_split_abun & sec_fit_split_abun < sec_fit_split_simple)
+close_enough(frac_fit_split_simple, frac_fit_split_abun)
+assert_that(frac_fit_split_dist < frac_fit_split_simple)
+```
+
 After testing the split strategy using a simple random sample to select the
@@
 dataset (Figure \ref{fig:results_split}).
-OTU quality in terms of MCC was similar with the simple and abundance-weighted
-sampling (median MCCs of `r mcc_fit_split_simple` and `r mcc_fit_split_abun`,
-respectively), but worse for similarity-weighted sampling (median MCC of
-`r mcc_fit_split_dist`).
+OTU quality in terms of MCC was similar across all three sampling methods
+(median MCC of `r mcc_fit_split_simple`).
 In closed-reference clustering mode, the fraction of sequences that mapped were
@@
 similarity-weighted samples with a median runtime of `r sec_fit_split_dist`
-seconds compared to `r sec_fit_split_simple` and `r sec_fit_split_abun` seconds
-for simple and abundance-weighted sampling, respectively.
+seconds compared to `r sec_fit_split_abun` and `r sec_fit_split_simple` seconds
+for abundance-weighted and simple random sampling, respectively.
 Thus, employing more complicated sampling strategies such as abundance-weighted
@@
 require reference OTUs.
+
+```{r}
+ref_seqs <- read_tsv(here('subworkflows', '0_prep_db', 'data', 
+                          'seq_counts.tsv')) %>% 
+  pivot_wider(names_from = ref, values_from = nseqs) %>% as.list()
+assert_that(ref_seqs[['rdp']] < ref_seqs[['silva']])
+assert_that(ref_seqs[['silva']] < ref_seqs[['gg']])
+```
 

### 2021-10-22 paper/paper.Rmd
Don't include assert_that() code chunks

@@
 
-```{r deps, message = FALSE}
+```{r deps, include=FALSE, message = FALSE}
 library(assertthat)
@@
 
-```{r db_names}
+```{r db_names, include = FALSE}
 db_names <- names(dn_dbs_mcc)
@@
 
-```{r}
+```{r, include = FALSE}
 assert_that(!is.unsorted(c(frac_fit_rdp, frac_fit_silva, frac_fit_gg)))
@@
 
-```{r}
+```{r, include = FALSE}
 assert_that(!is.unsorted(c(closed_fit_silva_human_sec, 
@@
 
-```{r}
+```{r, include = FALSE}
 assert_that(mcc_closed_fit_db_vs_vsearch > 10)
@@
 
-```{r}
+```{r, include = FALSE}
 close_enough(mcc_opticlust_vs_fit_split_simple, 0.02)
@@
 ```
-```{r}
+```{r, include = FALSE}
 close_enough(cv_fit_split_mcc_human_simple, 0, tol = 10^-3)
@@
 
-```{r}
+```{r, include = FALSE}
 close_enough(c(mcc_fit_split_simple, mcc_fit_split_abun), mcc_fit_split_dist)
@@
 
-```{r}
+```{r, include = FALSE}
 ref_seqs <- read_tsv(here('subworkflows', '0_prep_db', 'data', 

### 2021-10-24 paper/paper.Rmd
Minor edits / clarifications

@@
 VSEARCH against the Greengenes database, with median MCC scores of 
-`r open_fit_gg_mcc` and `r open_vsearch_mcc`, respectively).
+`r open_fit_gg_mcc` and `r open_vsearch_mcc`, respectively.
 In terms of run time, OptiFit outperformed VSEARCH in both closed and open
@@
 
-To generate reference OTUs from independent databases, we downloaded sequences
+To generate reference OTUs from public databases, we downloaded sequences
 from the Greengenes database (v13_8_99) [@desantis_greengenes_2006], SILVA
@@
 the unprocessed Greengenes 97% OTU reference alignment, since this method is how
-VSEARCH is typically used by the QIIME2 software reference-based clustering
+VSEARCH is typically used by the QIIME2 software for reference-based clustering
 [@bolyen_reproducible_2019; @noauthor_clustering_nodate].
@@
 
-For each dataset, a fraction of the sequences was selected to be clustered _de
+For each dataset, half of the sequences were selected to be clustered _de
 novo_ into reference OTUs with OptiClust.
-We used three methods for selecting the fraction of sequences to be used
+We used three methods for selecting the subset of sequences to be used
 as the reference: a simple random sample, weighting sequences by relative
 abundance, and weighting by similarity to other sequences in the dataset.
-Dataset splitting was repeated with reference fractions ranging from 10% to 90%
-of the dataset and for 100 random seeds.
+Dataset splitting was repeated with 100 random seeds.
+With the simple random sampling method, dataset splitting was also repeated with
+reference fractions ranging from 10% to 90% of the dataset.
 For each dataset split, the remaining query sequences were clustered into the
@@
 
-We implemented the analysis workflow in Snakemake [@koster_snakemake_2012] and
-wrote scripts in R [@r_core_team_r_2020], Python [@van_rossum_python_2009], and
-GNU bash [@noauthor_bash_nodate].
-Software used includes mothur v1.45.0 [@schloss_introducing_2009], VSEARCH
-v2.13.3 [@rognes_vsearch_2016], numpy [@harris_array_2020], the tidyverse
-metapackage [@wickham_welcome_2019], R Markdown [@xie_r_2018], ggtext
-[@wilke_ggtext_2020], the SRA toolkit [@noauthor_sra-tools_nodate], and the
-conda environment manager [@noauthor_anaconda_2016].
-The complete workflow, manuscript, and conda environment are available at
-https://github.com/SchlossLab/Sovacool_OptiFit_2021.
+We implemented the analysis workflow in Snakemake [@koster_snakemake_2012] 
+and wrote scripts in R [@r_core_team_r_2020], 
+Python [@van_rossum_python_2009], 
+and GNU bash [@noauthor_bash_nodate].
+Software used includes mothur v1.46.1 [@schloss_introducing_2009], 
+VSEARCH v2.15.2 [@rognes_vsearch_2016], 
+the tidyverse metapackage [@wickham_welcome_2019], 
+R Markdown [@xie_r_2018], 
+ggraph [@pederson_ggraph_2021], 
+ggtext [@wilke_ggtext_2020], 
+numpy [@harris_array_2020], 
+the SRA toolkit [@noauthor_sra-tools_nodate], 
+and conda [@noauthor_anaconda_2016].
+The complete workflow and supporting files required to reproduce this manuscript
+are available at https://github.com/SchlossLab/Sovacool_OptiFit_2021.
 <!-- TODO: update repo link with journal title -->

### 2021-10-24 paper/paper.Rmd
Update w/ latest optifit split results

@@
 close_enough(c(mcc_fit_split_simple, mcc_fit_split_abun), mcc_fit_split_dist)
-assert_that(sec_fit_split_dist < sec_fit_split_abun & sec_fit_split_abun < sec_fit_split_simple)
+assert_that(!is.unsorted(c(sec_fit_split_dist, 
+                           sec_fit_split_simple, 
+                           sec_fit_split_abun)))
 close_enough(frac_fit_split_simple, frac_fit_split_abun)
@@
 similarity-weighted samples with a median runtime of `r sec_fit_split_dist`
-seconds compared to `r sec_fit_split_abun` and `r sec_fit_split_simple` seconds
-for abundance-weighted and simple random sampling, respectively.
+seconds compared to `r sec_fit_split_simple` and `r sec_fit_split_abun`  seconds
+for simple and abundance-weighted sampling, respectively.
 Thus, employing more complicated sampling strategies such as abundance-weighted

### 2021-10-25 paper/paper.Rmd
Use msphere csl
it turns out that it's the same as the msystems csl

@@
     html_preview: false
-csl: msystems.csl
+csl: msphere.csl
 fontsize: 12pt
@@
 
-figures_meta <- yaml::read_yaml(here('config', 'figures.yaml'))
+figures_meta <- yaml::read_yaml(here('paper', 'figures.yaml'))
 figs <- function(fig_name, attribute) {

### 2021-10-26 paper/paper.Rmd
Fix figure paths for github markdown

@@
   github_document:
-    includes:
-      in_header: preamble.tex
-      before_body: head.tex
     html_preview: false
@@
 schtools::set_knitr_opts()
-knitr::opts_chunk$set(fig.path = here::here('figures/'),
+knitr::opts_knit$set(base.dir = here::here(), 
+                     base.url = "../")
+knitr::opts_chunk$set(fig.path = 'figures/',
                       dpi = 300)

### 2021-10-26 code/R/render.R
Render pdf & gh markdown in separate rules

@@ -1,8 +1,6 @@
-source(snakemake@input[["fcns"]])
-log_smk()
 rmarkdown::render(
   here::here(snakemake@input[["Rmd"]]),
   params = list(snakemake = snakemake),
-  output_format = "all",
-  output_dir = here::here(snakemake@params[["outdir"]])
+  output_format = snakemake@params[['format']],
+  output_file = here::here(snakemake@output[1])
 )

### 2021-10-29 paper/paper.Rmd
Repo url isn't needed

@@
 geometry: margin=1.0in
-repository_url: https://github.com/SchlossLab/OptiFitAnalysis
 params:

### 2021-10-29 exploratory/2021/vsearch-debug_2021-09/README.Rmd
Add dates to notebooks

---
title: "Why does VSEARCH perform worse than in Pat's 2015 PeerJ paper?"
output: github_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      fig.path = "figures/"
                      )
```

```{r deps}
library(glue)
library(here)
library(tidyverse)
theme_set(theme_bw())
```

Pat had MCC's of ~`0.7` for the MiSeq mouse dataset with VSEARCH DGC.
I'm getting MCC's of `0.5781`.

## VSEARCH version number

Pat used VSEARCH v1.5.0, I'm using v2.15.2.

I tried re-running my [Snakemake workflow on the mouse data but with VSEARCH
v1.5.0](https://github.com/SchlossLab/Sovacool_OptiFit_2021/blob/2f39f25fc31980fe3a9a349562486a3f972efc8b/subworkflows/4_vsearch/Snakefile#L40),
and still got the exact same MCC result as before.

## MiSeq data preprocessing

## Silva & RDP versions

### 2021-10-29 paper/paper.Rmd

@@
 }
+
 load(here("results", "stats.RData"))
@@
 This approach takes into account the distances between all pairs of sequences
-when assigning query sequences to OTUs, in contrast to other _de novo_
-methods such as the greedy clustering algorithms implemented in USEARCH and
-VSEARCH, which only consider the distance between the query sequence and a
-representative centroid sequence in the OTU [@edgar_search_2010;
-@rognes_vsearch_2016].
+when assigning query sequences to OTUs, in contrast to other _de novo_ methods
+such as the greedy clustering algorithms implemented in USEARCH and VSEARCH
+[@edgar_search_2010; @rognes_vsearch_2016].
 In methods employing greedy clustering algorithms, only the distance between
-each sequence and the centroid sequence is considered while clustering.
-As a result, distances between pairs of sequences in the same OTU are frequenty
+each sequence and a representative centroid sequence in the OTU is considered
+while clustering.
+As a result, distances between pairs of sequences in the same OTU are frequently
 larger than the specified threshold, i.e. they are false positives.
@@
 While other tools represent reference OTUs with a single sequence, OptiFit uses
-multiple sequences in existing OTUs as the reference and fits new sequences
+multiple sequences in existing OTUs as the reference and fits new sequences to 
 those reference OTUs.
@@
 in a better Matthews correlation coefficient (MCC) [@westcott_opticlust_2017].
-The MCC uses all values from a confusion matrix and ranges from zero to one,
-with a score of one occurring when all sequence pairs are true positives and
-true negatives, and a score of zero when all pairs are false positives and false
-negatives.
+The MCC uses all values from a confusion matrix and ranges from negative one to
+one, with a score of one occurring when all sequence pairs are true positives
+and true negatives and a score of negative one occurring when all pairs are
+false positives and false negatives.
 Sequence pairs that are similar to each other (i.e. within the distance
@@
 
-```{r db_names, include = FALSE}
-db_names <- names(dn_dbs_mcc)
-db_names_upper <- case_when(db_names == 'gg' ~ 'Greengenes',
-                           TRUE ~ toupper(db_names))
-close_enough(dn_dbs_mcc['gg'], dn_dbs_mcc['silva'])
-close_enough(dn_dbs_mcc['gg'], dn_dbs_mcc['rdp'], 10^-1)
+```{r assert_db_mccs, include = FALSE}
 mcc_decreasing <- c(closed_fit_rdp_mcc, closed_fit_silva_mcc, closed_fit_gg_mcc,
@@
 sequence (see Figure \ref{fig:workflow}).
-The _de novo_ MCC scores were `r dn_dbs_mcc` for `r db_names_upper`,
+After trimming to the V4 region, the databases contained 
+`r dn_dbs %>% pull('nseqs')` unique sequences and produced _de novo_ MCC scores 
+of `r dn_dbs %>% pull('med_mcc')` for `r dn_dbs %>% pull('refname')`, 
 respectively.
@@
 
-```{r, include = FALSE}
+```{r assert_db_frac, include = FALSE}
 assert_that(!is.unsorted(c(frac_fit_rdp, frac_fit_silva, frac_fit_gg)))
@@
 
-```{r, include = FALSE}
+```{r assert_db_sec, include = FALSE}
 assert_that(!is.unsorted(c(closed_fit_silva_human_sec,
@@
 
-```{r, include = FALSE}
+```{r assert_vsearch, include = FALSE}
 assert_that(mcc_closed_fit_db_vs_vsearch > 10)
@@
 is considered when assigning sequences to OTUs.
-<!--**TODO: delete or move this sentence:** _De novo_ clustering with OptiClust produced `r mcc_opticlust_vs_vsearch`%
-higher quality OTUs than VSEARCH in terms of MCC, but performed
-`r sec_opticlust_vs_vsearch`% slower than VSEARCH.-->
 In closed reference mode, OptiFit produced `r mcc_closed_fit_db_vs_vsearch`%
@@
 
-```{r, include = FALSE}
+```{r assert_split_simple, include = FALSE}
 close_enough(mcc_opticlust_vs_fit_split_simple, 0.02)
@@
 
-```{r, include = FALSE}
+```{r assert_split_strat, include = FALSE}
 close_enough(c(mcc_fit_split_simple, mcc_fit_split_abun), mcc_fit_split_dist)
@@
 
-```{r, include = FALSE}
-ref_seqs <- read_tsv(here('subworkflows', '0_prep_db', 'data',
-                          'seq_counts.tsv')) %>%
-  pivot_wider(names_from = ref, values_from = nseqs) %>% as.list()
-assert_that(ref_seqs[['rdp']] < ref_seqs[['silva']])
-assert_that(ref_seqs[['silva']] < ref_seqs[['gg']])
+```{r assert_ref_sizes, include = FALSE}
+ref_seqs <- dn_dbs %>% pull(nseqs) 
+names(ref_seqs) <- dn_dbs %>% pull(ref)
+assert_that(!is.unsorted(ref_seqs[['rdp']], 
+                         ref_seqs[['silva']], 
+                         ref_seqs[['gg']])
+            )
 ```

### 2021-10-29 code/R/calc_results_stats.R
🎨  Style R code

@@ -46,12 +46,17 @@ sec_opticlust_vs_vsearch <- abs(rel_diff(dn_vsearch_sec, opticlust_sec))
 
 dn_dbs <- read_tsv("subworkflows/2_fit_reference_db/results/denovo_dbs.tsv") %>%
   group_by(ref) %>%
-  summarize(med_mcc = median(mcc)) %>% 
-  full_join(read_tsv(here('subworkflows', '0_prep_db', 'data',
-                          'seq_counts.tsv')),
-            by = "ref") %>% 
-  mutate(refname = case_when(ref == 'gg' ~ 'Greengenes',
-                             TRUE ~ toupper(ref)))
+  summarize(med_mcc = median(mcc)) %>%
+  full_join(read_tsv(here(
+    "subworkflows", "0_prep_db", "data",
+    "seq_counts.tsv"
+  )),
+  by = "ref"
+  ) %>%
+  mutate(refname = case_when(
+    ref == "gg" ~ "Greengenes",
+    TRUE ~ toupper(ref)
+  ))
 
 ################################################################################
 # ref db open
@@ -355,7 +360,7 @@ frac_fit_split_human_simple_1 <- dat %>%
     tool == "mothur",
     dataset == "human",
     ref_weight == "simple",
-    method == 'closed',
+    method == "closed",
     ref_frac == 0.1
   ) %>%
   pull(frac_map_median) %>%
@@ -367,7 +372,7 @@ frac_fit_split_human_simple_9 <- dat %>%
     tool == "mothur",
     dataset == "human",
     ref_weight == "simple",
-    method == 'closed',
+    method == "closed",
     ref_frac == 0.9
   ) %>%
   pull(frac_map_median) %>%
@@ -412,7 +417,7 @@ frac_fit_split_simple <- dat %>%
     strategy == "self-split",
     tool == "mothur",
     ref_weight == "simple",
-    method == 'closed',
+    method == "closed",
     ref_frac == 0.5
   ) %>%
   pull(frac_map_median) %>%
@@ -423,7 +428,7 @@ frac_fit_split_abun <- dat %>%
     strategy == "self-split",
     tool == "mothur",
     ref_weight == "abundance",
-    method == 'closed',
+    method == "closed",
     ref_frac == 0.5
   ) %>%
   pull(frac_map_median) %>%
@@ -434,7 +439,7 @@ frac_fit_split_dist <- dat %>%
     strategy == "self-split",
     tool == "mothur",
[…]

### 2021-10-29 exploratory/2021/test-criteria_2021-04.Rmd

--- 2021-03-24 exploratory/2021/test_criteria.Rmd
@@
 
-## OTU Quality
-
-```{r crit_mcc}
+```{r crit_facet, fig.dim = c(5, 5)}
 dat %>% 
-  ggplot(aes(x = criteria, y = mcc, color = printref)) +
-  geom_jitter(alpha = 0.5, width = 0.2) 
+  pivot_longer(c(mcc, fraction_mapped, sec), 
+               names_to = "metric") %>% 
+  ggplot(aes(x = criteria, y = value, color = printref)) +
+  geom_jitter(alpha = 0.5, width = 0.2) +
+  scale_color_manual(values = c("darkorange", "cyan4")) +
+  facet_wrap("metric", scales = "free", nrow = 3) +
+  coord_flip()
 ```
-
-## Fraction mapped
-
-```{r crit_map}
-dat %>% 
-  ggplot(aes(x = criteria, y = fraction_mapped, color = printref)) +
-  geom_jitter(alpha = 0.5, width = 0.2)
-```
-
-## Runtime
-
-```{r crit_runtime}
-dat %>% 
-  ggplot(aes(x = criteria, y = sec, color = printref)) +
-  geom_jitter(alpha = 0.5, width = 0.2)
-```

### 2021-10-30 exploratory/2021/algorithm-diagram_2021-08.Rmd

--- 2021-07-17 exploratory/2021/algorithm_diagram.Rmd
@@
 ```{r ggraph_exp, fig.dim=c(6,5), dpi=300}
-plot_graph <- function(graph, title = '') {
+plot_graph <- function(graph, title = '', 
+                       hide_loops = FALSE) {
+  loop_dir <- 90
+  loop_color <- ifelse(hide_loops, 'white', 'black')
   create_layout(graph, 'linear', sort.by = id) %>% 
@@
                 ) +
-  geom_edge_loop(aes(span = 1, direction = 90, strength = 0.2,# direction = 270, 
+  geom_edge_loop(aes(span = 1, 
+                     direction = loop_dir, 
+                     strength = 0.5,
                      color = is_loop)) +
   geom_node_label(aes(label = name)) +
-    scale_edge_color_manual(values = c('black')#c('white')
+    scale_edge_color_manual(values = c(loop_color)
                             ) +
@@
 ## do backend in python until the actual plotting step
+```{python dist_mat}
+import pandas as pd
+dist_mat = pd.DataFrame.from_dict({
+    "seq1": ["D", "F", "G", "H", "I", "I", "J", "J", "N", "O", "P", "P", "P", "Q", "Q", 'X', 'X', 'X', 'X', 'Y', 'W', 'W', 'W'], 
+    "seq2": ["B", "E", "C", "A", "B", "D", "A", "H", "M", "L", "K", "L", "O", "E", "F", 'Y', 'C', 'G', 'N', 'C', 'M', 'N', 'F']})
+```
+```{r optifit}
+reticulate::source_python(here('code', 'py', 'algorithm_diagram.py'))
+optifit <- create_optifit() 
+#dist_array <- optifit$fitmap$dists_to_array %>% py_to_r()
 
-```{r pseudo_optifit}
-reticulate::source_python(here('code', 'py', 'algorithm_diagram.py'))
-optifit_iters <- run_optifit() %>% 
+optifit_iters <- optifit$iterate %>% 
   lapply(function(x) {
@@
                 edges = x[['edges']] %>% py_to_r() %>% 
-                 mutate(is_loop = from == to)))
+                  bind_rows(data.frame(from = 1, to = 1, mcc = NA)) %>% 
+                 mutate(is_loop = from == to,
+                        loop_dir = ifelse(from == 1, 270, 90))))
     })
@@
   tbl_graph(nodes = x$nodes, edges = x$edges) %>% 
-    plot_graph(title = glue('{i}) mcc = {x$edges %>% filter(is_loop) %>% pull(mcc)}'))
+    plot_graph(title = glue('{i}) mcc = {x$edges %>% filter(is_loop) %>% pull(mcc)}'),
+               hide_loops = TRUE)
 }) %>% 

### 2021-10-30 exploratory/2021/did-results-change_2021-09/README.Rmd

---
title: "Did MCC results change?"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      fig.path = "figures/"
                      )
```

```{r deps}
library(glue)
library(here)
library(tidyverse)
theme_set(theme_bw())
```

Why does it look like OptiClust performed ~4% worse than OptiFit with the split strategy?

## Did OptiClust results change?

[March 2021 commit: `0b80681`](https://github.com/SchlossLab/Sovacool_OptiFit_2021/commits/main/subworkflows/1_prep_samples/results/opticlust_results.tsv)

```{r opticlust}
opticlust <- c('Nov-2020_5d87338', 'Mar-2021_0b80681') %>% 
  map_dfr(function(file) {
    read_tsv(here('exploratory', '2021', 'did-results-change', 
                glue('opticlust_results_{file}.tsv'))) %>% 
    mutate(file = file)
  })

opticlust %>% group_by(file) %>% summarize(min_mcc = min(mcc),
                                           median_mcc = median(mcc),
                                           max_mcc = max(mcc))

opticlust %>% filter(dataset == 'human' | dataset == 'mouse') %>% 
  group_by(file, dataset) %>% 
  summarize(min_mcc = min(mcc),
            median_mcc = median(mcc),
            max_mcc = max(mcc))
```

Nope!

## Did OptiFit results change?

```{r optifit}
optifit <- c('Dec-2020_5fcc830',
             'Apr-2021_c6fcdbd', 
             'Jul-2021_4090079') %>% 
  map_dfr(function(file) {
    read_tsv(here('exploratory', '2021', 'did-results-change', 
                glue('optifit_split_results_{file}.tsv'))) %>% 
    mutate(file = file)
  })

optifit %>% filter(dataset == 'human' | dataset == 'mouse') %>% 
  group_by(file, dataset) %>% 
  summarize(min_mcc = min(mcc),
            median_mcc = median(mcc),
            max_mcc = max(mcc))

optifit %>% 
  filter(dataset == 'human' | dataset == 'mouse') %>% 
  ggplot(aes(x = file, y = mcc, color = dataset)) +
  geom_boxplot()
```

Yeah! OptiFit results improved. What changed? I don't think I've updated mothur since December.

### 2021-10-30 paper/paper.Rmd
Manually set fig2 dimensions to remove gray bar
dot didn't output it as exactly 6x4, so Rmd was adding a gray bar to make up for it.
Corrected it by adding in the actual dimensions of the tiff file.
Use imagemagick to fix gray bar in workflow figure

--- 2021-10-29 paper/paper.Rmd
@@
 
-```{r workflow, fig.dim=figs('workflow','dim'), fig.cap=figs('workflow','cap')}
+```{r workflow, fig.dim=figs('workflow', 'dim'), fig.cap=figs('workflow','cap')}
 figs('workflow', 'path') %>%

### 2021-11-04 paper/paper.Rmd

@@
 
-Since OptiClust and OptiFit employ a random number generator to break ties
-when OTU assignments are of equal quality, they produce slightly different OTU
-assignments when repeated with different random seeds.
+OptiClust and OptiFit randomize the order of query sequences prior to clustering
+and employ a random number generator to break ties when OTU assignments are of
+equal quality.
+As a result, they produce slightly different OTU assignments when repeated with
+different random seeds.
 To capture any variation in OTU quality or execution time, clustering was

### 2021-11-07 paper/paper.Rmd

@@
 
+# check whether all elements of two numeric vectors are close enough
 close_enough <- function(x, y, tol = 10^-2) {
   assert_that(all(near(x, y, tol = tol)))
+}
+
+# check whether all elements given are sorted in non-descending order
+is_nondesc <- function(...) {
+  things <- c(...)
+  if (length(things) < 1) {
+    warning('Zero elements were given to `is_nondesc()`')
+  }
+  first <- things[1]
+  last <- things[length(things)]
+  !is.unsorted(things) & first <= last
 }
@@
 ```{r assert_db_mccs, include = FALSE}
-mcc_decreasing <- c(closed_fit_rdp_mcc, closed_fit_silva_mcc, closed_fit_gg_mcc,
+mcc_ascending <- c(closed_fit_rdp_mcc, closed_fit_silva_mcc, closed_fit_gg_mcc,
                     open_fit_silva_mcc, open_fit_gg_mcc, open_fit_rdp_mcc,
                     opticlust_mcc)
-assert_that(!is.unsorted(mcc_decreasing))
+is_nondesc(mcc_ascending)
 ```
@@
 ```{r assert_db_frac, include = FALSE}
-assert_that(!is.unsorted(c(frac_fit_rdp, frac_fit_silva, frac_fit_gg)))
+is_nondesc(frac_fit_rdp, frac_fit_silva, frac_fit_gg)
 assert_that((frac_fit_gg - frac_fit_silva) < 10)
@@
 ```{r assert_db_sec, include = FALSE}
-assert_that(!is.unsorted(c(closed_fit_silva_human_sec,
+is_nondesc(closed_fit_silva_human_sec,
            opticlust_human_sec,
-                           open_fit_silva_human_sec)))
+           open_fit_silva_human_sec)
 ```
@@
 ```{r assert_split_simple, include = FALSE}
-close_enough(mcc_opticlust_vs_fit_split_simple, 0.02)
+assert_that(mcc_opticlust_vs_fit_split_simple < 0.09)
 assert_that((frac_fit_split - frac_fit_gg) > 20)
-assert_that(!is.unsorted(c(sec_closed_fit_split_vs_db, sec_closed_fit_db_vs_clust, sec_open_fit_split_vs_clust, sec_open_fit_split_vs_db)))
+is_nondesc(sec_closed_fit_split_vs_db, sec_closed_fit_db_vs_clust, sec_open_fit_split_vs_clust, sec_open_fit_split_vs_db)
 ```
@@
 close_enough(c(mcc_fit_split_simple, mcc_fit_split_abun), mcc_fit_split_dist)
-assert_that(!is.unsorted(c(sec_fit_split_dist,
-                           sec_fit_split_simple,
-                           sec_fit_split_abun)))
-close_enough(frac_fit_split_simple, frac_fit_split_abun)
-assert_that(frac_fit_split_dist < frac_fit_split_simple)
+is_nondesc(sec_fit_split_dist,
+           sec_fit_split_abun,
+           sec_fit_split_simple)
+is_nondesc(frac_fit_split_dist,
+           frac_fit_split_abun,
+           frac_fit_split_simple)
 ```
@@
 names(ref_seqs) <- dn_dbs %>% pull(ref)
-assert_that(!is.unsorted(ref_seqs[['rdp']], 
+is_nondesc(ref_seqs[['rdp']],
            ref_seqs[['silva']],
            ref_seqs[['gg']])
-            )
 ```
@@
 and GNU bash [@noauthor_bash_nodate].
-Software used includes mothur v1.46.1 [@schloss_introducing_2009],
+Software used includes mothur v1.47.0 [@schloss_introducing_2009],
 VSEARCH v2.15.2 [@rognes_vsearch_2016],

### 2021-11-09 paper/paper.Rmd
Cherry-pick: Reduce abstract to 250 words

@@
 params:
-    snakemake: "none"
+  include_figures: false
 ---
@@
 schtools::set_knitr_opts()
-knitr::opts_knit$set(base.dir = here::here(), 
-                     base.url = "../")
+#knitr::opts_knit$set(base.dir = here::here(),
+#                     base.url = "../")
 knitr::opts_chunk$set(fig.path = 'figures/',
@@
 set.seed(20210624)
-snakemake <- params$snakemake
-if (is_character(snakemake)) {
-  stop("You need to knit this R Markdown document from the Snakemake workflow")
-}
 
@@
 figures_meta <- yaml::read_yaml(here('paper', 'figures.yaml'))
-figs <- function(fig_name, attribute) {
+
+figs <- function(fig_name, attribute, include_figures = params$include_figures) {
   value <- figures_meta[[fig_name]][[attribute]]
@@
     value <- glue("\\label{fig:[fig_name]}[value]", .open='[', .close = ']')
+  } else if (attribute == 'dim' & !isTRUE(include_figures)) {
+    value <- c(0.1,0.1)
   } else {
@@
   return(value)
+}
+
+render_figure <- function(fig_name, include_figures = params$include_figures) {
+  if (isTRUE(include_figures)) {
+    figs(fig_name, 'path') %>%
+      tiff::readTIFF() %>%
+      grid::grid.raster()
+  } else {
+    ggplot() + theme_void()
+  }
 }
@@
 
-Assigning amplicon sequences to operational taxonomic units (OTUs) is often an
-important step in characterizing the composition of microbial communities across
-large datasets.
-OptiClust, a _de novo_ OTU clustering method, has been
-shown to produce higher quality OTU assignments than other methods and at
-comparable or faster speeds.
+Assigning amplicon sequences to operational taxonomic units (OTUs) is an
+important step in characterizing microbial communities across large datasets.
+OptiClust, a _de novo_ OTU clustering method, produces higher
+quality OTUs than other methods at comparable or faster speeds.
 A notable difference between _de novo_ clustering and database-dependent
 reference clustering methods is that OTU assignments from _de novo_ methods may
-change when new sequences are added to a dataset.
-However, in some cases one may wish to incorporate new samples into a previously
-clustered dataset without performing clustering again on all sequences, such as
-when comparing across datasets or deploying machine learning models where OTUs
-are features.
-Existing reference-based clustering methods produce consistent OTUs, but they
-only consider the similarity of each query sequence to a single reference
-sequence in an OTU, thus resulting in OTU assignments that are significantly
-worse than those generated by _de novo_ methods.
-To provide an efficient and robust method to fit amplicon sequence data to
-existing OTUs, we developed the OptiFit algorithm.
-Inspired by OptiClust, OptiFit considers the similarity of all pairs of reference and
-query sequences in an OTU to produce OTUs of the best possible quality.
-We tested OptiFit using four microbiome datasets with two different strategies:
-by clustering to an external reference database or by splitting the dataset into a
-reference and query set and clustering the query sequences to the reference set
-after clustering it using OptiClust.
-The result is an improved implementation of closed and open-reference clustering.
-OptiFit produces OTUs of similar quality as OptiClust and at faster speeds when
-using the split dataset strategy, although the OTU quality and processing speed
-depends on the database chosen when using the external database strategy.
-OptiFit provides a suitable option for users who require consistent OTU
+change when new sequences are added.
+However, one may wish to incorporate new samples to previously clustered
+datasets without clustering all sequences again, such as when
+comparing across datasets or deploying machine learning models.
+Existing reference-based methods produce consistent OTUs, but only consider the
+similarity of each query sequence to a single reference sequence in an OTU,
+resulting in assignments that are worse than those generated by _de novo_
+methods.
+To provide an efficient method to fit sequences to existing OTUs, we developed
+the OptiFit algorithm.
+Like OptiClust, OptiFit considers the similarity of all pairs of reference and
+query sequences to produce OTUs of the best quality.
+We tested OptiFit using four datasets with two strategies: 1) clustering to a
+reference database or 2) splitting the dataset into a reference and query set,
+clustering the references using OptiClust, then clustering the queries
+to the references.
+The result is an improved implementation of reference-based clustering.
+OptiFit produces similar quality OTUs as OptiClust at faster speeds when
+using the split dataset strategy.
+OptiFit provides a suitable option for users requiring consistent OTU
 assignments at the same quality afforded by _de novo_ clustering methods.
@@
 ```{r algorithm, fig.dim=figs('algorithm','dim'), fig.cap=figs('algorithm','cap')}
-figs('algorithm', 'path') %>%
-  tiff::readTIFF() %>%
-  grid::grid.raster()
+render_figure('algorithm')
 ```
@@
 ```{r workflow, fig.dim=figs('workflow', 'dim'), fig.cap=figs('workflow','cap')}
-figs('workflow', 'path') %>%
-  tiff::readTIFF() %>%
-  grid::grid.raster()
+render_figure('workflow')
 ```
@@
 ```{r results_sum, fig.dim=figs('results_sum', 'dim'), fig.cap=figs('results_sum', 'cap')}
-figs('results_sum', 'path') %>%
-  tiff::readTIFF() %>%
-  grid::grid.raster()
+render_figure('results_sum')
 ```
@@
 ```{r results_split, fig.dim=figs('results_split', 'dim'), fig.cap=figs('results_split', 'cap')}
-figs('results_split', 'path') %>%
-  tiff::readTIFF() %>%
-  grid::grid.raster()
+render_figure('results_split')
 ```
-```{r, include = FALSE}
+
+```{r assert_split_fracs, include = FALSE}
 close_enough(cv_fit_split_mcc_human_simple, 0, tol = 10^-3)

### 2021-11-09 code/py/abstract_word_count.py
🎨  Style Python code 🐍

@@ -1,34 +1,41 @@
 def word_count(infilename, starter, stopper):
-    with open(infilename, 'r') as infile:
+    with open(infilename, "r") as infile:
         words = []
         line = next(infile)
         while starter != line:
             line = next(infile)
-        line = next(infile) # make sure the first parsed line is not the starter
+        line = next(infile)  # make sure the first parsed line is not the starter
         while stopper != line:
             words += [word for word in line.strip().split()]
             line = next(infile)
     return len(words)
 
+
 def check_wc(section_name, num_words, word_limit):
     if num_words > word_limit:
-        raise ValueError(f"The {section_name} section is {num_words} words. You need to cut {num_words - word_limit} words.")
+        raise ValueError(
+            f"The {section_name} section is {num_words} words. You need to cut {num_words - word_limit} words."
+        )
+
 
 def main(src_filename, log_filename):
-    wc_abs = word_count(src_filename, '## Abstract\n', '### Importance\n')
-    check_wc('abstract', wc_abs, 250)
+    wc_abs = word_count(src_filename, "## Abstract\n", "### Importance\n")
+    check_wc("abstract", wc_abs, 250)
     word_limit_imp = 150
-    wc_imp = word_count(src_filename, '### Importance\n', "\\newpage\n")
-    check_wc('importance', wc_imp, 150)
-    with open(log_filename, 'w') as outfile:
-        outfile.write('section\tword_count\n')
-        for section, word_limit, starter, stopper in zip(['abstract', 'importance'], 
-                                                        [250, 150],
-                                    ['## Abstract\n', '### Importance\n'],  ['### Importance\n', "\\newpage\n"]):
+    wc_imp = word_count(src_filename, "### Importance\n", "\\newpage\n")
+    check_wc("importance", wc_imp, 150)
+    with open(log_filename, "w") as outfile:
+        outfile.write("section\tword_count\n")
+        for section, word_limit, starter, stopper in zip(
+            ["abstract", "importance"],
+            [250, 150],
+            ["## Abstract\n", "### Importance\n"],
+            ["### Importance\n", "\\newpage\n"],
+        ):
             wc = word_count(src_filename, starter, stopper)
             check_wc(section, wc, word_limit)
             outfile.write(f"{section}\t{wc}\n")
 
 
 if __name__ == "__main__":
-    main('paper/paper.Rmd', 'log/count_words_abstract.log')
+    main("paper/paper.Rmd", "log/count_words_abstract.log")

### 2021-12-15 paper/paper.Rmd
Allow paper to be knitted outside snakemake

--- 2021-11-09 paper/paper.Rmd
@@
 schtools::set_knitr_opts()
-#knitr::opts_knit$set(base.dir = here::here(),
-#                     base.url = "../")
+knitr::opts_knit$set(base.dir = here::here(), 
+                     base.url = "../")
 knitr::opts_chunk$set(fig.path = 'figures/',
@@
 figures_meta <- yaml::read_yaml(here('paper', 'figures.yaml'))
-
-figs <- function(fig_name, attribute, include_figures = params$include_figures) {
+figs <- function(fig_name, attribute) {
   value <- figures_meta[[fig_name]][[attribute]]
@@
     value <- glue("\\label{fig:[fig_name]}[value]", .open='[', .close = ']')
-  } else if (attribute == 'dim' & !isTRUE(include_figures)) {
-    value <- c(0.1,0.1)
   } else {
@@
   return(value)
-}
-
-render_figure <- function(fig_name, include_figures = params$include_figures) {
-  if (isTRUE(include_figures)) {
-    figs(fig_name, 'path') %>%
-      tiff::readTIFF() %>%
-      grid::grid.raster()
-  } else {
-    ggplot() + theme_void()
-  }
 }
@@
 
-```{r algorithm, fig.dim=figs('algorithm','dim'), fig.cap=figs('algorithm','cap')}
-render_figure('algorithm')
+```{r algorithm, eval=params$include_figures, fig.dim=figs('algorithm','dim'), fig.cap=figs('algorithm','cap')}
+figs('algorithm', 'path') %>%
+  tiff::readTIFF() %>%
+  grid::grid.raster()
 ```
@@
 
-```{r workflow, fig.dim=figs('workflow', 'dim'), fig.cap=figs('workflow','cap')}
-render_figure('workflow')
+```{r workflow, eval=params$include_figures, fig.dim=figs('workflow', 'dim'), fig.cap=figs('workflow','cap')}
+figs('workflow', 'path') %>%
+  tiff::readTIFF() %>%
+  grid::grid.raster()
 ```
@@
 
-```{r results_sum, fig.dim=figs('results_sum', 'dim'), fig.cap=figs('results_sum', 'cap')}
-render_figure('results_sum')
+```{r results_sum, eval=params$include_figures, fig.dim=figs('results_sum', 'dim'), fig.cap=figs('results_sum', 'cap')}
+figs('results_sum', 'path') %>%
+  tiff::readTIFF() %>%
+  grid::grid.raster()
 ```
@@
 
-```{r results_split, fig.dim=figs('results_split', 'dim'), fig.cap=figs('results_split', 'cap')}
-render_figure('results_split')
+```{r results_split, eval=params$include_figures, fig.dim=figs('results_split', 'dim'), fig.cap=figs('results_split', 'cap')}
+figs('results_split', 'path') %>%
+  tiff::readTIFF() %>%
+  grid::grid.raster()
 ```
-
 ```{r assert_split_fracs, include = FALSE}

### 2021-12-15 code/R/render.R

@@ -1,6 +1,6 @@
 rmarkdown::render(
   here::here(snakemake@input[["Rmd"]]),
-  params = list(snakemake = snakemake),
+  params = list(include_figures = TRUE),
   output_format = snakemake@params[["format"]],
   output_file = here::here(snakemake@output[1])
 )

### 2021-12-15 paper/paper.Rmd
Revert "Reduce abstract to 250 words"
This reverts commit 27d08bd30efe0fda235487ebb03b76014a53fb82.

Because I need to show track changes against the version we
originally submitted to the journal.

@@
 
-Assigning amplicon sequences to operational taxonomic units (OTUs) is an
-important step in characterizing microbial communities across large datasets.
-OptiClust, a _de novo_ OTU clustering method, produces higher
-quality OTUs than other methods at comparable or faster speeds.
+Assigning amplicon sequences to operational taxonomic units (OTUs) is often an
+important step in characterizing the composition of microbial communities across
+large datasets.
+OptiClust, a _de novo_ OTU clustering method, has been
+shown to produce higher quality OTU assignments than other methods and at
+comparable or faster speeds.
 A notable difference between _de novo_ clustering and database-dependent
 reference clustering methods is that OTU assignments from _de novo_ methods may
-change when new sequences are added.
-However, one may wish to incorporate new samples to previously clustered
-datasets without clustering all sequences again, such as when
-comparing across datasets or deploying machine learning models.
-Existing reference-based methods produce consistent OTUs, but only consider the
-similarity of each query sequence to a single reference sequence in an OTU,
-resulting in assignments that are worse than those generated by _de novo_
-methods.
-To provide an efficient method to fit sequences to existing OTUs, we developed
-the OptiFit algorithm.
-Like OptiClust, OptiFit considers the similarity of all pairs of reference and
-query sequences to produce OTUs of the best quality.
-We tested OptiFit using four datasets with two strategies: 1) clustering to a
-reference database or 2) splitting the dataset into a reference and query set,
-clustering the references using OptiClust, then clustering the queries
-to the references.
-The result is an improved implementation of reference-based clustering.
-OptiFit produces similar quality OTUs as OptiClust at faster speeds when
-using the split dataset strategy.
-OptiFit provides a suitable option for users requiring consistent OTU
+change when new sequences are added to a dataset.
+However, in some cases one may wish to incorporate new samples into a previously
+clustered dataset without performing clustering again on all sequences, such as
+when comparing across datasets or deploying machine learning models where OTUs
+are features.
+Existing reference-based clustering methods produce consistent OTUs, but they
+only consider the similarity of each query sequence to a single reference
+sequence in an OTU, thus resulting in OTU assignments that are significantly
+worse than those generated by _de novo_ methods.
+To provide an efficient and robust method to fit amplicon sequence data to
+existing OTUs, we developed the OptiFit algorithm.
+Inspired by OptiClust, OptiFit considers the similarity of all pairs of reference and
+query sequences in an OTU to produce OTUs of the best possible quality.
+We tested OptiFit using four microbiome datasets with two different strategies:
+by clustering to an external reference database or by splitting the dataset into a
+reference and query set and clustering the query sequences to the reference set
+after clustering it using OptiClust.
+The result is an improved implementation of closed and open-reference clustering.
+OptiFit produces OTUs of similar quality as OptiClust and at faster speeds when
+using the split dataset strategy, although the OTU quality and processing speed
+depends on the database chosen when using the external database strategy.
+OptiFit provides a suitable option for users who require consistent OTU
 assignments at the same quality afforded by _de novo_ clustering methods.

### 2021-12-15 paper/paper_v1.Rmd
Get original submitted version w/o figures

--- 2021-12-15 paper/paper.Rmd
@@
     value <- glue("\\label{fig:[fig_name]}[value]", .open='[', .close = ']')
+  } else if (attribute == 'dim' & isFALSE(include_figures)) {
+    value <- c(0.1,0.1)
   } else {
@@
 
-```{r algorithm, eval=params$include_figures, fig.dim=figs('algorithm','dim'), fig.cap=figs('algorithm','cap')}
-figs('algorithm', 'path') %>%
-  tiff::readTIFF() %>%
-  grid::grid.raster()
+```{r algorithm, fig.dim=figs('algorithm','dim'), fig.cap=figs('algorithm','cap')}
+render_figure('algorithm')
 ```
@@
 
-```{r workflow, eval=params$include_figures, fig.dim=figs('workflow', 'dim'), fig.cap=figs('workflow','cap')}
-figs('workflow', 'path') %>%
-  tiff::readTIFF() %>%
-  grid::grid.raster()
+```{r workflow, fig.dim=figs('workflow', 'dim'), fig.cap=figs('workflow','cap')}
+render_figure('workflow')
 ```
@@
 
-```{r results_sum, eval=params$include_figures, fig.dim=figs('results_sum', 'dim'), fig.cap=figs('results_sum', 'cap')}
-figs('results_sum', 'path') %>%
-  tiff::readTIFF() %>%
-  grid::grid.raster()
+```{r results_sum, fig.dim=figs('results_sum', 'dim'), fig.cap=figs('results_sum', 'cap')}
+render_figure('results_sum')
 ```
@@
 
-```{r results_split, eval=params$include_figures, fig.dim=figs('results_split', 'dim'), fig.cap=figs('results_split', 'cap')}
-figs('results_split', 'path') %>%
-  tiff::readTIFF() %>%
-  grid::grid.raster()
+```{r results_split, fig.dim=figs('results_split', 'dim'), fig.cap=figs('results_split', 'cap')}
+render_figure('results_split')
 ```
+
 ```{r assert_split_fracs, include = FALSE}

### 2021-12-15 paper/paper_before-review.Rmd
Rename manuscript version from before review

--- 2021-12-15 paper/paper.Rmd
@@
 schtools::set_knitr_opts()
-knitr::opts_knit$set(base.dir = here::here(),
-                     base.url = "../")
+#knitr::opts_knit$set(base.dir = here::here(),
+#                     base.url = "../")
 knitr::opts_chunk$set(fig.path = 'figures/',

### 2021-12-15 paper/response-to-reviewers.md
Paste reviewer comments

Reviewer comments:

Reviewer #1 (Comments for the Author):

In this manuscript, Sovacool et al. describe a new method for fitting amplicon sequences to existing datasets. This is an interesting problem that emerges in microbial ecology: how do you compare new sequence data to existing results without reprocessing all of the sequences together and creating new de novo sequences. Early (haha) in microbial ecology, one proposed method of generating OTUs was to align new data to an existing reference database. This was done by either a closed (only fitting to the existing data) or open (fitting as much as possible to the existing data but then generating additional de novo OTUs as needed) appraoch. Due to many drawbacks, this method is not widely used these days. However, one benefit it had was the ability to add new data to an existing set of OTUs. This allowed researchers to make comparisons across projects or even to conduct meta-type-analyses. In this manuscript, the authors describe a new method that uses the strengths of modern clustering methods (e.g., OptiClust) and the ability to perform cross study comparisons through reference mapping. The others do a very thorough job of describing the new method, comparing it to existing methods, exploring the potential user options, and summarizing everything in a way that gives the reader an understanding of exactly why they may want to use the new tool, how to use it, and what benefits it will have.

The manuscript is well written and describes a new computational tool without computer science jargon. I think this manuscript and the method will be very well received because it does fill a gap in the computation analysis of amplicon sequences. A have only a few comments and suggestions as outlined below:

OTU quality: The main tool for comparing OTU quality in this manuscript is MCC. This metric is central to de novo OTU clustering in the OptiClust approach, and I think it greatly benefits amplicon-based approaches. The MCC metric is briefly described on Lines 54-55. However, since it is so central to the comparisons in the current study, I wonder if the reader would benefit from a bit more detail and, I think the reader needs to be reminded about MCC when interpreting the results. When reviewing Fig 1: the reader should know exactly why an MCC of 0.91 is better than 0.78 and have a way to conceptualizing the difference. Also, it may be helpful to remind the reader how quality was determined (see L172).

OptiClust: On lines 78-80 the authors state that "previous studies" found OptiClust generate the highest quality OTUs, but then they have only a single reference which is the original paper describing the method (and from the corresponding authors research group). Full disclosure, I use OptiClust. However, I think the authors need to either say "A previous study" or provide additional references (and perhaps one from outside the research group).

Computational Resources: Throughout the manuscript, the authors often use compute time as a measure of the computational resource demand. This is an important metric, but I also wonder about RAM. Specifically, at the end of the manuscript the authors state "we recommend using OptiClust for de novo sequencing over the split strategy with OptiFit since OptiClust is simpler to execute and performs similarly in terms of both run time and OTU quality" (L279-281). Earlier they also stated that OptiClust was faster (L244 and L257). Regardless of time comparisons, I'm also wondering if the OptiFit split strategy could be an added benefit for individuals that lack computation performance (e.g., RAM). I don't remember from the OptiClust paper how much RAM is needed or if it scaled-up with data size, but I thought it was interesting that 10% of a sequencing run could be used to generate de novo sequences with OptiClust and then OptiFit could be used to match to other 90% without sacrificing quality. Could this be a benefit for research groups that may be computationally limited?

Other Reference Based Methods: I think the authors make a good point when the compare VSEARCH and OptiFit (L261-271). Specifically, I found it interesting to think about how VSEARCH can allow more dissimilar sequences to be clustered into the reference OTU. I think this is something that is worth turning into a stronger statement. For example, since VSEARCH could cluster 3% dissimilar to a centroid without considering the other sequences being clustered, does that mean a maximum of 6% dissimilar across all reference clustered sequences is theoretically possible? In other words, is VSEARCH clustering more in the closed approach because the within-OTU maximum dissimilarity is actually very high? This was just a though, but perhaps another way to strengthen the discussion.

Reviewer #2 (Comments for the Author):

The manuscript introduces optiFit, an algorithm that builds on OptiClust that aims to provide an efficient and robust method to fit amplicon data to existing OTUs. The authors utilize OptiClust output as their starting list of reference OTUs, and aim to cluster their new query sequences to the existing output. The results section is logical and very well written: It describes how the algorithm works, the benchmarks its performance with various reference databases (SILVA, GreenGenes, and RDP), compare the results with QIME2 clustering approaches.

The reviewer is by no means an expert in computational biology, and merely has a microbial ecology based background. As such, issues regarding performance metrics, algorithm strategy, coefficient, etc.. are beyond my expertise. Judging merely based on the need for such algorithm, the overall strategy, and the apparent results and benchmarking using reference datasets and databases, I find the results convincing. The need for such addition to the suite of approaches offered by MOTHUR is also well explained. I encourage the authors to define what "high quality OTUs" and "low quality OTUs" means and how they are assessed early on in the introduction.

Reviewer #3 (Comments for the Author):

In this manuscript, Sovacool et. al. describe OptiFit, a new reference-based clustering method for assigning (amplicon) read sequences to representative OTUs. This method (OptiFit) builds upon the concept of OptiClust, a de novo clustering and OTU picking method previously developed by co-author Westcott. When carrying out clustering of a given read sequence, OptiClust utilizes information about all of the sequences within an OTU, instead of only the representative or centroid sequence for that OTU cluster, an approach used by many other clustering methods. This enables OptiClust to generate higher quality OTUs when compared to other de novo methods. OptiFit uses OptiClust's wholistic method (considering all sequences within a given reference OTU) when performing reference-based clustering/OTU picking. All query sequences start as their own OTUs, and the OptiFit algorithm reassigns those sequences to one of the reference OTUs according to which assignment would maximize the Matthews Correlation Coefficent (MCC) for the dataset; this approach takes into account each of the query sequences' similarities to all of the sequences within each reference OTU. OptiFit is capable of carrying out both open and closed reference-based clustering.

The authors demonstrate OptiFit to be a powerful reference-based clustering method. The quality of OTU assignments obtained by OptiFit is similar to those generated by OptiClust, although there is a run-time penalty for OptiFit when operating open reference mode, although it does not appear to be prohibitive. As the authors note, the quality of OTU assignments is highly dependent upon the nature of the samples being studied, as well as the reference databases being utilized, particularly when carrying out closed reference-based clustering. The inclusion of an option to perform open reference-based clustering was a beneficial decision made by the authors. The authors also compared the performance of OptiFit (and OptiClust) to VSEARCH. OptiFit outperforms VSEARCH in terms of both speed (execution time) and OTU quality (quantified using median MCC scores), although VSEARCH is generally able to classify a larger percent of reads to an OTU in closed reference mode; as the authors note, this is due to VSEARCH only considering an OTU's centroid sequence, as opposed to all of the sequences within that particular OTU, making OptiFit more selective.

The OptiFit algorithm is a natural extension of the OptiClust approach to sequence classification, albeit for situations where a reference-based approach is either required or beneficial. One of the more appealing aspects of OptiFit is that it does not change OTU assignments for samples/data that have been already analyzed when adding new information to a dataset. This functionality would be particularly useful for investigators who are carrying out meta-analysis type studies or who have very large datasets to which new samples are added regularly and reanalysis would be computationally burdensome.

A general question I have is whether the authors have considered suggesting that users make use of the data-splitting approach that they outline in their own paper. OTU classification is a duel task problem (identifying OTUs in a dataset and assigning each read sequence to an appropriate OTU). With de novo classification approaches the same data is used to both identify and classify the reads. From my perspective, the development of OptiFit makes a perfect partner for OptiClust in a Train/Test (Train/Classify) splitting situation, where the users would split data, train using OptiClust and then classify the remaining reads using OptiFit. Of course, pursuing this would be well beyond the scope of this manuscript, but the authors may consider discussing such a possibility for future work.

The manuscript is generally clear, concise, and well written. There are only so many times that one can write "clustered"/"clustering" (or other variants) in a sentence or paragraph and still have it make sense. I can tell (and appreciate) that the authors took great care in the writing process to properly convey their message.

Specific comments:

Line 4 (Abstract): The sentence discussing OptiClust appears to be tangential to the opening of the abstract. Perhaps it would be better to add this context further down in the abstract (line 16-ish).

Line 83-85: "While other tools represent reference OTUs with a single sequence, OptiFit
84 uses multiple sequences in existing OTUs as the reference and fits new sequences to
85 those reference OTUs." Should "multiple" be "all" in this sentence?

Lines 129 - 131: Please include references for Greengenes and SILVA, and indicate what versions of all three (Greengenes, SILVA, and RDP) were used, even though this information is included in the Methods section.

Lines 136-138: "Clustering sequences to ..." I think it would serve the reader to reiterate/emphasize that OptiFit was used for clustering, i.e. "Clustering query sequences with OptiFit to ...".

Lines 136-138: I suggest adding the corresponding Figure 3 labels (in parentheses), similar to what was done for the "self-split" tests described in Lines 196-198. This will help guide the reader, as there is quite a bit of information being displayed in Figure 3 as well as what is being described in this paragraph.

Lines 141-142: "Thus, open reference OptiFit produced OTUs of very similar quality as de novo clustering, ..." please add "with OptiClust" to be more specific about what the comparison being made is.

Figure 1: W-Z are designated as the query sequences in the caption. Z does not appear in the distance threshold table at the top of the figure and is not assigned to an OTU at the end of Cycle 4. From Line 115 ("Alternatively, a sequence will remained unassigned...") the reader can infer that sequence Z is different enough from the others that it was designated a singleton (outside of the distance threshold discussed in lines 111-113). If I'm correct, I encourage the authors to make this point explicitly in Figure 1's caption.

Figure 3: I suggest changing the label for "de novo" to "OptiClust (de novo)" or similar to emphasize that OptiClust was being used for those benchmark tests.

Figure 3: The second sentence in the caption ("Each dataset...") is difficult to follow due to its structure. I suggest rewriting to make it clear that de novo clustering was performed with OptiClust, and 4 different reference based trials were performed (self-split (referencing the OptiClust-ed reference subset), referencing GG, SILVA, & RDP).

### 2021-12-15 paper/paper_before-review.Rmd
Hard-code the date we submitted it

--- 2021-12-15 paper/paper.Rmd
@@
 title: "OptiFit: an improved method for fitting amplicon sequences to existing OTUs"
-date: '`r Sys.Date()`'
+date: '2021-11-08'
 bibliography: references.bib
@@
 schtools::set_knitr_opts()
-knitr::opts_knit$set(base.dir = here::here(),
-                     base.url = "../")
 knitr::opts_chunk$set(fig.path = 'figures/',

### 2021-12-15 paper/response-to-reviewers.Rmd
Bold reviewer comments & use R Markdown

--- 2021-12-15 paper/response-to-reviewers.md
@@
-Reviewer comments:
+---
+date: '`r Sys.Date()`'
+output:
+  pdf_document:
+    latex_engine: xelatex
+  github_document:
+    html_preview: false
+header_includes:
+  - \usepackage{helvet} % Helvetica font
+  - \renewcommand*\familydefault{\sfdefault}
+fontsize: 12pt
+geometry: margin=1.0in
+---
 
-Reviewer #1 (Comments for the Author):
+```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
+schtools::set_knitr_opts()
+```
 
-In this manuscript, Sovacool et al.
-describe a new method for fitting amplicon sequences to existing datasets.
-This is an interesting problem that emerges in microbial ecology: how do you compare new sequence data to existing results without reprocessing all of the sequences together and creating new de novo sequences.
-Early (haha) in microbial ecology, one proposed method of generating OTUs was to align new data to an existing reference database.
-This was done by either a closed (only fitting to the existing data) or open (fitting as much as possible to the existing data but then generating additional de novo OTUs as needed) appraoch.
-Due to many drawbacks, this method is not widely used these days.
-However, one benefit it had was the ability to add new data to an existing set of OTUs.
-This allowed researchers to make comparisons across projects or even to conduct meta-type-analyses.
-In this manuscript, the authors describe a new method that uses the strengths of modern clustering methods (e.g., OptiClust) and the ability to perform cross study comparisons through reference mapping.
-The others do a very thorough job of describing the new method, comparing it to existing methods, exploring the potential user options, and summarizing everything in a way that gives the reader an understanding of exactly why they may want to use the new tool, how to use it, and what benefits it will have.
+# Response to Reviewers
 
-The manuscript is well written and describes a new computational tool without computer science jargon.
-I think this manuscript and the method will be very well received because it does fill a gap in the computation analysis of amplicon sequences.
-A have only a few comments and suggestions as outlined below:
+## Reviewer #1 (Comments for the Author):
 
-OTU quality: The main tool for comparing OTU quality in this manuscript is MCC.
-This metric is central to de novo OTU clustering in the OptiClust approach, and I think it greatly benefits amplicon-based approaches.
-The MCC metric is briefly described on Lines 54-55.
-However, since it is so central to the comparisons in the current study, I wonder if the reader would benefit from a bit more detail and, I think the reader needs to be reminded about MCC when interpreting the results.
-When reviewing Fig 1: the reader should know exactly why an MCC of 0.91 is better than 0.78 and have a way to conceptualizing the difference.
-Also, it may be helpful to remind the reader how quality was determined (see L172).
+**In this manuscript, Sovacool et al. describe a new method for fitting amplicon
+sequences to existing datasets. This is an interesting problem that emerges in
+microbial ecology: how do you compare new sequence data to existing results
+without reprocessing all of the sequences together and creating new de novo
+sequences. Early (haha) in microbial ecology, one proposed method of generating
+OTUs was to align new data to an existing reference database. This was done by
+either a closed (only fitting to the existing data) or open (fitting as much as
+possible to the existing data but then generating additional de novo OTUs as
+needed) appraoch. Due to many drawbacks, this method is not widely used these
+days. However, one benefit it had was the ability to add new data to an existing
+set of OTUs. This allowed researchers to make comparisons across projects or
+even to conduct meta-type-analyses. In this manuscript, the authors describe a
+new method that uses the strengths of modern clustering methods (e.g.,
+OptiClust) and the ability to perform cross study comparisons through reference
+mapping. The others do a very thorough job of describing the new method,
+comparing it to existing methods, exploring the potential user options, and
+summarizing everything in a way that gives the reader an understanding of
+exactly why they may want to use the new tool, how to use it, and what benefits
+it will have.**
 
-OptiClust: On lines 78-80 the authors state that "previous studies" found OptiClust generate the highest quality OTUs, but then they have only a single reference which is the original paper describing the method (and from the corresponding authors research group).
-Full disclosure, I use OptiClust.
-However, I think the authors need to either say "A previous study" or provide additional references (and perhaps one from outside the research group).
+**The manuscript is well written and describes a new computational tool without
+computer science jargon. I think this manuscript and the method will be very
+well received because it does fill a gap in the computation analysis of amplicon
+sequences. A have only a few comments and suggestions as outlined below:**
 
-Computational Resources: Throughout the manuscript, the authors often use compute time as a measure of the computational resource demand.
-This is an important metric, but I also wonder about RAM.
-Specifically, at the end of the manuscript the authors state "we recommend using OptiClust for de novo sequencing over the split strategy with OptiFit since OptiClust is simpler to execute and performs similarly in terms of both run time and OTU quality" (L279-281).
-Earlier they also stated that OptiClust was faster (L244 and L257).
-Regardless of time comparisons, I'm also wondering if the OptiFit split strategy could be an added benefit for individuals that lack computation performance (e.g., RAM).
-I don't remember from the OptiClust paper how much RAM is needed or if it scaled-up with data size, but I thought it was interesting that 10% of a sequencing run could be used to generate de novo sequences with OptiClust and then OptiFit could be used to match to other 90% without sacrificing quality.
-Could this be a benefit for research groups that may be computationally limited?
+**OTU quality: The main tool for comparing OTU quality in this manuscript is
+MCC. This metric is central to de novo OTU clustering in the OptiClust approach,
+and I think it greatly benefits amplicon-based approaches. The MCC metric is
+briefly described on Lines 54-55. However, since it is so central to the
+comparisons in the current study, I wonder if the reader would benefit from a
+bit more detail and, I think the reader needs to be reminded about MCC when
+interpreting the results. When reviewing Fig 1: the reader should know exactly
+why an MCC of 0.91 is better than 0.78 and have a way to conceptualizing the
+difference. Also, it may be helpful to remind the reader how quality was
+determined (see L172).**
 
-Other Reference Based Methods: I think the authors make a good point when the compare VSEARCH and OptiFit (L261-271).
-Specifically, I found it interesting to think about how VSEARCH can allow more dissimilar sequences to be clustered into the reference OTU.
-I think this is something that is worth turning into a stronger statement.
-For example, since VSEARCH could cluster 3% dissimilar to a centroid without considering the other sequences being clustered, does that mean a maximum of 6% dissimilar across all reference clustered sequences is theoretically possible?
-In other words, is VSEARCH clustering more in the closed approach because the within-OTU maximum dissimilarity is actually very high?
-This was just a though, but perhaps another way to strengthen the discussion.
+**OptiClust: On lines 78-80 the authors state that "previous studies" found
+OptiClust generate the highest quality OTUs, but then they have only a single
+reference which is the original paper describing the method (and from the
+corresponding authors research group). Full disclosure, I use OptiClust.
+However, I think the authors need to either say "A previous study" or provide
+additional references (and perhaps one from outside the research group).**
 
-Reviewer #2 (Comments for the Author):
+**Computational Resources: Throughout the manuscript, the authors often use
+compute time as a measure of the computational resource demand. This is an
+important metric, but I also wonder about RAM. Specifically, at the end of the
+manuscript the authors state "we recommend using OptiClust for de novo
+sequencing over the split strategy with OptiFit since OptiClust is simpler to
+execute and performs similarly in terms of both run time and OTU quality"
+(L279-281). Earlier they also stated that OptiClust was faster (L244 and L257).
+Regardless of time comparisons, I'm also wondering if the OptiFit split strategy
+could be an added benefit for individuals that lack computation performance
+(e.g., RAM). I don't remember from the OptiClust paper how much RAM is needed or
+if it scaled-up with data size, but I thought it was interesting that 10% of a
+sequencing run could be used to generate de novo sequences with OptiClust and
+then OptiFit could be used to match to other 90% without sacrificing quality.
+Could this be a benefit for research groups that may be computationally
+limited?**
 
-The manuscript introduces optiFit, an algorithm that builds on OptiClust that aims to provide an efficient and robust method to fit amplicon data to existing OTUs.
-The authors utilize OptiClust output as their starting list of reference OTUs, and aim to cluster their new query sequences to the existing output.
-The results section is logical and very well written: It describes how the algorithm works, the benchmarks its performance with various reference databases (SILVA, GreenGenes, and RDP), compare the results with QIME2 clustering approaches.
+**Other Reference Based Methods: I think the authors make a good point when the
+compare VSEARCH and OptiFit (L261-271). Specifically, I found it interesting to
+think about how VSEARCH can allow more dissimilar sequences to be clustered into
+the reference OTU. I think this is something that is worth turning into a
+stronger statement. For example, since VSEARCH could cluster 3% dissimilar to a
+centroid without considering the other sequences being clustered, does that mean
+a maximum of 6% dissimilar across all reference clustered sequences is
+theoretically possible? In other words, is VSEARCH clustering more in the closed
+approach because the within-OTU maximum dissimilarity is actually very high?
+This was just a though, but perhaps another way to strengthen the discussion.**
 
-The reviewer is by no means an expert in computational biology, and merely has a microbial ecology based background.
-As such, issues regarding performance metrics, algorithm strategy, coefficient, etc..
-are beyond my expertise.
-Judging merely based on the need for such algorithm, the overall strategy, and the apparent results and benchmarking using reference datasets and databases, I find the results convincing.
-The need for such addition to the suite of approaches offered by MOTHUR is also well explained.
-I encourage the authors to define what "high quality OTUs" and "low quality OTUs" means and how they are assessed early on in the introduction.
+## Reviewer #2 (Comments for the Author):
 
-Reviewer #3 (Comments for the Author):
+**The manuscript introduces optiFit, an algorithm that builds on OptiClust that
+aims to provide an efficient and robust method to fit amplicon data to existing
+OTUs. The authors utilize OptiClust output as their starting list of reference
+OTUs, and aim to cluster their new query sequences to the existing output. The
+results section is logical and very well written: It describes how the algorithm
+works, the benchmarks its performance with various reference databases (SILVA,
+GreenGenes, and RDP), compare the results with QIME2 clustering approaches.**
 
-In this manuscript, Sovacool et.
-al.
-describe OptiFit, a new reference-based clustering method for assigning (amplicon) read sequences to representative OTUs.
-This method (OptiFit) builds upon the concept of OptiClust, a de novo clustering and OTU picking method previously developed by co-author Westcott.
-When carrying out clustering of a given read sequence, OptiClust utilizes information about all of the sequences within an OTU, instead of only the representative or centroid sequence for that OTU cluster, an approach used by many other clustering methods.
-This enables OptiClust to generate higher quality OTUs when compared to other de novo methods.
-OptiFit uses OptiClust's wholistic method (considering all sequences within a given reference OTU) when performing reference-based clustering/OTU picking.
-All query sequences start as their own OTUs, and the OptiFit algorithm reassigns those sequences to one of the reference OTUs according to which assignment would maximize the Matthews Correlation Coefficent (MCC) for the dataset; this approach takes into account each of the query sequences' similarities to all of the sequences within each reference OTU.
-OptiFit is capable of carrying out both open and closed reference-based clustering.
+**The reviewer is by no means an expert in computational biology, and merely has
+a microbial ecology based background. As such, issues regarding performance
+metrics, algorithm strategy, coefficient, etc.. are beyond my expertise. Judging
+merely based on the need for such algorithm, the overall strategy, and the
+apparent results and benchmarking using reference datasets and databases, I find
+the results convincing. The need for such addition to the suite of approaches
+offered by MOTHUR is also well explained. I encourage the authors to define what
+"high quality OTUs" and "low quality OTUs" means and how they are assessed early
+on in the introduction.**
 
-The authors demonstrate OptiFit to be a powerful reference-based clustering method.
-The quality of OTU assignments obtained by OptiFit is similar to those generated by OptiClust, although there is a run-time penalty for OptiFit when operating open reference mode, although it does not appear to be prohibitive.
-As the authors note, the quality of OTU assignments is highly dependent upon the nature of the samples being studied, as well as the reference databases being utilized, particularly when carrying out closed reference-based clustering.
-The inclusion of an option to perform open reference-based clustering was a beneficial decision made by the authors.
-The authors also compared the performance of OptiFit (and OptiClust) to VSEARCH.
-OptiFit outperforms VSEARCH in terms of both speed (execution time) and OTU quality (quantified using median MCC scores), although VSEARCH is generally able to classify a larger percent of reads to an OTU in closed reference mode; as the authors note, this is due to VSEARCH only considering an OTU's centroid sequence, as opposed to all of the sequences within that particular OTU, making OptiFit more selective.
+## Reviewer #3 (Comments for the Author):
 
-The OptiFit algorithm is a natural extension of the OptiClust approach to sequence classification, albeit for situations where a reference-based approach is either required or beneficial.
-One of the more appealing aspects of OptiFit is that it does not change OTU assignments for samples/data that have been already analyzed when adding new information to a dataset.
-This functionality would be particularly useful for investigators who are carrying out meta-analysis type studies or who have very large datasets to which new samples are added regularly and reanalysis would be computationally burdensome.
+**In this manuscript, Sovacool et. al. describe OptiFit, a new reference-based
+clustering method for assigning (amplicon) read sequences to representative
+OTUs. This method (OptiFit) builds upon the concept of OptiClust, a de novo
+clustering and OTU picking method previously developed by co-author Westcott.
+When carrying out clustering of a given read sequence, OptiClust utilizes
+information about all of the sequences within an OTU, instead of only the
+representative or centroid sequence for that OTU cluster, an approach used by
+many other clustering methods. This enables OptiClust to generate higher quality
+OTUs when compared to other de novo methods. OptiFit uses OptiClust's wholistic
+method (considering all sequences within a given reference OTU) when performing
+reference-based clustering/OTU picking. All query sequences start as their own
+OTUs, and the OptiFit algorithm reassigns those sequences to one of the
+reference OTUs according to which assignment would maximize the Matthews
+Correlation Coefficent (MCC) for the dataset; this approach takes into account
+each of the query sequences' similarities to all of the sequences within each
+reference OTU. OptiFit is capable of carrying out both open and closed
+reference-based clustering.**
 
-A general question I have is whether the authors have considered suggesting that users make use of the data-splitting approach that they outline in their own paper.
-OTU classification is a duel task problem (identifying OTUs in a dataset and assigning each read sequence to an appropriate OTU).
-With de novo classification approaches the same data is used to both identify and classify the reads.
-From my perspective, the development of OptiFit makes a perfect partner for OptiClust in a Train/Test (Train/Classify) splitting situation, where the users would split data, train using OptiClust and then classify the remaining reads using OptiFit.
-Of course, pursuing this would be well beyond the scope of this manuscript, but the authors may consider discussing such a possibility for future work.
+**The authors demonstrate OptiFit to be a powerful reference-based clustering
+method. The quality of OTU assignments obtained by OptiFit is similar to those
+generated by OptiClust, although there is a run-time penalty for OptiFit when
+operating open reference mode, although it does not appear to be prohibitive. As
+the authors note, the quality of OTU assignments is highly dependent upon the
+nature of the samples being studied, as well as the reference databases being
+utilized, particularly when carrying out closed reference-based clustering. The
+inclusion of an option to perform open reference-based clustering was a
+beneficial decision made by the authors. The authors also compared the
+performance of OptiFit (and OptiClust) to VSEARCH. OptiFit outperforms VSEARCH
+in terms of both speed (execution time) and OTU quality (quantified using median
+MCC scores), although VSEARCH is generally able to classify a larger percent of
+reads to an OTU in closed reference mode; as the authors note, this is due to
+VSEARCH only considering an OTU's centroid sequence, as opposed to all of the
+sequences within that particular OTU, making OptiFit more selective.**
 
-The manuscript is generally clear, concise, and well written.
-There are only so many times that one can write "clustered"/"clustering" (or other variants) in a sentence or paragraph and still have it make sense.
-I can tell (and appreciate) that the authors took great care in the writing process to properly convey their message.
+**The OptiFit algorithm is a natural extension of the OptiClust approach to
+sequence classification, albeit for situations where a reference-based approach
+is either required or beneficial. One of the more appealing aspects of OptiFit
+is that it does not change OTU assignments for samples/data that have been
+already analyzed when adding new information to a dataset. This functionality
+would be particularly useful for investigators who are carrying out
+meta-analysis type studies or who have very large datasets to which new samples
+are added regularly and reanalysis would be computationally burdensome.**
 
-Specific comments:
+**A general question I have is whether the authors have considered suggesting
+that users make use of the data-splitting approach that they outline in their
+own paper. OTU classification is a duel task problem (identifying OTUs in a
+dataset and assigning each read sequence to an appropriate OTU). With de novo
+classification approaches the same data is used to both identify and classify
+the reads. From my perspective, the development of OptiFit makes a perfect
+partner for OptiClust in a Train/Test (Train/Classify) splitting situation,
+where the users would split data, train using OptiClust and then classify the
+remaining reads using OptiFit. Of course, pursuing this would be well beyond the
+scope of this manuscript, but the authors may consider discussing such a
+possibility for future work.**
 
-Line 4 (Abstract): The sentence discussing OptiClust appears to be tangential to the opening of the abstract. Perhaps it would be better to add this context further down in the abstract (line 16-ish).
+**The manuscript is generally clear, concise, and well written. There are only
+so many times that one can write "clustered"/"clustering" (or other variants) in
+a sentence or paragraph and still have it make sense. I can tell (and
+appreciate) that the authors took great care in the writing process to properly
+convey their message.**
 
-Line 83-85: "While other tools represent reference OTUs with a single sequence, OptiFit
-84 uses multiple sequences in existing OTUs as the reference and fits new sequences to
-85 those reference OTUs." Should "multiple" be "all" in this sentence?
+**Specific comments:**
 
-Lines 129 - 131: Please include references for Greengenes and SILVA, and indicate what versions of all three (Greengenes, SILVA, and RDP) were used, even though this information is included in the Methods section.
+**Line 4 (Abstract): The sentence discussing OptiClust appears to be tangential
+to the opening of the abstract. Perhaps it would be better to add this context
+further down in the abstract (line 16-ish).**
 
-Lines 136-138: "Clustering sequences to ..." I think it would serve the reader to reiterate/emphasize that OptiFit was used for clustering, i.e. "Clustering query sequences with OptiFit to ...".
+**Line 83-85: "While other tools represent reference OTUs with a single
+sequence, OptiFit 84 uses multiple sequences in existing OTUs as the reference
+and fits new sequences to 85 those reference OTUs." Should "multiple" be "all"
+in this sentence?**
 
-Lines 136-138: I suggest adding the corresponding Figure 3 labels (in parentheses), similar to what was done for the "self-split" tests described in Lines 196-198.
-This will help guide the reader, as there is quite a bit of information being displayed in Figure 3 as well as what is being described in this paragraph.
+**Lines 129 - 131: Please include references for Greengenes and SILVA, and
+indicate what versions of all three (Greengenes, SILVA, and RDP) were used, even
+though this information is included in the Methods section.**
 
-Lines 141-142: "Thus, open reference OptiFit produced OTUs of very similar quality as de novo clustering, ..." please add "with OptiClust" to be more specific about what the comparison being made is.
+**Lines 136-138: "Clustering sequences to ..." I think it would serve the reader
+to reiterate/emphasize that OptiFit was used for clustering, i.e. "Clustering
+query sequences with OptiFit to ...".**
 
-Figure 1: W-Z are designated as the query sequences in the caption.
-Z does not appear in the distance threshold table at the top of the figure and is not assigned to an OTU at the end of Cycle 4.
-From Line 115 ("Alternatively, a sequence will remained unassigned...") the reader can infer that sequence Z is different enough from the others that it was designated a singleton (outside of the distance threshold discussed in lines 111-113).
-If I'm correct, I encourage the authors to make this point explicitly in Figure 1's caption.
+**Lines 136-138: I suggest adding the corresponding Figure 3 labels (in
+parentheses), similar to what was done for the "self-split" tests described in
+Lines 196-198. This will help guide the reader, as there is quite a bit of
+information being displayed in Figure 3 as well as what is being described in
+this paragraph.**
 
-Figure 3: I suggest changing the label for "de novo" to "OptiClust (de novo)" or similar to emphasize that OptiClust was being used for those benchmark tests.
+**Lines 141-142: "Thus, open reference OptiFit produced OTUs of very similar
+quality as de novo clustering, ..." please add "with OptiClust" to be more
+specific about what the comparison being made is.**
 
-Figure 3: The second sentence in the caption ("Each dataset...") is difficult to follow due to its structure.
-I suggest rewriting to make it clear that de novo clustering was performed with OptiClust, and 4 different reference based trials were performed (self-split (referencing the OptiClust-ed reference subset), referencing GG, SILVA, & RDP).
+**Figure 1: W-Z are designated as the query sequences in the caption. Z does not
+appear in the distance threshold table at the top of the figure and is not
+assigned to an OTU at the end of Cycle 4. From Line 115 ("Alternatively, a
+sequence will remained unassigned...") the reader can infer that sequence Z is
+different enough from the others that it was designated a singleton (outside of
+the distance threshold discussed in lines 111-113). If I'm correct, I encourage
+the authors to make this point explicitly in Figure 1's caption.**
+
+**Figure 3: I suggest changing the label for "de novo" to "OptiClust (de novo)"
+or similar to emphasize that OptiClust was being used for those benchmark
+tests.**
+
+**Figure 3: The second sentence in the caption ("Each dataset...") is difficult
+to follow due to its structure. I suggest rewriting to make it clear that de
+novo clustering was performed with OptiClust, and 4 different reference based
+trials were performed (self-split (referencing the OptiClust-ed reference
+subset), referencing GG, SILVA, & RDP).**

### 2021-12-15 paper/response-to-reviewers.md
Bold reviewer comments & use R Markdown

@@
-Reviewer comments:
+2021-12-15
 
-Reviewer #1 (Comments for the Author):
+# Response to Reviewers
 
-In this manuscript, Sovacool et al.
-describe a new method for fitting amplicon sequences to existing datasets.
-This is an interesting problem that emerges in microbial ecology: how do you compare new sequence data to existing results without reprocessing all of the sequences together and creating new de novo sequences.
-Early (haha) in microbial ecology, one proposed method of generating OTUs was to align new data to an existing reference database.
-This was done by either a closed (only fitting to the existing data) or open (fitting as much as possible to the existing data but then generating additional de novo OTUs as needed) appraoch.
-Due to many drawbacks, this method is not widely used these days.
-However, one benefit it had was the ability to add new data to an existing set of OTUs.
-This allowed researchers to make comparisons across projects or even to conduct meta-type-analyses.
-In this manuscript, the authors describe a new method that uses the strengths of modern clustering methods (e.g., OptiClust) and the ability to perform cross study comparisons through reference mapping.
-The others do a very thorough job of describing the new method, comparing it to existing methods, exploring the potential user options, and summarizing everything in a way that gives the reader an understanding of exactly why they may want to use the new tool, how to use it, and what benefits it will have.
+## Reviewer #1 (Comments for the Author):
 
-The manuscript is well written and describes a new computational tool without computer science jargon.
-I think this manuscript and the method will be very well received because it does fill a gap in the computation analysis of amplicon sequences.
-A have only a few comments and suggestions as outlined below:
+**In this manuscript, Sovacool et al. describe a new method for fitting
+amplicon sequences to existing datasets. This is an interesting problem
+that emerges in microbial ecology: how do you compare new sequence data
+to existing results without reprocessing all of the sequences together
+and creating new de novo sequences. Early (haha) in microbial ecology,
+one proposed method of generating OTUs was to align new data to an
+existing reference database. This was done by either a closed (only
+fitting to the existing data) or open (fitting as much as possible to
+the existing data but then generating additional de novo OTUs as needed)
+appraoch. Due to many drawbacks, this method is not widely used these
+days. However, one benefit it had was the ability to add new data to an
+existing set of OTUs. This allowed researchers to make comparisons
+across projects or even to conduct meta-type-analyses. In this
+manuscript, the authors describe a new method that uses the strengths of
+modern clustering methods (e.g., OptiClust) and the ability to perform
+cross study comparisons through reference mapping. The others do a very
+thorough job of describing the new method, comparing it to existing
+methods, exploring the potential user options, and summarizing
+everything in a way that gives the reader an understanding of exactly
+why they may want to use the new tool, how to use it, and what benefits
+it will have.**
 
-OTU quality: The main tool for comparing OTU quality in this manuscript is MCC.
-This metric is central to de novo OTU clustering in the OptiClust approach, and I think it greatly benefits amplicon-based approaches.
-The MCC metric is briefly described on Lines 54-55.
-However, since it is so central to the comparisons in the current study, I wonder if the reader would benefit from a bit more detail and, I think the reader needs to be reminded about MCC when interpreting the results.
-When reviewing Fig 1: the reader should know exactly why an MCC of 0.91 is better than 0.78 and have a way to conceptualizing the difference.
-Also, it may be helpful to remind the reader how quality was determined (see L172).
+**The manuscript is well written and describes a new computational tool
+without computer science jargon. I think this manuscript and the method
+will be very well received because it does fill a gap in the computation
+analysis of amplicon sequences. A have only a few comments and
+suggestions as outlined below:**
 
-OptiClust: On lines 78-80 the authors state that "previous studies" found OptiClust generate the highest quality OTUs, but then they have only a single reference which is the original paper describing the method (and from the corresponding authors research group).
-Full disclosure, I use OptiClust.
-However, I think the authors need to either say "A previous study" or provide additional references (and perhaps one from outside the research group).
+**OTU quality: The main tool for comparing OTU quality in this
+manuscript is MCC. This metric is central to de novo OTU clustering in
+the OptiClust approach, and I think it greatly benefits amplicon-based
+approaches. The MCC metric is briefly described on Lines 54-55. However,
+since it is so central to the comparisons in the current study, I wonder
+if the reader would benefit from a bit more detail and, I think the
+reader needs to be reminded about MCC when interpreting the results.
+When reviewing Fig 1: the reader should know exactly why an MCC of 0.91
+is better than 0.78 and have a way to conceptualizing the difference.
+Also, it may be helpful to remind the reader how quality was determined
+(see L172).**
 
-Computational Resources: Throughout the manuscript, the authors often use compute time as a measure of the computational resource demand.
-This is an important metric, but I also wonder about RAM.
-Specifically, at the end of the manuscript the authors state "we recommend using OptiClust for de novo sequencing over the split strategy with OptiFit since OptiClust is simpler to execute and performs similarly in terms of both run time and OTU quality" (L279-281).
-Earlier they also stated that OptiClust was faster (L244 and L257).
-Regardless of time comparisons, I'm also wondering if the OptiFit split strategy could be an added benefit for individuals that lack computation performance (e.g., RAM).
-I don't remember from the OptiClust paper how much RAM is needed or if it scaled-up with data size, but I thought it was interesting that 10% of a sequencing run could be used to generate de novo sequences with OptiClust and then OptiFit could be used to match to other 90% without sacrificing quality.
-Could this be a benefit for research groups that may be computationally limited?
+**OptiClust: On lines 78-80 the authors state that “previous studies”
+found OptiClust generate the highest quality OTUs, but then they have
+only a single reference which is the original paper describing the
+method (and from the corresponding authors research group). Full
+disclosure, I use OptiClust. However, I think the authors need to either
+say “A previous study” or provide additional references (and perhaps one
+from outside the research group).**
 
-Other Reference Based Methods: I think the authors make a good point when the compare VSEARCH and OptiFit (L261-271).
-Specifically, I found it interesting to think about how VSEARCH can allow more dissimilar sequences to be clustered into the reference OTU.
-I think this is something that is worth turning into a stronger statement.
-For example, since VSEARCH could cluster 3% dissimilar to a centroid without considering the other sequences being clustered, does that mean a maximum of 6% dissimilar across all reference clustered sequences is theoretically possible?
-In other words, is VSEARCH clustering more in the closed approach because the within-OTU maximum dissimilarity is actually very high?
-This was just a though, but perhaps another way to strengthen the discussion.
+**Computational Resources: Throughout the manuscript, the authors often
+use compute time as a measure of the computational resource demand. This
+is an important metric, but I also wonder about RAM. Specifically, at
+the end of the manuscript the authors state “we recommend using
+OptiClust for de novo sequencing over the split strategy with OptiFit
+since OptiClust is simpler to execute and performs similarly in terms of
+both run time and OTU quality” (L279-281). Earlier they also stated that
+OptiClust was faster (L244 and L257). Regardless of time comparisons,
+I’m also wondering if the OptiFit split strategy could be an added
+benefit for individuals that lack computation performance (e.g., RAM). I
+don’t remember from the OptiClust paper how much RAM is needed or if it
+scaled-up with data size, but I thought it was interesting that 10% of a
+sequencing run could be used to generate de novo sequences with
+OptiClust and then OptiFit could be used to match to other 90% without
+sacrificing quality. Could this be a benefit for research groups that
+may be computationally limited?**
 
-Reviewer #2 (Comments for the Author):
+**Other Reference Based Methods: I think the authors make a good point
+when the compare VSEARCH and OptiFit (L261-271). Specifically, I found
+it interesting to think about how VSEARCH can allow more dissimilar
+sequences to be clustered into the reference OTU. I think this is
+something that is worth turning into a stronger statement. For example,
+since VSEARCH could cluster 3% dissimilar to a centroid without
+considering the other sequences being clustered, does that mean a
+maximum of 6% dissimilar across all reference clustered sequences is
+theoretically possible? In other words, is VSEARCH clustering more in
+the closed approach because the within-OTU maximum dissimilarity is
+actually very high? This was just a though, but perhaps another way to
+strengthen the discussion.**
 
-The manuscript introduces optiFit, an algorithm that builds on OptiClust that aims to provide an efficient and robust method to fit amplicon data to existing OTUs.
-The authors utilize OptiClust output as their starting list of reference OTUs, and aim to cluster their new query sequences to the existing output.
-The results section is logical and very well written: It describes how the algorithm works, the benchmarks its performance with various reference databases (SILVA, GreenGenes, and RDP), compare the results with QIME2 clustering approaches.
+## Reviewer #2 (Comments for the Author):
 
-The reviewer is by no means an expert in computational biology, and merely has a microbial ecology based background.
-As such, issues regarding performance metrics, algorithm strategy, coefficient, etc..
-are beyond my expertise.
-Judging merely based on the need for such algorithm, the overall strategy, and the apparent results and benchmarking using reference datasets and databases, I find the results convincing.
-The need for such addition to the suite of approaches offered by MOTHUR is also well explained.
-I encourage the authors to define what "high quality OTUs" and "low quality OTUs" means and how they are assessed early on in the introduction.
+**The manuscript introduces optiFit, an algorithm that builds on
+OptiClust that aims to provide an efficient and robust method to fit
+amplicon data to existing OTUs. The authors utilize OptiClust output as
+their starting list of reference OTUs, and aim to cluster their new
+query sequences to the existing output. The results section is logical
+and very well written: It describes how the algorithm works, the
+benchmarks its performance with various reference databases (SILVA,
+GreenGenes, and RDP), compare the results with QIME2 clustering
+approaches.**
 
-Reviewer #3 (Comments for the Author):
+**The reviewer is by no means an expert in computational biology, and
+merely has a microbial ecology based background. As such, issues
+regarding performance metrics, algorithm strategy, coefficient, etc..
+are beyond my expertise. Judging merely based on the need for such
+algorithm, the overall strategy, and the apparent results and
+benchmarking using reference datasets and databases, I find the results
+convincing. The need for such addition to the suite of approaches
+offered by MOTHUR is also well explained. I encourage the authors to
+define what “high quality OTUs” and “low quality OTUs” means and how
+they are assessed early on in the introduction.**
 
-In this manuscript, Sovacool et.
-al.
-describe OptiFit, a new reference-based clustering method for assigning (amplicon) read sequences to representative OTUs.
-This method (OptiFit) builds upon the concept of OptiClust, a de novo clustering and OTU picking method previously developed by co-author Westcott.
-When carrying out clustering of a given read sequence, OptiClust utilizes information about all of the sequences within an OTU, instead of only the representative or centroid sequence for that OTU cluster, an approach used by many other clustering methods.
-This enables OptiClust to generate higher quality OTUs when compared to other de novo methods.
-OptiFit uses OptiClust's wholistic method (considering all sequences within a given reference OTU) when performing reference-based clustering/OTU picking.
-All query sequences start as their own OTUs, and the OptiFit algorithm reassigns those sequences to one of the reference OTUs according to which assignment would maximize the Matthews Correlation Coefficent (MCC) for the dataset; this approach takes into account each of the query sequences' similarities to all of the sequences within each reference OTU.
-OptiFit is capable of carrying out both open and closed reference-based clustering.
+## Reviewer #3 (Comments for the Author):
 
-The authors demonstrate OptiFit to be a powerful reference-based clustering method.
-The quality of OTU assignments obtained by OptiFit is similar to those generated by OptiClust, although there is a run-time penalty for OptiFit when operating open reference mode, although it does not appear to be prohibitive.
-As the authors note, the quality of OTU assignments is highly dependent upon the nature of the samples being studied, as well as the reference databases being utilized, particularly when carrying out closed reference-based clustering.
-The inclusion of an option to perform open reference-based clustering was a beneficial decision made by the authors.
-The authors also compared the performance of OptiFit (and OptiClust) to VSEARCH.
-OptiFit outperforms VSEARCH in terms of both speed (execution time) and OTU quality (quantified using median MCC scores), although VSEARCH is generally able to classify a larger percent of reads to an OTU in closed reference mode; as the authors note, this is due to VSEARCH only considering an OTU's centroid sequence, as opposed to all of the sequences within that particular OTU, making OptiFit more selective.
+**In this manuscript, Sovacool et. al. describe OptiFit, a new
+reference-based clustering method for assigning (amplicon) read
+sequences to representative OTUs. This method (OptiFit) builds upon the
+concept of OptiClust, a de novo clustering and OTU picking method
+previously developed by co-author Westcott. When carrying out clustering
+of a given read sequence, OptiClust utilizes information about all of
+the sequences within an OTU, instead of only the representative or
+centroid sequence for that OTU cluster, an approach used by many other
+clustering methods. This enables OptiClust to generate higher quality
+OTUs when compared to other de novo methods. OptiFit uses OptiClust’s
+wholistic method (considering all sequences within a given reference
+OTU) when performing reference-based clustering/OTU picking. All query
+sequences start as their own OTUs, and the OptiFit algorithm reassigns
+those sequences to one of the reference OTUs according to which
+assignment would maximize the Matthews Correlation Coefficent (MCC) for
+the dataset; this approach takes into account each of the query
+sequences’ similarities to all of the sequences within each reference
+OTU. OptiFit is capable of carrying out both open and closed
+reference-based clustering.**
 
-The OptiFit algorithm is a natural extension of the OptiClust approach to sequence classification, albeit for situations where a reference-based approach is either required or beneficial.
-One of the more appealing aspects of OptiFit is that it does not change OTU assignments for samples/data that have been already analyzed when adding new information to a dataset.
-This functionality would be particularly useful for investigators who are carrying out meta-analysis type studies or who have very large datasets to which new samples are added regularly and reanalysis would be computationally burdensome.
+**The authors demonstrate OptiFit to be a powerful reference-based
+clustering method. The quality of OTU assignments obtained by OptiFit is
+similar to those generated by OptiClust, although there is a run-time
+penalty for OptiFit when operating open reference mode, although it does
+not appear to be prohibitive. As the authors note, the quality of OTU
+assignments is highly dependent upon the nature of the samples being
+studied, as well as the reference databases being utilized, particularly
+when carrying out closed reference-based clustering. The inclusion of an
+option to perform open reference-based clustering was a beneficial
+decision made by the authors. The authors also compared the performance
+of OptiFit (and OptiClust) to VSEARCH. OptiFit outperforms VSEARCH in
+terms of both speed (execution time) and OTU quality (quantified using
+median MCC scores), although VSEARCH is generally able to classify a
+larger percent of reads to an OTU in closed reference mode; as the
+authors note, this is due to VSEARCH only considering an OTU’s centroid
+sequence, as opposed to all of the sequences within that particular OTU,
+making OptiFit more selective.**
 
-A general question I have is whether the authors have considered suggesting that users make use of the data-splitting approach that they outline in their own paper.
-OTU classification is a duel task problem (identifying OTUs in a dataset and assigning each read sequence to an appropriate OTU).
-With de novo classification approaches the same data is used to both identify and classify the reads.
-From my perspective, the development of OptiFit makes a perfect partner for OptiClust in a Train/Test (Train/Classify) splitting situation, where the users would split data, train using OptiClust and then classify the remaining reads using OptiFit.
-Of course, pursuing this would be well beyond the scope of this manuscript, but the authors may consider discussing such a possibility for future work.
+**The OptiFit algorithm is a natural extension of the OptiClust approach
+to sequence classification, albeit for situations where a
+reference-based approach is either required or beneficial. One of the
+more appealing aspects of OptiFit is that it does not change OTU
+assignments for samples/data that have been already analyzed when adding
+new information to a dataset. This functionality would be particularly
+useful for investigators who are carrying out meta-analysis type studies
+or who have very large datasets to which new samples are added regularly
+and reanalysis would be computationally burdensome.**
 
-The manuscript is generally clear, concise, and well written.
-There are only so many times that one can write "clustered"/"clustering" (or other variants) in a sentence or paragraph and still have it make sense.
-I can tell (and appreciate) that the authors took great care in the writing process to properly convey their message.
+**A general question I have is whether the authors have considered
+suggesting that users make use of the data-splitting approach that they
+outline in their own paper. OTU classification is a duel task problem
+(identifying OTUs in a dataset and assigning each read sequence to an
+appropriate OTU). With de novo classification approaches the same data
+is used to both identify and classify the reads. From my perspective,
+the development of OptiFit makes a perfect partner for OptiClust in a
+Train/Test (Train/Classify) splitting situation, where the users would
+split data, train using OptiClust and then classify the remaining reads
+using OptiFit. Of course, pursuing this would be well beyond the scope
+of this manuscript, but the authors may consider discussing such a
+possibility for future work.**
 
-Specific comments:
+**The manuscript is generally clear, concise, and well written. There
+are only so many times that one can write “clustered”/“clustering” (or
+other variants) in a sentence or paragraph and still have it make sense.
+I can tell (and appreciate) that the authors took great care in the
+writing process to properly convey their message.**
 
-Line 4 (Abstract): The sentence discussing OptiClust appears to be tangential to the opening of the abstract. Perhaps it would be better to add this context further down in the abstract (line 16-ish).
+**Specific comments:**
 
-Line 83-85: "While other tools represent reference OTUs with a single sequence, OptiFit
-84 uses multiple sequences in existing OTUs as the reference and fits new sequences to
-85 those reference OTUs." Should "multiple" be "all" in this sentence?
+**Line 4 (Abstract): The sentence discussing OptiClust appears to be
+tangential to the opening of the abstract. Perhaps it would be better to
+add this context further down in the abstract (line 16-ish).**
 
-Lines 129 - 131: Please include references for Greengenes and SILVA, and indicate what versions of all three (Greengenes, SILVA, and RDP) were used, even though this information is included in the Methods section.
+**Line 83-85: “While other tools represent reference OTUs with a single
+sequence, OptiFit 84 uses multiple sequences in existing OTUs as the
+reference and fits new sequences to 85 those reference OTUs.” Should
+“multiple” be “all” in this sentence?**
 
-Lines 136-138: "Clustering sequences to ..." I think it would serve the reader to reiterate/emphasize that OptiFit was used for clustering, i.e. "Clustering query sequences with OptiFit to ...".
+**Lines 129 - 131: Please include references for Greengenes and SILVA,
+and indicate what versions of all three (Greengenes, SILVA, and RDP)
+were used, even though this information is included in the Methods
+section.**
 
-Lines 136-138: I suggest adding the corresponding Figure 3 labels (in parentheses), similar to what was done for the "self-split" tests described in Lines 196-198.
-This will help guide the reader, as there is quite a bit of information being displayed in Figure 3 as well as what is being described in this paragraph.
+**Lines 136-138: “Clustering sequences to …” I think it would serve the
+reader to reiterate/emphasize that OptiFit was used for clustering,
+i.e. “Clustering query sequences with OptiFit to …”.**
 
-Lines 141-142: "Thus, open reference OptiFit produced OTUs of very similar quality as de novo clustering, ..." please add "with OptiClust" to be more specific about what the comparison being made is.
+**Lines 136-138: I suggest adding the corresponding Figure 3 labels (in
+parentheses), similar to what was done for the “self-split” tests
+described in Lines 196-198. This will help guide the reader, as there is
+quite a bit of information being displayed in Figure 3 as well as what
+is being described in this paragraph.**
 
-Figure 1: W-Z are designated as the query sequences in the caption.
-Z does not appear in the distance threshold table at the top of the figure and is not assigned to an OTU at the end of Cycle 4.
-From Line 115 ("Alternatively, a sequence will remained unassigned...") the reader can infer that sequence Z is different enough from the others that it was designated a singleton (outside of the distance threshold discussed in lines 111-113).
-If I'm correct, I encourage the authors to make this point explicitly in Figure 1's caption.
+**Lines 141-142: “Thus, open reference OptiFit produced OTUs of very
+similar quality as de novo clustering, …” please add “with OptiClust” to
+be more specific about what the comparison being made is.**
 
-Figure 3: I suggest changing the label for "de novo" to "OptiClust (de novo)" or similar to emphasize that OptiClust was being used for those benchmark tests.
+**Figure 1: W-Z are designated as the query sequences in the caption. Z
+does not appear in the distance threshold table at the top of the figure
+and is not assigned to an OTU at the end of Cycle 4. From Line 115
+(“Alternatively, a sequence will remained unassigned…”) the reader can
+infer that sequence Z is different enough from the others that it was
+designated a singleton (outside of the distance threshold discussed in
+lines 111-113). If I’m correct, I encourage the authors to make this
+point explicitly in Figure 1’s caption.**
 
-Figure 3: The second sentence in the caption ("Each dataset...") is difficult to follow due to its structure.
-I suggest rewriting to make it clear that de novo clustering was performed with OptiClust, and 4 different reference based trials were performed (self-split (referencing the OptiClust-ed reference subset), referencing GG, SILVA, & RDP).
+**Figure 3: I suggest changing the label for “de novo” to “OptiClust (de
+novo)” or similar to emphasize that OptiClust was being used for those
+benchmark tests.**
+
+**Figure 3: The second sentence in the caption (“Each dataset…”) is
+difficult to follow due to its structure. I suggest rewriting to make it
+clear that de novo clustering was performed with OptiClust, and 4
+different reference based trials were performed (self-split (referencing
+the OptiClust-ed reference subset), referencing GG, SILVA, & RDP).**

### 2021-12-15 paper/paper.Rmd
Bold reviewer comments & use R Markdown

--- 2021-12-15 paper/paper.Rmd
@@
 schtools::set_knitr_opts()
-knitr::opts_knit$set(base.dir = here::here(), 
-                     base.url = "../")
 knitr::opts_chunk$set(fig.path = 'figures/',

### 2021-12-15 paper/response-to-reviewers.Rmd
Use Helvetica

@@
 ---
-date: '`r Sys.Date()`'
 output:
@@
     html_preview: false
-header_includes:
-  - \usepackage{helvet} % Helvetica font
-  - \renewcommand*\familydefault{\sfdefault}
+mainfont: "Helvetica"
 fontsize: 12pt

### 2021-12-16 code/R/plot_results_split.R
Show memory usage

@@ -48,7 +48,7 @@ dat <- list(optifit_split, opticlust) %>%
     method %>% as.character() != "closed" ~ NA_real_,
     TRUE ~ fraction_mapped
   )) %>%
-  pivot_longer(c(mcc, fraction_mapped, sec),
+  pivot_longer(c(mcc, fraction_mapped, sec, mem_gb),
     names_to = "metric"
   ) %>%
   mutate(
@@ -57,9 +57,10 @@ dat <- list(optifit_split, opticlust) %>%
         metric == "mcc" ~ "MCC",
         metric == "fraction_mapped" ~ "Fraction Mapped",
         metric == "sec" ~ "Runtime (sec)",
+        metric == 'mem_gb' ~ "Memory Usage (GB)",
         TRUE ~ metric
       ),
-      levels = c("MCC", "Fraction Mapped", "Runtime (sec)")
+      levels = c("MCC", "Fraction Mapped", "Runtime (sec)", "Memory Usage (GB)")
     ),
     strategy = factor(
       case_when(

### 2021-12-16 code/R/plot_results_sum.R

@@ -53,7 +53,7 @@ mothur_vsearch <- list(optifit_all, opticlust, vsearch) %>%
     method %>% as.character() != "closed" ~ NA_real_,
     TRUE ~ fraction_mapped
   )) %>%
-  pivot_longer(c(mcc, fraction_mapped, sec),
+  pivot_longer(c(mcc, fraction_mapped, sec, mem_gb),
     names_to = "metric"
   ) %>%
   mutate(
@@ -62,9 +62,10 @@ mothur_vsearch <- list(optifit_all, opticlust, vsearch) %>%
         metric == "mcc" ~ "MCC",
         metric == "fraction_mapped" ~ "Fraction Mapped",
         metric == "sec" ~ "Runtime (sec)",
+        metric == 'mem_gb' ~ "Memory Usage (GB)",
         TRUE ~ metric
       ),
-      levels = c("MCC", "Fraction Mapped", "Runtime (sec)")
+      levels = c("MCC", "Fraction Mapped", "Runtime (sec)", "Memory Usage (GB)")
     ),
     strategy = factor(
       case_when(

### 2021-12-16 paper/paper.Rmd

@@
 The complete workflow and supporting files required to reproduce this manuscript
-are available at https://github.com/SchlossLab/Sovacool_OptiFit_2021.
-<!-- TODO: update repo link with journal title -->
+are available at https://github.com/SchlossLab/Sovacool_OptiFit_mSphere_2022.
 

### 2021-12-16 paper/paper_before-review.Rmd
Lock in stats from before the review

@@
     fig_caption: yes
+    keep_tex: false
   github_document:
@@
 
-load(here("results", "stats.RData"))
+load(here("results", "stats_before-review.RData"))
 

### 2022-01-12 exploratory/2022/sensspec_2022-01.Rmd
Plot sensspec & confusion matrix

---
date: '`r Sys.Date()`'
output: 
  github_document:
    html_preview: false
---

# Sensitivity & Specificity

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      fig.path = "figures/",
                      dpi = 300
                      )
```
```{r deps, warning=FALSE, message=FALSE}
set.seed(2020112)
library(glue)
library(here)
library(tidyverse)
theme_set(theme_bw())
```

```{r dat}
dat_agg <- read_tsv(here('results', 'aggregated.tsv'))
```

```{r sensspec}
dat_agg %>% 
  filter(dataset == 'human', 
         ref == 'gg' | method == 'de_novo') %>% 
  ggplot(aes(sensitivity, specificity, color = tool, shape = method)) +
  stat_summary(
    geom = "point",
    fun = median,
    size = 2
  ) +
  scale_shape_manual(values = list(open = 1, closed = 19, de_novo = 17)) +
  ylim(0, 1) + xlim(0, 1)
```

```{r conf_matrix, fig.dim=c(6.87,6.87)}
dat_agg %>% 
  filter(dataset == 'human', 
         ref == 'gg' | strategy != 'database') %>%
  pivot_longer(c(tp, tn, fp, fn), 
               names_to = 'conf_mat_quad', 
               values_to = 'conf_mat_val') %>%
  mutate(conf_mat_quad = factor(conf_mat_quad, 
                                levels = c('tp', 'fn', 'fp', 'tn'))) %>% 
  # mutate(tool_method = factor(glue("{tool}_{method}"),
  #                             levels = c("mothur_de_novo", "vsearch_de_novo",
  #                                        "mothur_closed", "vsearch_closed",
  #                                        "mothur_open", "vsearch_open"))
  #        ) %>% 
  ggplot(aes(
    strategy,
    conf_mat_val,
    color = tool,
    shape = method
  )) +
  facet_wrap("conf_mat_quad", 
             scales = 'free_y',
             nrow = 2) +
  stat_summary(
    geom = "point",
    fun = median,
    size = 2,
    position = position_dodge(width = 0.4)
  ) +
  scale_shape_manual(values = list(
    open = 1,
    closed = 19,
    de_novo = 17
  )) +
  labs(y = '', x = '', title = 'Confusion Matrix for the human dataset')
```

### 2022-01-12 code/R/calc_results_stats.R
🎨  Style R code

@@ -11,7 +11,7 @@ coeff_var <- function(x) {
 
 
 dat <- read_tsv(here("results", "summarized.tsv"))
-agg <- read_tsv(here('results', 'aggregated.tsv'))
+agg <- read_tsv(here("results", "aggregated.tsv"))
 ################################################################################
 # de novo datasets
 opticlust_mcc <- agg %>%
@@ -33,7 +33,7 @@ opticlust_mem <- agg %>%
     strategy == "de_novo",
     tool == "mothur"
   ) %>%
-  pull(mem_gb) %>% 
+  pull(mem_gb) %>%
   median()
 dn_vsearch_mcc <- agg %>%
   filter(strategy == "de_novo", tool == "vsearch") %>%
@@ -286,10 +286,12 @@ mcc_closed_fit_db_vs_vsearch <- rel_diff(closed_fit_db_mcc, closed_vsearch_mcc) 
 ################################################################################
 # fit split
 cv_fit_split_mcc <- agg %>%
-  filter(strategy == "self-split",
-         tool == "mothur",
-         ref_weight == "simple",
-         ref_frac == 0.5) %>%
+  filter(
+    strategy == "self-split",
+    tool == "mothur",
+    ref_weight == "simple",
+    ref_frac == 0.5
+  ) %>%
   pull(mcc) %>%
   coeff_var()
 
@@ -354,20 +356,23 @@ mem_closed_fit_split_vs_clust <- rel_diff(closed_fit_split_mem, opticlust_mem)
 mem_open_fit_split_vs_clust <- rel_diff(open_fit_split_mem, opticlust_mem)
 
 cv_fit_split_mcc_human_simple <- agg %>%
-  filter(strategy == "self-split",
-         tool == "mothur",
-         dataset == "human",
-         ref_weight == "simple") %>%
+  filter(
+    strategy == "self-split",
+    tool == "mothur",
+    dataset == "human",
+    ref_weight == "simple"
+  ) %>%
   pull(mcc) %>%
   coeff_var()
 
-cv_fit_split_mem_human_simple <- agg %>% filter(
-  strategy == "self-split",
-  tool == "mothur",
-  dataset == "human",
-  ref_weight == "simple"
-) %>%
-  pull(mem_gb) %>% 
+cv_fit_split_mem_human_simple <- agg %>%
+  filter(
+    strategy == "self-split",
+    tool == "mothur",
+    dataset == "human",
[…]

### 2022-01-13 paper/paper.Rmd

@@
 `r sec_open_fit_split_vs_clust`%, respectively.
+Random access memory (RAM) usage was similar, with OptiFit requiring slightly 
+more RAM in gigabytes than OptiFit. 
+Open and closed reference OptiFit required `r mem_open_fit_split_vs_clust`% and
+`r mem_closed_fit_split_vs_clust`% more RAM than OptiClust, respectively 
+(data not shown).
 The split dataset strategy also performed `r sec_closed_fit_split_vs_db`% faster
@@
 ```{r assert_split_fracs, include = FALSE}
-close_enough(cv_fit_split_mcc_human_simple, 0, tol = 10^-3)
+close_enough(cv_fit_split_mcc_human_simple, 0, tol = 10^-2)
 assert_that(sec_fit_split_human_simple_1 > sec_fit_split_human_simple_9)
@@
 with 90% of sequences in the reference.
+The RAM usage was virtually the same across reference fraction sizes, with a
+coefficient of variation of `r cv_fit_split_mem_human_simple` for the human
+dataset (data not shown).
 In closed reference mode, the fraction of sequences that mapped increased as the
@@
 These trends held for the other datasets as well.
-Thus, the reference fraction did not affect OTU quality in terms of MCC score,
-but did affect the run time and the fraction of sequences that mapped during the
-closed reference clustering.
+Thus, the reference fraction did not affect OTU quality in terms of MCC score
+nor the memory usage, but did affect the run time and the fraction of sequences
+that mapped during the closed reference clustering.
 

### 2022-01-13 code/R/plot_algorithm_diagram.R
🎨  Style R code

@@ -9,7 +9,7 @@ library(reticulate)
 library(tidygraph)
 library(tidyverse)
 set.seed(20200308)
-#use_python('/usr/local/bin/python3')
+# use_python('/usr/local/bin/python3')
 source_python(here("code", "py", "algorithm_diagram.py"))
 optifit <- create_optifit()
 optifit_iters <- optifit$iterate %>%

### 2022-01-13 code/R/plot_results_sum.R

@@ -107,10 +107,11 @@ color_labels <- lapply(
   }
 ) %>% unlist()
 
-plot_results_sum <- mothur_vsearch %>% 
-  mutate(tool = case_when(tool == "vsearch" ~ "VSEARCH",
-                          tool == "mothur" ~ "OptiClust (_de novo_) or OptiFit")
-         ) %>% 
+plot_results_sum <- mothur_vsearch %>%
+  mutate(tool = case_when(
+    tool == "vsearch" ~ "VSEARCH",
+    tool == "mothur" ~ "OptiClust (_de novo_) or OptiFit"
+  )) %>%
   ggplot(aes(value, strategy, color = tool, shape = method)) +
   # stat_summary(geom = "linerange",
   #              fun.data = med_iqr,
@@ -139,13 +140,13 @@ plot_results_sum <- mothur_vsearch %>%
     legend.text = element_markdown(),
     legend.position = "top",
     legend.margin = margin(t = 0, r = 0, b = 0, l = 0, unit = "pt"),
-    legend.spacing.x = unit(0.5, 'pt'),
+    legend.spacing.x = unit(0.5, "pt"),
     plot.margin = unit(x = c(0, 0, 0, 0), units = "pt")
   ) +
   guides(
     shape = guide_legend(order = 1),
     colour = guide_legend(
-       override.aes = list(size = -1)
+      override.aes = list(size = -1)
     )
   )
 

### 2022-01-13 code/py/algorithm_diagram.py
🎨  Style Python code 🐍

@@ -3,7 +3,7 @@
 from collections import defaultdict, Counter
 from copy import deepcopy
 from itertools import combinations
-from math import sqrt, factorial#, comb
+from math import sqrt, factorial  # , comb
 import numpy
 import pandas
 
@@ -32,10 +32,12 @@ def create_optifit():
     optifit = OptiFit(ref_otus, query_seqs, query_dist_mat, n_seqs=54)
     return optifit
 
-def write_iters(optifit_iters, outdir = "figures/algorithm_steps"):
+
+def write_iters(optifit_iters, outdir="figures/algorithm_steps"):
     for i, iter_dir in enumerate(optifit_iters):
-        iter_dir['nodes'].to_csv(f"{outdir}/{i}_nodes")
-        iter_dir['edges'].to_csv(f"{outdir}/{i}_edges")
+        iter_dir["nodes"].to_csv(f"{outdir}/{i}_nodes")
+        iter_dir["edges"].to_csv(f"{outdir}/{i}_edges")
+
 
 def opticlust_example():
     dist_frame = {
@@ -446,6 +448,7 @@ class OptiIter:
     @property
     def to_dict(self):
         return {"nodes": self.nodes, "edges": self.edges}
+
 
 class OptiOption:
     def __init__(self, curr_fitmap, curr_seq, sim_seq):

### 2022-01-13 paper/paper.Rmd
Make remaining edits suggested by reviewers

@@
 schtools::set_knitr_opts()
-knitr::opts_chunk$set(fig.path = 'figures/',
-                      dpi = 300)
+knitr::opts_chunk$set(fig.path = dplyr::if_else(params$include_figures, 
+                                                'figures/',
+                                                'blanks/'),
+                      dpi = 300
+                      )
 ```
@@
 important step in characterizing microbial communities across large datasets.
-OptiClust, a _de novo_ OTU clustering method, produces higher
-quality OTUs than other methods at comparable or faster speeds.
 A notable difference between _de novo_ clustering and database-dependent
@@
 the OptiFit algorithm.
-Like OptiClust, OptiFit considers the similarity of all pairs of reference and
-query sequences to produce OTUs of the best quality.
+Inspired by the _de novo_ OptiClust algorithm, OptiFit considers the similarity 
+of all pairs of reference and query sequences to produce OTUs of the best 
+possible quality.
 We tested OptiFit using four datasets with two strategies: 1) clustering to a
@@
 @westcott_novo_2015].
+OTU quality can be conceptualized as how well the OTU assignments match the
+definition set by the distance threshold, i.e. whether sequence pairs that are
+at least as similar as the distance threshold are assigned to the same OTU and
+sequence pairs that are more dissimilar than the distance threshold are assigned
+to different OTUs.
 
@@
 thus less willing to take on false positives.
+
 A limitation of _de novo_ clustering is that different OTU assignments will be
@@
 they are not represented by a similar sequence in the database.
-Previous studies found that the OptiClust _de novo_ clustering algorithm created
+We previously found that the OptiClust _de novo_ clustering algorithm created
 the highest quality OTU assignments of all clustering methods [@westcott_opticlust_2017].
@@
 While other tools represent reference OTUs with a single sequence, OptiFit uses
-multiple sequences in existing OTUs as the reference and fits new sequences to 
+all sequences in existing OTUs as the reference and fits new sequences to 
 those reference OTUs.
@@
 a different OTU or remain in its current OTU, choosing the option that results
-in a better Matthews correlation coefficient (MCC) [@westcott_opticlust_2017].
+in a better MCC score [@westcott_opticlust_2017].
 The MCC uses all values from a confusion matrix and ranges from negative one to
 one, with a score of one occurring when all sequence pairs are true positives
-and true negatives and a score of negative one occurring when all pairs are
-false positives and false negatives.
+and true negatives, a score of negative one occurring when all pairs are
+false positives and false negatives, and a score of zero when there are equal
+numbers of true and false assignments (i.e. no better than random guessing).
 Sequence pairs that are similar to each other (i.e. within the distance
@@
 OTU.
+Thus, a pair of sequences is considered correctly assigned when their OTU
+assignment matches the OTU definition set by the distance threshold.
 OptiClust iterations continue until the MCC stabilizes or until a maximum number
@@
 There are two strategies for generating OTUs with OptiFit:
-1) cluster the query sequences to reference OTUs generated by _de novo_ clustering an
-independent database, or
+1) cluster the query sequences to reference OTUs generated by _de novo_
+clustering an independent database, or
 2) split the dataset into a reference and query fraction, cluster the reference
@@
 To test how OptiFit performs for reference-based clustering, we clustered each
-dataset to three databases of reference OTUs: the Greengenes database, the SILVA
-non-redundant database, and the Ribosomal Database Project (RDP)
-[@desantis_greengenes_2006; @quast_silva_2013; @cole_ribosomal_2014].
+dataset to three databases of reference OTUs: 
+the Greengenes database v13_8_99 [@desantis_greengenes_2006], 
+the SILVA non-redundant database v132 [@quast_silva_2013], 
+and the Ribosomal Database Project (RDP) v16 [@cole_ribosomal_2014].
 Reference OTUs for each database were created by performing _de novo_ clustering
@@
 respectively.
-Clustering sequences to Greengenes and SILVA in closed reference mode performed
-similarly, with median MCC scores of `r closed_fit_gg_mcc` and
-`r closed_fit_silva_mcc` respectively, while the median MCC
-was `r closed_fit_rdp_mcc` when clustering to RDP
-(Figure \ref{fig:results_sum}).
+Clustering query sequences with OptiFit to Greengenes and SILVA in closed
+reference mode performed similarly, with median MCC scores of 
+`r closed_fit_gg_mcc` and `r closed_fit_silva_mcc` respectively, while the median
+MCC was `r closed_fit_rdp_mcc` when clustering to RDP (Figure
+\ref{fig:results_sum}; "db: Greengenes", "db: SILVA", and "db: RDP").
 For comparison, clustering datasets with OptiClust produced an average MCC score
-of `r opticlust_mcc`.
+of `r opticlust_mcc` (Figure \ref{fig:results_sum}; "_de novo_").
 This gap in OTU quality mostly disappeared when clustering in open reference
@@
 Thus, open reference OptiFit produced OTUs of very similar quality as _de novo_
-clustering, and closed reference OptiFit followed closely behind as long as a
+clustering with OptiClust, and closed reference OptiFit followed closely behind as long as a
 suitable reference database was chosen.
@@
 In closed reference mode, OptiFit produced `r mcc_closed_fit_db_vs_vsearch`%
-higher quality OTUs than VSEARCH, but VSEARCH was able to cluster
-`r frac_vsearch_vs_fit`% more query sequences than OptiFit to the Greengenes
-reference database (Figure \ref{fig:results_sum}).
+higher quality OTUs than VSEARCH in terms of MCC score, but VSEARCH was able to
+cluster `r frac_vsearch_vs_fit`% more query sequences than OptiFit to the
+Greengenes reference database (Figure \ref{fig:results_sum}).
 This is because VSEARCH only considers the distances between each query sequence
@@
 the reference by a simple random sample (a 50% split)
-(Figure \ref{fig:results_sum}; self-split).
+(Figure \ref{fig:results_sum}; "self-split").
 OTU quality was similar to that from OptiClust regardless of
@@
 In contrast, VSEARCH only requires each query sequence to be similar to the
-single centroid sequence that seeded the OTU.
+single centroid sequence that seeded the OTU, thus allowing pairs of query
+sequences to be less similar to each other than the threshold specified.
 Because of this, VSEARCH sacrifices OTU quality by allowing more dissimilar
-sequences to be clustered into OTUs.
+sequences to be clustered into the same OTUs.
 

### 2022-01-13 paper/paper.Rmd
Fix behavior of PDFs without figures

@@
                                                 'figures/',
-                                                'blanks/'),
+                                                'figures/blanks/'),
                       dpi = 300

### 2022-01-13 paper/paper_before-review.Rmd
Fix behavior of PDFs without figures

@@
 schtools::set_knitr_opts()
-knitr::opts_chunk$set(fig.path = 'figures/',
+knitr::opts_chunk$set(fig.path = dplyr::if_else(params$include_figures, 
+                                                'figures/',
+                                                'figures/blanks/'),
                       dpi = 300)

### 2022-01-13 paper/response-to-reviewers.md
Render everything

@@
-2021-12-15
-
 # Response to Reviewers
@@
 
+Thank you for your enthusiasm!
+
 **OTU quality: The main tool for comparing OTU quality in this
@@
 
+We describe MCC in more detail under the description of the OptiFit
+algorithm, which we expect readers would read just before looking at
+Figure 1. We added a few more details to the MCC description in this
+section (L111-113 and L117-118), In the caption of Figure 1 we also
+added examples of the confusion matrix values for the first and last MCC
+scores.
+
+We revised the sentence formerly at L172 to note that we mean “quality”
+in terms of MCC score (L184).
+
 **OptiClust: On lines 78-80 the authors state that “previous studies”
@@
 from outside the research group).**
+
+Thank you for catching this. We have revised this sentence to say “We
+previously found” in order to clarify that this claim is supported by a
+single study from our group (L86).
 
@@
 
+Saving computational resources is an interesting thought! However, we
+found that the RAM usage was similar between OptiClust and OptiFit
+regardless of the reference fraction size used in the split strategy.
+Both OptiClust and OptiFit consider the distances between all pairs of
+sequences, so changing the reference fraction and using the remaining
+sequences as the query doesn’t change the total number of distances that
+are considered in the MCC. We have added statements to note how the RAM
+usage of OptiFit & OptiClust compared (L217-219) and how the RAM usage
+compared across reference fraction sizes for the split dataset strategy
+(L233-235). We chose not to include the RAM data in the figures since we
+found RAM to be similar, and the figures already contain a lot of
+information.
+
 **Other Reference Based Methods: I think the authors make a good point
@@
 strengthen the discussion.**
+
+We agree, it is theoretically possible that when using VSEARCH for
+reference clustering, query sequences may be 3% dissimilar to a centroid
+sequence but possibly up to 6% dissimilar to each other. Since we did
+not investigate this directly, we hesitate to make *too* strong of a
+statement in the discussion, but we did add a statement (L289) to better
+emphasize this point.
 
@@
 they are assessed early on in the introduction.**
+
+Thank you for your feedback!
+
+We added a sentence in the introduction to clarify the meaning of
+quality (L54-57), and provided additional details of the MCC score under
+the description of the OptiFit algorithm (L111-113 and L117-118).
 
@@
 
+We are pleased that you made this connection! We agree that it is well
+beyond the scope of this manuscript, and we are already working on this
+for another forthcoming paper. Thus, we chose to merely mention machine
+learning alongside the other potential applications in order to keep
+this paper narrowly focused on the OptiFit algorithm.
+
 **The manuscript is generally clear, concise, and well written. There
@@
 
+Thank you!
+
 **Specific comments:**
+
+Thank you for your thorough reading of the manuscript and the detailed
+comments you provided below.
 
@@
 add this context further down in the abstract (line 16-ish).**
+
+We have moved the mention of OptiClust near where you suggested (L17).
+We also trimmed down the abstract for brevity.
 
@@
 
+Yes, thank you! (L92)
+
 **Lines 129 - 131: Please include references for Greengenes and SILVA,
@@
 
+References to all three were included at the end of the sentence in the
+order they were mentioned. To clear up any confusion, we moved each
+database’s reference to immediately after its mention, and also noted
+the version numbers here as you suggested (L141-143).
+
 **Lines 136-138: “Clustering sequences to …” I think it would serve the
@@
 i.e. “Clustering query sequences with OptiFit to …”.**
+
+We agree! (L147-148)
 
@@
 
+Good suggestion. (L150)
+
 **Lines 141-142: “Thus, open reference OptiFit produced OTUs of very
@@
 be more specific about what the comparison being made is.**
+
+Agreed. (L155)
 
@@
 
+That’s correct, we added a sentence to the caption as you suggested.
+
 **Figure 3: I suggest changing the label for “de novo” to “OptiClust (de
@@
 benchmark tests.**
+
+In Figure 3, we show results from *de novo* clustering with both the
+OptiClust algorithm in mothur (red triangles) and VSEARCH (blue
+triangles). Changing the label to “OptiClust (*de novo*)” would not be
+accurate, as VSEARCH does not use the OptiClust algorithm. Instead, we
+changed the red color label from “mothur” to “OptiClust (\_de novo) or
+OptiFit” in Figure 3.
 
@@
 the OptiClust-ed reference subset), referencing GG, SILVA, & RDP).**
+
+We edited the caption to make the workflow more clear.

### 2022-01-13 code/py/algorithm_diagram.py
🎨  Style Python code 🐍

@@ -403,9 +403,9 @@ class OptiFit:
                     iteration = OptiIter(curr_fitmap, seq, self.query_seqs)
                     iterations.append(iteration.to_dict)
                     curr_fitmap = iteration.best_map
-        return 
+        return
 
-    @property   
+    @property
     def iterate_obj(self):
         iterations = list()
         curr_fitmap = self.fitmap

### 2022-01-14 paper/response-to-reviewers.Rmd

@@
 
+Thank you for your enthusiasm!
+
 **OTU quality: The main tool for comparing OTU quality in this manuscript is
@@
 
+We describe MCC in more detail under the description of the OptiFit algorithm, 
+which we expect readers would read just before looking at Figure 1.
+We added a few more details to the MCC description in this section (L111-113 and
+L117-118), 
+In the caption of Figure 1 we also added examples of the confusion matrix 
+values for the first and last MCC scores.
+
+We revised the sentence formerly at L172 to note that we mean "quality" in terms
+of MCC score (L184).
+
 **OptiClust: On lines 78-80 the authors state that "previous studies" found
@@
 additional references (and perhaps one from outside the research group).**
+
+Thank you for catching this.
+We have revised this sentence to say "We previously found" in order to clarify
+that this claim is supported by a single study from our group (L86).
 
@@
 
+Saving computational resources is an interesting thought! 
+However, we found that the RAM usage was similar between OptiClust and OptiFit
+regardless of the reference fraction size used in the split strategy.
+Both OptiClust and OptiFit consider the distances between all pairs of
+sequences, so changing the reference fraction and using the remaining sequences
+as the query doesn't change the total number of distances that are considered in
+the MCC.
+We have added statements to note how the RAM usage of OptiFit & OptiClust
+compared (L217-219) and how the RAM usage compared across reference fraction
+sizes for the split dataset strategy (L233-235).
+We chose not to include the RAM data in the figures since we found RAM to be
+similar, and the figures already contain a lot of information.
+
 **Other Reference Based Methods: I think the authors make a good point when the
@@
 This was just a though, but perhaps another way to strengthen the discussion.**
+
+We agree, it is theoretically possible that when using VSEARCH for reference
+clustering, query sequences may be 3% dissimilar to a centroid sequence but
+possibly up to 6% dissimilar to each other.
+Since we did not investigate this directly, we hesitate to make *too* strong of 
+a statement in the discussion, but we did add a statement (L289) to
+better emphasize this point.
 
@@
 on in the introduction.**
+
+Thank you for your feedback!
+
+We added a sentence in the introduction to clarify the meaning of quality
+(L54-57), and provided additional details of the MCC score under the description
+of the OptiFit algorithm (L111-113 and L117-118).
 
@@
 
+We are pleased that you made this connection! 
+We agree that it is well beyond the scope of this manuscript, and we are already
+working on this for another forthcoming paper.
+Thus, we chose to merely mention machine learning alongside the other potential
+applications in order to keep this paper narrowly focused on the OptiFit
+algorithm.
+
 **The manuscript is generally clear, concise, and well written. There are only
@@
 
+Thank you!
+
 **Specific comments:**
+
+Thank you for your thorough reading of the manuscript and the detailed comments
+you provided below.
 
@@
 further down in the abstract (line 16-ish).**
+
+We have moved the mention of OptiClust near where you suggested (L17).
+We also trimmed down the abstract for brevity.
 
@@
 
+Yes, thank you! (L92)
+
 **Lines 129 - 131: Please include references for Greengenes and SILVA, and
@@
 
+References to all three were included at the end of the sentence in the order
+they were mentioned.
+To clear up any confusion, we moved each database's reference to immediately
+after its mention, and also noted the version numbers here as you suggested (L141-143).
+
 **Lines 136-138: "Clustering sequences to ..." I think it would serve the reader
@@
 query sequences with OptiFit to ...".**
+
+We agree! (L147-148)
 
@@
 
+Good suggestion. (L150)
+
 **Lines 141-142: "Thus, open reference OptiFit produced OTUs of very similar
@@
 specific about what the comparison being made is.**
+
+Agreed. (L155)
 
@@
 
+That's correct, we added a sentence to the caption as you suggested.
+
 **Figure 3: I suggest changing the label for "de novo" to "OptiClust (de novo)"
@@
 tests.**
+
+In Figure 3, we show results from _de novo_ clustering with both the OptiClust
+algorithm in mothur (red triangles) and VSEARCH (blue triangles). 
+Changing the label to "OptiClust (_de novo_)" would not be accurate, as VSEARCH 
+does not use the OptiClust algorithm.
+Instead, we changed the red color label from "mothur" to "OptiClust (_de novo)
+or OptiFit" in Figure 3.
 
@@
 subset), referencing GG, SILVA, & RDP).**
+
+We edited the caption to make the workflow more clear.

### 2022-01-14 exploratory/2022/sensspec_2022-01.Rmd

@@
 
-```{r conf_matrix, fig.dim=c(6.87,6.87)}
+```{r conf_matrix, fig.dim=c(6.87,5)}
 dat_agg %>% 

### 2022-01-18 paper/response-to-reviewers.Rmd
Fix typo
fix typooooo

@@
 We added a few more details to the MCC description in this section (L111-113 and
-L117-118), 
+L117-118).
 In the caption of Figure 1 we also added examples of the confusion matrix 
@@
 does not use the OptiClust algorithm.
-Instead, we changed the red color label from "mothur" to "OptiClust (_de novo)
+Instead, we changed the red color label from "mothur" to "OptiClust (_de novo_)
 or OptiFit" in Figure 3.

### 2022-01-18 paper/paper_before-review.Rmd
Move figures to end, just before references
Make sure figure captions are included in latexdiff

@@
 
-figures_meta <- yaml::read_yaml(here('paper', 'figures.yaml'))
-
-figs <- function(fig_name, attribute, include_figures = params$include_figures) {
+figures_meta <- yaml::read_yaml(here('paper', 'figures_before-review.yaml'))
+include_figures = params$include_figures
+figs <- function(fig_name, attribute) {
   value <- figures_meta[[fig_name]][[attribute]]
-  if (attribute == 'cap') {
-    value <- glue("\\label{fig:[fig_name]}[value]", .open='[', .close = ']')
-  } else if (attribute == 'dim' & !isTRUE(include_figures)) {
-    value <- c(0.1,0.1)
-  } else {
+  if (attribute != 'cap') {
     value <- eval(parse(text = value))
@@
 
-render_figure <- function(fig_name, include_figures = params$include_figures) {
-  if (isTRUE(include_figures)) {
+render_figure <- function(fig_name) {
     figs(fig_name, 'path') %>%
@@
       grid::grid.raster()
-  } else {
-    ggplot() + theme_void()
-  }
 }
@@
 
-```{r algorithm, fig.dim=figs('algorithm','dim'), fig.cap=figs('algorithm','cap')}
-render_figure('algorithm')
-```
-
 OptiFit begins where OptiClust ends, starting with a list of reference OTUs and
@@
 the sequence pairs that are within the distance threshold (e.g. 0.03)
-(Figure \ref{fig:algorithm}).
+(Figure 1).
 Initially, all query sequences are placed into separate OTUs.
@@
 with OptiClust at a distance threshold of 3% using the V4 region of each
-sequence (see Figure \ref{fig:workflow}).
+sequence (see Figure 2).
 After trimming to the V4 region, the databases contained 
@@
 `r closed_fit_silva_mcc` respectively, while the median MCC
-was `r closed_fit_rdp_mcc` when clustering to RDP
-(Figure \ref{fig:results_sum}).
+was `r closed_fit_rdp_mcc` when clustering to RDP (Figure 3).
 For comparison, clustering datasets with OptiClust produced an average MCC score
@@
 suitable reference database was chosen.
-
-```{r workflow, fig.dim=figs('workflow', 'dim'), fig.cap=figs('workflow','cap')}
-render_figure('workflow')
-```
 
@@
 (`r frac_fit_gg`%) than with SILVA (`r frac_fit_silva`%) or with the RDP
-(`r frac_fit_rdp`%) (Figure \ref{fig:results_sum}).
+(`r frac_fit_rdp`%) (Figure 3).
 This mirrored the result reported above that Greengenes produced better OTUs in
@@
 Across all dataset and database combinations, closed reference OptiFit
-outperformed both OptiClust and open reference OptiFit (Figure \ref{fig:results_sum}).
+outperformed both OptiClust and open reference OptiFit (Figure 3).
 For example, with the human dataset fit to SILVA reference OTUs, the average run
@@
 clustering sacrifices OTU quality for execution speed.
-
-```{r results_sum, fig.dim=figs('results_sum', 'dim'), fig.cap=figs('results_sum', 'cap')}
-render_figure('results_sum')
-```
 
@@
 `r frac_vsearch_vs_fit`% more query sequences than OptiFit to the Greengenes
-reference database (Figure \ref{fig:results_sum}).
+reference database (Figure 3).
 This is because VSEARCH only considers the distances between each query sequence
@@
 the reference by a simple random sample (a 50% split)
-(Figure \ref{fig:results_sum}; self-split).
+(Figure 3; self-split).
 OTU quality was similar to that from OptiClust regardless of
@@
 
-```{r results_split, fig.dim=figs('results_split', 'dim'), fig.cap=figs('results_split', 'cap')}
-render_figure('results_split')
-```
-
 ```{r assert_split_fracs, include = FALSE}
@@
 sequences were created, with the remaining sequences used for the query
-(Figure \ref{fig:results_split}).
+(Figure 4).
 OTU quality was remarkably consistent across reference fraction sizes.
@@
 relative abundance, and weighting by similarity to other sequences in the
-dataset (Figure \ref{fig:results_split}).
+dataset (Figure 4).
 OTU quality in terms of MCC was similar across all three sampling methods
@@
 
+## Figures
+
+```{r algorithm, fig.dim=figs('algorithm','dim'), fig.cap=figs('algorithm','cap'), eval=include_figures}
+render_figure('algorithm')
+```
+
+```{r fig1cap, results='asis', eval=!include_figures}
+cat(paste('\\textbf{Figure 1}:', figs('algorithm', 'cap')))
+```
+
+```{r workflow, fig.dim=figs('workflow', 'dim'), fig.cap=figs('workflow','cap'), eval=include_figures}
+render_figure('workflow')
+```
+
+```{r fig2cap, results='asis', eval=!include_figures}
+cat(paste('\\textbf{Figure 2}:', figs('workflow', 'cap')))
+```
+
+```{r results_sum, fig.dim=figs('results_sum', 'dim'), fig.cap=figs('results_sum', 'cap'), eval=include_figures}
+render_figure('results_sum')
+```
+
+```{r fig3cap, results='asis', eval=!include_figures}
+cat(paste('\\textbf{Figure 3}:', figs('results_sum', 'cap')))
+```
+
+```{r results_split, fig.dim=figs('results_split', 'dim'), fig.cap=figs('results_split', 'cap'), eval=include_figures}
+render_figure('results_split')
+```
+
+```{r fig4cap, results='asis', eval=!include_figures}
+cat(paste('\\textbf{Figure 4}:', figs('results_split', 'cap')))
+```
+
 ## References

### 2022-01-18 paper/response-to-reviewers.md
Fix typo

@@
 Figure 1. We added a few more details to the MCC description in this
-section (L111-113 and L117-118), In the caption of Figure 1 we also
+section (L111-113 and L117-118). In the caption of Figure 1 we also
 added examples of the confusion matrix values for the first and last MCC
@@
 accurate, as VSEARCH does not use the OptiClust algorithm. Instead, we
-changed the red color label from “mothur” to “OptiClust (\_de novo) or
+changed the red color label from “mothur” to “OptiClust (*de novo*) or
 OptiFit” in Figure 3.

### 2022-01-18 paper/paper.Rmd
Fix typo: OptiFit uses more RAM than OptiClust

@@
 figures_meta <- yaml::read_yaml(here('paper', 'figures.yaml'))
-
-figs <- function(fig_name, attribute, include_figures = params$include_figures) {
+include_figures <- params$include_figures
+figs <- function(fig_name, attribute, incl_figs = include_figures) {
   value <- figures_meta[[fig_name]][[attribute]]
-  if (attribute == 'cap') {
+  if (attribute == 'cap' | attribute == 'caption') {
     value <- glue("\\label{fig:[fig_name]}[value]", .open='[', .close = ']')
-  } else if (attribute == 'dim' & !isTRUE(include_figures)) {
+  } else if (attribute == 'dim' & !isTRUE(incl_figs)) {
     value <- c(0.1,0.1)
@@
 `r closed_fit_gg_mcc` and `r closed_fit_silva_mcc` respectively, while the median
-MCC was `r closed_fit_rdp_mcc` when clustering to RDP (Figure
-\ref{fig:results_sum}; "db: Greengenes", "db: SILVA", and "db: RDP").
+MCC was `r closed_fit_rdp_mcc` when clustering to RDP 
+(Figure 3; "db: Greengenes", "db: SILVA", and "db: RDP").
 For comparison, clustering datasets with OptiClust produced an average MCC score
-of `r opticlust_mcc` (Figure \ref{fig:results_sum}; "_de novo_").
+of `r opticlust_mcc` (Figure 3; "_de novo_").
 This gap in OTU quality mostly disappeared when clustering in open reference
@@
 cluster `r frac_vsearch_vs_fit`% more query sequences than OptiFit to the
-Greengenes reference database (Figure \ref{fig:results_sum}).
+Greengenes reference database (Figure 3).
 This is because VSEARCH only considers the distances between each query sequence
@@
 using the split dataset strategy with half of the sequences selected for
-the reference by a simple random sample (a 50% split)
-(Figure \ref{fig:results_sum}; "self-split").
+the reference by a simple random sample (a 50% split) (Figure 3; "self-split").
 OTU quality was similar to that from OptiClust regardless of
@@
 Random access memory (RAM) usage was similar, with OptiFit requiring slightly 
-more RAM in gigabytes than OptiFit. 
+more RAM in gigabytes than OptiClust. 
 Open and closed reference OptiFit required `r mem_open_fit_split_vs_clust`% and
@@
 query sequences than the database strategy.
-
-```{r results_split, fig.dim=figs('results_split', 'dim'), fig.cap=figs('results_split', 'cap')}
-render_figure('results_split')
-```
 

### 2022-01-18 paper/response-to-reviewers.Rmd
Move figures after references & cite in response

@@
 In the caption of Figure 1 we also added examples of the confusion matrix 
-values for the first and last MCC scores.
+values for the first and last MCC scores (L439-443, L449-450).
 
@@
 
-That's correct, we added a sentence to the caption as you suggested.
+That's correct, we added a sentence to the caption as you suggested (L448-449).
 
@@
 
-We edited the caption to make the workflow more clear.
+We edited the caption to make the workflow more clear (L464-469).

### 2022-01-19 paper/response-to-reviewers.md

@@
 added examples of the confusion matrix values for the first and last MCC
-scores.
+scores (L439-443, L449-450).
 
@@
 
-That’s correct, we added a sentence to the caption as you suggested.
+That’s correct, we added a sentence to the caption as you suggested
+(L448-449).
 

### 2022-02-03 paper/paper.Rmd
Move figures after references & cite in response
Fix typo
Hopefully mSphere fixes it in the published version soon too

@@
 Sequence pairs that are not similar to each other are true negatives if they are
-not clustered into the same OTU, and false positives if they are not in the same
+not clustered into the same OTU, and false positives if they are in the same
 OTU.
@@
 
-### Data Processing Steps
+### Data processing steps
 
@@
 
-### Data and Code Availability
+### Data and code availability
 
@@
 
-## Figures
+## References
+
+\setlength{\parindent}{-0.25in}
+\setlength{\leftskip}{0.25in}
+\noindent
+
+<div id="refs"></div>
+
+\setlength{\parindent}{0in}
+\setlength{\leftskip}{0in}
+
+\newpage
+
+```{r figheader, eval=!include_figures, results='asis'}
+cat('## Figure Captions\n')
+```
 
@@
 ```
-
-## References
